### 2024-07-30 _manuscript/_tex/index.tex
start putting things into the paper

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{jds}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={My wonderful paper},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{My wonderful paper}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
This is the abstract
\end{abstract}

\section{Introduction}\label{introduction}

In this paper, a concept called analysis plan is proposed to describe
the logical structure of a data analysis. An analysis plan is a set of
analysis steps plus their expected outcomes. It is a formal
representation of the analysis process and can be used to guide the
analysis process, to communicate and compare the analysis process to
others, and to evaluate the analysis process. The concept of analysis
plan is illustrated with examples and the implications of the concept
for data analysis practice is discussed.

The analysis plan described in this paper should be differentiated from
the pre-specifies analysis plan document often used in biostatistics to
specifies the hypothesis, data collection mechanism, statistical
procedures etc of randomized experiments.

\section{Analysis plan}\label{analysis-plan}

describe/ define what analysis plan is

analysis plan and outcome plan

analysis plan as unit tests to divide the ``result universe'', which
allows us to answer questions like:

\begin{itemize}
\tightlist
\item
  how would the results change if the value of a unit test change
\item
  whether our outcome expectation aligns with the plan expectation,
  meaning whether ``it is possible for the combination of plan
  expectations to produce the outcome expectation''
\end{itemize}

\section{Examples}\label{examples}

\subsection{A toy example}\label{a-toy-example}

\subsection{Linear regression}\label{linear-regression}

Consider a linear regression model to study the effect of PM10 on
mortality (provide context of using PM10 to study mortality). From the
literature, analysts may expect a significant PM10 coefficient in the
linear model from the literature. This is the outcome expectation:
p-value of PM10 coefficient is less than 0.05. There are multiple
factors that can affect the outcome expectation of this analysis, for
example, 1) sample size, 2) model specification, and 3) correlation
structure between variables. Temperature is often an important
confounder to consider in such study (add reference).

To construct the result universe, we can simulate datasets with
different configerations of the above factors. Here, sample sizes of 50,
100, 500, and 1000 are considered. Two model specifications are
included: 1) linear model with PM10 as the only covariate
(\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
temperature as covariates
(\(\text{mortality} \sim \text{PM10} + \text{temp}\). A grid-based
approach is used to simulate correlation structure. Reasonable ranges of
correlation between the three variables are

\begin{align*}
\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
\end{align*}

\includegraphics{index_files/figure-pdf/unnamed-chunk-2-1.pdf}

\phantomsection\label{cell-fig-result-universe}
\begin{figure}[H]

\centering{

\captionsetup{labelsep=none}\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}

}

\caption{\label{fig-result-universe}}

\end{figure}%

Figure~\ref{fig-result-universe} shows that result universe of the
linear regression model with a) colored by whether the p-value of PM10
is significant (less than 0.05), b) highlighting the effect of adding
temperature to the model for a fixed correlation structure, and c)
highlighting the effect of increasing sample size.

\end{document}

### 2024-07-30 index.qmd
start putting things into the paper

---
title: "My wonderful paper"
abstract: "This is the abstract"
bibliography: references.bib
documentclass: jds
format: pdf
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

# Introduction

In this paper, a concept called analysis plan is proposed to describe the logical structure of a data analysis. An analysis plan is a set of analysis steps plus their expected outcomes. It is a formal representation of the analysis process and can be used to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process. The concept of analysis plan is illustrated with examples and the implications of the concept for data analysis practice is discussed.

The analysis plan described in this paper should be differentiated from the pre-specifies analysis plan document often used in biostatistics to specifies the hypothesis, data collection mechanism, statistical procedures etc of randomized experiments.

# Analysis plan

describe/ define what analysis plan is

analysis plan and outcome plan

analysis plan as unit tests to divide the "result universe", which allows us to answer questions like:

- how would the results change if the value of a unit test change
- whether our outcome expectation aligns with the plan expectation, meaning whether "it is possible for the combination of plan expectations to produce the outcome expectation"

# Examples

## A toy example

## Linear regression

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature. There are multiple factors that can affect the outcome expectation of linear regression as such, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables. Temperature is often an important confounder to consider in such study (add reference).

To construct the result universe, we can simulate datasets with different configerations of the above factors. Here, sample sizes of 50, 100, 500, and 1000 are considered. Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$. A grid-based approach is used to simulate correlation structure. Reasonable ranges of correlation between the three variables are

\begin{align*}
\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
\end{align*}

```{r}
# Generate correlation matrices 
corr_grid <- expand.grid(seq(-0.01, -0.001, by = 0.001), 
                         seq(-0.6, -0.2, 0.05), seq(0.2, 0.6, 0.05))

# Function to compute correlation matrix for each combination 
gen_corr_mtx <- function(r1, r2, r3) {   
  cor_matrix <- matrix(c(1, r1, r2,                          
                         r1, 1, r3,                          
                         r2, r3, 1), nrow = 3, byrow = TRUE)          
  # Store the matrix in the list   
  if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
  }

# Plan for parallel processing
corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
sample_size <- c(50, 100, 500, 1000) 
model <- c("mortality ~ pm10 + temp", "mortality ~ pm10") |> map(as.formula)

generate_data <- function(n, mtx, seed = 123) {   
  mu <- c(0, 0, 0)   
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  set.seed(seed)      
  tibble(mortality = qpois(U[,1], 182), # assume distribution
         pm10 = qnorm(U[,2], mean = 26, sd = 12),           
         temp = qnorm(U[,3], mean = 55, sd = 16))   
  }

res <- tibble(corr_mtx = corr_mtx) |> 
  mutate(id = row_number()) |> 
  crossing(sample_size, model) |> 
  rowwise() |>   
  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx)),          
         fit = list(summary(lm(model, data))$coefficients))  
```

```{r}
#| fig.width = 10,
#| fig.height = 5
dt <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    xy_correlation = as.numeric(corr_mtx[1, 2]),     
    xz_correlation = as.numeric(corr_mtx[1, 3]),     
    yz_correlation = as.numeric(corr_mtx[2, 3]),     
    fml = deparse(model) |> as.factor(),     
    expect = ifelse(p_value < 0.05, 1, 0) |> as.factor(),     
    ) |>    
  ungroup()

code_tbl <- crossing(V1 = c("signifpm10", NA),           
                     V2 = c("smallsample", NA), 
                     V3 = c("withtemp", NA)) |>      
  mutate(col1 = paste0(V1,"_", V2, "_", V3))

lookup_tbl <- tibble(   
  value = crossing(x = c(1, 0), y = c(1, 0), z = c(1, 0)) |>      
    mutate(col = paste0(x,"_", y, "_", z)) |> 
    arrange(-x, -y, -z) |>      
    pull(col),   
  plan1 = unique(code_tbl$col1) )

dt2 <- dt |>    
  mutate(T1 = ifelse(sample_size < 200, 1, 0),          
         T2 = ifelse(fml == "mortality ~ pm10 + temp", 1, 0),          
         pp1 = paste0(expect, "_", T1, "_", T2),) |>    
  left_join(lookup_tbl, by = c("pp1" = "value"))

p0 <- dt |> ggplot() +   
  geom_point(aes(x = p_value , y = coef), color = "grey", size = 0.5) +  
  labs(x = "p-value", y = "coefficient") +   
  scale_color_brewer(palette = "Dark2", direction = -1) +   
  theme_bw() +    
  theme(aspect.ratio = 1, 
        panel.grid.minor = element_blank())

p0 + 
  geom_point(aes(x = p_value , y = coef), size = 0.5) +  
  facet_grid(as.character(model) ~ sample_size) +    
  labs(title = "Coefficient vs. p-value")
```

```{r}
#| label = "fig-result-universe",
#| fig.width = 10,
#| fig.height = 5
df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
#df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec

p1 <- p0 + 
  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.5) +  
  labs(title = "Coefficient vs. p-value") + 
  theme(legend.position = 'none')

p3 <- p0 + 
  geom_point(aes(x = p_value, y = coef, color = fml), size = 0.5) +
  labs(title = "Effect of adding temperature")

df00 <-  dt |> filter(xy_correlation == -0.005, 
                     between(xz_correlation, -0.46, -0.43),
                     yz_correlation == 0.4) |> arrange(expect)
df4 <- df00 |> filter((fml == "mortality ~ pm10 + temp"))

p5 <- p0 + 
  geom_point(data = df4, aes(x = p_value, y = coef, color = expect), size = 1) +
  geom_path(data = df4, aes(x = p_value, y = coef), color = "black",
            arrow = arrow(type = "open", angle = 30, length = unit(0.1, "inches"))) +
  scale_color_brewer(palette = "Dark2", direction = -1) +
  labs(title = "Effect of sample size")

(p1 | p3 | p5) + plot_annotation(tag_levels = "a") + 
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom') 
```

@fig-result-universe shows that result universe of the linear regression model with a) colored by whether the p-value of PM10 is significant (less than 0.05), b) highlighting the effect of adding temperature to the model for a fixed correlation structure, and c) highlighting the effect of increasing sample size.

### 2024-08-01 _manuscript/_tex/index.tex
add more content

@@
 
+(background)
+
 In this paper, a concept called analysis plan is proposed to describe
@@
 others, and to evaluate the analysis process. The concept of analysis
-plan is illustrated with examples and the implications of the concept
-for data analysis practice is discussed.
+plan is illustrated with examples. The implications of the concept for
+data analysis practice is discussed.
 
@@
 
-\section{Analysis plan}\label{analysis-plan}
+The rest of the paper is organized as follows: Section~\ref{sec-plan}
+describes the concept of analysis plan in detail.
+Section~\ref{sec-examples} provides examples of analysis plan {[}more
+details{]}. (need another section here or before examples?)
+Section~\ref{sec-conclusion} concludes the paper.
+
+\section{Analysis plan}\label{sec-plan}
 
@@
 
-\section{Examples}\label{examples}
+\section{Examples}\label{sec-examples}
+
+Three examples are presented to illustrate how the concept of analysis
+plan can be applied to data analysis. {[}toy example{]}.
+Section~\ref{sec-linear-reg} illustrates how constructing the result
+universe in a linear regression model of PM10 on mortality can help
+understand the impact of sample size, model specification, and variable
+correlation structure on data analysis. {[}example three{]}
 
@@
 
-\subsection{Linear regression}\label{linear-regression}
+\subsection{Linear regression}\label{sec-linear-reg}
 
 Consider a linear regression model to study the effect of PM10 on
-mortality (provide context of using PM10 to study mortality). From the
-literature, analysts may expect a significant PM10 coefficient in the
-linear model from the literature. This is the outcome expectation:
-p-value of PM10 coefficient is less than 0.05. There are multiple
-factors that can affect the outcome expectation of this analysis, for
-example, 1) sample size, 2) model specification, and 3) correlation
-structure between variables. Temperature is often an important
-confounder to consider in such study (add reference).
+mortality (provide context of using PM10 to study mortality). Analysts
+may expect a significant (p-value \(\le\) 0.05) PM10 coefficient in the
+linear model from the literature. This is the \emph{outcome
+expectation}. There are multiple factors that can affect the outcome
+expectation of linear regression, which here is called \emph{plan
+expectation}, for example, 1) sample size, 2) model specification, and
+3) correlation structure between variables. Adequate sample size is
+required to achieve the desired power to detect the significance of PM10
+on mortality. Temperature is often an important confounder to consider
+in such study (add reference). From some domain knowledge, an analyst
+may expect that the significance of PM10 coefficient can be attained by
+adding temperature to the model. Analysts may also expect certain
+correlation structure between PM10, temperature, and mortality, and the
+distribution of each variable.
 
-To construct the result universe, we can simulate datasets with
-different configerations of the above factors. Here, sample sizes of 50,
-100, 500, and 1000 are considered. Two model specifications are
-included: 1) linear model with PM10 as the only covariate
+To build the result universe, datasets can be simulated to either meet
+and fail these plan expectations, allowing the analysts to observe the
+significance of PM10 coefficient. Here, sample sizes of 50, 100, 500,
+and 1000 are considered. Two model specifications are included: 1)
+linear model with PM10 as the only covariate
 (\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
 temperature as covariates
-(\(\text{mortality} \sim \text{PM10} + \text{temp}\). A grid-based
+(\(\text{mortality} \sim \text{PM10} + \text{temp}\)). A grid-based
 approach is used to simulate correlation structure. Reasonable ranges of
 correlation between the three variables are
+\(\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]\),
+\(\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]\),
+and \(\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]\).
 
-\begin{align*}
-\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
-\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
-\end{align*}
+Figure~\ref{fig-result-universe} shows that result universe of the
+linear regression model and how a change of decision in one of the plan
+expectations above affect the outcome expectation. Panel a) is colored
+by the outcome expectation -- whether a significant p-value is found in
+the PM10 coefficient. Panel b) shows the effect of adding temperature to
+the model and the results show that the significance of PM10 coefficient
+can be achieved by adding temperature to the model for a sample size of
+500. Panel c) shows that increasing sample size from 50 to 100 enhances
+the significance of p-value for PM10 and the significance remains with
+further increases in sample size.
 
-\includegraphics{index_files/figure-pdf/unnamed-chunk-2-1.pdf}
+A result universe constructed in this example can be presented to
+analysts to answer the what-if questions raised in the data analysis.
+What if the sample size is increased? What if temperature is added to
+the model? What would the results expect to be changed when the
+correlation structure is different? For analysts, expectations can be
+used as unit tests to divide the result universe, which allows them to
+understand the impact of each factor on the outcome expectation.
+
+The result universe also provides a holistic view of how the results
+obtained by the analysts are situated in all possible results. This can
+be seen as a direct towards trustworthy data analysis, where the
+audience of the analysis to exercise their own cognitive model
+{[}@grolemund\_cognitive\_2014{]} to evaluate the results reported.
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
 
@@
 
-\captionsetup{labelsep=none}\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}
+\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}
 
@@
 
-\caption{\label{fig-result-universe}}
+\caption{\label{fig-result-universe}The result universe of linear
+regression model to study the effect of PM10 on mortality: a) colored by
+whether the p-value of PM10is significant (less than 0.05), b) the
+effect of adding temperature to the model for a sample size of 500, c)
+the effect of increasing sample size for a fixed correlation structure.}
 
@@
 
-Figure~\ref{fig-result-universe} shows that result universe of the
-linear regression model with a) colored by whether the p-value of PM10
-is significant (less than 0.05), b) highlighting the effect of adding
-temperature to the model for a fixed correlation structure, and c)
-highlighting the effect of increasing sample size.
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
+
+\section{Conclusion}\label{sec-conclusion}
+
+\section{References}\label{references}
 

### 2024-08-01 index.qmd
add more content

@@
 abstract: "This is the abstract"
-bibliography: references.bib
+#bibliography: references.bib
 documentclass: jds
-format: pdf
+format: 
+  pdf: 
+   keep-tex: true
 preamble: >
@@
 
+(background)
+
 In this paper, a concept called analysis plan is proposed to describe the logical structure of a data analysis.
@@
 It is a formal representation of the analysis process and can be used to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
-The concept of analysis plan is illustrated with examples and the implications of the concept for data analysis practice is discussed.
+The concept of analysis plan is illustrated with examples.
+The implications of the concept for data analysis practice is discussed.
 
@@
 
-# Analysis plan
+The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail.
+@sec-examples provides examples of analysis plan [more details].
+(need another section here or before examples?) @sec-conclusion concludes the paper.
+
+# Analysis plan {#sec-plan}
 
@@
 
-# Examples
+# Examples {#sec-examples}
+
+Three examples are presented to illustrate how the concept of analysis plan can be applied to data analysis.
+[toy example].
+@sec-linear-reg illustrates how constructing  the result universe in a linear regression model of PM10 on mortality can help understand the impact of sample size, model specification, and variable correlation structure on data analysis.
+[example three]
 
@@
 
-## Linear regression
+## Linear regression {#sec-linear-reg}
 
@@
 Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature.
-There are multiple factors that can affect the outcome expectation of linear regression as such, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables.
+This is the *outcome expectation*.
+There are multiple factors that can affect the outcome expectation of linear regression, which here is called *plan expectation*, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables.
+Adequate sample size is required to achieve the desired power to detect the significance of PM10 on mortality.
 Temperature is often an important confounder to consider in such study (add reference).
+From some domain knowledge, an analyst may expect that the significance of PM10 coefficient can be attained by adding temperature to the model.
+Analysts may also expect certain correlation structure between PM10, temperature, and mortality, and the distribution of each variable.
 
-To construct the result universe, we can simulate datasets with different configerations of the above factors.
+To build the result universe, datasets can be simulated to either meet and fail these plan expectations, allowing the analysts to observe the significance of PM10 coefficient.
 Here, sample sizes of 50, 100, 500, and 1000 are considered.
-Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$.
+Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$).
 A grid-based approach is used to simulate correlation structure.
-Reasonable ranges of correlation between the three variables are
+Reasonable ranges of correlation between the three variables are $\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]$, $\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]$, and $\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]$.
 
-\begin{align*}
-\text{cor}(\text{mortality}, \text{PM10}) &\in [-0.01, 0] \\ 
-\text{cor}(\text{mortality}, \text{temperature}) &\in [-0.6, -0.2] \\ \text{cor}(\text{PM10}, \text{temperature}) &\in [0.2, 0.6] 
-\end{align*}
+@fig-result-universe shows that result universe of the linear regression model and how a change of decision in one of the plan expectations above affect the outcome expectation.
+Panel a) is colored by the outcome expectation -- whether a significant p-value is found in the PM10 coefficient.
+Panel b) shows the effect of adding temperature to the model and the results show that the significance of PM10 coefficient can be achieved by adding temperature to the model for a sample size of 500.
+Panel c) shows that increasing sample size from 50 to 100 enhances the significance of p-value for PM10 and the significance remains with further increases in sample size.
 
-```{r}
+A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis.
+What if the sample size is increased?
+What if temperature is added to the model?
+What would the results expect to be changed when the correlation structure is different?
+For analysts, expectations can be used as unit tests to divide the result universe, which allows them to understand the impact of each factor on the outcome expectation.
+
+The result universe also provides a holistic view of how the results obtained by the analysts are situated in all possible results.
+This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.
+
+```{r sim-linear-res}
 # Generate correlation matrices 
@@
 corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
-  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
+  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) 
+  }) 
 corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
@@
   tibble(mortality = qpois(U[,1], 182), # assume distribution
-         pm10 = qnorm(U[,2], mean = 26, sd = 12),           
+         pm10 = qgamma(U[,2], shape = 6, rate = 0.2),           
          temp = qnorm(U[,3], mean = 55, sd = 16))   
   }
+
+# test for dist fit
+# fitdist(pm10$pm10, "gamma") -> pm10_fit
+# plot(pm10_fit) # gamma is better than normal for the fit of pm10
+#fitdist(pm10$temp, "norm") -> temp_fit
+# plot(temp_fit)
 
@@
 ```{r}
-#| fig.width = 10,
-#| fig.height = 5
+#| label: fig-result-universe
+#| fig-cap: "The result universe of linear regression model to study the effect of PM10 on mortality: a) colored by whether the p-value of PM10is significant (less than 0.05), b) the effect of adding temperature to the model for a sample size of 500, c) the effect of increasing sample size for a fixed correlation structure."
+#| fig.width: 12
+#| fig.height: 12
 dt <- res |>   
@@
   labs(x = "p-value", y = "coefficient") +   
-  scale_color_brewer(palette = "Dark2", direction = -1) +   
   theme_bw() +    
@@
 
-p0 + 
-  geom_point(aes(x = p_value , y = coef), size = 0.5) +  
-  facet_grid(as.character(model) ~ sample_size) +    
-  labs(title = "Coefficient vs. p-value")
-```
-
-```{r}
-#| label = "fig-result-universe",
-#| fig.width = 10,
-#| fig.height = 5
 df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
@@
                      yz_correlation == 0.4) |> arrange(expect)
-#df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec
+df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec
 
 p1 <- p0 + 
-  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.5) +  
+  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.3) +  
+  scale_color_brewer(palette = "Dark2", direction = -1) +
   labs(title = "Coefficient vs. p-value") + 
@@
 
-p3 <- p0 + 
-  geom_point(aes(x = p_value, y = coef, color = fml), size = 0.5) +
+p2 <- p0 + 
+  geom_point(data = df0, aes(x = p_value, y = coef, color = expect), size = 2) +
+  geom_path(data = df0, aes(x = p_value, y = coef), color = "black",
+            arrow = arrow(type = "open", angle = 30, 
+                          length = unit(0.1, "inches"))) +
+  ggrepel::geom_label_repel(
+    data = df0, aes(x = p_value, y = coef, label = fml, color = expect)
+    ) +
+  scale_color_brewer(palette = "Dark2", direction = -1) +
   labs(title = "Effect of adding temperature")
@@
                      yz_correlation == 0.4) |> arrange(expect)
-df4 <- df00 |> filter((fml == "mortality ~ pm10 + temp"))
+df <- df00 |> filter((fml == "mortality ~ pm10 + temp"))
 
-p5 <- p0 + 
-  geom_point(data = df4, aes(x = p_value, y = coef, color = expect), size = 1) +
-  geom_path(data = df4, aes(x = p_value, y = coef), color = "black",
-            arrow = arrow(type = "open", angle = 30, length = unit(0.1, "inches"))) +
+p3 <- p0 + 
+  geom_point(data = df, aes(x = p_value, y = coef, color = expect), size = 2) +
+  geom_path(data = df, aes(x = p_value, y = coef), color = "black",
+            arrow = arrow(type = "open", angle = 30, 
+                          length = unit(0.1, "inches"))) +
+  ggrepel::geom_label_repel(
+    data = df, aes(x = p_value, y = coef, label = sample_size, color = expect)
+    ) +
   scale_color_brewer(palette = "Dark2", direction = -1) +
@@
 
-(p1 | p3 | p5) + plot_annotation(tag_levels = "a") + 
+(p1 | p2) / (p3 | plot_spacer()) + 
+  plot_annotation(tag_levels = "a") + 
   plot_layout(guides = 'collect') &
@@
 
-@fig-result-universe shows that result universe of the linear regression model with a) colored by whether the p-value of PM10 is significant (less than 0.05), b) highlighting the effect of adding temperature to the model for a fixed correlation structure, and c) highlighting the effect of increasing sample size.
+# Conclusion {#sec-conclusion}
+
+# References

### 2024-08-01 _manuscript/_tex/index.tex
notes after the meeting

@@
 \fi
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
 \usepackage{bookmark}
@@
 
-(background)
+{[}background{]}
 
@@
 
-describe/ define what analysis plan is
-
-analysis plan and outcome plan
-
-analysis plan as unit tests to divide the ``result universe'', which
-allows us to answer questions like:
+\begin{itemize}
+\item
+  describe/ define what analysis plan is, expectation (outcome and plan)
+\item
+  why it is useful to be explicit about expectation?
 
@@
   \item
-  how would the results change if the value of a unit test change
-\item
-  whether our outcome expectation aligns with the plan expectation,
-  meaning whether ``it is possible for the combination of plan
-  expectations to produce the outcome expectation''
+    expectations can be used to formulate unit tests, which helps to
+    divide the ``result universe''
+  \end{itemize}
 \end{itemize}
@@
 
+\begin{itemize}
+\item
+  add a paragraph to describe the simulation process
+\item
+  add a fourth panel to describe the comparison of a right/ wrong
+  expectation, i.e.~correlation on PM10 and mortality
+\end{itemize}
+
 Figure~\ref{fig-result-universe} shows that result universe of the
@@
 the significance of p-value for PM10 and the significance remains with
-further increases in sample size.
+further increases in sample size. {[}note: weave the ``actual data''
+into the example linear regression model{]}
 
-A result universe constructed in this example can be presented to
-analysts to answer the what-if questions raised in the data analysis.
-What if the sample size is increased? What if temperature is added to
-the model? What would the results expect to be changed when the
-correlation structure is different? For analysts, expectations can be
-used as unit tests to divide the result universe, which allows them to
-understand the impact of each factor on the outcome expectation.
+(these two paragraphs may go to a new section \emph{discussion}) A
+result universe constructed in this example can be presented to analysts
+to answer the what-if questions raised in the data analysis. What if the
+sample size is increased? What if temperature is added to the model?
+What would the results expect to be changed when the correlation
+structure is different? For analysts, expectations can be used as unit
+tests to divide the result universe, which allows them to understand the
+impact of each factor on the outcome expectation.
 
@@
 audience of the analysis to exercise their own cognitive model
-{[}@grolemund\_cognitive\_2014{]} to evaluate the results reported.
+\citep{grolemund_cognitive_2014} to evaluate the results reported.
 
@@
 
-\section{References}\label{references}
+\renewcommand\refname{References}
+  \bibliography{references.bib}
 

### 2024-08-01 index.qmd
notes after the meeting

@@
 abstract: "This is the abstract"
-#bibliography: references.bib
+cite-method: natbib
+bibliography: references.bib
 documentclass: jds
@@
 
-(background)
+[background]
 
@@
 
-describe/ define what analysis plan is
+* describe/ define what analysis plan is, expectation (outcome and plan)
 
-analysis plan and outcome plan
+* why it is useful to be explicit about expectation?
 
-analysis plan as unit tests to divide the "result universe", which allows us to answer questions like:
-
-- how would the results change if the value of a unit test change
-- whether our outcome expectation aligns with the plan expectation, meaning whether "it is possible for the combination of plan expectations to produce the outcome expectation"
+    * expectations can be used to formulate unit tests, which helps to divide the "result universe"
 
@@
 
+ * add a paragraph to describe the simulation process
+
+ * add a fourth panel to describe the comparison of a right/ wrong expectation, i.e. correlation on PM10 and mortality
+
 @fig-result-universe shows that result universe of the linear regression model and how a change of decision in one of the plan expectations above affect the outcome expectation.
@@
 Panel c) shows that increasing sample size from 50 to 100 enhances the significance of p-value for PM10 and the significance remains with further increases in sample size.
+[note: weave the "actual data" into the example linear regression model]
 
-A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis.
+(these two paragraphs may go to a new section *discussion*) A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis.
 What if the sample size is increased?

### 2024-09-12 _manuscript/_tex/index.tex
the toy example + some notesg

@@
 % Make \paragraph and \subparagraph free-standing
+\makeatletter
 \ifx\paragraph\undefined\else
   \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
 \fi
@@
   \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
 \fi
+\makeatother
 
@@
 \makeatother
+
 \ifLuaTeX
@@
 
-{[}background{]}
+{[}background - In a data analysis workflow{]}
+
+For expert analysts, the expectation of the analysis process is often
+implicit. They may have a mental model of the analysis process and the
+expected results. However, for non-expert analysts, the expectation of
+the analysis process may not be clear. They may not know what to expect
+from the analysis process, or what to look for when an analysis process
+fails the expectation. An analysis plan can help to make the expectation
+of the analysis process explicit. It can help to guide the analysis
+process, to communicate and compare the analysis process to others, and
+to evaluate the analysis process.
 
@@
 \begin{itemize}
+\tightlist
 \item
   describe/ define what analysis plan is, expectation (outcome and plan)
+
+  \begin{itemize}
+  \tightlist
+  \item
+    expectations can be used to formulate unit tests, they can be used
+    to guide to diagnose in the analysis process: data issue,
+    unreasonable expectation -- expectation is ``wrong''
+  \item
+    some expectations can be about data: e.g.~check temporal gaps in
+    time series data. Some others are statistical checks: sample size,
+    outliers, model specification, or checks based on domain knowledge:
+    include temperature in the mortality \textasciitilde{} PM10 model.
+  \end{itemize}
+\end{itemize}
+
+Q: Whether we should formulate these concept with math notation?? A:
+only if it helps
+
+An analysis plan is a set of analysis steps plus their expected
+outcomes. An analysis plan consists of two parts: outcome expectation
+and plan expectation. Outcome expectation is the expected result of the
+analysis, for example, a significant p-value or a coefficient estimate
+within a certain range. Plan expectation is the expected results from
+the intermediate steps of the analysis, for example, the expected
+distribution of the data, or the expected correlation structure between
+variables.
+
+While there is typically one major result expectation in an analysis,
+multiple plan expectations can be formulated as unit tests to guide the
+analysis process. These plan expectations serve as road marks to test,
+at various stages of the analysis, whether the results are within
+expectation. When the outcome is out of expectation, these plan
+expectations can serve as unit tests to diagnose the analysis process.
+
+\begin{itemize}
+\tightlist
 \item
@@
   \item
-    expectations can be used to formulate unit tests, which helps to
-    divide the ``result universe''
+    with simulation, the unit tests can be used to construct the
+    ``result universe''
+  \item
+    With formulated expectation, one can generate multiple version of
+    the dataset for teaching statistics
   \end{itemize}
@@
 
-\section{Examples}\label{sec-examples}
+When the expectation is formulated, we can also construct values fulfill
+or not fulfill the expectation. With simulation, this allows to
+construct the set of all possible results that can be obtained from the
+analysis process, which we called the result universe. With such a
+result universe, we can see how the results obtained by the analysts are
+situated in all possible results. When accompanied by the result
+universe, it is clearer how different factors in the analysis affects
+the main results. This can be seen as a direct towards trustworthy data
+analysis, where the audience of the analysis to exercise their own
+cognitive model \citep{grolemund_cognitive_2014} to evaluate the results
+reported.
+
+\subsection{A toy example}\label{a-toy-example}
+
+Let's think about a 5-day step count. You make a resolution to walk on
+average 5000 steps a day (your expectation) and using an app to record
+your step count. After 5 days, the app tells you've walked on average
+8000 steps.
+
+It is easy to come up with reasons why an 8000 average step is resulted
+based on common sense:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  you may run a 10k on day 1, resulting a high step count on the day
+  (outlier on the right).
+\item
+  you left your phone at home on day 3, resulting a zero or minimal step
+  count on the day (outlier on the left).
+\item
+  you may realise the step count may increase since you were in a hiking
+  trip in the last five days (average shift).
+\end{enumerate}
+
+Based on these reasons, you may devise a set of unit tests to check the
+step count data, i.e.~check the maximum and minimum step count, check
+the difference between each day.
+
+\begin{itemize}
+\tightlist
+\item
+  If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the
+  maximum check will flag the data for investigate the maximum. The
+  difference between days test will also flag the data
+\item
+  If the daily count looks like c(20000, 20000, 20000, 20000, 20000),
+  the maximum check will flag the data for investigate the maximum.
+\end{itemize}
+
+Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is
+flagged by the minimal tests but won't cause an average of 8000 average
+step.
+
+The statistical procedure of averaging 5 numbers ``around 5000'' to get
+a mean of 5000 is \emph{consistent} meaning if all the numbers are
+around 5000, we are guaranteed to get a mean around 5000. We could
+devise 5 unit tests to check each number. Since you're more familiar
+with your daily life, you may realise the step count may increase since
+you were in a hiking trip in the last two days. This may prompt you to
+check the step count.
+
+In a data analysis, it is not practical to check every entry of the
+data, a similar strategy of devising tests to check for
+
+\begin{itemize}
+\tightlist
+\item
+  The combination of unit tests are not unique
+\item
+  The unit tests provide guidance for diagnosing the results, but are
+  not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the
+  max diff test but receive a within expectation mean.
+\end{itemize}
+
+\section{Applications}\label{sec-examples}
 
@@
 correlation structure on data analysis. {[}example three{]}
-
-\subsection{A toy example}\label{a-toy-example}
 
@@
 
-(these two paragraphs may go to a new section \emph{discussion}) A
-result universe constructed in this example can be presented to analysts
-to answer the what-if questions raised in the data analysis. What if the
-sample size is increased? What if temperature is added to the model?
-What would the results expect to be changed when the correlation
-structure is different? For analysts, expectations can be used as unit
-tests to divide the result universe, which allows them to understand the
-impact of each factor on the outcome expectation.
-
-The result universe also provides a holistic view of how the results
-obtained by the analysts are situated in all possible results. This can
-be seen as a direct towards trustworthy data analysis, where the
-audience of the analysis to exercise their own cognitive model
-\citep{grolemund_cognitive_2014} to evaluate the results reported.
-
 \textsubscript{Source:

### 2024-09-12 index.qmd
the toy example + some notesg

@@
 
-[background]
+[background - In a data analysis workflow]
+
+For expert analysts, the expectation of the analysis process is often implicit.
+They may have a mental model of the analysis process and the expected results.
+However, for non-expert analysts, the expectation of the analysis process may not be clear.
+They may not know what to expect from the analysis process, or what to look for when an analysis process fails the expectation.
+An analysis plan can help to make the expectation of the analysis process explicit.
+It can help to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
 
@@
 * describe/ define what analysis plan is, expectation (outcome and plan)
+    * expectations can be used to formulate unit tests, they can be used to guide to diagnose in the analysis process: data issue, unreasonable expectation -- expectation is "wrong" 
+    * some expectations can be about data: e.g.
+check temporal gaps in time series data.
+Some others are statistical checks: sample size, outliers, model specification, or checks based on domain knowledge: include temperature in the mortality ~ PM10 model.
+
+Q: Whether we should formulate these concept with math notation?? A: only if it helps
+
+An analysis plan is a set of analysis steps plus their expected outcomes.
+An analysis plan consists of two parts: outcome expectation and plan expectation.
+Outcome expectation is the expected result of the analysis, for example, a significant p-value or a coefficient estimate within a certain range.
+Plan expectation is the expected results from the intermediate steps of the analysis, for example, the expected distribution of the data, or the expected correlation structure between variables.
+
+While there is typically one major result expectation in an analysis, multiple plan expectations can be formulated as unit tests to guide the analysis process.
+These plan expectations serve as road marks to test, at various stages of the analysis, whether the results are within expectation.
+When the outcome is out of expectation, these plan expectations can serve as unit tests to diagnose the analysis process.
 
 * why it is useful to be explicit about expectation? 
+    * with simulation, the unit tests can be used to construct the "result universe" 
+    * With formulated expectation, one can generate multiple version of the dataset for teaching statistics
 
-    * expectations can be used to formulate unit tests, which helps to divide the "result universe"
+When the expectation is formulated, we can also construct values fulfill or not fulfill the expectation.
+With simulation, this allows to construct the set of all possible results that can be obtained from the analysis process, which we called the result universe.
+With such a result universe, we can see how the results obtained by the analysts are situated in all possible results.
+When accompanied by the result universe, it is clearer how different factors in the analysis affects the main results.
+This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.
 
-# Examples {#sec-examples}
+## A toy example
+
+Let's think about a 5-day step count.
+You make a resolution to walk on average 5000 steps a day (your expectation) and using an app to record your step count.
+After 5 days, the app tells you've walked on average 8000 steps.
+
+It is easy to come up with reasons why an 8000 average step is resulted based on common sense:
+
+1. you may run a 10k on day 1, resulting a high step count on the day (outlier on the right).
+2. you left your phone at home on day 3, resulting a zero or minimal step count on the day (outlier on the left).
+3. you may realise the step count may increase since you were in a hiking trip in the last five days (average shift).
+
+Based on these reasons, you may devise a set of unit tests to check the step count data, i.e. check the maximum and minimum step count, check the difference between each day.
+
+- If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the maximum check will flag the data for investigate the maximum. The difference between days test will also flag the data
+- If the daily count looks like c(20000, 20000, 20000, 20000, 20000), the maximum check will flag the data for investigate the maximum.
+
+Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is flagged by the minimal tests but won't cause an average of 8000 average step.
+
+The statistical procedure of averaging 5 numbers "around 5000" to get a mean of 5000 is *consistent* meaning if all the numbers are around 5000, we are guaranteed to get a mean around 5000.
+We could devise 5 unit tests to check each number.
+Since you're more familiar with your daily life, you may realise the step count may increase since you were in a hiking trip in the last two days.
+This may prompt you to check the step count.
+
+In a data analysis, it is not practical to check every entry of the data, a similar strategy of devising tests to check for
+
+- The combination of unit tests are not unique
+- The unit tests provide guidance for diagnosing the results, but are not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the max diff test but receive a within expectation mean.
+
+# Applications {#sec-examples}
 
@@
 [example three]
-
-## A toy example
 
@@
 [note: weave the "actual data" into the example linear regression model]
-
-(these two paragraphs may go to a new section *discussion*) A result universe constructed in this example can be presented to analysts to answer the what-if questions raised in the data analysis.
-What if the sample size is increased?
-What if temperature is added to the model?
-What would the results expect to be changed when the correlation structure is different?
-For analysts, expectations can be used as unit tests to divide the result universe, which allows them to understand the impact of each factor on the outcome expectation.
-
-The result universe also provides a holistic view of how the results obtained by the analysts are situated in all possible results.
-This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.
 

### 2024-09-13 _manuscript/_tex/index.tex
more definition and explanation on expectations and the current gap

@@
 
-\begin{itemize}
-\tightlist
-\item
-  describe/ define what analysis plan is, expectation (outcome and plan)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    expectations can be used to formulate unit tests, they can be used
-    to guide to diagnose in the analysis process: data issue,
-    unreasonable expectation -- expectation is ``wrong''
-  \item
-    some expectations can be about data: e.g.~check temporal gaps in
-    time series data. Some others are statistical checks: sample size,
-    outliers, model specification, or checks based on domain knowledge:
-    include temperature in the mortality \textasciitilde{} PM10 model.
-  \end{itemize}
-\end{itemize}
-
 Q: Whether we should formulate these concept with math notation?? A:
@@
 
-An analysis plan is a set of analysis steps plus their expected
-outcomes. An analysis plan consists of two parts: outcome expectation
-and plan expectation. Outcome expectation is the expected result of the
-analysis, for example, a significant p-value or a coefficient estimate
-within a certain range. Plan expectation is the expected results from
-the intermediate steps of the analysis, for example, the expected
-distribution of the data, or the expected correlation structure between
-variables.
+TODO: more about analysis plan itself here
 
-While there is typically one major result expectation in an analysis,
-multiple plan expectations can be formulated as unit tests to guide the
-analysis process. These plan expectations serve as road marks to test,
-at various stages of the analysis, whether the results are within
-expectation. When the outcome is out of expectation, these plan
-expectations can serve as unit tests to diagnose the analysis process.
+TODO: what if the expectation is ``wrong''
+
+An analysis plan is a set of analysis steps combined with expectations.
+Expectations represent our belief about certain aspects of the analysis,
+independent of the analysis itself. It can be divided into two types:
+\emph{outcome expectation} and \emph{plan expectation}. Outcome
+expectation refers to what we anticipate from the main result of the
+analysis based on prior knowledge. They shape how we interpret the
+results and assess whether they are consistent with existing knowledge
+or indicate the need for updates \citep{grolemund_cognitive_2014}. For
+example, in public health, prior research shows the average increase in
+mortality rate per unit increase in PM10 is about 0.50\%
+\citep{liu2019ambient}. This serves as an expectation for similar future
+studies. Plan expectations concern the intermediate steps within the
+analysis rather than the final outcome. They serve as checkpoints to
+detect deflection in the analysis process For example, we may expect
+temporal data to be ordered with no gaps and duplicates, or expect that
+temperature will be a significant covariate in the linear regression
+model of PM10 on mortality.
+
+(might be useful) Analysis plans can be constructed at various
+granularities, at the highest level, one may only has a plan of the
+specific method used for analysing data and the expected outcome. This
+provides little guidance when a deviation from expectation occurs. At
+the lowest level, one may have a plan for each data entry and every data
+handling steps. This provides too much detail and may not be practical
+in practice.
+
+Experienced analysts often have implicit expectation about the outcome
+and rely on a few ``directional signs'' to check when the outcome
+deviate from those expectation. However, these expectations are rarely
+made explicit within the analysis workflow. This makes it challenging
+for consumers of the analysis to evaluate the results, since it becomes
+difficult to disentangle whether discrepancies arise from differing
+expectations or from the use of statistical technique, without running
+the analysis themselves. Non-expert analysts, lacking prior knowledge or
+instinct, may not have clear expectations of the results. This can lead
+to reduced confidence of the analysis and makes it more difficult and
+time-consuming to diagnose the cause of the deviation when the results
+don't align with expectations.
+
+By explicitly formulating these expectations, an analysis plan can guide
+the analysis process, facilitate the communication and comparison with
+other analyses and evaluate the validity of the results.
 
@@
   \tightlist
+  \item
+    and can be used to guide the analysis process
   \item

### 2024-09-13 index.qmd
more definition and explanation on expectations and the current gap

@@
 
-* describe/ define what analysis plan is, expectation (outcome and plan)
-    * expectations can be used to formulate unit tests, they can be used to guide to diagnose in the analysis process: data issue, unreasonable expectation -- expectation is "wrong" 
-    * some expectations can be about data: e.g.
-check temporal gaps in time series data.
-Some others are statistical checks: sample size, outliers, model specification, or checks based on domain knowledge: include temperature in the mortality ~ PM10 model.
-
 Q: Whether we should formulate these concept with math notation?? A: only if it helps
 
-An analysis plan is a set of analysis steps plus their expected outcomes.
-An analysis plan consists of two parts: outcome expectation and plan expectation.
-Outcome expectation is the expected result of the analysis, for example, a significant p-value or a coefficient estimate within a certain range.
-Plan expectation is the expected results from the intermediate steps of the analysis, for example, the expected distribution of the data, or the expected correlation structure between variables.
+TODO: more about analysis plan itself here
 
-While there is typically one major result expectation in an analysis, multiple plan expectations can be formulated as unit tests to guide the analysis process.
-These plan expectations serve as road marks to test, at various stages of the analysis, whether the results are within expectation.
-When the outcome is out of expectation, these plan expectations can serve as unit tests to diagnose the analysis process.
+TODO: what if the expectation is "wrong"
+
+An analysis plan is a set of analysis steps combined with expectations.
+Expectations represent our belief about certain aspects of the analysis, independent of the analysis itself.
+It can be divided into two types: *outcome expectation* and *plan expectation*.
+Outcome expectation refers to what we anticipate from the main result of the analysis based on prior knowledge.
+They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates [@grolemund_cognitive_2014].
+For example, in public health, prior research shows the average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient].
+This serves as an expectation for similar future studies.
+Plan expectations concern the intermediate steps within the analysis rather than the final outcome.
+They serve as checkpoints to detect deflection in the analysis process For example, we may expect temporal data to be ordered with no gaps and duplicates, or expect that temperature will be a significant covariate in the linear regression model of PM10 on mortality.
+
+(might be useful) Analysis plans can be constructed at various granularities, at the highest level, one may only has a plan of the specific method used for analysing data and the expected outcome.
+This provides little guidance when a deviation from expectation occurs.
+At the lowest level, one may have a plan for each data entry and every data handling steps.
+This provides too much detail and may not be practical in practice.
+
+Experienced analysts often have implicit expectation about the outcome and rely on a few "directional signs" to check when the outcome deviate from those expectation.
+However, these expectations are rarely made explicit within the analysis workflow.
+This makes it challenging for consumers of  the analysis to evaluate the results, since it becomes difficult to disentangle whether discrepancies arise from differing expectations or from the use of statistical technique, without running the analysis themselves.
+Non-expert analysts, lacking prior knowledge or instinct, may not have clear expectations of the results.
+This can lead to reduced confidence of the analysis and makes it more difficult and time-consuming to diagnose the cause of the deviation when the results don't align with expectations.
+
+By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and comparison with other analyses and evaluate the validity of the results.
 
 * why it is useful to be explicit about expectation? 
+    * and can be used to guide the analysis process
     * with simulation, the unit tests can be used to construct the "result universe" 

### 2024-09-13 _manuscript/_tex/index.tex
explain result universeg

@@
 
+The whole point being how can be build trust in other's analysis. Some
+provides scripts to run the analysis but that's not enough. We need to
+know what to expect from the analysis and that is often not explicitly
+specified.
+
 An analysis plan is a set of analysis steps combined with expectations.
@@
 time-consuming to diagnose the cause of the deviation when the results
-don't align with expectations.
+don't align with expectations. By explicitly formulating these
+expectations, an analysis plan can guide the analysis process,
+facilitate the communication and evaluate the validity of the results.
 
-By explicitly formulating these expectations, an analysis plan can guide
-the analysis process, facilitate the communication and comparison with
-other analyses and evaluate the validity of the results.
+The expectations can be thought of as a set of unit tests used to
+validate the results of data analysis. By specifying a range of values
+for these tests, multiple versions of the dataset can be generated to
+satisfy different sets of plan expectations. This allows us to present
+what we called the ``result universe'' -- the complete set of possible
+results that can be obtained from one data analysis process. By
+visualizing the result universe, data analysis consumers can observe how
+changes in expectations affect the results and the range of alternative
+outcomes that could arise under different conditions. This enables them
+to evaluate the outcomes based on their own plan expectations and gain a
+broader perspective on how the actual results produced by analysts fit
+within this spectrum of possibilities, promoting transparency and trust
+in analysis.
 
-\begin{itemize}
-\tightlist
-\item
-  why it is useful to be explicit about expectation?
-
-  \begin{itemize}
-  \tightlist
-  \item
-    and can be used to guide the analysis process
-  \item
-    with simulation, the unit tests can be used to construct the
-    ``result universe''
-  \item
-    With formulated expectation, one can generate multiple version of
-    the dataset for teaching statistics
-  \end{itemize}
-\end{itemize}
-
-When the expectation is formulated, we can also construct values fulfill
-or not fulfill the expectation. With simulation, this allows to
-construct the set of all possible results that can be obtained from the
-analysis process, which we called the result universe. With such a
-result universe, we can see how the results obtained by the analysts are
-situated in all possible results. When accompanied by the result
-universe, it is clearer how different factors in the analysis affects
-the main results. This can be seen as a direct towards trustworthy data
-analysis, where the audience of the analysis to exercise their own
-cognitive model \citep{grolemund_cognitive_2014} to evaluate the results
-reported.
+Furthermore, by generating multiple versions of the data, we can emulate
+various scenarios within the same context for students to exercise
+judgement when conducting data analysis in a classroom setting.
 

### 2024-09-13 index.qmd
explain result universeg

@@
 
+The whole point being how can be build trust in other's analysis.
+Some provides scripts to run the analysis but that's not enough.
+We need to know what to expect from the analysis and that is often not explicitly specified.
+
 An analysis plan is a set of analysis steps combined with expectations.
@@
 This can lead to reduced confidence of the analysis and makes it more difficult and time-consuming to diagnose the cause of the deviation when the results don't align with expectations.
+By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and evaluate the validity of the results.
 
-By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and comparison with other analyses and evaluate the validity of the results.
+The expectations can be thought of as a set of unit tests used to validate the results of data analysis.
+By specifying a range of values for these tests, multiple versions of the dataset can be generated to satisfy different sets of plan expectations.
+This allows us to present what we called the "result universe" -- the complete set of possible results that can be obtained from one data analysis process.
+By visualizing the result universe, data analysis consumers can observe how changes in expectations affect the results and the range of alternative outcomes that could arise under different conditions.
+This enables them to evaluate the outcomes based on their own plan expectations and gain a broader perspective on how the actual results produced by analysts fit within this spectrum of possibilities, promoting transparency and trust in analysis.
 
-* why it is useful to be explicit about expectation? 
-    * and can be used to guide the analysis process
-    * with simulation, the unit tests can be used to construct the "result universe" 
-    * With formulated expectation, one can generate multiple version of the dataset for teaching statistics
-
-When the expectation is formulated, we can also construct values fulfill or not fulfill the expectation.
-With simulation, this allows to construct the set of all possible results that can be obtained from the analysis process, which we called the result universe.
-With such a result universe, we can see how the results obtained by the analysts are situated in all possible results.
-When accompanied by the result universe, it is clearer how different factors in the analysis affects the main results.
-This can be seen as a direct towards trustworthy data analysis, where the audience of the analysis to exercise their own cognitive model [@grolemund_cognitive_2014] to evaluate the results reported.
+Furthermore, by generating multiple versions of the data, we can emulate various scenarios within the same context for students to exercise judgement when conducting data analysis in a classroom setting.
 

### 2024-10-17 _manuscript/_tex/index.tex
add some prompts to the paper

@@
 {[}background - In a data analysis workflow{]}
+
+In a data analysis workflow -\textgreater{} going back to diagnose the
+descripency between the expectation and the result is often
+time-consuming and challenging.
+
+For consumers of data analysis, it is often difficult to evaluate the
+analysis outcome. Not always have the script to run, even difficult when
+different expectations are held.
+
+We propose here a concept called analysis plan to make the expectation
+of the analysis process explicit. It can help to guide the analysis
+process, to communicate and compare the analysis process to others, and
+to evaluate the analysis process. (think more closer here how our
+approach help to solve the problem)
 
@@
 
+\section{Method}\label{method}
+
+\subsection{Toy example revisited}\label{toy-example-revisited}
+
+\begin{itemize}
+\item
+  provide interpretation at different scenarios:
+
+  \begin{enumerate}
+  \def\labelenumi{\arabic{enumi})}
+  \tightlist
+  \item
+    one test is flagged, the prediction is as expected:
+  \item
+    multiple tests are flagged, the prediction is unexpected,
+  \item
+    no test is flagged, the prediction is unexpected,
+  \item
+    no test is flagged, the prediction is as expected
+  \end{enumerate}
+\end{itemize}
+
 \section{Applications}\label{sec-examples}
@@
 
+\section{Discussion}\label{discussion}
+
+\begin{itemize}
+\item
+  how to systematically simulate data is still unknown, sensitivity of
+  the simulation to the resutls
+\item
+  currently no automated way to generate unit tests
+\end{itemize}
+
 \section{Conclusion}\label{sec-conclusion}

### 2024-10-17 index.qmd
add some prompts to the paper

@@
 [background - In a data analysis workflow]
+
+In a data analysis workflow -> going back to diagnose the descripency between the expectation and the result is often time-consuming and challenging.
+
+For consumers of data analysis, it is often difficult to evaluate the analysis outcome. Not always have the script to run, even difficult when different expectations are held.
+
+We propose here a concept called analysis plan to make the expectation of the analysis process explicit.
+It can help to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
+(think more closer here how our approach help to solve the problem)
 
@@
 - The unit tests provide guidance for diagnosing the results, but are not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the max diff test but receive a within expectation mean.
+
+# Method
+
+## Toy example revisited
+
+* provide interpretation at different scenarios:
+
+  1) one test is flagged, the prediction is as expected: 
+  2) multiple tests are flagged, the prediction is unexpected, 
+  3) no test is flagged, the prediction is unexpected, 
+  4) no test is flagged, the prediction is as expected
 
@@
 
+# Discussion
+
+* how to systematically simulate data is still unknown, sensitivity of the simulation to the resutls
+
+* currently no automated way to generate unit tests
+
 # Conclusion {#sec-conclusion}

### 2024-10-24 _manuscript/_tex/index.tex
some dot points

@@
 
+\begin{itemize}
+\item
+  multiple tests can be generated to diagnose different aspects of an
+  analysis
+\item
+  good tests
+
+  \begin{enumerate}
+  \def\labelenumi{\arabic{enumi})}
+  \tightlist
+  \item
+    are responsive to an unexpected result - accuracy,
+  \item
+    motivate actions of analysts to investigate the results. This points
+    towards a smaller set of independent tests
+  \end{enumerate}
+\item
+  Since both the plan/ outcome expectations can be phrased as unit
+  tests, which output binary outcomes, consider using the plan
+  expectations as predictors of the outcome expectations. This allows to
+  generate the confusion matrix of predicting the outcome with the plan
+  expectation.
+\item
+  For an analysis, ideally, good plan expectations should maximize the
+  detection of unexpected outcomes while minimize the false positive
+  discovery, which suggest the use of precision and recall (REF) for
+  evaluating the performance of the plan expectations.
+
+  \begin{itemize}
+  \tightlist
+  \item
+    precision: the proportion of unexpected results (TP) out of all the
+    predicted unexpected results (TP + FP)
+  \item
+    recall: the proportion of unexpected results (TP) out of all the
+    actual unexpected results (TP + FN)
+  \end{itemize}
+\item
+  A logic regression (ref) is used to model the relationship between the
+  plan and outcome expectations. (justify the use of Logic regression)
+\item
+  on independence of the tests
+\end{itemize}
+
 \subsection{Toy example revisited}\label{toy-example-revisited}

### 2024-10-24 index.qmd
some dot points

@@
 # Method
+
+- multiple tests can be generated to diagnose different aspects of an analysis 
+- good tests 
+    1) are responsive to an unexpected result - accuracy, 
+    2) motivate actions of analysts to investigate the results. This points towards a smaller set of independent tests
+
+- Since both the plan/ outcome expectations can be phrased as unit tests, which output binary outcomes, consider using the plan expectations as predictors of the outcome expectations.
+This allows to generate the confusion matrix of predicting the outcome with the plan expectation.
+
+- For an analysis, ideally, good plan expectations should maximize the detection of unexpected outcomes while minimize the false positive discovery, which suggest the use of precision and recall (REF) for evaluating the performance of the plan expectations.
+
+  * precision: the proportion of unexpected results (TP) out of all the predicted unexpected results (TP + FP)
+  * recall: the proportion of unexpected results (TP) out of all the actual unexpected results (TP + FN)
+
+* A logic regression (ref) is used to model the relationship between the plan and outcome expectations. (justify the use of Logic regression)
+
+* on independence of the tests
 

### 2024-10-31 _manuscript/_tex/index.tex
add dot points

@@
 
+\renewcommand*\contentsname{Table of contents}
+{
+\hypersetup{linkcolor=}
+\setcounter{tocdepth}{3}
+\tableofcontents
+}
+
+\newpage
+
 \section{Introduction}\label{introduction}
@@
 
+\section{Literature review}\label{literature-review}
+
+\subsection{Diagnosing unexpected outcomes in data
+analysis}\label{diagnosing-unexpected-outcomes-in-data-analysis}
+
+\subsection{Unit tests}\label{unit-tests}
+
+\subsection{Data quality checks}\label{data-quality-checks}
+
 \section{Analysis plan}\label{sec-plan}
+
+\subsection{Framing checks into unit
+tests}\label{framing-checks-into-unit-tests}
 
@@
 \section{Method}\label{method}
+
+\subsection{A workflow to assess the quality of unit
+tests}\label{a-workflow-to-assess-the-quality-of-unit-tests}
 
@@
   how to systematically simulate data is still unknown, sensitivity of
-  the simulation to the resutls
+  the simulation to the results
+\item
+  plotting is a critical way to check data and they can still be frame
+  into a unit test. it is a open problem to how to encode the
+  visualization into the unit tests. Maybe a procedure like confirm plot
+  (this looks alright to you) and then press the button to continue
 \item

### 2024-10-31 index.qmd
add dot points

@@
    keep-tex: true
+   toc: true
 preamble: >
@@
 ```
+
+\newpage
 
@@
 
+# Literature review
+
+## Diagnosing unexpected outcomes in data analysis
+
+## Unit tests
+
+## Data quality checks
+
 # Analysis plan {#sec-plan}
+
+## Framing checks into unit tests
 
@@
 # Method
+
+## A workflow to assess the quality of unit tests
 
@@
 
-* how to systematically simulate data is still unknown, sensitivity of the simulation to the resutls
+* how to systematically simulate data is still unknown, sensitivity of the simulation to the results
+
+* plotting is a critical way to check data and they can still be frame into a unit test.
+it is a open problem to how to encode the visualization into the unit tests.
+Maybe a procedure like confirm plot (this looks alright to you) and then press the button to continue
 

### 2024-10-31 _manuscript/_tex/index.tex
second version of the introduction

@@
 \begin{abstract}
-This is the abstract
+Unexpected results in data analysis can prompt researchers to examine
+data quality and analysis steps. While some researchers can typically
+diagnose issues effectively with a few checks, others may struggle to
+identify appropriate checks to diagnose the problem. {[}an example of
+check{]} These checks are often informal and difficult to trace and
+discuss in publications, resulting in others questioning the
+trustworthiness of the analysis. To address this, we formalize the
+informal checks into an \emph{analysis plan} that encompasses the
+analysis steps and (the unit tests): one test for whether the result
+meets expectations and multiple tests for checking the analysis. We then
+present a procedure to assess the quality of these unit tests based on
+their accuracy and redundancy on simulated versions of the original
+data. The accuracy is assessed using binary classification metrics,
+\emph{i.e.}, precision and recall, while redundancy is calculated using
+mutual information. This procedure can be used to conduct a sensitivity
+analysis, compare different analysis plans, and to identify the optimal
+cutoff point for the unit tests.
 \end{abstract}
@@
 
-{[}background - In a data analysis workflow{]}
+In a data analysis, an experienced researcher can often quickly assess
+whether a result meets their expectation, while others may be unsure of
+what to anticipate from the analysis or how to recognize when results
+diverge form expected norms. (These expectations are often based on
+prior knowledge, domain expertise, or common sense and they can often be
+framed into a Boolean question that can be answered by a unit test. For
+example, in a linear regression model, one may expect the p-value of the
+coefficient to be less than 0.05. On top of the expectation on the final
+results, researchers may also have expectations on the intermediate
+steps of the analysis process. For example, one may expect the data
+doesn't contain outliers, etc. These expectations are often checked
+early on to prevent an unexpected propagate through the analysis
+process.)
 
-In a data analysis workflow -\textgreater{} going back to diagnose the
-descripency between the expectation and the result is often
-time-consuming and challenging.
+This expectation-setting process is often implicit and rarely documented
+or discussed in publications. (There are most publication requires open
+source code for publication, however, it is no instruction or guidance
+on checking \ldots). Yet, making them explicit is crucial for 1) helping
+junior researchers interpret results and diagnose the unexpected, 2)
+providing checkpoints that support independent replication or
+application of methods to new data, 3) aligning assumptino across
+researchers from different backgrounds.
 
-For consumers of data analysis, it is often difficult to evaluate the
-analysis outcome. Not always have the script to run, even difficult when
-different expectations are held.
-
-We propose here a concept called analysis plan to make the expectation
-of the analysis process explicit. It can help to guide the analysis
-process, to communicate and compare the analysis process to others, and
-to evaluate the analysis process. (think more closer here how our
-approach help to solve the problem)
-
-For expert analysts, the expectation of the analysis process is often
-implicit. They may have a mental model of the analysis process and the
-expected results. However, for non-expert analysts, the expectation of
-the analysis process may not be clear. They may not know what to expect
-from the analysis process, or what to look for when an analysis process
-fails the expectation. An analysis plan can help to make the expectation
-of the analysis process explicit. It can help to guide the analysis
-process, to communicate and compare the analysis process to others, and
-to evaluate the analysis process.
-
-In this paper, a concept called analysis plan is proposed to describe
-the logical structure of a data analysis. An analysis plan is a set of
-analysis steps plus their expected outcomes. It is a formal
-representation of the analysis process and can be used to guide the
-analysis process, to communicate and compare the analysis process to
-others, and to evaluate the analysis process. The concept of analysis
-plan is illustrated with examples. The implications of the concept for
-data analysis practice is discussed.
-
-The analysis plan described in this paper should be differentiated from
-the pre-specifies analysis plan document often used in biostatistics to
-specifies the hypothesis, data collection mechanism, statistical
-procedures etc of randomized experiments.
+In this paper, we formalize these informal checks into an \emph{analysis
+plan} comprising the analysis steps along with the associated unit
+tests. This formalization allows us to examine assumptions made about
+the data during analysis and to compare different unit tests in
+diagnosing unexpected outcomes. We then introduce a procedure to assess
+the quality of these unit tests based on their accuracy and redundancy
+across simulated version of the original data. The accuracy is assessed
+using binary classification metrics -- precision and recall -- while
+redundancy is measured using mutual information. We illustrate these
+concepts with a simple step count example and a scenario of diagnosing
+unexpected regression results when studying the effect of PM10 on
+mortality.
 
@@
 
+TODO: what if the expectation is ``wrong''
+
 \subsection{Unit tests}\label{unit-tests}
@@
 \subsection{Data quality checks}\label{data-quality-checks}
+
+\begin{itemize}
+\tightlist
+\item
+  \texttt{pointblank}, \texttt{assertthat}
+\end{itemize}
 
@@
 only if it helps
-
-TODO: more about analysis plan itself here
-
-TODO: what if the expectation is ``wrong''
-
-The whole point being how can be build trust in other's analysis. Some
-provides scripts to run the analysis but that's not enough. We need to
-know what to expect from the analysis and that is often not explicitly
-specified.
 

### 2024-10-31 index.qmd
second version of the introduction

@@
 title: "My wonderful paper"
-abstract: "This is the abstract"
+abstract: "Unexpected results in data analysis can prompt researchers to examine data quality and analysis steps.
+While some researchers can typically diagnose issues effectively with a few checks, others may struggle to identify appropriate checks to diagnose the problem.
+[an example of check] These checks are often informal and difficult to trace and discuss in publications, resulting in others questioning the trustworthiness of the analysis.
+To address this, we formalize the informal checks into an *analysis plan* that encompasses the analysis steps and (the unit tests): one test for whether the result meets expectations and multiple tests for checking the analysis.
+We then present a procedure to assess the quality of these unit tests based on their accuracy and redundancy on simulated versions of the original data.
+The accuracy is assessed using binary classification metrics, *i.e.*, precision and recall, while redundancy is calculated using mutual information.
+This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
 cite-method: natbib
@@
 
-[background - In a data analysis workflow]
+In a data analysis, an experienced researcher can often quickly assess whether a result meets their expectation, while others may be unsure of what to anticipate from the analysis or how to recognize when results diverge form expected norms.
+(These expectations are often based on prior knowledge, domain expertise, or common sense and they can often be framed into a Boolean question that can be answered by a unit test.
+For example, in a linear regression model, one may expect the p-value of the coefficient to be less than 0.05.
+On top of the expectation on the final results, researchers may also have expectations on the intermediate steps of the analysis process.
+For example, one may expect the data doesn't contain outliers, etc.
+These expectations are often checked early on to prevent an unexpected propagate through the analysis process.)
 
-In a data analysis workflow -> going back to diagnose the descripency between the expectation and the result is often time-consuming and challenging.
+This expectation-setting process is often implicit and rarely documented or discussed in publications.
+(There are most publication requires open source code for publication, however, it is no instruction or guidance on checking ...).
+Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumptino across researchers from different backgrounds.
 
-For consumers of data analysis, it is often difficult to evaluate the analysis outcome. Not always have the script to run, even difficult when different expectations are held.
-
-We propose here a concept called analysis plan to make the expectation of the analysis process explicit.
-It can help to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
-(think more closer here how our approach help to solve the problem)
-
-For expert analysts, the expectation of the analysis process is often implicit.
-They may have a mental model of the analysis process and the expected results.
-However, for non-expert analysts, the expectation of the analysis process may not be clear.
-They may not know what to expect from the analysis process, or what to look for when an analysis process fails the expectation.
-An analysis plan can help to make the expectation of the analysis process explicit.
-It can help to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
-
-In this paper, a concept called analysis plan is proposed to describe the logical structure of a data analysis.
-An analysis plan is a set of analysis steps plus their expected outcomes.
-It is a formal representation of the analysis process and can be used to guide the analysis process, to communicate and compare the analysis process to others, and to evaluate the analysis process.
-The concept of analysis plan is illustrated with examples.
-The implications of the concept for data analysis practice is discussed.
-
-The analysis plan described in this paper should be differentiated from the pre-specifies analysis plan document often used in biostatistics to specifies the hypothesis, data collection mechanism, statistical procedures etc of randomized experiments.
+In this paper, we formalize these informal checks into an *analysis plan* comprising the analysis steps along with the associated unit tests.
+This formalization allows us to examine assumptions made about the data during analysis and to compare different unit tests in diagnosing unexpected outcomes.
+We then introduce a procedure to assess the quality of these unit tests based on their accuracy and redundancy across simulated version of the original data.
+The accuracy is assessed using binary classification metrics -- precision and recall -- while redundancy is measured using mutual information.
+We illustrate these concepts with a simple step count example and a scenario of diagnosing unexpected regression results when studying the effect of PM10 on mortality.
 
@@
 
+TODO: what if the expectation is "wrong"
+
 ## Unit tests
@@
 ## Data quality checks
+
+- `pointblank`, `assertthat`
 
@@
 
-TODO: more about analysis plan itself here
-
-TODO: what if the expectation is "wrong"
-
-The whole point being how can be build trust in other's analysis.
-Some provides scripts to run the analysis but that's not enough.
-We need to know what to expect from the analysis and that is often not explicitly specified.
+<!-- The analysis plan described in this paper should be differentiated from the pre-specifies analysis plan document often used in biostatistics to specifies the hypothesis, data collection mechanism, statistical procedures etc of randomized experiments.
+-->
 

### 2024-11-05 _manuscript/_tex/index.tex
notes on the literature review

@@
 providing checkpoints that support independent replication or
-application of methods to new data, 3) aligning assumptino across
+application of methods to new data, 3) aligning assumption across
 researchers from different backgrounds.
@@
 
+\citep{peng_diagnosing_2021}
+
 TODO: what if the expectation is ``wrong''
 
-\subsection{Unit tests}\label{unit-tests}
-
-\subsection{Data quality checks}\label{data-quality-checks}
+\subsection{Data analysis checks in statistical
+software}\label{data-analysis-checks-in-statistical-software}
 
@@
 \item
-  \texttt{pointblank}, \texttt{assertthat}
+  The \texttt{assertr} package \citep{assertr} provides a set of
+  functions to check the quality of data, including the presence of
+  missing values, the distribution of variables, and the relationship
+  between variables.
+\item
+  Similarly, the \texttt{pointblank} package \citep{pointblank} is a
+  data validation tool that allows users to define data quality checks.
+\end{itemize}
+
+For model checks,
+
+\begin{itemize}
+\tightlist
+\item
+  the \texttt{performance} package \citep{performance} provides a set of
+  functions to assess the performance of a model, including residual
+  analysis, goodness-of-fit tests.
+\item
+  The \texttt{yardstick} package \citep{yardstick} provides the
+  calculation of model performance metrics in the \texttt{tidymodels}
+  \citep{tidymodels} ecosystem.
+\end{itemize}
+
+\subsection{Unit tests}\label{unit-tests}
+
+Why develop unit tests in software engineering? Because it is difficult
+to predict the outcome of a program, especially when the program is
+complex.
+
+Software testing relies on ``oracles'' to provide the expected output
+necessary for verifying test results. For example, to test whether the
+program correctly calculates 1 + 1, one need to supply the correct
+answer, 2. However, establishing this ``correct'' output can sometimes
+be challenging, where obtaining a solution may be difficult without the
+program itself. This situation leads to the oracle problem
+\citep{barr2014oracle}. In data analysis, the similar oracle problem can
+happen, as the ``truth'' of an outcome, the expectation, depends on the
+underlying theory or interpretation. For example, in a linear regression
+model, the significance of a coefficient may be expected or unexpected
+based on the theory, making it challenging for researcher with a
+different theory to assess the results and the analysis.
+
+Existing tools for checks in data analysis including those check for
+data quality and model diagnostics. In software development,
+
+\begin{itemize}
+\tightlist
+\item
+  the \texttt{testthat} package \citep{testthat} provides the
+  infrastructure for testing R code. It allows users to write unit tests
+  for functions and packages.
+\item
+  The \texttt{assertthat} package \citep{assertthat} helps to write
+  assertion statements that are supposed to be true at a certain point
+  in the code for defensive programming.
 \end{itemize}

### 2024-11-05 index.qmd
notes on the literature review

@@
 (There are most publication requires open source code for publication, however, it is no instruction or guidance on checking ...).
-Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumptino across researchers from different backgrounds.
+Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumption across researchers from different backgrounds.
 
@@
 
+[@peng_diagnosing_2021]
+
 TODO: what if the expectation is "wrong"
+
+## Data analysis checks in statistical software
+
+  * The `assertr` package [@assertr] provides a set of functions to check the quality of data, including the presence of missing values, the distribution of variables, and the relationship between variables.
+  * Similarly, the `pointblank` package [@pointblank] is a data validation tool that allows users to define data quality checks.
+
+  For model checks,
+
+  * the `performance` package [@performance] provides a set of functions to assess the performance of a model, including residual analysis, goodness-of-fit tests. 
+  * The `yardstick` package [@yardstick] provides the calculation of model performance metrics in the `tidymodels` [@tidymodels] ecosystem.
 
@@
 
-## Data quality checks
+Why develop unit tests in software engineering?
+Because it is difficult to predict the outcome of a program, especially when the program is complex.
+Because the loss associated with a program failure is costly.
 
-- `pointblank`, `assertthat`
+Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
+For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2.
+However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself.
+This situation leads to the oracle problem [@barr2014oracle].
+In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation.
+For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.
+
+Existing tools for checks in data analysis including those check for data quality and model diagnostics. In software development,
+
+* the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
+* The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.
 

### 2024-11-06 _manuscript/_tex/index.tex
more writing on the method

@@
 
+In data analysis, checks are often performed to ensure the quality of
+the data and the validity of the model. For data checks,
+
 \begin{itemize}
@@
 
+These checks test on specific aspect of the data or the model but
+doesn't reflect the underlying assumption of the data analysts about the
+data or the problem context. It may be obvious to see the assumption if
+someone is testing for data quality on whether a column in the data
+frame is between two values. In a much complex analysis, it is difficult
+to trace the assumptions made by the data analysts. Currently, most of
+these assumptions are not documented or discussed in the analysis
+process, or at least in a machine readable format.
+
 \subsection{Unit tests}\label{unit-tests}
@@
 to predict the outcome of a program, especially when the program is
-complex.
+complex. Because the loss associated with a program failure is costly.
 
@@
 
-\begin{itemize}
-\item
-  multiple tests can be generated to diagnose different aspects of an
-  analysis
-\item
-  good tests
+In real-world applications, it is rare to create a set of unit tests can
+fully guarantee expected results. On one hand, there is the cost of
+effort involved in manually developing all these tests; on the other,
+there is the inherent complexity of the problem. (can we - if we're
+thriving for detecting 95\% of the cause?). However, the quality of unit
+tests can be evaluated by simulated data. One set of tests is considered
+better than another if a small set of tests can reliably detect
+unexpected outcomes, which brings two criteria in the evaluation metric:
+accuracy and parsimony.
 
-  \begin{enumerate}
-  \def\labelenumi{\arabic{enumi})}
-  \tightlist
-  \item
-    are responsive to an unexpected result - accuracy,
-  \item
-    motivate actions of analysts to investigate the results. This points
-    towards a smaller set of independent tests
-  \end{enumerate}
-\item
-  Since both the plan/ outcome expectations can be phrased as unit
-  tests, which output binary outcomes, consider using the plan
-  expectations as predictors of the outcome expectations. This allows to
-  generate the confusion matrix of predicting the outcome with the plan
-  expectation.
-\item
-  For an analysis, ideally, good plan expectations should maximize the
-  detection of unexpected outcomes while minimize the false positive
-  discovery, which suggest the use of precision and recall (REF) for
-  evaluating the performance of the plan expectations.
+Accuracy refers to a set of tests' ability to accurately detect
+unexpected outcomes while minimizing false positives and false
+negatives. A false positive can indicate (caution or skepticism on
+checking the data), whereas a false negative suggests the tests may lack
+sensitivity to unexpected outcomes. Since both plan and outcome
+expectations can be framed as unit tests with binary outcomes, one
+approach is to predict the outcome expectation based on the plan
+expectation. Performance of the tests can then be evaluated using
+precision and recall metrics from the confusion matrix:
 
@@
 \item
-    precision: the proportion of unexpected results (TP) out of all the
-    predicted unexpected results (TP + FP)
+  precision: the proportion of correctly identified unexpected results
+  (true positives) out of all the predicted unexpected results (true
+  positives + false positives)
 \item
-    recall: the proportion of unexpected results (TP) out of all the
-    actual unexpected results (TP + FN)
+  recall: the proportion of correctly identified unexpected results
+  (true positives) out of all the actual unexpected results (true
+  positives + false negatives)
 \end{itemize}
+
+The second criteria is parsimony in the tests. While tests may score
+high on accuracy, they may be less effective at explaining the reasons
+behind unexpected results. This could happen if a set of tests are all
+tangentially related to the cause of the unexpected results, but none
+addressing the root cause. It may also occur if the tests are correlated
+with one another, leading to redundancy.
+
+\begin{itemize}
+\item
+  explain mutual information
+\item
+  explain logic regression
+\item
+  explain combining precision, recall, and independence together through
+  ``means''
 \item
@@
   plan and outcome expectations. (justify the use of Logic regression)
-\item
-  on independence of the tests
 \end{itemize}
@@
 \item
-  currently no automated way to generate unit tests
+  currently no automated way to generate unit tests. It is interesting
+  to see the automation of generating unit tests, although it requires
+  the inputs from experts across a wide array of common scenarios.
 \end{itemize}

### 2024-11-06 index.qmd
more writing on the method

@@
 
+  In data analysis, checks are often performed to ensure the quality of the data and the validity of the model. For data checks,
+
   * The `assertr` package [@assertr] provides a set of functions to check the quality of data, including the presence of missing values, the distribution of variables, and the relationship between variables.
@@
   * The `yardstick` package [@yardstick] provides the calculation of model performance metrics in the `tidymodels` [@tidymodels] ecosystem.
+
+These checks test on specific aspect of the data or the model but doesn't reflect the underlying assumption of the data analysts about the data or the problem context.
+It may be obvious to see the assumption if someone is testing for data quality on whether a column in the data frame is between two values.
+In a much complex analysis, it is difficult to trace the assumptions made by the data analysts.
+Currently, most of these assumptions are not documented or discussed in the analysis process, or at least in a machine readable format.
 
@@
 
-- multiple tests can be generated to diagnose different aspects of an analysis 
-- good tests 
-    1) are responsive to an unexpected result - accuracy, 
-    2) motivate actions of analysts to investigate the results. This points towards a smaller set of independent tests
+In real-world applications, it is rare to create a set of unit tests can fully guarantee expected results.
+On one hand, there is the cost of effort involved in manually developing all these tests; on the other, there is the inherent complexity of the problem.
+(can we - if we're thriving for detecting 95% of the cause?).
+However, the quality of unit tests can be evaluated by simulated data.
+One set of tests is considered better than another if a small set of tests can reliably detect unexpected outcomes, which brings two criteria in the evaluation metric: accuracy and parsimony.
 
-- Since both the plan/ outcome expectations can be phrased as unit tests, which output binary outcomes, consider using the plan expectations as predictors of the outcome expectations.
-This allows to generate the confusion matrix of predicting the outcome with the plan expectation.
+Accuracy refers to a set of tests' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
+A false positive can indicate (caution or skepticism on checking the data), whereas a false negative suggests the tests may lack sensitivity to unexpected outcomes.
+Since both plan and outcome expectations can be framed as unit tests with binary outcomes, one approach is to predict the outcome expectation based on the plan expectation.
+Performance of the tests can then be evaluated using precision and recall metrics from the confusion matrix:
 
-- For an analysis, ideally, good plan expectations should maximize the detection of unexpected outcomes while minimize the false positive discovery, which suggest the use of precision and recall (REF) for evaluating the performance of the plan expectations.
+  * precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
+  * recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)
 
-  * precision: the proportion of unexpected results (TP) out of all the predicted unexpected results (TP + FP)
-  * recall: the proportion of unexpected results (TP) out of all the actual unexpected results (TP + FN)
+The second criteria is parsimony in the tests.
+While tests may score high on accuracy, they may be less effective at explaining the reasons behind unexpected results.
+This could happen if a set of tests are all tangentially related to the cause of the unexpected results, but none addressing the root cause.
+It may also occur if the tests are correlated with one another, leading to redundancy.
+
+- explain mutual information
+- explain logic regression 
+- explain combining precision, recall, and independence together through "means"
 
 * A logic regression (ref) is used to model the relationship between the plan and outcome expectations. (justify the use of Logic regression)
-
-* on independence of the tests
 
@@
 
-* currently no automated way to generate unit tests
+* currently no automated way to generate unit tests.
+It is interesting to see the automation of generating unit tests, although it requires the inputs from experts across a wide array of common scenarios.
 

### 2024-11-07 _manuscript/_tex/index.tex
more on data analysis check literature

@@
 
-\citep{peng_diagnosing_2021}
+\citep{peng_diagnosing_2021} describes three pedagogical exercises of
+introducing diagnosing unexpected outcome into a classroom setting.
 
@@
 
-In data analysis, checks are often performed to ensure the quality of
-the data and the validity of the model. For data checks,
+Currently, little has been done on how to computationally incorporate
+this diagnostic approach into data analysis workflow or software. Most
+of the diagnostic tools focus on defining user-specified rules, such as
+data quality checks or producing metrics to summarize model performance,
+as in model diagnostics. For example, the \texttt{assertr} package
+\citep{assertr} and the \texttt{pointblank} package \citep{pointblank}
+provide data validation tools to allow users to define data quality
+checks. In contrast, packages provide model checks tools like
+\texttt{performance} \citep{performance} and \texttt{yardstick}
+\citep{yardstick}, from the \texttt{tidymodels} \citep{tidymodels}
+ecosystem, offers goodness-of-fit, residual analysis, and performance
+metrics.
 
-\begin{itemize}
-\tightlist
-\item
-  The \texttt{assertr} package \citep{assertr} provides a set of
-  functions to check the quality of data, including the presence of
-  missing values, the distribution of variables, and the relationship
-  between variables.
-\item
-  Similarly, the \texttt{pointblank} package \citep{pointblank} is a
-  data validation tool that allows users to define data quality checks.
-\end{itemize}
-
-For model checks,
-
-\begin{itemize}
-\tightlist
-\item
-  the \texttt{performance} package \citep{performance} provides a set of
-  functions to assess the performance of a model, including residual
-  analysis, goodness-of-fit tests.
-\item
-  The \texttt{yardstick} package \citep{yardstick} provides the
-  calculation of model performance metrics in the \texttt{tidymodels}
-  \citep{tidymodels} ecosystem.
-\end{itemize}
-
-These checks test on specific aspect of the data or the model but
-doesn't reflect the underlying assumption of the data analysts about the
-data or the problem context. It may be obvious to see the assumption if
-someone is testing for data quality on whether a column in the data
-frame is between two values. In a much complex analysis, it is difficult
-to trace the assumptions made by the data analysts. Currently, most of
-these assumptions are not documented or discussed in the analysis
-process, or at least in a machine readable format.
+These packages provide the tools to conduct diagnostics but still don't
+reflect the mental process of how data analysts diagnose the unexpected
+output. For example, when an unexpected output occurs, an analyst may
+check on whether a column in the data frame is between two values for
+data quality. However, it is not documented what motivates the analyst
+to conduct this check -- whether it also applies to other researchers
+analyzing new data in the same contexts, whether it is a common practice
+in the field, or whether it is a reaction to this particular data or
+scenario. Currently, most of these assumptions are not discussed in the
+publication or captured by tools themselves. While one might be able to
+infer some of the mental process of the analysts from external sources,
+such as talking to them or watching screencast videos produced by the
+analysts e.g.~TidyTuesday screencast videos, these insights are not
+systematically documented or made machine-readable. This gap highlights
+the need for tools that provide higher level documentation of reasoning
+behind the checks, facilitating a more transparent and interpretable
+analysis process.
 

### 2024-11-07 index.qmd
more on data analysis check literature

@@
 
-[@peng_diagnosing_2021]
+[@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.
 
@@
 
-  In data analysis, checks are often performed to ensure the quality of the data and the validity of the model. For data checks,
+Currently, little has been done on how to computationally incorporate this diagnostic approach into data analysis workflow or software.
+Most of the diagnostic tools focus on defining user-specified rules, such as data quality checks or producing metrics to summarize model performance, as in model diagnostics.
+For example, the `assertr` package [@assertr] and the `pointblank` package [@pointblank] provide data validation tools to allow users to define data quality checks.
+In contrast, packages provide model checks tools like `performance` [@performance] and `yardstick` [@yardstick], from the `tidymodels` [@tidymodels] ecosystem, offers goodness-of-fit, residual analysis, and model performance metrics.
 
-  * The `assertr` package [@assertr] provides a set of functions to check the quality of data, including the presence of missing values, the distribution of variables, and the relationship between variables.
-  * Similarly, the `pointblank` package [@pointblank] is a data validation tool that allows users to define data quality checks.
-
-  For model checks,
-
-  * the `performance` package [@performance] provides a set of functions to assess the performance of a model, including residual analysis, goodness-of-fit tests. 
-  * The `yardstick` package [@yardstick] provides the calculation of model performance metrics in the `tidymodels` [@tidymodels] ecosystem.
-
-These checks test on specific aspect of the data or the model but doesn't reflect the underlying assumption of the data analysts about the data or the problem context.
-It may be obvious to see the assumption if someone is testing for data quality on whether a column in the data frame is between two values.
-In a much complex analysis, it is difficult to trace the assumptions made by the data analysts.
-Currently, most of these assumptions are not documented or discussed in the analysis process, or at least in a machine readable format.
+These packages provide the tools to conduct diagnostics but still don't reflect the mental process of how data analysts diagnose the unexpected output.
+For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality.
+However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario.
+Currently, most of these assumptions are not discussed in the publication or captured by tools themselves.
+While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g.
+TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable.
+This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.
 

### 2024-11-07 _manuscript/_tex/index.tex
add more dot points

@@
 
+\begin{itemize}
+\tightlist
+\item
+  emphasize trustworthy data science since it is the theme of the
+  special issue
+\end{itemize}
+
 In a data analysis, an experienced researcher can often quickly assess
@@
 using binary classification metrics -- precision and recall -- while
-redundancy is measured using mutual information. We illustrate these
-concepts with a simple step count example and a scenario of diagnosing
-unexpected regression results when studying the effect of PM10 on
-mortality.
+redundancy is measured using mutual information. The workflow provides
+numerical guarantee of the consistency of the expectation and the data
+for the analysis plan. This means that given the assumption of the data
+generating mechanism, the analysis will provide the outcome we expect.
+In data analysis, analysts are not necessarily collecting the data and
+they can only make reasonable assumption about the quality of the data.
+This method we propose provides a numerical guarantee, based on
+simulation, that the analysis will produce the outcome we expect, given
+the assumption of the data generating mechanism. This provides
+transparency of the analysis and build trustworthy data analysis.
 
@@
 \citep{yardstick}, from the \texttt{tidymodels} \citep{tidymodels}
-ecosystem, offers goodness-of-fit, residual analysis, and performance
-metrics.
+ecosystem, offers goodness-of-fit, residual analysis, and model
+performance metrics.
 
@@
 complex. Because the loss associated with a program failure is costly.
-
-Software testing relies on ``oracles'' to provide the expected output
-necessary for verifying test results. For example, to test whether the
-program correctly calculates 1 + 1, one need to supply the correct
-answer, 2. However, establishing this ``correct'' output can sometimes
-be challenging, where obtaining a solution may be difficult without the
-program itself. This situation leads to the oracle problem
-\citep{barr2014oracle}. In data analysis, the similar oracle problem can
-happen, as the ``truth'' of an outcome, the expectation, depends on the
-underlying theory or interpretation. For example, in a linear regression
-model, the significance of a coefficient may be expected or unexpected
-based on the theory, making it challenging for researcher with a
-different theory to assess the results and the analysis.
 
@@
   the inputs from experts across a wide array of common scenarios.
+\item
+  There are cost and benefit on what granularity to set expectation.
+  Setting them on every step of the analysis requires more work from the
+  analysts while for complex analysis, the diagnostics may require
+  divide the analysis into different components and diagnose it in a
+  hierarchical structure. The method proposed in this paper should be
+  applicable for both fine and coarse analysis.
+\item
+  (only mention it) Software testing relies on ``oracles'' to provide
+  the expected output necessary for verifying test results. For example,
+  to test whether the program correctly calculates 1 + 1, one need to
+  supply the correct answer, 2. However, establishing this ``correct''
+  output can sometimes be challenging, where obtaining a solution may be
+  difficult without the program itself. This situation leads to the
+  oracle problem \citep{barr2014oracle}. In data analysis, the similar
+  oracle problem can happen, as the ``truth'' of an outcome, the
+  expectation, depends on the underlying theory or interpretation. For
+  example, in a linear regression model, the significance of a
+  coefficient may be expected or unexpected based on the theory, making
+  it challenging for researcher with a different theory to assess the
+  results and the analysis.
 \end{itemize}

### 2024-11-07 index.qmd
add more dot points

@@
 
+- emphasize trustworthy data science since it is the theme of the special issue
+
 In a data analysis, an experienced researcher can often quickly assess whether a result meets their expectation, while others may be unsure of what to anticipate from the analysis or how to recognize when results diverge form expected norms.
@@
 The accuracy is assessed using binary classification metrics -- precision and recall -- while redundancy is measured using mutual information.
-We illustrate these concepts with a simple step count example and a scenario of diagnosing unexpected regression results when studying the effect of PM10 on mortality.
+The workflow provides numerical guarantee of the consistency of the expectation and the data for the analysis plan.
+This means that given the assumption of the data generating mechanism, the analysis will provide the outcome we expect.
+In data analysis, analysts are not necessarily collecting the data and they can only make reasonable assumption about the quality of the data.
+This method we propose provides a numerical guarantee, based on simulation, that the analysis will produce the outcome we expect, given the assumption of the data generating mechanism.
+This provides transparency of the analysis and build trustworthy data analysis.
 
@@
 Because the loss associated with a program failure is costly.
-
-Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
-For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2.
-However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself.
-This situation leads to the oracle problem [@barr2014oracle].
-In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation.
-For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.
 
@@
 
+* There are cost and benefit on what granularity to set expectation.
+Setting them on every step of the analysis requires more work from the analysts while for complex analysis, the diagnostics may require divide the analysis into different components and diagnose it in a hierarchical structure.
+The method proposed in this paper should be applicable for both fine and coarse analysis.
+
+* (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
+For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2.
+However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself.
+This situation leads to the oracle problem [@barr2014oracle].
+In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation.
+For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.
+
 # Conclusion {#sec-conclusion}

### 2024-11-07 _manuscript/_tex/index.tex
remove the result universe example

@@
 
-(might be useful) Analysis plans can be constructed at various
-granularities, at the highest level, one may only has a plan of the
-specific method used for analysing data and the expected outcome. This
-provides little guidance when a deviation from expectation occurs. At
-the lowest level, one may have a plan for each data entry and every data
-handling steps. This provides too much detail and may not be practical
-in practice.
-
 Experienced analysts often have implicit expectation about the outcome
@@
 in analysis.
-
-Furthermore, by generating multiple versions of the data, we can emulate
-various scenarios within the same context for students to exercise
-judgement when conducting data analysis in a classroom setting.
 
@@
 
-To build the result universe, datasets can be simulated to either meet
-and fail these plan expectations, allowing the analysts to observe the
-significance of PM10 coefficient. Here, sample sizes of 50, 100, 500,
-and 1000 are considered. Two model specifications are included: 1)
-linear model with PM10 as the only covariate
-(\(\text{mortality} \sim \text{PM10}\)), 2) linear model with PM10 and
-temperature as covariates
-(\(\text{mortality} \sim \text{PM10} + \text{temp}\)). A grid-based
-approach is used to simulate correlation structure. Reasonable ranges of
-correlation between the three variables are
-\(\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]\),
-\(\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]\),
-and \(\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]\).
-
 \begin{itemize}
+\tightlist
 \item
   add a paragraph to describe the simulation process
-\item
-  add a fourth panel to describe the comparison of a right/ wrong
-  expectation, i.e.~correlation on PM10 and mortality
 \end{itemize}
-
-Figure~\ref{fig-result-universe} shows that result universe of the
-linear regression model and how a change of decision in one of the plan
-expectations above affect the outcome expectation. Panel a) is colored
-by the outcome expectation -- whether a significant p-value is found in
-the PM10 coefficient. Panel b) shows the effect of adding temperature to
-the model and the results show that the significance of PM10 coefficient
-can be achieved by adding temperature to the model for a sample size of
-500. Panel c) shows that increasing sample size from 50 to 100 enhances
-the significance of p-value for PM10 and the significance remains with
-further increases in sample size. {[}note: weave the ``actual data''
-into the example linear regression model{]}
-
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
-
-\phantomsection\label{cell-fig-result-universe}
-\begin{figure}[H]
-
-\centering{
-
-\includegraphics{index_files/figure-pdf/fig-result-universe-1.pdf}
-
-}
-
-\caption{\label{fig-result-universe}The result universe of linear
-regression model to study the effect of PM10 on mortality: a) colored by
-whether the p-value of PM10is significant (less than 0.05), b) the
-effect of adding temperature to the model for a sample size of 500, c)
-the effect of increasing sample size for a fixed correlation structure.}
-
-\end{figure}%
 
@@
 \item
-  There are cost and benefit on what granularity to set expectation.
-  Setting them on every step of the analysis requires more work from the
-  analysts while for complex analysis, the diagnostics may require
-  divide the analysis into different components and diagnose it in a
-  hierarchical structure. The method proposed in this paper should be
-  applicable for both fine and coarse analysis.
+  There are cost and benefit on setting expectation on different
+  granularity. At the lowest level, one may have a plan for each data
+  entry and every data handling steps. This requires more work from the
+  analysts and may not be practical in practice. For more complex
+  analyses, analysts may divide the analysis into sections and set
+  expectations for each. They can then focus on the specific sections
+  flagged by the tests and sub-divide the sections to set expectation
+  and diagnose the analysis in a hierarchical manner.
 \item

### 2024-11-07 index.qmd
remove the result universe example

@@
 
-(might be useful) Analysis plans can be constructed at various granularities, at the highest level, one may only has a plan of the specific method used for analysing data and the expected outcome.
-This provides little guidance when a deviation from expectation occurs.
-At the lowest level, one may have a plan for each data entry and every data handling steps.
-This provides too much detail and may not be practical in practice.
-
 Experienced analysts often have implicit expectation about the outcome and rely on a few "directional signs" to check when the outcome deviate from those expectation.
@@
 This enables them to evaluate the outcomes based on their own plan expectations and gain a broader perspective on how the actual results produced by analysts fit within this spectrum of possibilities, promoting transparency and trust in analysis.
-
-Furthermore, by generating multiple versions of the data, we can emulate various scenarios within the same context for students to exercise judgement when conducting data analysis in a classroom setting.
 
@@
 
-To build the result universe, datasets can be simulated to either meet and fail these plan expectations, allowing the analysts to observe the significance of PM10 coefficient.
-Here, sample sizes of 50, 100, 500, and 1000 are considered.
-Two model specifications are included: 1) linear model with PM10 as the only covariate ($\text{mortality} \sim \text{PM10}$), 2) linear model with PM10 and temperature as covariates ($\text{mortality} \sim \text{PM10} + \text{temp}$).
-A grid-based approach is used to simulate correlation structure.
-Reasonable ranges of correlation between the three variables are $\text{cor}(\text{mortality}, \text{PM10}) \in [-0.01, 0]$, $\text{cor}(\text{mortality}, \text{temperature}) \in [-0.6, -0.2]$, and $\text{cor}(\text{PM10}, \text{temperature}) \in [0.2, 0.6]$.
-
  * add a paragraph to describe the simulation process
-
- * add a fourth panel to describe the comparison of a right/ wrong expectation, i.e. correlation on PM10 and mortality
-
-@fig-result-universe shows that result universe of the linear regression model and how a change of decision in one of the plan expectations above affect the outcome expectation.
-Panel a) is colored by the outcome expectation -- whether a significant p-value is found in the PM10 coefficient.
-Panel b) shows the effect of adding temperature to the model and the results show that the significance of PM10 coefficient can be achieved by adding temperature to the model for a sample size of 500.
-Panel c) shows that increasing sample size from 50 to 100 enhances the significance of p-value for PM10 and the significance remains with further increases in sample size.
-[note: weave the "actual data" into the example linear regression model]
 
@@
 
-# test for dist fit
-# fitdist(pm10$pm10, "gamma") -> pm10_fit
-# plot(pm10_fit) # gamma is better than normal for the fit of pm10
-#fitdist(pm10$temp, "norm") -> temp_fit
-# plot(temp_fit)
-
 res <- tibble(corr_mtx = corr_mtx) |> 
@@
          fit = list(summary(lm(model, data))$coefficients))  
-```
-
-```{r}
-#| label: fig-result-universe
-#| fig-cap: "The result universe of linear regression model to study the effect of PM10 on mortality: a) colored by whether the p-value of PM10is significant (less than 0.05), b) the effect of adding temperature to the model for a sample size of 500, c) the effect of increasing sample size for a fixed correlation structure."
-#| fig.width: 12
-#| fig.height: 12
-dt <- res |>   
-  mutate(     
-    p_value = fit[2,4],     
-    coef = fit[2,1],     
-    xy_correlation = as.numeric(corr_mtx[1, 2]),     
-    xz_correlation = as.numeric(corr_mtx[1, 3]),     
-    yz_correlation = as.numeric(corr_mtx[2, 3]),     
-    fml = deparse(model) |> as.factor(),     
-    expect = ifelse(p_value < 0.05, 1, 0) |> as.factor(),     
-    ) |>    
-  ungroup()
-
-code_tbl <- crossing(V1 = c("signifpm10", NA),           
-                     V2 = c("smallsample", NA), 
-                     V3 = c("withtemp", NA)) |>      
-  mutate(col1 = paste0(V1,"_", V2, "_", V3))
-
-lookup_tbl <- tibble(   
-  value = crossing(x = c(1, 0), y = c(1, 0), z = c(1, 0)) |>      
-    mutate(col = paste0(x,"_", y, "_", z)) |> 
-    arrange(-x, -y, -z) |>      
-    pull(col),   
-  plan1 = unique(code_tbl$col1) )
-
-dt2 <- dt |>    
-  mutate(T1 = ifelse(sample_size < 200, 1, 0),          
-         T2 = ifelse(fml == "mortality ~ pm10 + temp", 1, 0),          
-         pp1 = paste0(expect, "_", T1, "_", T2),) |>    
-  left_join(lookup_tbl, by = c("pp1" = "value"))
-
-p0 <- dt |> ggplot() +   
-  geom_point(aes(x = p_value , y = coef), color = "grey", size = 0.5) +  
-  labs(x = "p-value", y = "coefficient") +   
-  theme_bw() +    
-  theme(aspect.ratio = 1, 
-        panel.grid.minor = element_blank())
-
-df0 <-  dt |> filter(sample_size == 500, xy_correlation == -0.005, 
-                     between(xz_correlation, -0.46, -0.43),
-                     yz_correlation == 0.4) |> arrange(expect)
-df2 <- df0 |> filter(fml == "mortality ~ pm10 + temp") # test for model spec
-
-p1 <- p0 + 
-  geom_point(aes(x = p_value , y = coef, color = expect), size = 0.3) +  
-  scale_color_brewer(palette = "Dark2", direction = -1) +
-  labs(title = "Coefficient vs. p-value") + 
-  theme(legend.position = 'none')
-
-p2 <- p0 + 
-  geom_point(data = df0, aes(x = p_value, y = coef, color = expect), size = 2) +
-  geom_path(data = df0, aes(x = p_value, y = coef), color = "black",
-            arrow = arrow(type = "open", angle = 30, 
-                          length = unit(0.1, "inches"))) +
-  ggrepel::geom_label_repel(
-    data = df0, aes(x = p_value, y = coef, label = fml, color = expect)
-    ) +
-  scale_color_brewer(palette = "Dark2", direction = -1) +
-  labs(title = "Effect of adding temperature")
-
-df00 <-  dt |> filter(xy_correlation == -0.005, 
-                     between(xz_correlation, -0.46, -0.43),
-                     yz_correlation == 0.4) |> arrange(expect)
-df <- df00 |> filter((fml == "mortality ~ pm10 + temp"))
-
-p3 <- p0 + 
-  geom_point(data = df, aes(x = p_value, y = coef, color = expect), size = 2) +
-  geom_path(data = df, aes(x = p_value, y = coef), color = "black",
-            arrow = arrow(type = "open", angle = 30, 
-                          length = unit(0.1, "inches"))) +
-  ggrepel::geom_label_repel(
-    data = df, aes(x = p_value, y = coef, label = sample_size, color = expect)
-    ) +
-  scale_color_brewer(palette = "Dark2", direction = -1) +
-  labs(title = "Effect of sample size")
-
-(p1 | p2) / (p3 | plot_spacer()) + 
-  plot_annotation(tag_levels = "a") + 
-  plot_layout(guides = 'collect') &
-  theme(legend.position = 'bottom') 
 ```
@@
 
-* There are cost and benefit on what granularity to set expectation.
-Setting them on every step of the analysis requires more work from the analysts while for complex analysis, the diagnostics may require divide the analysis into different components and diagnose it in a hierarchical structure.
-The method proposed in this paper should be applicable for both fine and coarse analysis.
+* There are cost and benefit on setting expectation on different granularity.
+At the lowest level, one may have a plan for each data entry and every data handling steps.
+This requires more work from the analysts and may not be practical in practice.
+For more complex analyses, analysts may divide the analysis into sections and set expectations for each.
+They can then focus on the specific sections flagged by the tests and sub-divide the sections to set expectation and diagnose the analysis in a hierarchical manner.
 

### 2024-11-13 _manuscript/_tex/index.tex
Section 3.2 toy example'

@@
 
-The expectations can be thought of as a set of unit tests used to
-validate the results of data analysis. By specifying a range of values
-for these tests, multiple versions of the dataset can be generated to
-satisfy different sets of plan expectations. This allows us to present
-what we called the ``result universe'' -- the complete set of possible
-results that can be obtained from one data analysis process. By
-visualizing the result universe, data analysis consumers can observe how
-changes in expectations affect the results and the range of alternative
-outcomes that could arise under different conditions. This enables them
-to evaluate the outcomes based on their own plan expectations and gain a
-broader perspective on how the actual results produced by analysts fit
-within this spectrum of possibilities, promoting transparency and trust
-in analysis.
-
 \subsection{A toy example}\label{a-toy-example}
 
-Let's think about a 5-day step count. You make a resolution to walk on
-average 5000 steps a day (your expectation) and using an app to record
-your step count. After 5 days, the app tells you've walked on average
-8000 steps.
+Consider a 30-day step count goal. Suppose you resolve to walk at least
+8,000 steps each day, using an app to record your daily step count. Your
+target average is 9,000 steps per day, with some ``lows'' day, where you
+walk around 4,000 steps and ``high'' days where you reach about 12,000
+steps. After 30 days, you check how many times your step count fall
+below 8,000 and aim for no more than five days under this threshold.
 
-It is easy to come up with reasons why an 8000 average step is resulted
-based on common sense:
+To simulation this data, three normal distributions with different means
+are used for the daily step counts: \(\mathcal{N}(4000, 200)\) for low
+days, \(\mathcal{N}(12000, 200)\) for high days, and
+\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
+days can be simulated from a Poisson distribution with \(\lambda = 4\).
+Figure~\ref{fig-step-count} displays the number of days with fewer than
+8,000 steps across 300 simulated 30-day periods.
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  you may run a 10k on day 1, resulting a high step count on the day
-  (outlier on the right).
-\item
-  you left your phone at home on day 3, resulting a zero or minimal step
-  count on the day (outlier on the left).
-\item
-  you may realise the step count may increase since you were in a hiking
-  trip in the last five days (average shift).
-\end{enumerate}
+In this scenario, the outcome expectation is that the number of days
+with a step count below 8,000 will be no more than five. To diagnose
+potential reasons why this outcome expectation might fail, we can
+establish a few plan expectations. For example, if the average step
+count is too low, this may suggest there are too many low days,
+potentially lead to an unexpected outcome. Similarly, we can also check
+the quantile of the step count, if more than a third of the days fall
+below 8,000, this could indicate an excess of low-count days.
+Additionally, we may may expect the standard deviation of the step count
+not to be overly large.
 
-Based on these reasons, you may devise a set of unit tests to check the
-step count data, i.e.~check the maximum and minimum step count, check
-the difference between each day.
+These considerations yield the following three unit tests as plan
+expectations:
 
@@
 \item
-  If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the
-  maximum check will flag the data for investigate the maximum. The
-  difference between days test will also flag the data
+  test1: the test fails if the mean step count is below 8,200
 \item
-  If the daily count looks like c(20000, 20000, 20000, 20000, 20000),
-  the maximum check will flag the data for investigate the maximum.
+  test2: the test fails if the 30th percentile of the step counts is
+  below 8,200
+\item
+  test3: the test fails if the standard deviation of the step
+  countsexceeds 2,500.
 \end{itemize}
 
-Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is
-flagged by the minimal tests but won't cause an average of 8000 average
-step.
+\phantomsection\label{cell-fig-step-count}
+\begin{figure}[H]
 
-The statistical procedure of averaging 5 numbers ``around 5000'' to get
-a mean of 5000 is \emph{consistent} meaning if all the numbers are
-around 5000, we are guaranteed to get a mean around 5000. We could
-devise 5 unit tests to check each number. Since you're more familiar
-with your daily life, you may realise the step count may increase since
-you were in a hiking trip in the last two days. This may prompt you to
-check the step count.
+\centering{
 
-In a data analysis, it is not practical to check every entry of the
-data, a similar strategy of devising tests to check for
+\includegraphics{index_files/figure-pdf/fig-step-count-1.pdf}
 
-\begin{itemize}
-\tightlist
-\item
-  The combination of unit tests are not unique
-\item
-  The unit tests provide guidance for diagnosing the results, but are
-  not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the
-  max diff test but receive a within expectation mean.
-\end{itemize}
+}
+
+\caption{\label{fig-step-count}Number of days with fewer than 8,000
+steps across 300 simulated 30-day periods. The orange bars indicate
+instances where the count exeeds five days, representing an unexpected
+outcome in this scenario.}
+
+\end{figure}%
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
 

### 2024-11-13 index.qmd
Section 3.2 toy example'

@@
 
-The expectations can be thought of as a set of unit tests used to validate the results of data analysis.
+<!-- The expectations can be thought of as a set of unit tests used to validate the results of data analysis.
 By specifying a range of values for these tests, multiple versions of the dataset can be generated to satisfy different sets of plan expectations.
@@
 This enables them to evaluate the outcomes based on their own plan expectations and gain a broader perspective on how the actual results produced by analysts fit within this spectrum of possibilities, promoting transparency and trust in analysis.
+-->
 
@@
 
-Let's think about a 5-day step count.
-You make a resolution to walk on average 5000 steps a day (your expectation) and using an app to record your step count.
-After 5 days, the app tells you've walked on average 8000 steps.
+Consider a 30-day step count goal.
+Suppose you resolve to walk at least 8,000 steps each day, using an app to record your daily step count.
+Your target average is 9,000 steps per day, with some "lows" day, where you walk around 4,000 steps and "high" days  where you reach about 12,000 steps.
+After 30 days, you check how many times your step count fall below 8,000 and aim for no more than five days under this threshold.
 
-It is easy to come up with reasons why an 8000 average step is resulted based on common sense:
+To simulation this data, three normal distributions with different means are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
+The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$.
+@fig-step-count displays the number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
 
-1. you may run a 10k on day 1, resulting a high step count on the day (outlier on the right).
-2. you left your phone at home on day 3, resulting a zero or minimal step count on the day (outlier on the left).
-3. you may realise the step count may increase since you were in a hiking trip in the last five days (average shift).
+In this scenario, the outcome expectation is that the number of days with a step count below 8,000 will be no more than five.
+To diagnose potential reasons why this outcome expectation might fail, we can establish a few plan expectations.
+For example, if the average step count is too low, this may suggest there are too many low days, potentially lead to an unexpected outcome.
+Similarly, we can also check the quantile of the step count, if more than a third of the days fall below 8,000, this could indicate an excess of low-count days.
+Additionally, we may may expect the standard deviation of the step count not to be overly large.
 
-Based on these reasons, you may devise a set of unit tests to check the step count data, i.e. check the maximum and minimum step count, check the difference between each day.
+These considerations yield the following three unit tests as plan expectations:
 
-- If the daily count looks like c(4000, 5000, 5500, 5500, 20000), the maximum check will flag the data for investigate the maximum. The difference between days test will also flag the data
-- If the daily count looks like c(20000, 20000, 20000, 20000, 20000), the maximum check will flag the data for investigate the maximum.
+  - test1: the test fails if the mean step count is below 8,200
+  - test2: the test fails if the 30th percentile of the step counts is below 8,200 
+  - test3: the test fails if the standard deviation of the step countsexceeds 2,500.
 
-Some part of the space is impossible: c(0, 4000, 5000, 5500, 5500) is flagged by the minimal tests but won't cause an average of 8000 average step.
+```{r}
+#| label: fig-step-count
+#| fig-cap: Number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
+The orange bars indicate instances where the count exeeds five days, representing an unexpected outcome in this scenario.
+set.seed(1234)
+step_count <- tibble(id = 1:300) |> 
+  rowwise() |> 
+  mutate(
+    small_n = rpois(1, 4),
+    large_n = rpois(1, 4),
+    norm_sd = rlogis(1, 800, scale = 100),
+    norm_n = 30 - small_n - large_n,
+    step = list(round(c(rnorm(small_n, 4000, 200), 
+                        rnorm(large_n, 12000, 200), 
+                        rnorm(norm_n, 9000, 300)
+                        ), 0)),
+    less_than_8000 = sum(step < 8000), 
+    unexpect = ifelse(less_than_8000 > 5, 1, 0),
+    test1 = ifelse(mean(step) < 8200, 1, 0),
+    test2 = ifelse(quantile(step, 0.3, na.rm = TRUE) < 8300, 1, 0),
+    test3 = ifelse(sd(step) > 2500, 1, 0),
+    ) |>
+  ungroup()
 
-The statistical procedure of averaging 5 numbers "around 5000" to get a mean of 5000 is *consistent* meaning if all the numbers are around 5000, we are guaranteed to get a mean around 5000.
-We could devise 5 unit tests to check each number.
-Since you're more familiar with your daily life, you may realise the step count may increase since you were in a hiking trip in the last two days.
-This may prompt you to check the step count.
+# step_count |> 
+#   dplyr::select(id, step) |> 
+#   unnest(step) |>
+#   mutate(day = rep(1:30, 300)) |> 
+#   ggplot(aes(x = step, y = id)) + 
+#   geom_point(alpha = 0.5) + 
+#   theme_bw()
 
-In a data analysis, it is not practical to check every entry of the data, a similar strategy of devising tests to check for
-
-- The combination of unit tests are not unique
-- The unit tests provide guidance for diagnosing the results, but are not red flags: c(2000, 2000, 5000, 8000, 8000) will likely to fail the max diff test but receive a within expectation mean.
+step_count |> 
+  ggplot(aes(x = less_than_8000, fill = as.factor(unexpect))) + 
+  geom_histogram(binwidth = 0.5) + 
+  scale_x_continuous(breaks = seq(0, 13, 1)) +
+  scale_fill_brewer(palette = "Dark2") + 
+  xlab("Number of days with step count less than 8000") + 
+  theme_bw() + 
+  theme(legend.position = "none")
+```
 

### 2024-11-15 _manuscript/_tex/index.tex
methodology

@@
 unexpected outcomes, which brings two criteria in the evaluation metric:
-accuracy and parsimony.
+accuracy and parsimony. Figure~\ref{fig-metric-calc} illustrates the
+workflow for calculating the metrics.
+
+\phantomsection\label{cell-fig-metric-calc}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics[width=6.73in,height=\textheight]{figures/metric-calc.png}
+
+}
+
+\caption{\label{fig-metric-calc}this is the cap}
+
+\end{figure}%
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
 
@@
 checking the data), whereas a false negative suggests the tests may lack
-sensitivity to unexpected outcomes. Since both plan and outcome
-expectations can be framed as unit tests with binary outcomes, one
-approach is to predict the outcome expectation based on the plan
-expectation. Performance of the tests can then be evaluated using
-precision and recall metrics from the confusion matrix:
+sensitivity to unexpected outcomes. To model the relationship between
+the plan and outcome expectation (binary-binary), a logic regression
+model is used \citep{ruczinski_logic_2003}. Originally developed for SNP
+microarray data, logic regression constructs Boolean combinations of
+binary variables to solve regression problems {[}more introduction on
+logic regression{]}. Compared to other tree-based methods or machine
+learning methods for binary-binary prediction, logic regression produces
+Boolean combinations, or meta rules, that combines unit tests to solve
+the prediction problem. {[}need rewrite here{]}. The performance of the
+tests can be evaluated using precision and recall metrics derived from
+the confusion matrix of the logic regression prediction
 
@@
 
-\begin{itemize}
-\item
-  explain mutual information
-\item
-  explain logic regression
-\item
-  explain combining precision, recall, and independence together through
-  ``means''
-\item
-  A logic regression (ref) is used to model the relationship between the
-  plan and outcome expectations. (justify the use of Logic regression)
-\end{itemize}
+To quantify redundancy, the concept of mutual information is used.
+Mutual information \(I(x, y)\) measures the amount of information shared
+between two random variables and is defined as the KL-distance
+\(D(p \parallel q)\) between the joint distribution of the two variables
+and the product of the marginal distributions:
+
+\[I(x,y) = D\big(p(x,y) \parallel p(x)p(y)\big) = \sum_x \sum_y p(x,y) \log \frac{p(x,y)}{p(x)p(y)}\]
+
+This concept extends naturally to multiple variables through total
+correlation {[}ref{]}, \(C(X_1, X_2, \cdots, X_n)\), which captures
+redundancy across a set of \(n\) variables:
+
+\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}\]
+
+A high mutual information value indicates redundancy among the tests,
+while a low value suggests that the tests are independent and provide
+unique information to diagnose the unexpected outcome. To standardize
+this measure, the total correlation \emph{per observation} is
+calculated, and an independence score, ranging between 0 and 1, is
+defined as 1 - mutual information.
+
+To combine precision, recall, and independence into a single metric,
+various mathematical means, such as arithmetic mean, harmonic mean, and
+quadratic mean, can be used. The differences among these means are
+minimal when the three metrics are similar. However, as the differences
+among the metrics increases, the harmonic mean tends to produce the
+smallest overall score, as it penalizes low values, while the quadratic
+mean tends to produce the largest score by rewarding higher values more.
+For simple interpretation of the score, the arithmetic mean is
+preferred, while in applications where the difference between precision,
+recall, and independence need to be penalized or rewarded more, the
+harmonic and quadratic mean should be considered.
 
@@
 
-\begin{itemize}
-\item
-  provide interpretation at different scenarios:
+In the step count example, we can use the logic regression model to
+evaluate the quality of the unit tests. The logic regression model is
+fitted to the three unit tests (test1, test2, test3) and the outcome
+expectation (unexpect) as the response variable. The model is then used
+to predict the outcome expectation based on the unit tests. The
+prediction is then compared to the actual outcome expectation to
+calculate the precision and recall of the tests. The independence of the
+tests is also calculated to assess the redundancy of the tests.
+Figure~\ref{fig-logic-reg} presents the suggested logic regression model
+as a combination of test 1 and test 3 with an OR operator.
 
-  \begin{enumerate}
-  \def\labelenumi{\arabic{enumi})}
-  \tightlist
-  \item
-    one test is flagged, the prediction is as expected:
-  \item
-    multiple tests are flagged, the prediction is unexpected,
-  \item
-    no test is flagged, the prediction is unexpected,
-  \item
-    no test is flagged, the prediction is as expected
-  \end{enumerate}
-\end{itemize}
+Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
+independence for the three individual tests and the combined test rule
+(test1 OR test3) from the logic regression. The harmonic and arithmetic
+means are included to evaluate the quality of the unit tests in
+diagnosing unexpected step counts. The results show that the combined
+test rule (test1 OR test3) has the highest precision, recall, and
+independence, suggesting that it is the most effective test for
+diagnosing unexpected step counts. We also include the metric calculated
+from fitting a regression tree model to the data to compare the
+performance of the logic regression model. The regression tree produces
+a similar model of first split on test1 and then split on test3, and
+results in the same accuracy and overall score as the logic regression
+model.
+
+\phantomsection\label{cell-fig-logic-reg}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{index_files/figure-pdf/fig-logic-reg-1.pdf}
+
+}
+
+\caption{\label{fig-logic-reg}Logic regression model fitted to the three
+unit tests (test1, test2, test3) and the outcome expectation (unexpect)
+as the response variable. The model suggests using an OR rule to combine
+test1 and test3 to predict the outcome expectation.}
+
+\end{figure}%
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2424}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1515}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1061}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1970}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+
+\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
+parsimony (independence) metrics for each individual unit test and for
+the combined test rule (test1 OR test3) derived from the logic
+regression model. The harmonic and arithmetic means of the three metrics
+are included to evaluate the quality of the unit tests in diagnosing
+unexpected step counts (more than five days with fewer than 8,000
+steps).}
+
+\tabularnewline
+
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+tests
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+precision
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+recall
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+independence
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+harmonic
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+arithmetic
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+test1 & 0.48 & 0.96 & 1 & 0.73 & 0.82 \\
+test2 & 0.21 & 1.00 & 1 & 0.45 & 0.74 \\
+test3 & 0.59 & 0.80 & 1 & 0.76 & 0.80 \\
+test1 OR test3 & 0.82 & 0.84 & 1 & 0.88 & 0.89 \\
+regression tree & 0.82 & 0.84 & 1 & 0.88 & 0.89 \\
+
+\end{longtable}
+
+\textsubscript{Source:
+\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
+Notebook}}
 

### 2024-11-15 index.qmd
methodology

@@
 library(ggh4x)
+library(adtoolbox)
+library(broom)
+library(knitr)
+library(rpart)
+library(infotheo)
 pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
@@
 One set of tests is considered better than another if a small set of tests can reliably detect unexpected outcomes, which brings two criteria in the evaluation metric: accuracy and parsimony.
+@fig-metric-calc illustrates the workflow for calculating the metrics.
+
+```{r}
+#| label: fig-metric-calc
+#| fig-cap: "this is the cap"
+knitr::include_graphics(here::here("figures", "metric-calc.png"))
+```
 
@@
 A false positive can indicate (caution or skepticism on checking the data), whereas a false negative suggests the tests may lack sensitivity to unexpected outcomes.
-Since both plan and outcome expectations can be framed as unit tests with binary outcomes, one approach is to predict the outcome expectation based on the plan expectation.
-Performance of the tests can then be evaluated using precision and recall metrics from the confusion matrix:
+To model the relationship between the plan and outcome expectation (binary-binary), a logic regression model is used [@ruczinski_logic_2003].
+Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems [more introduction on logic regression].
+Compared to other tree-based methods or machine learning methods for binary-binary prediction, logic regression produces Boolean combinations, or meta rules, that combines unit tests to solve the prediction problem.
+[need rewrite here].
+The performance of the tests can be evaluated using precision and recall metrics derived from the confusion matrix of the logic regression prediction
 
@@
 
-- explain mutual information
-- explain logic regression 
-- explain combining precision, recall, and independence together through "means"
+To quantify redundancy, the concept of mutual information is used.
+Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the KL-distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:
 
-* A logic regression (ref) is used to model the relationship between the plan and outcome expectations. (justify the use of Logic regression)
+$$I(x,y) = D\big(p(x,y) \parallel p(x)p(y)\big) = \sum_x \sum_y p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$$
+
+This concept extends naturally to multiple variables through total correlation [ref], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
+
+$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}$$
+
+A high mutual information value indicates redundancy among the tests, while a low value suggests that the tests are independent and provide unique information to diagnose the unexpected outcome.
+To standardize this measure, the total correlation *per observation* is calculated, and an independence score, ranging between 0 and 1, is defined as 1 - mutual information.
+
+To combine precision, recall, and independence into a single metric, various mathematical means, such as arithmetic mean, harmonic mean, and quadratic mean, can be used.
+The differences among these means are minimal when the three metrics are similar.
+However, as the differences among the metrics increases, the harmonic mean tends to produce the smallest overall score, as it penalizes low values, while the quadratic mean tends to produce the largest score by rewarding higher values more.
+For simple interpretation of the score, the arithmetic mean is preferred, while in applications where the difference between precision, recall, and independence need to be penalized or rewarded more, the harmonic and quadratic mean should be considered.
 
@@
   4) no test is flagged, the prediction is as expected
+
+```{r}
+#| label: fig-logic-reg
+#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable.
+The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
+fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
+plot(fit)
+```
+
+```{r}
+#| label: tbl-logic-reg
+list(tibble(.fitted = step_count$test1, unexpect = step_count$unexpect),
+     tibble(.fitted = step_count$test2, unexpect = step_count$unexpect),
+     tibble(.fitted = step_count$test3, unexpect = step_count$unexpect)
+     ) |>
+  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
+            mutate(independence = 1) |> 
+            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "tests") |> 
+  mutate(tests = c("test1", "test2", "test3")) |>
+  bind_rows(augment(fit) |> 
+              calc_miscla_rate(unexpect, .fitted) |> 
+              calc_independence() |> 
+              calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
+              mutate(tests = "test1 OR test3") |> 
+              dplyr::select(-overlapping)) |>
+  kable(digits = 3, 
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual unit test and for the combined test rule (test1 OR test3) derived from the logic regression model.
+The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the unit tests in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
+```
 

### 2024-11-15 _manuscript/_tex/index.tex
toy example revisit

@@
 \endlastfoot
-test1 & 0.48 & 0.96 & 1 & 0.73 & 0.82 \\
-test2 & 0.21 & 1.00 & 1 & 0.45 & 0.74 \\
-test3 & 0.59 & 0.80 & 1 & 0.76 & 0.80 \\
-test1 OR test3 & 0.82 & 0.84 & 1 & 0.88 & 0.89 \\
-regression tree & 0.82 & 0.84 & 1 & 0.88 & 0.89 \\
+test1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 \\
+test2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 \\
+test3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 \\
+test1 OR test3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 \\
+regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 \\
 

### 2024-11-15 index.qmd
toy example revisit

@@
 
-* provide interpretation at different scenarios:
+In the step count example, we can use the logic regression model to evaluate the quality of the unit tests.
+The logic regression model is fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable.
+The model is then used to predict the outcome expectation based on the unit tests.
+The prediction is then compared to the actual outcome expectation to calculate the precision and recall of the tests.
+The independence of the tests is also calculated to assess the redundancy of the tests.
+@fig-logic-reg presents the suggested logic regression model as a combination of test 1 and test 3 with an OR operator.
 
-  1) one test is flagged, the prediction is as expected:
-  2) multiple tests are flagged, the prediction is unexpected,
-  3) no test is flagged, the prediction is unexpected,
-  4) no test is flagged, the prediction is as expected
+@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
+The harmonic and arithmetic means are included to evaluate the quality of the unit tests in diagnosing unexpected step counts.
+The results show that the combined test rule (test1 OR test3) has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected step counts.
+We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
+The regression tree produces a similar model of first split on test1 and then split on test3, and results in the same accuracy and overall score as the logic regression model.
+
+<!-- * provide interpretation at different scenarios: -->
+
+  <!--   1) one test is flagged, the prediction is as expected:  -->
+  <!--   2) multiple tests are flagged, the prediction is unexpected,  -->
+  <!--   3) no test is flagged, the prediction is unexpected,  -->
+  <!--   4) no test is flagged, the prediction is as expected -->
 
@@
 plot(fit)
+```
+
+```{r}
+fit_regtree <- rpart(unexpect ~ ., data = step_count |> dplyr::select(unexpect:test3), method = "class", 
+                     control = rpart.control(cp = 0.01)) 
+pred_vec <- predict(fit_regtree) |> as_tibble() |> mutate(.fitted = ifelse(`0` > `1`, 0, 1)) |> dplyr::select(.fitted)
+regtree_df <- step_count |> dplyr::select(unexpect) |> bind_cols(pred_vec) |> 
+  calc_miscla_rate(unexpect, .fitted) |> 
+  mutate(overlapping = multiinformation(step_count[,c(9, 11)])/nrow(step_count), 
+         independence = 1 - overlapping) |> 
+  calc_metrics(metrics = c("harmonic", "arithmetic"))
 ```
@@
               dplyr::select(-overlapping)) |> 
+  bind_rows(regtree_df |> 
+              dplyr::select(-overlapping) |> 
+              mutate(tests = "regression tree")) |>
   kable(digits = 3, 

### 2024-12-03 _manuscript/_tex/index.tex
working on the linear regression application

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \makeatletter
@@
 
-\section{Analysis plan}\label{sec-plan}
+\section{xxxxx (not analysis plan)}\label{sec-plan}
 
@@
 
+TODO: add one or two scenarios on why step count could go up or down: 1)
+you take off your watch in the middle of the day, 2) have an workout
+
+TODO: phrase it as Researcher measuring someone's step count rather than
+measuring your own step count
+
 These considerations yield the following three unit tests as plan
@@
 
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
-
 \section{Method}\label{method}
@@
 \end{figure}%
-
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
 
@@
 
+top-down and bottom up (regression tree + logic tree): more naturally
+useful summary of the say it is organized. (put down the plot)
+
 \phantomsection\label{cell-fig-logic-reg}
@@
 \end{figure}%
-
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
-
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
 
@@
 
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
-
 \section{Applications}\label{sec-examples}
 
-Three examples are presented to illustrate how the concept of analysis
-plan can be applied to data analysis. {[}toy example{]}.
-Section~\ref{sec-linear-reg} illustrates how constructing the result
-universe in a linear regression model of PM10 on mortality can help
-understand the impact of sample size, model specification, and variable
-correlation structure on data analysis. {[}example three{]}
+A regression example is produced to illustrate the test selection
+process for analyzing the effect of PM10 on mortality. The example
+demonstrates how the test process can be used to select cutoff value in
+the unit tests and how the test can be used to diagnose the unexpected
+results.
 
@@
 Consider a linear regression model to study the effect of PM10 on
-mortality (provide context of using PM10 to study mortality). Analysts
-may expect a significant (p-value \(\le\) 0.05) PM10 coefficient in the
-linear model from the literature. This is the \emph{outcome
-expectation}. There are multiple factors that can affect the outcome
-expectation of linear regression, which here is called \emph{plan
-expectation}, for example, 1) sample size, 2) model specification, and
-3) correlation structure between variables. Adequate sample size is
-required to achieve the desired power to detect the significance of PM10
-on mortality. Temperature is often an important confounder to consider
-in such study (add reference). From some domain knowledge, an analyst
-may expect that the significance of PM10 coefficient can be attained by
-adding temperature to the model. Analysts may also expect certain
-correlation structure between PM10, temperature, and mortality, and the
-distribution of each variable.
+mortality (provide context of using PM10 to study mortality). From the
+literature, analysts may expect a significant (p-value \(\le\) 0.05)
+PM10 coefficient in the linear model after considering the temperature
+confounding {[}TODO add reference{]}. This expectation can be framed
+into a unit test: the test fails, labelled as 1, if the p-value of the
+PM10 coefficient is greater than 0.05. And we have devised three unit
+tests to consider the 1) sample size, 2) strength of correlation between
+mortality and PM10, and 3) strength of correlation between mortality and
+temperature. {[}sample size matters because \ldots, correlation between
+mortality and PM10 matters because\ldots{]}. However, it is not clear
+how to set the cutoff value for these tests to determine a failure. Here
+a list of tests are created in Table~\ref{tbl-tests}.
+
+\begin{longtable}[]{@{}l@{}}
+\caption{A list of unit tests created for the linear regression model of
+mortality on PM10 and temperature. The tests are based on the sample
+size, correlation between mortality and PM10, and correlation between
+mortality and temperature. The cutoff values for each test are specified
+to determine a failure.}\label{tbl-tests}\tabularnewline
+\toprule\noalign{}
+the test fails if \ldots{} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+the test fails if \ldots{} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+sample size less than or equal to 200 \\
+sample size less than or equal to 400 \\
+sample size less than or equal to 600 \\
+sample size less than or equal to 800 \\
+mortality-PM10 correlation less than -0.03 \\
+mortality-PM10 correlation less than -0.04 \\
+mortality-PM10 correlation less than -0.05 \\
+mortality-PM10 correlation less than -0.06 \\
+mortality-temperature correlation greater than -0.3 \\
+mortality-temperature correlation greater than -0.35 \\
+mortality-temperature correlation greater than -0.4 \\
+mortality-temperature correlation greater than -0.45 \\
+\end{longtable}
+
+To generate replicate of the data, we first simulate the correlation
+matrix for the three variables: mortality, PM10, and temperature with
+the following vector:
 
@@
 \item
-  add a paragraph to describe the simulation process
+  mortality and PM10: -0.03 to -0.1
+\item
+  mortality and temperature: -0.2 to -0.4
+\item
+  PM10 and temperature: 0.2 to 0.4
 \end{itemize}
 
-\textsubscript{Source:
-\href{https://huizezhang-sherry.github.io/paper-analysis-plan/index.qmd.html}{Article
-Notebook}}
+We then use a Gaussian copula to generate a multivariate normal
+distribution based on the specified correlation matrix and sample size.
+The multivariate normal distribution is transformed using the normal
+CDF, followed by the inverse CDF of the assumed distributions of the
+three variables. Various distributions are fitted to the variables,
+including poisson and negative binomial for mortality; gamma,
+log-normal, exponential, weibull, and normal for pm10 and temperature;
+and beta for pm10 after rescaling the data to \([0-1]\). AIC is used to
+determine the best distribution fit for each variable with qq plot
+presented in Figure~\ref{fig-dist-fit} to evaluate the fit. The
+following distributions and parameters, estimated from the original
+data, are used to simulate the three variables:
+
+\begin{itemize}
+\tightlist
+\item
+  mortality, negative binomial distribution with size = 74 and mu = 183
+\item
+  PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then
+  multiply by 100
+\item
+  temperature, weibull distribution with shape = 3.8 and scale = 61
+\end{itemize}
+
+\phantomsection\label{cell-fig-dist-fit}
+\begin{figure}[H]
+
+\centering{
+
+\includegraphics{index_files/figure-pdf/fig-dist-fit-1.pdf}
+
+}
+
+\caption{\label{fig-dist-fit}QQ-plot of the distribution fit for
+mortality, PM10, and temperature based on the fitted distribution from
+the original data. The fitted distribution is compared to the observed
+data to assess the distribution fit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{index_files/figure-pdf/unnamed-chunk-8-1.pdf}
+
+}
+
+\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
+twelve unit tests and the outcome expectation (unexpect) as the response
+variable. The model suggests the relationship:
+(\(\text{cor}(\text{m}, \text{tmp}) > -0.35\) AND
+\(\text{smpl\_size} \le 600\)) OR
+\(\text{cor}(\text{m},  \text{PM10}) < - 0.05\)}
+
+\end{figure}%
+
+\begin{longtable}[]{@{}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0882}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1471}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1029}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1765}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1912}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1324}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1618}}@{}}
+
+\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
+parsimony (independence) metrics derived from the logic regression
+model, along with harmonic and arithmetic means, for individual unit
+tests (1: sample size, 2: mortality-PM10 correlation, 3:
+mortality-temperature correlation), and the combined test rule 4:
+(sample size AND mortality-temperature correlation) OR mortality-PM10
+correlation.}
+
+\tabularnewline
+
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedleft
+tests
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+precision
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+recall
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+overlapping
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+independence
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+harmonic
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+arithmetic
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+1 & 0.713 & 0.662 & 0 & 1 & 0.767 & 0.792 \\
+2 & 0.906 & 0.748 & 0 & 1 & 0.872 & 0.885 \\
+3 & 0.789 & 0.733 & 0 & 1 & 0.826 & 0.841 \\
+4 & 0.908 & 0.794 & 0 & 1 & 0.893 & 0.901 \\
+
+\end{longtable}
+
+A total of 12 tests with four cutoff values each for the three types of
+tests are used to fit the logic regression and calculate the precision
+and recall score for predictability of an unexpected outcome (p value
+\textless{} 0.05). Figure~\ref{fig-linear-reg-tree} shows the logic
+regression tree from the fitted model and Table~\ref{tbl-linear-reg}
+prints the numerical summary of three selected single test and their
+combined test found by the logic regression. The logic regression model
+is capable of picking up the best cutoff value for each type of tests
+and yields the following three tests:
+
+\begin{itemize}
+\tightlist
+\item
+  sample size less than or equal to 600
+\item
+  mortality-temperature correlation greater than -0.35
+\item
+  mortality-PM10 correlation less than -0.05
+\end{itemize}
+
+The combined test rule, (mortality-tempearture correlation AND sample
+size) OR mortality-PM10 correlation, has the highest precision, recall,
+and independence, suggesting that it is the most effective test for
+diagnosing unexpected results in the linear regression model given our
+assumption on the data. The tree structure suggests obtaining the
+mortality-PM10 correlation is as important as obtaining a combined
+relationship between mortality-temperature correlation and sample size
+in diagnosing unexpected results in the linear regression model. And
+based on our assumption of the variable distribution from the data
+simulation, we can obtain a precision of 0.908 and a recall of 0.794 for
+the unexpected outcome. These provides check points for the analysts to
+diagnose the unexpected results and to improve the model.
 

### 2024-12-03 index.qmd
working on the linear regression application

@@
    toc: true
+   notebook-links: false
 preamble: >
@@
 The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail.
-@sec-examples provides examples of analysis plan [more details].
+@sec-examples provides examples of analysis plan \[more details\].
 (need another section here or before examples?) @sec-conclusion concludes the paper.
@@
 
-* the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
-* The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.
+-   the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
+-   The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.
 
-# Analysis plan {#sec-plan}
+# xxxxx (not analysis plan) {#sec-plan}
 
@@
 Additionally, we may may expect the standard deviation of the step count not to be overly large.
+
+TODO: add one or two scenarios on why step count could go up or down: 1) you take off your watch in the middle of the day, 2) have an workout
+
+TODO: phrase it as Researcher measuring someone's step count rather than measuring your own step count
 
@@
 To model the relationship between the plan and outcome expectation (binary-binary), a logic regression model is used [@ruczinski_logic_2003].
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems [more introduction on logic regression].
+Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems \[more introduction on logic regression\].
 Compared to other tree-based methods or machine learning methods for binary-binary prediction, logic regression produces Boolean combinations, or meta rules, that combines unit tests to solve the prediction problem.
-[need rewrite here].
+\[need rewrite here\].
 The performance of the tests can be evaluated using precision and recall metrics derived from the confusion matrix of the logic regression prediction
 
-  * precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
-  * recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)
+-   precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
+-   recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)
 
@@
 
-This concept extends naturally to multiple variables through total correlation [ref], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
+This concept extends naturally to multiple variables through total correlation \[ref\], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
 
@@
 
+top-down and bottom up (regression tree + logic tree): more naturally useful summary of the say it is organized. (put down the plot)
+
 <!-- * provide interpretation at different scenarios: -->
 
-  <!--   1) one test is flagged, the prediction is as expected:  -->
-  <!--   2) multiple tests are flagged, the prediction is unexpected,  -->
-  <!--   3) no test is flagged, the prediction is unexpected,  -->
-  <!--   4) no test is flagged, the prediction is as expected -->
+<!--   1) one test is flagged, the prediction is as expected:  --> <!--   2) multiple tests are flagged, the prediction is unexpected,  --> <!--   3) no test is flagged, the prediction is unexpected,  --> <!--   4) no test is flagged, the prediction is as expected -->
 
@@
 #| label: fig-logic-reg
+#| fig.width: 6
+#| fig.heigh: 2
 #| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable.
@@
 fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
-plot(fit)
+plot_logicreg(fit)
 ```
@@
 
-Three examples are presented to illustrate how the concept of analysis plan can be applied to data analysis.
-[toy example].
-@sec-linear-reg illustrates how constructing  the result universe in a linear regression model of PM10 on mortality can help understand the impact of sample size, model specification, and variable correlation structure on data analysis.
-[example three]
+A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality.
+The example demonstrates how the test process can be used to select cutoff value in the unit tests and how the test can be used to diagnose the unexpected results.
 
@@
 Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality).
-Analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model from the literature.
-This is the *outcome expectation*.
-There are multiple factors that can affect the outcome expectation of linear regression, which here is called *plan expectation*, for example, 1) sample size, 2) model specification, and 3) correlation structure between variables.
-Adequate sample size is required to achieve the desired power to detect the significance of PM10 on mortality.
-Temperature is often an important confounder to consider in such study (add reference).
-From some domain knowledge, an analyst may expect that the significance of PM10 coefficient can be attained by adding temperature to the model.
-Analysts may also expect certain correlation structure between PM10, temperature, and mortality, and the distribution of each variable.
+From the literature, analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model after considering the temperature confounding \[TODO add reference\].
+This expectation can be framed into a unit test: the test fails, labelled as 1, if the p-value of the PM10 coefficient is greater than 0.05.
+And we have devised three unit tests to consider the 1) sample size, 2) strength of correlation between mortality and PM10, and 3) strength of correlation between mortality and temperature.
+\[sample size matters because ..., correlation between mortality and PM10 matters because...\].
+However, it is not clear how to set the cutoff value for these tests to determine a failure.
+Here a list of tests are created in @tbl-tests.
 
- * add a paragraph to describe the simulation process
+| the test fails if ... | 
+|--------------| 
+| sample size less than or equal to 200 |
+| sample size less than or equal to 400 |
+| sample size less than or equal to 600 |
+| sample size less than or equal to 800 |
+|  mortality-PM10 correlation less than -0.03 |
+|  mortality-PM10 correlation less than -0.04 |
+|  mortality-PM10 correlation less than -0.05 |
+|  mortality-PM10 correlation less than -0.06 |
+| mortality-temperature correlation greater than -0.3 | 
+| mortality-temperature correlation greater than -0.35 | 
+| mortality-temperature correlation greater than -0.4 | 
+| mortality-temperature correlation greater than -0.45 |
 
-```{r sim-linear-res}
+: A list of unit tests created for the linear regression model of mortality on PM10 and temperature.
+The tests are based on the sample size, correlation between mortality and PM10, and correlation between mortality and temperature.
+The cutoff values for each test are specified to determine a failure.
+{#tbl-tests}
+
+To generate replicate of the data, we first simulate the correlation matrix for the three variables: mortality, PM10, and temperature with the following vector:
+
+-   mortality and PM10: -0.03 to -0.1
+-   mortality and temperature: -0.2 to -0.4
+-   PM10 and temperature: 0.2 to 0.4
+
+We then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size.
+The multivariate normal distribution is transformed using the normal CDF, followed by the inverse CDF of the assumed distributions of the three variables.
+Various distributions are fitted to the variables, including poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$.
+AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit.
+The following distributions and parameters, estimated from the original data, are used to simulate the three variables:
+
+-   mortality, negative binomial distribution with size = 74 and mu = 183
+-   PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then multiply by 100
+-   temperature, weibull distribution with shape = 3.8 and scale = 61
+
+```{r}
+#| label: fig-dist-fit
+#| fig-cap: QQ-plot of the distribution fit for mortality, PM10, and temperature based on the fitted distribution from the original data.
+The fitted distribution is compared to the observed data to assess the distribution fit.
+#| fig.height: 4
+#| fig.width: 10
+aa2 <- fitdistrplus::fitdist(pm10$mortality, "nbinom")
+p1 <- tibble(fitted= rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
+       observed= pm10$mortality) %>%
+  ggplot(aes(sample = fitted)) +
+  stat_qq() +
+  stat_qq_line() + 
+  theme_bw() + 
+  theme(aspect.ratio = 1) + 
+  ggtitle("mortality: \nNB(size = 74, mu = 183)")
+
+aa6 <- fitdistrplus::fitdist(pm10$pm10[pm10$pm10 < 90]/100, "beta")
+p2 <- tibble(fitted = rbeta(557, shape1 = aa6$estimate[1], shape2 = aa6$estimate[2]) * 100,
+       observed = pm10$pm10[pm10$pm10 < 90]) %>%
+  ggplot(aes(sample = fitted)) +
+  stat_qq() +
+  stat_qq_line() + 
+  theme_bw() + 
+  theme(aspect.ratio = 1) + 
+  ggtitle("pm10: \n100 * beta(shp1 = 4.21, shp2 = 11.67)")
+
+aa4 <- fitdistrplus::fitdist(pm10$temp, "weibull")
+p3 <- tibble(fitted= rweibull(558, aa4$estimate[1], aa4$estimate[2]),
+       observed= pm10$temp) %>%
+  ggplot(aes(sample = fitted)) +
+  stat_qq() +
+  stat_qq_line() + 
+  theme_bw() + 
+  theme(aspect.ratio = 1) + 
+  ggtitle("temperature: \nweibull(shape = 3.8, scale = 61)")
+
+p1 | p2 | p3
+```
+
+```{r}
+# cor(pm10$pm10, pm10$mortality)
+# cor(pm10$temp, pm10$mortality)
+# cor(pm10$temp, pm10$pm10)
 # Generate correlation matrices 
-corr_grid <- expand.grid(seq(-0.01, -0.001, by = 0.001), 
-                         seq(-0.6, -0.2, 0.05), seq(0.2, 0.6, 0.05))
+corr_grid <- expand.grid(seq(0.01, 0.15, by = 0.02),                         
+                         seq(0.25, 0.5, by = 0.05),                           
+                         seq(0.25, 0.45, by = 0.05))
 
@@
 gen_corr_mtx <- function(r1, r2, r3) {   
-  cor_matrix <- matrix(c(1, r1, r2,                          
-                         r1, 1, r3,                          
-                         r2, r3, 1), nrow = 3, byrow = TRUE)          
+  # correlation between mortality and pm10 are negative - r1
+  # correlation between mortality and temp are negative - r2
+  cor_matrix <- matrix(c(1, -r1, -r2,                          
+                         -r1, 1, r3,                          
+                         -r2, r3, 1), nrow = 3, byrow = TRUE)          
   # Store the matrix in the list   
@@
 corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
-  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) 
-  }) 
+  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
 corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
-sample_size <- c(50, 100, 500, 1000) 
-model <- c("mortality ~ pm10 + temp", "mortality ~ pm10") |> map(as.formula)
+sample_size <- c(400, 600, 800) 
+model <- c("mortality ~ pm10 + temp") |> map(as.formula)
 
-generate_data <- function(n, mtx, seed = 123) {   
+generate_data <- function(n, mtx, seed = 123, outlier = FALSE) {   
   mu <- c(0, 0, 0)   
+  set.seed(seed)  
   data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
   U <- pnorm(data, mean = 0, sd = 1)   
-  set.seed(seed)      
-  tibble(mortality = qpois(U[,1], 182), # assume distribution
-         pm10 = qgamma(U[,2], shape = 6, rate = 0.2),           
-         temp = qnorm(U[,3], mean = 55, sd = 16))   
+  tibble(mortality = qnbinom(U[,1], size = 74, mu = 183),           
+         pm10 = qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100,           
+         temp = qweibull(U[,3], shape = 3.8, scale = 61))   
 }
@@
          fit = list(summary(lm(model, data))$coefficients))
+
+pm10_df <- res |>   
+  mutate(     
+    p_value = fit[2,4],     
+    coef = fit[2,1],     
+    mor_pm10_correlation = corr_mtx[1, 2],     
+    mor_temp_correlation = corr_mtx[1, 3],     
+    temp_pm10_yz_correlation = corr_mtx[2, 3],     
+    fml = deparse(model) |> as.factor(),     
+    unexpect = ifelse(p_value < 0.05, 0, 1),
+    #unexpect = ifelse(between(coef, 0, 0.15), 0, 1),
+    `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
+    `smpl_size <= 400` = ifelse(sample_size <= 400, 1, 0),
+    `smpl_size <= 600` = ifelse(sample_size <= 600, 1, 0),
+    `smpl_size <= 800` = ifelse(sample_size <= 800, 1, 0),
+    `cor(m, PM10) < -0.03` = ifelse(mor_pm10_correlation < -0.03, 1, 0),
+    `cor(m, PM10) < -0.04` = ifelse(mor_pm10_correlation < -0.04, 1, 0),
+    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
+    `cor(m, PM10) < -0.06` = ifelse(mor_pm10_correlation < -0.06, 1, 0),
+    `cor(m, tmp) > -0.3` = ifelse(mor_temp_correlation > -0.3, 1, 0),
+    `cor(m, tmp) > -0.35` = ifelse(mor_temp_correlation > -0.35, 1, 0),
+    `cor(m, tmp) > -0.4` = ifelse(mor_temp_correlation > -0.4, 1, 0),
+    `cor(m, tmp) > -0.45` = ifelse(mor_temp_correlation > -0.45, 1, 0)
+  ) |>    
+  ungroup()
+
 ```
+
+::: {#fig-linear-reg-tree}
+
+```{r}
+#| fig.width: 8
+#| fig.heigh: 3
+fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`cor(m, tmp) > -0.45`, 
+                                seed = 4, nleaves = 3)
+plot_logicreg(fit) 
+```
+<!-- The caption with some LaTeX like this $\lambda$. -->
+
+Logic regression model fitted to the twelve unit tests and the outcome expectation (unexpect) as the response variable. The model suggests the relationship:
+($\text{cor}(\text{m}, \text{tmp}) > -0.35$ AND $\text{smpl\_size} \le 600$) OR $\text{cor}(\text{m},  \text{PM10}) < - 0.05$
+:::
+
+```{r}
+#| label: tbl-linear-reg
+library(kableExtra)
+df <- list(
+  tibble(.fitted = pm10_df$`smpl_size <= 600`, 
+         unexpect = pm10_df$unexpect),
+  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.04`, 
+         unexpect = pm10_df$unexpect),
+  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.45`, 
+         unexpect = pm10_df$unexpect),
+  augment(fit)) |>
+  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
+            mutate(overlapping = 0, independence = 1) |> 
+            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")
+
+tbl <- tibble(tests = c(1,2,3,4)) |> bind_cols(df |> dplyr::select(-id)) 
+tbl |> 
+  kable(digits = 3, escape = FALSE,
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation), and the combined test rule 4: (sample size AND mortality-temperature correlation) OR mortality-PM10 correlation.")
+
+```
+
+A total of `r ncol(pm10_df) - ncol(res) - 7` tests with four cutoff values each for the three types of tests are used to fit the logic regression and calculate the precision and recall score for predictability of an unexpected outcome (p value \< 0.05).
+@fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of three selected single test and their combined test found by the logic regression.
+The logic regression model is capable of picking up the best cutoff value for each type of tests and yields the following three tests:
+
+-   sample size less than or equal to 600
+-   mortality-temperature correlation greater than -0.35
+-   mortality-PM10 correlation less than -0.05
+
+The combined test rule, (mortality-tempearture correlation AND sample size) OR mortality-PM10 correlation, has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected results in the linear regression model given our assumption on the data.
+The tree structure suggests obtaining the mortality-PM10 correlation is as important as obtaining a combined relationship between mortality-temperature correlation and sample size in diagnosing unexpected results in the linear regression model.
+And based on our assumption of the variable distribution from the data simulation, we can obtain a precision of 0.908 and a recall of 0.794 for the unexpected outcome.
+These provides check points for the analysts to diagnose the unexpected results and to improve the model.
 
@@
 
-* how to systematically simulate data is still unknown, sensitivity of the simulation to the results
+-   how to systematically simulate data is still unknown, sensitivity of the simulation to the results
 
-* plotting is a critical way to check data and they can still be frame into a unit test.
+-   plotting is a critical way to check data and they can still be frame into a unit test.
 it is a open problem to how to encode the visualization into the unit tests.
@@
 
-* currently no automated way to generate unit tests.
+-   currently no automated way to generate unit tests.
 It is interesting to see the automation of generating unit tests, although it requires the inputs from experts across a wide array of common scenarios.
 
-* There are cost and benefit on setting expectation on different granularity.
+-   There are cost and benefit on setting expectation on different granularity.
 At the lowest level, one may have a plan for each data entry and every data handling steps.
@@
 
-* (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
+-   (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
 For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2.

### 2024-12-04 _manuscript/_tex/index.tex
literature review on data analysis checks

@@
 
-\subsection{Data analysis checks in statistical
-software}\label{data-analysis-checks-in-statistical-software}
+\subsection{Data analysis checks}\label{data-analysis-checks}
 
-Currently, little has been done on how to computationally incorporate
-this diagnostic approach into data analysis workflow or software. Most
-of the diagnostic tools focus on defining user-specified rules, such as
-data quality checks or producing metrics to summarize model performance,
-as in model diagnostics. For example, the \texttt{assertr} package
-\citep{assertr} and the \texttt{pointblank} package \citep{pointblank}
-provide data validation tools to allow users to define data quality
-checks. In contrast, packages provide model checks tools like
-\texttt{performance} \citep{performance} and \texttt{yardstick}
-\citep{yardstick}, from the \texttt{tidymodels} \citep{tidymodels}
-ecosystem, offers goodness-of-fit, residual analysis, and model
-performance metrics.
+A substantial body of literature has addressed the definition of data
+quality \citep[more]{8642813} and developed frameworks, that includes
+dimensions, attributes, and measures to evaluate and improve data
+quality
+\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
+These frameworks are often used information system and database
+management and support business decision-making in the industry. For
+research purposes, high-quality data ensures the credibility of
+scientific findings and supports reproducibility and reusability in
+future studies {[}ref{]}. With the growing prevalence of open data in
+scientific research, the consumers or users of the data typically are no
+longer the data producers or collectors who have the full knowledge of
+data in hand, prompting more interest towards the data quality checks in
+the data analysis process. In R, there are some packages,
+like\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
+provides basic data screening and reporting, while another class of
+packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
+\citep{validate}, and \texttt{pointblank} \citep{pointblank}, focuses on
+providing data validation tools, allowing users to define customized
+data quality checks based on the applications.
 
-These packages provide the tools to conduct diagnostics but still don't
-reflect the mental process of how data analysts diagnose the unexpected
-output. For example, when an unexpected output occurs, an analyst may
-check on whether a column in the data frame is between two values for
-data quality. However, it is not documented what motivates the analyst
-to conduct this check -- whether it also applies to other researchers
-analyzing new data in the same contexts, whether it is a common practice
-in the field, or whether it is a reaction to this particular data or
-scenario. Currently, most of these assumptions are not discussed in the
-publication or captured by tools themselves. While one might be able to
-infer some of the mental process of the analysts from external sources,
-such as talking to them or watching screencast videos produced by the
-analysts e.g.~TidyTuesday screencast videos, these insights are not
-systematically documented or made machine-readable. This gap highlights
-the need for tools that provide higher level documentation of reasoning
-behind the checks, facilitating a more transparent and interpretable
-analysis process.
+While data validation focuses on verifying or subsetting data based on
+predefined rules \citep{zio2015methodology}, data analysis checks
+specify assumptions made on the data required for the analysis process
+to obtain the reported results. These checks often involve evaluating
+aspects such as data distribution and outliers, relationships between
+variables, and the validity of model assumptions.
+
+However, there are other checks that reflect the mental process of of
+how data analysts diagnose the unexpected output that are not documented
+in this process. For example, when an unexpected output occurs, an
+analyst may check on whether a column in the data frame is between two
+values for data quality. However, it is not documented what motivates
+the analyst to conduct this check -- whether it also applies to other
+researchers analyzing new data in the same contexts, whether it is a
+common practice in the field, or whether it is a reaction to this
+particular data or scenario. Currently, most of these assumptions are
+not discussed in the publication or captured by tools themselves. While
+one might be able to infer some of the mental process of the analysts
+from external sources, such as talking to them or watching screencast
+videos produced by the analysts e.g.~TidyTuesday screencast videos,
+these insights are not systematically documented or made
+machine-readable. A question to consider here is whether it is possible
+to deliver this type of quality checks as structured data in a data
+analysis. This gap highlights the need for tools that provide higher
+level documentation of reasoning behind the checks, facilitating a more
+transparent and interpretable analysis process.
 

### 2024-12-04 index.qmd
literature review on data analysis checks

@@
 
-## Data analysis checks in statistical software
+## Data analysis checks
 
-Currently, little has been done on how to computationally incorporate this diagnostic approach into data analysis workflow or software.
-Most of the diagnostic tools focus on defining user-specified rules, such as data quality checks or producing metrics to summarize model performance, as in model diagnostics.
-For example, the `assertr` package [@assertr] and the `pointblank` package [@pointblank] provide data validation tools to allow users to define data quality checks.
-In contrast, packages provide model checks tools like `performance` [@performance] and `yardstick` [@yardstick], from the `tidymodels` [@tidymodels] ecosystem, offers goodness-of-fit, residual analysis, and model performance metrics.
+A substantial body of literature has addressed the definition of data quality [@8642813, more] and developed frameworks, that includes dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
+These frameworks are often used information system and database management and support business decision-making in the industry.
+For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and reusability in future studies [ref].
+With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process.
+In R, there are some packages, like`skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g.
+`assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank], focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.
 
-These packages provide the tools to conduct diagnostics but still don't reflect the mental process of how data analysts diagnose the unexpected output.
+While data validation focuses on verifying or subsetting data based on predefined rules [@zio2015methodology], data analysis checks specify assumptions made on the data required for the analysis process to obtain the reported results.
+These checks often involve evaluating aspects such as data distribution and outliers, relationships between variables, and the validity of model assumptions.
+
+However, there are other checks that reflect the mental process of of how data analysts diagnose the unexpected output that are not documented in this process.
 For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality.
@@
 TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable.
+A question to consider here is whether it is possible to deliver this type of quality checks as structured data in a data analysis.
 This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.

### 2024-12-04 _manuscript/_tex/index.tex
refine introduction

@@
 \item
-  emphasize trustworthy data science since it is the theme of the
+  TODO: emphasize trustworthy data science since it is the theme of the
   special issue
@@
 
-In a data analysis, an experienced researcher can often quickly assess
-whether a result meets their expectation, while others may be unsure of
-what to anticipate from the analysis or how to recognize when results
-diverge form expected norms. (These expectations are often based on
-prior knowledge, domain expertise, or common sense and they can often be
-framed into a Boolean question that can be answered by a unit test. For
-example, in a linear regression model, one may expect the p-value of the
-coefficient to be less than 0.05. On top of the expectation on the final
-results, researchers may also have expectations on the intermediate
-steps of the analysis process. For example, one may expect the data
-doesn't contain outliers, etc. These expectations are often checked
-early on to prevent an unexpected propagate through the analysis
-process.)
+In data analysis, experienced researchers often rely on their prior
+knowledge or domain expertise to quickly assess whether results align
+with their expectations. When outcomes deviate, they can quickly iterate
+through the analysis cycle to identify potential issues, refine methods,
+or revise assumptions. However, this iterative process they employ is
+rarely explicit, documented, or discussed in publications. As a result,
+readers are typically presented with the final outcomes of the analysis
+cycle where the results and expectations are aligned -- achieved either
+by refining the analysis or updating the expectations based on
+statistical evidence. These missing pieces of information provides
+little guidance for diagnosing issues in the analysis when the same
+methodology is applied to a new dataset that produces different
+outcomes. Similarly, when researchers with different background
+knowledge it becomes unclear whether variations in results arise from
+differences in the assumptions or in the analysis itself.
 
-This expectation-setting process is often implicit and rarely documented
-or discussed in publications. (There are most publication requires open
-source code for publication, however, it is no instruction or guidance
-on checking \ldots). Yet, making them explicit is crucial for 1) helping
-junior researchers interpret results and diagnose the unexpected, 2)
-providing checkpoints that support independent replication or
-application of methods to new data, 3) aligning assumption across
-researchers from different backgrounds.
+In this paper, we frame these expectations as checks within an
+\emph{analysis plan}, which allows us to examine these implicit
+assumptions made about the data during analysis and to compare different
+checks for diagnosing unexpected outcomes. We then introduce a procedure
+to evaluate the quality of these checks, either individually or in
+combination through logic regression, based on their accuracy and
+redundancy across simulation of the original data. Accuracy is assessed
+using binary classification metrics, precision and recall, while
+redundancy is measured using mutual information. The proposed workflow
+offers a numerical guarantee, based on simulation, that the analysis
+will produce the expected results, assuming the data generating
+mechanism holds. {[}something about trustworthy data science{]}
 
-In this paper, we formalize these informal checks into an \emph{analysis
-plan} comprising the analysis steps along with the associated unit
-tests. This formalization allows us to examine assumptions made about
-the data during analysis and to compare different unit tests in
-diagnosing unexpected outcomes. We then introduce a procedure to assess
-the quality of these unit tests based on their accuracy and redundancy
-across simulated version of the original data. The accuracy is assessed
-using binary classification metrics -- precision and recall -- while
-redundancy is measured using mutual information. The workflow provides
-numerical guarantee of the consistency of the expectation and the data
-for the analysis plan. This means that given the assumption of the data
-generating mechanism, the analysis will provide the outcome we expect.
-In data analysis, analysts are not necessarily collecting the data and
-they can only make reasonable assumption about the quality of the data.
-This method we propose provides a numerical guarantee, based on
-simulation, that the analysis will produce the outcome we expect, given
-the assumption of the data generating mechanism. This provides
-transparency of the analysis and build trustworthy data analysis.
-
-The rest of the paper is organized as follows: Section~\ref{sec-plan}
-describes the concept of analysis plan in detail.
-Section~\ref{sec-examples} provides examples of analysis plan {[}more
-details{]}. (need another section here or before examples?)
+The rest of the paper is organized as follows:
+Section~\ref{sec-lit-review} reviews the concept of diagnosing
+unexpected outcomes and data quality checks. Section~\ref{sec-plan}
+introduces the concept of analysis plan, illustrated with a toy example
+on fitness step count. Section~\ref{sec-method} describes the procedure
+for selecting optimal check combinations to predict unexpected outcomes.
+Section~\ref{sec-examples} applies this selection procedure to a linear
+regression example examining the effect of PM10 on mortality.
+Section~\ref{sec-discussion} discusses a few key considerations and
 Section~\ref{sec-conclusion} concludes the paper.
 
-\section{Literature review}\label{literature-review}
+\section{Literature review}\label{sec-lit-review}
 
@@
 
+Yet, making them explicit is crucial for 1) helping junior researchers
+interpret results and diagnose the unexpected, 2) providing checkpoints
+that support independent replication or application of methods to new
+data, 3) aligning assumption across researchers from different
+backgrounds.
+
 \subsection{A toy example}\label{a-toy-example}
@@
 
-\section{Method}\label{method}
+\section{Method}\label{sec-method}
 
@@
 
-\section{Discussion}\label{discussion}
+\section{Discussion}\label{sec-discussion}
 

### 2024-12-04 index.qmd
refine introduction

@@
 
--   emphasize trustworthy data science since it is the theme of the special issue
+-   TODO: emphasize trustworthy data science since it is the theme of the special issue
 
-In a data analysis, an experienced researcher can often quickly assess whether a result meets their expectation, while others may be unsure of what to anticipate from the analysis or how to recognize when results diverge form expected norms.
-(These expectations are often based on prior knowledge, domain expertise, or common sense and they can often be framed into a Boolean question that can be answered by a unit test.
-For example, in a linear regression model, one may expect the p-value of the coefficient to be less than 0.05.
-On top of the expectation on the final results, researchers may also have expectations on the intermediate steps of the analysis process.
-For example, one may expect the data doesn't contain outliers, etc.
-These expectations are often checked early on to prevent an unexpected propagate through the analysis process.)
+In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations.
+When outcomes deviate, they can quickly iterate through the analysis cycle to identify potential issues, refine methods, or revise assumptions.
+However, this iterative process they employ is rarely explicit, documented, or discussed in publications.
+As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence.
+These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
+Similarly, when researchers with different background knowledge it becomes unclear whether variations in results arise from differences in the assumptions or in the analysis itself.
 
-This expectation-setting process is often implicit and rarely documented or discussed in publications.
-(There are most publication requires open source code for publication, however, it is no instruction or guidance on checking ...).
-Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumption across researchers from different backgrounds.
+In this paper, we frame these expectations as checks within an *analysis plan*, which allows us to examine these implicit assumptions made about the data during analysis and to compare different checks for diagnosing unexpected outcomes.
+We then introduce a procedure to evaluate the quality of these checks, either individually or in combination through logic regression, based on their accuracy and redundancy across simulation of the original data.
+Accuracy is assessed using binary classification metrics, precision and recall, while redundancy is measured using mutual information.
+The proposed workflow offers a numerical guarantee, based on simulation, that the analysis will produce the expected results, assuming the data generating mechanism holds.
+[something about trustworthy data science]
 
-In this paper, we formalize these informal checks into an *analysis plan* comprising the analysis steps along with the associated unit tests.
-This formalization allows us to examine assumptions made about the data during analysis and to compare different unit tests in diagnosing unexpected outcomes.
-We then introduce a procedure to assess the quality of these unit tests based on their accuracy and redundancy across simulated version of the original data.
-The accuracy is assessed using binary classification metrics -- precision and recall -- while redundancy is measured using mutual information.
-The workflow provides numerical guarantee of the consistency of the expectation and the data for the analysis plan.
-This means that given the assumption of the data generating mechanism, the analysis will provide the outcome we expect.
-In data analysis, analysts are not necessarily collecting the data and they can only make reasonable assumption about the quality of the data.
-This method we propose provides a numerical guarantee, based on simulation, that the analysis will produce the outcome we expect, given the assumption of the data generating mechanism.
-This provides transparency of the analysis and build trustworthy data analysis.
+The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks.
+@sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count.
+@sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes.
+@sec-examples applies this selection procedure to a linear regression example examining the effect of PM10 on mortality.
+@sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.
 
-The rest of the paper is organized as follows: @sec-plan describes the concept of analysis plan in detail.
-@sec-examples provides examples of analysis plan \[more details\].
-(need another section here or before examples?) @sec-conclusion concludes the paper.
-
-# Literature review
+# Literature review {#sec-lit-review}
 
@@
 By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and evaluate the validity of the results.
+
+Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumption across researchers from different backgrounds.
 
@@
 
-# Method
+# Method {#sec-method}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-12-05 _manuscript/_tex/index.tex
small changes

@@
 
+\section{Analysis validation checks}\label{sec-plan}
+
+\subsection{Framing expectations as
+checks}\label{framing-expectations-as-checks}
+
+Q: Whether we should formulate these concept with math notation?? A:
+only if it helps
+
 While data validation focuses on verifying or subsetting data based on
@@
 transparent and interpretable analysis process.
-
-\subsection{Unit tests}\label{unit-tests}
-
-Why develop unit tests in software engineering? Because it is difficult
-to predict the outcome of a program, especially when the program is
-complex. Because the loss associated with a program failure is costly.
-
-Existing tools for checks in data analysis including those check for
-data quality and model diagnostics. In software development,
-
-\begin{itemize}
-\tightlist
-\item
-  the \texttt{testthat} package \citep{testthat} provides the
-  infrastructure for testing R code. It allows users to write unit tests
-  for functions and packages.
-\item
-  The \texttt{assertthat} package \citep{assertthat} helps to write
-  assertion statements that are supposed to be true at a certain point
-  in the code for defensive programming.
-\end{itemize}
-
-\section{xxxxx (not analysis plan)}\label{sec-plan}
-
-\subsection{Framing checks into unit
-tests}\label{framing-checks-into-unit-tests}
-
-Q: Whether we should formulate these concept with math notation?? A:
-only if it helps
 
@@
 
-Yet, making them explicit is crucial for 1) helping junior researchers
-interpret results and diagnose the unexpected, 2) providing checkpoints
-that support independent replication or application of methods to new
-data, 3) aligning assumption across researchers from different
-backgrounds.
+Yet, making them explicit is crucial for
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  helping junior researchers interpret results and diagnose the
+  unexpected,
+\item
+  providing checkpoints that support independent replication or
+  application of methods to new data, and
+\item
+  aligning assumption across researchers from different backgrounds.
+\end{enumerate}
 

### 2024-12-05 index.qmd
small changes

@@
 
+# Analysis validation checks {#sec-plan}
+
+## Framing expectations as checks
+
+Q: Whether we should formulate these concept with math notation?? A: only if it helps
+
 While data validation focuses on verifying or subsetting data based on predefined rules [@zio2015methodology], data analysis checks specify assumptions made on the data required for the analysis process to obtain the reported results.
@@
 This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.
-
-## Unit tests
-
-Why develop unit tests in software engineering?
-Because it is difficult to predict the outcome of a program, especially when the program is complex.
-Because the loss associated with a program failure is costly.
-
-Existing tools for checks in data analysis including those check for data quality and model diagnostics. In software development,
-
--   the `testthat` package [@testthat] provides the infrastructure for testing R code. It allows users to write unit tests for functions and packages.
--   The `assertthat` package [@assertthat] helps to write assertion statements that are supposed to be true at a certain point in the code for defensive programming.
-
-# xxxxx (not analysis plan) {#sec-plan}
-
-## Framing checks into unit tests
-
-Q: Whether we should formulate these concept with math notation?? A: only if it helps
-
-<!-- The analysis plan described in this paper should be differentiated from the pre-specifies analysis plan document often used in biostatistics to specifies the hypothesis, data collection mechanism, statistical procedures etc of randomized experiments.
--->
 
@@
 
-Yet, making them explicit is crucial for 1) helping junior researchers interpret results and diagnose the unexpected, 2) providing checkpoints that support independent replication or application of methods to new data, 3) aligning assumption across researchers from different backgrounds.
+Yet, making them explicit is crucial for
+
+  1) helping junior researchers interpret results and diagnose the unexpected, 
+  2) providing checkpoints that support independent replication or application of methods to new data, and
+  3) aligning assumption across researchers from different backgrounds.
 

### 2024-12-05 _manuscript/_tex/index.tex
update the linear regression example

@@
 for selecting optimal check combinations to predict unexpected outcomes.
-Section~\ref{sec-examples} applies this selection procedure to a linear
-regression example examining the effect of PM10 on mortality.
+Section~\ref{sec-pm10-mortality} applies this selection procedure to a
+linear regression example examining the effect of PM10 on mortality.
 Section~\ref{sec-discussion} discusses a few key considerations and
@@
 process for analyzing the effect of PM10 on mortality. The example
-demonstrates how the test process can be used to select cutoff value in
-the unit tests and how the test can be used to diagnose the unexpected
-results.
+demonstrates how the test process can be used to select cutoff values in
+the unit tests and how the test can be used to diagnose an unexpected
+PM10 coefficient from the generalized linear model.
 
-\subsection{Linear regression}\label{sec-linear-reg}
+\subsection{Effect of PM10 on mortality}\label{sec-pm10-mortality}
 
-Consider a linear regression model to study the effect of PM10 on
-mortality (provide context of using PM10 to study mortality). From the
-literature, analysts may expect a significant (p-value \(\le\) 0.05)
-PM10 coefficient in the linear model after considering the temperature
-confounding {[}TODO add reference{]}. This expectation can be framed
-into a unit test: the test fails, labelled as 1, if the p-value of the
-PM10 coefficient is greater than 0.05. And we have devised three unit
-tests to consider the 1) sample size, 2) strength of correlation between
-mortality and PM10, and 3) strength of correlation between mortality and
-temperature. {[}sample size matters because \ldots, correlation between
-mortality and PM10 matters because\ldots{]}. However, it is not clear
-how to set the cutoff value for these tests to determine a failure. Here
-a list of tests are created in Table~\ref{tbl-tests}.
+Consider a generalized linear model (GLM) to study the effect of PM10 on
+mortality {[}TODO: provide more context of the mortality-PM10 study{]}.
+Analysts may expect a PM10 coefficient between {[}0, 0.005{]} after
+considering the temperature confounding {[}TODO: reference{]}. This
+expectation can be framed into a check that fails, labelled as 1, if the
+PM10 coefficient is outside the range {[}0, 0.005{]}. Multiple factors
+can impact the PM10 coefficient, such as the sample size, the strength
+of the correlation between mortality and PM10, and the strength of the
+correlation between mortality and temperature. Analysts may expect a
+reasonable sample size to ensure the reliability of the coefficient
+estimate. Outliers in the three variables can also leverage the
+coefficient. While these are possible factors that could affect the
+analysis result, it is not clear the cutoff values for these checks to
+determine a failure. Here a list of checks are created in
+Table~\ref{tbl-checks}.
 
 \begin{longtable}[]{@{}l@{}}
-\caption{A list of unit tests created for the linear regression model of
-mortality on PM10 and temperature. The tests are based on the sample
-size, correlation between mortality and PM10, and correlation between
-mortality and temperature. The cutoff values for each test are specified
-to determine a failure.}\label{tbl-tests}\tabularnewline
+\caption{A list of checks considered for the generalized linear model of
+mortality on PM10 and temperature. The checks are based on the sample
+size, correlation between mortality and PM10, correlation between
+mortality and temperature, and univariate outlier detection. Multiple
+cutoff values are specified for each check to determine a
+failure.}\label{tbl-checks}\tabularnewline
 \toprule\noalign{}
-the test fails if \ldots{} \\
+the check fails if \ldots{} \\
 \midrule\noalign{}
@@
 \toprule\noalign{}
-the test fails if \ldots{} \\
+the check fails if \ldots{} \\
 \midrule\noalign{}
@@
 \endlastfoot
-sample size less than or equal to 200 \\
-sample size less than or equal to 400 \\
-sample size less than or equal to 600 \\
-sample size less than or equal to 800 \\
-mortality-PM10 correlation less than -0.03 \\
-mortality-PM10 correlation less than -0.04 \\
-mortality-PM10 correlation less than -0.05 \\
-mortality-PM10 correlation less than -0.06 \\
-mortality-temperature correlation greater than -0.3 \\
-mortality-temperature correlation greater than -0.35 \\
-mortality-temperature correlation greater than -0.4 \\
-mortality-temperature correlation greater than -0.45 \\
+Sample size less than or equal to 200 \\
+Sample size less than or equal to 400 \\
+Sample size less than or equal to 600 \\
+Sample size less than or equal to 800 \\
+Mortality-PM10 correlation less than -0.03 \\
+Mortality-PM10 correlation less than -0.04 \\
+Mortality-PM10 correlation less than -0.05 \\
+Mortality-PM10 correlation less than -0.06 \\
+Mortality-temperature correlation greater than -0.3 \\
+Mortality-temperature correlation greater than -0.35 \\
+Mortality-temperature correlation greater than -0.4 \\
+Mortality-temperature correlation greater than -0.45 \\
+Outlier(s) are presented in the variable PM10 \\
+Outlier(s) are presented in the variable mortality \\
 \end{longtable}
@@
 To generate replicate of the data, we first simulate the correlation
-matrix for the three variables: mortality, PM10, and temperature with
-the following vector:
-
-\begin{itemize}
-\tightlist
-\item
-  mortality and PM10: -0.03 to -0.1
-\item
-  mortality and temperature: -0.2 to -0.4
-\item
-  PM10 and temperature: 0.2 to 0.4
-\end{itemize}
-
-We then use a Gaussian copula to generate a multivariate normal
-distribution based on the specified correlation matrix and sample size.
-The multivariate normal distribution is transformed using the normal
-CDF, followed by the inverse CDF of the assumed distributions of the
-three variables. Various distributions are fitted to the variables,
-including poisson and negative binomial for mortality; gamma,
-log-normal, exponential, weibull, and normal for pm10 and temperature;
-and beta for pm10 after rescaling the data to \([0-1]\). AIC is used to
-determine the best distribution fit for each variable with qq plot
-presented in Figure~\ref{fig-dist-fit} to evaluate the fit. The
-following distributions and parameters, estimated from the original
-data, are used to simulate the three variables:
-
-\begin{itemize}
-\tightlist
-\item
-  mortality, negative binomial distribution with size = 74 and mu = 183
-\item
-  PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then
-  multiply by 100
-\item
-  temperature, weibull distribution with shape = 3.8 and scale = 61
-\end{itemize}
+matrix of the three variables in a grid and then use a Gaussian copula
+to generate a multivariate normal distribution based on the specified
+correlation matrix and sample size. The multivariate normal distribution
+is transformed using the normal CDF before the inverse CDF of the
+assumed distributions of the three variables is applied. To determine
+the appropriate distribution of each variable, various distributions are
+fitted and compared. This includes poisson and negative binomial for
+mortality; gamma, log-normal, exponential, weibull, and normal for pm10
+and temperature; and beta for pm10 after rescaling the data to
+\([0-1]\). AIC is used to determine the best distribution fit for each
+variable with qq plot presented in Figure~\ref{fig-dist-fit} to evaluate
+the fit. AIC suggests a negative binomial distributio nfor mortality, a
+beta distribution for PM10 (multiple by 100 to obtain the original
+scale), and a Weibull distribution for temperature. To include the
+effect of outlier, we add a single outlier to the data for mortality and
+PM10 {[}more details{]}.
 
@@
 \endlastfoot
-1 & 0.713 & 0.662 & 0 & 1 & 0.767 & 0.792 \\
-2 & 0.906 & 0.748 & 0 & 1 & 0.872 & 0.885 \\
-3 & 0.789 & 0.733 & 0 & 1 & 0.826 & 0.841 \\
-4 & 0.908 & 0.794 & 0 & 1 & 0.893 & 0.901 \\
+1 & 0.087 & 0.215 & 0 & 1 & 0.175 & 0.434 \\
+2 & 0.988 & 0.610 & 0 & 1 & 0.822 & 0.866 \\
+3 & 0.392 & 0.581 & 0 & 1 & 0.569 & 0.658 \\
+4 & 0.649 & 0.641 & 0 & 1 & 0.732 & 0.763 \\
+5 & 0.760 & 0.880 & 0 & 1 & 0.869 & 0.880 \\
 
@@
 
-A total of 12 tests with four cutoff values each for the three types of
-tests are used to fit the logic regression and calculate the precision
-and recall score for predictability of an unexpected outcome (p value
-\textless{} 0.05). Figure~\ref{fig-linear-reg-tree} shows the logic
-regression tree from the fitted model and Table~\ref{tbl-linear-reg}
-prints the numerical summary of three selected single test and their
-combined test found by the logic regression. The logic regression model
-is capable of picking up the best cutoff value for each type of tests
-and yields the following three tests:
+A logic regression is fitted to predict whether the PM10 coefficient is
+unexpected (outside the range of {[}0, 0.005{]}) using the checks listed
+in Table~\ref{tbl-checks}. Precision, recall, and independence score,
+along with their harmonic and arithmetic mean are calculated.
+Figure~\ref{fig-linear-reg-tree} shows the logic regression tree from
+the fitted model and Table~\ref{tbl-linear-reg} prints the numerical
+summary of four selected single test and their combined test found by
+the logic regression. The logic regression model picks up the following
+cutoff value for each type of check:
 
@@
 \item
-  sample size less than or equal to 600
+  sample size \emph{larger than} 200
 \item
@@
 \item
-  mortality-PM10 correlation less than -0.05
+  mortality-PM10 correlation less than -0.04
+\item
+  mortality contain outliers that are detected by the univariate outlier
+  detection
 \end{itemize}
 
-The combined test rule, (mortality-tempearture correlation AND sample
-size) OR mortality-PM10 correlation, has the highest precision, recall,
-and independence, suggesting that it is the most effective test for
-diagnosing unexpected results in the linear regression model given our
-assumption on the data. The tree structure suggests obtaining the
-mortality-PM10 correlation is as important as obtaining a combined
-relationship between mortality-temperature correlation and sample size
-in diagnosing unexpected results in the linear regression model. And
-based on our assumption of the variable distribution from the data
-simulation, we can obtain a precision of 0.908 and a recall of 0.794 for
-the unexpected outcome. These provides check points for the analysts to
-diagnose the unexpected results and to improve the model.
+The tree structure suggests checking mortality-PM10 correlation and a
+sample size larger than 200 with an additional check of either outlier
+on mortality or correlation between mortality and temperature. This
+combined check rule generates a 0.76 precision and a 0.88 recall for
+predicting the unexpected PM10 coefficient. The single check,
+\texttt{cor(m,\ PM10)\ \textless{}\ -0.03}, is also powerful with a high
+precision of 0.988, but the low recall value of 0.61 suggests its high
+false positive rate, as compared to the combined rule suggested by the
+logic regression.
 

### 2024-12-05 index.qmd
update the linear regression example

@@
 @sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes.
-@sec-examples applies this selection procedure to a linear regression example examining the effect of PM10 on mortality.
+@sec-pm10-mortality applies this selection procedure to a linear regression example examining the effect of PM10 on mortality.
 @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.
@@
 A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality.
-The example demonstrates how the test process can be used to select cutoff value in the unit tests and how the test can be used to diagnose the unexpected results.
+The example demonstrates how the test process can be used to select cutoff values in the unit tests and how the test can be used to diagnose an unexpected PM10 coefficient from the generalized linear model.
 
-## Linear regression {#sec-linear-reg}
+## Effect of PM10 on mortality {#sec-pm10-mortality}
 
-Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality).
-From the literature, analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model after considering the temperature confounding \[TODO add reference\].
-This expectation can be framed into a unit test: the test fails, labelled as 1, if the p-value of the PM10 coefficient is greater than 0.05.
-And we have devised three unit tests to consider the 1) sample size, 2) strength of correlation between mortality and PM10, and 3) strength of correlation between mortality and temperature.
-\[sample size matters because ..., correlation between mortality and PM10 matters because...\].
-However, it is not clear how to set the cutoff value for these tests to determine a failure.
-Here a list of tests are created in @tbl-tests.
+Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study].
+Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference].
+This expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005].
+Multiple factors can impact the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
+Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate.
+Outliers in the three variables can also leverage the coefficient.
+While these are possible factors that could affect the analysis result, it is not clear the cutoff values for these checks to determine a failure.
+Here a list of checks are created in @tbl-checks.
 
-| the test fails if ... | 
+| the check fails if ... | 
 |--------------| 
-| sample size less than or equal to 200 |
-| sample size less than or equal to 400 |
-| sample size less than or equal to 600 |
-| sample size less than or equal to 800 |
-|  mortality-PM10 correlation less than -0.03 |
-|  mortality-PM10 correlation less than -0.04 |
-|  mortality-PM10 correlation less than -0.05 |
-|  mortality-PM10 correlation less than -0.06 |
-| mortality-temperature correlation greater than -0.3 | 
-| mortality-temperature correlation greater than -0.35 | 
-| mortality-temperature correlation greater than -0.4 | 
-| mortality-temperature correlation greater than -0.45 |
+| Sample size less than or equal to 200 |
+| Sample size less than or equal to 400 |
+| Sample size less than or equal to 600 |
+| Sample size less than or equal to 800 |
+| Mortality-PM10 correlation less than -0.03 |
+| Mortality-PM10 correlation less than -0.04 |
+| Mortality-PM10 correlation less than -0.05 |
+| Mortality-PM10 correlation less than -0.06 |
+| Mortality-temperature correlation greater than -0.3 | 
+| Mortality-temperature correlation greater than -0.35 | 
+| Mortality-temperature correlation greater than -0.4 | 
+| Mortality-temperature correlation greater than -0.45 | 
+| Outlier(s) are presented in the variable PM10 | 
+| Outlier(s) are presented in the variable mortality |
 
-: A list of unit tests created for the linear regression model of mortality on PM10 and temperature.
-The tests are based on the sample size, correlation between mortality and PM10, and correlation between mortality and temperature.
-The cutoff values for each test are specified to determine a failure.
-{#tbl-tests}
+: A list of checks considered for the generalized linear model of mortality on PM10 and temperature.
+The checks are based on the sample size, correlation between mortality and PM10, correlation between mortality and temperature, and univariate outlier detection.
+Multiple cutoff values are specified for each check to determine a failure.
+{#tbl-checks}
 
-To generate replicate of the data, we first simulate the correlation matrix for the three variables: mortality, PM10, and temperature with the following vector:
-
--   mortality and PM10: -0.03 to -0.1
--   mortality and temperature: -0.2 to -0.4
--   PM10 and temperature: 0.2 to 0.4
-
-We then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size.
-The multivariate normal distribution is transformed using the normal CDF, followed by the inverse CDF of the assumed distributions of the three variables.
-Various distributions are fitted to the variables, including poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$.
+To generate replicate of the data, we first simulate the correlation matrix of the three variables in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size.
+The multivariate normal distribution is transformed using the normal CDF before the inverse CDF of the assumed distributions of the three variables is applied.
+To determine the appropriate distribution of each variable, various distributions are fitted and compared.
+This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$.
 AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit.
-The following distributions and parameters, estimated from the original data, are used to simulate the three variables:
-
--   mortality, negative binomial distribution with size = 74 and mu = 183
--   PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then multiply by 100
--   temperature, weibull distribution with shape = 3.8 and scale = 61
+AIC suggests a negative binomial distributio nfor mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature.
+To include the effect of outlier, we add a single outlier to the data for mortality and PM10 [more details].
 
@@
 # cor(pm10$temp, pm10$pm10)
-# Generate correlation matrices 
-corr_grid <- expand.grid(seq(0.01, 0.15, by = 0.02),                         
+corr_grid <- expand.grid(seq(0.01, 0.2, by = 0.02),                         
                          seq(0.25, 0.5, by = 0.05),                           
@@
 
-# Function to compute correlation matrix for each combination 
 gen_corr_mtx <- function(r1, r2, r3) {   
@@
                          -r2, r3, 1), nrow = 3, byrow = TRUE)          
-  # Store the matrix in the list   
   if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
@@
 
-# Plan for parallel processing
 corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
@@
 corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
-sample_size <- c(400, 600, 800) 
-model <- c("mortality ~ pm10 + temp") |> map(as.formula)
+sample_size <- c(200, 400, 600, 800, 1000)
+outlier <- c(TRUE, FALSE)
 
@@
   U <- pnorm(data, mean = 0, sd = 1)
+
+  if (!outlier) {
    tibble(mortality = qnbinom(U[,1], size = 74, mu = 183),           
@@
          temp = qweibull(U[,3], shape = 3.8, scale = 61))    
+  } else{
+    pm10_vec <- qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100
+    tibble(
+      mortality = c(qnbinom(U[,1], size = 74, mu = 183)[-1],
+                    rnorm(n = 1, mean = 300, sd = 10)),           
+      pm10 = c(pm10_vec[-1], rnorm(n = 1, mean = 100, sd = 10)),           
+      temp = qweibull(U[,3], shape = 3.8, scale = 61)
+      )    
+  }
+
 }
@@
   mutate(id = row_number()) |> 
-  crossing(sample_size, model) |> 
+  crossing(sample_size, outlier) |> 
   rowwise() |>   
-  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx)),          
-         fit = list(summary(lm(model, data))$coefficients))
+  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx, outlier = outlier)),          
+         fit = list(summary(glm(mortality ~ pm10 + temp, family = "poisson", data))$coefficients))
 
@@
     temp_pm10_yz_correlation = corr_mtx[2, 3],     
-    fml = deparse(model) |> as.factor(),     
-    unexpect = ifelse(p_value < 0.05, 0, 1),
-    #unexpect = ifelse(between(coef, 0, 0.15), 0, 1),
+    unexpect = ifelse(between(coef, 0, 0.005), 0, 1),
     `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
@@
 
+pm10_df <- pm10_df |> 
+  rowwise() |> 
+  mutate(`PM10 outlier` = ifelse(any(scale(data$pm10) > 4), 1, 0),
+         `mortality_outlier` = ifelse(any(scale(data$mortality) > 4), 1, 0)
+         ) |> 
+  ungroup()
 ```
@@
 #| fig.heigh: 3
-fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`cor(m, tmp) > -0.45`, 
-                                seed = 4, nleaves = 3)
+fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
+                                seed = 6, nleaves = 4)
 plot_logicreg(fit) 
@@
 df <- list(
-  tibble(.fitted = pm10_df$`smpl_size <= 600`, 
+  tibble(.fitted = pm10_df$`smpl_size <= 200`, 
          unexpect = pm10_df$unexpect),
-  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.04`, 
+  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.03`, 
          unexpect = pm10_df$unexpect),
-  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.45`, 
+  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.35`, 
+         unexpect = pm10_df$unexpect),
+  tibble(.fitted = pm10_df$mortality_outlier, 
          unexpect = pm10_df$unexpect),
@@
 
-tbl <- tibble(tests = c(1,2,3,4)) |> bind_cols(df |> dplyr::select(-id)) 
+tbl <- tibble(tests = c(1,2,3,4, 5)) |> bind_cols(df |> dplyr::select(-id)) 
 tbl |> 
@@
 
-A total of `r ncol(pm10_df) - ncol(res) - 7` tests with four cutoff values each for the three types of tests are used to fit the logic regression and calculate the precision and recall score for predictability of an unexpected outcome (p value \< 0.05).
-@fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of three selected single test and their combined test found by the logic regression.
-The logic regression model is capable of picking up the best cutoff value for each type of tests and yields the following three tests:
+A logic regression is fitted to predict whether the PM10 coefficient is unexpected (outside the range of [0, 0.005]) using the checks listed in @tbl-checks.
+Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated.
+@fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of four selected single test and their combined test found by the logic regression.
+The logic regression model picks up the following cutoff value for each type of check:
 
--   sample size less than or equal to 600
+-   sample size *larger than* 200
 -   mortality-temperature correlation greater than -0.35
--   mortality-PM10 correlation less than -0.05
+-   mortality-PM10 correlation less than -0.04
+-   mortality contain outliers that are detected by the univariate outlier detection
 
-The combined test rule, (mortality-tempearture correlation AND sample size) OR mortality-PM10 correlation, has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected results in the linear regression model given our assumption on the data.
-The tree structure suggests obtaining the mortality-PM10 correlation is as important as obtaining a combined relationship between mortality-temperature correlation and sample size in diagnosing unexpected results in the linear regression model.
-And based on our assumption of the variable distribution from the data simulation, we can obtain a precision of 0.908 and a recall of 0.794 for the unexpected outcome.
-These provides check points for the analysts to diagnose the unexpected results and to improve the model.
+<!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->
+
+The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature.
+This combined check rule generates a `r round(as.numeric(df[5, "precision"]), 3)` precision and a `r round(as.numeric(df[5, "recall"]), 3)` recall for predicting the unexpected PM10 coefficient.
+The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.
 

### 2024-12-05 _manuscript/_tex/index.tex
formality

@@
 }
+
+README:
+
+\begin{itemize}
+\tightlist
+\item
+  \emph{Check for TODOs}
+\item
+  Literature review: Section \emph{diagnosing unexpected outcomes in
+  data analysis}
+\item
+  \emph{Discussion} and \emph{Conclusion}
+\item
+  In Section \emph{Application}: provide more context on the
+  PM10-mortality study and add reference of the {[}0, 0.005{]} PM10
+  coefficient
+\item
+  Roger's Orcid???
+\end{itemize}
 
@@
 
+TODO
+
 \begin{itemize}
 \item
-  how to systematically simulate data is still unknown, sensitivity of
-  the simulation to the results
+  how to systematically simulate data is still unknown
 \item
   plotting is a critical way to check data and they can still be frame
-  into a unit test. it is a open problem to how to encode the
-  visualization into the unit tests. Maybe a procedure like confirm plot
-  (this looks alright to you) and then press the button to continue
+  into checks. i.e.~the density/ histogram suggests there are outliers.
+  It is a open problem to how to encode the visualization into checks.
 \item
-  currently no automated way to generate unit tests. It is interesting
-  to see the automation of generating unit tests, although it requires
-  the inputs from experts across a wide array of common scenarios.
+  currently no automated way to generate checks. It is interesting to
+  see how check generation can be automated, although it requires the
+  inputs from experts across a wide array of common scenarios.
+\item
+  checks are also closely related to the concept of unit tests in
+  software engineering. While unit tests are designed to isolate and
+  test specific aspects of the code, it is difficult for analysis
+  validation check to do so.
 \item
@@
   and diagnose the analysis in a hierarchical manner.
-\item
-  (only mention it) Software testing relies on ``oracles'' to provide
-  the expected output necessary for verifying test results. For example,
-  to test whether the program correctly calculates 1 + 1, one need to
-  supply the correct answer, 2. However, establishing this ``correct''
-  output can sometimes be challenging, where obtaining a solution may be
-  difficult without the program itself. This situation leads to the
-  oracle problem \citep{barr2014oracle}. In data analysis, the similar
-  oracle problem can happen, as the ``truth'' of an outcome, the
-  expectation, depends on the underlying theory or interpretation. For
-  example, in a linear regression model, the significance of a
-  coefficient may be expected or unexpected based on the theory, making
-  it challenging for researcher with a different theory to assess the
-  results and the analysis.
 \end{itemize}
@@
 \section{Conclusion}\label{sec-conclusion}
+
+TODO
+
+\section{Acknowledgement}\label{acknowledgement}
+
+The article is created using Quarto \citep{Allaire_Quarto_2022} in R
+\citep{R}. The source code for reproducing the work reported in this
+paper can be found at:
+\url{https://github.com/huizezhang-sherry/paper-avc}.
 

### 2024-12-05 index.qmd
formality

@@
 ```
+
+README:
+
+* *Check for TODOs*
+- Literature review: Section *diagnosing unexpected outcomes in data analysis* 
+- *Discussion* and *Conclusion*
+- In Section *Application*: provide more context on the PM10-mortality study and add reference of the [0, 0.005] PM10 coefficient
+- Roger's Orcid???
 
@@
 
--   how to systematically simulate data is still unknown, sensitivity of the simulation to the results
+TODO
 
--   plotting is a critical way to check data and they can still be frame into a unit test.
-it is a open problem to how to encode the visualization into the unit tests.
-Maybe a procedure like confirm plot (this looks alright to you) and then press the button to continue
+-   how to systematically simulate data is still unknown
 
--   currently no automated way to generate unit tests.
-It is interesting to see the automation of generating unit tests, although it requires the inputs from experts across a wide array of common scenarios.
+-   plotting is a critical way to check data and they can still be frame into checks.
+i.e.
+the density/ histogram suggests there are outliers.
+It is a open problem to how to encode the visualization into checks.
+
+-   currently no automated way to generate checks. It is interesting to see how check generation can be automated, although it requires the inputs from experts across a wide array of common scenarios.
+
+- checks are also closely related to the concept of unit tests in software engineering.
+While unit tests are designed to isolate and test specific aspects of the code, it is difficult for analysis validation check to do so.
 
@@
 
--   (only mention it) Software testing relies on "oracles" to provide the expected output necessary for verifying test results.
-For example, to test whether the program correctly calculates 1 + 1, one need to supply the correct answer, 2.
-However, establishing this "correct" output can sometimes be challenging, where obtaining a solution may be difficult without the program itself.
-This situation leads to the oracle problem [@barr2014oracle].
-In data analysis, the similar oracle problem can happen, as the "truth" of an outcome, the expectation, depends on the underlying theory or interpretation.
-For example, in a linear regression model, the significance of a coefficient may be expected or unexpected based on the theory, making it challenging for researcher with a different theory to assess the results and the analysis.
-
 # Conclusion {#sec-conclusion}
 
+TODO
+
+# Acknowledgement
+
+The article is created using Quarto [@Allaire_Quarto_2022] in R [@R].
+The source code for reproducing the work reported in this paper can be
+found at: <https://github.com/huizezhang-sherry/paper-avc>.
+
 # References

### 2024-12-06 _manuscript/_tex/index.tex
change document class

@@
 \documentclass[
-]{jds}
+  12pt,
+]{interact}
 
 \usepackage{amsmath,amssymb}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 \usepackage{xcolor}
+\usepackage{orcidlink}
 \makeatletter
@@
   pdftitle={My wonderful paper},
+  pdfauthor={H. Sherry Zhang; Roger D. Peng},
   colorlinks=true,
@@
 \title{My wonderful paper}
-\author{}
-\date{}
+\author{H. Sherry Zhang$\textsuperscript{1}$, Roger D.
+Peng$\textsuperscript{1}$}
 
+\thanks{CONTACT: H. Sherry
+Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
+D.
+Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
 \begin{document}
+\captionsetup{labelsep=space}
 \maketitle
+\textsuperscript{1}  University of Texas at Austin,  
 \begin{abstract}
@@
 
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
-
+\setstretch{2}
 README:
@@
   coefficient
-\item
-  Roger's Orcid???
 \end{itemize}

### 2024-12-06 index.qmd
change document class

@@
 This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
+author:
+  - name: "H. Sherry Zhang"
+    affiliation: "University of Texas at Austin"
+    email: "huize.zhang@austin.utexas.edu"
+    attributes:
+      corresponding: true
+  - name: "Roger D. Peng"
+    affiliation: "University of Texas at Austin"
+    email: "roger.peng@austin.utexas.edu"
 cite-method: natbib
 bibliography: references.bib
-documentclass: jds
 format:
-  pdf:
+ tandf-pdf:
    keep-tex: true
-   toc: true
+   fontsize: 12pt
+   linestretch: 2
    notebook-links: false
@@
 - In Section *Application*: provide more context on the PM10-mortality study and add reference of the [0, 0.005] PM10 coefficient
-- Roger's Orcid???
 

### 2024-12-06 index.qmd
author

@@
 author:
-  - name: "H. Sherry Zhang"
-    affiliation: "University of Texas at Austin"
-    email: "huize.zhang@austin.utexas.edu"
+  - name: H. Sherry Zhang
+    email: huize.zhang@austin.utexas.edu
+    affiliations: 
+        - id: 1
+          name: University of Texas at Austin
+          department: Department of Statistics and Data Sciences
+          city: Texas
+          country: United States
+          postal-code: 78705
     attributes:
         corresponding: true
-  - name: "Roger D. Peng"
-    affiliation: "University of Texas at Austin"
-    email: "roger.peng@austin.utexas.edu"
+  - name: Roger D. Peng
+    email: roger.peng@austin.utexas.edu
+    affiliations: 
+        - id: 1
+          name: University of Texas at Austin
+          department: Department of Statistics and Data Sciences
+          city: Texas
+          country: United States
+          postal-code: 78705
 cite-method: natbib

### 2024-12-06 _manuscript/_tex/index.tex
author
rewrite Section 3.1

@@
 \maketitle
-\textsuperscript{1}  University of Texas at Austin,  
+\textsuperscript{1} Department of Statistics and Data
+Sciences, University of Texas at Austin, Texas, United States
 \begin{abstract}
@@
 knowledge or domain expertise to quickly assess whether results align
-with their expectations. When outcomes deviate, they can quickly iterate
-through the analysis cycle to identify potential issues, refine methods,
-or revise assumptions. However, this iterative process they employ is
-rarely explicit, documented, or discussed in publications. As a result,
-readers are typically presented with the final outcomes of the analysis
-cycle where the results and expectations are aligned -- achieved either
-by refining the analysis or updating the expectations based on
-statistical evidence. These missing pieces of information provides
-little guidance for diagnosing issues in the analysis when the same
-methodology is applied to a new dataset that produces different
+with their expectations. When a result falls outside of this interval,
+it prompts the researchers to investigate backwards on the data quality,
+the analysis steps, or the assumptions made during the analysis process.
+This mental process of where to diagnose unexpected outcomes is often
+difficult to trace and discuss in publications. As a result, readers are
+typically presented with the final outcomes of the analysis cycle where
+the results and expectations are aligned -- achieved either by refining
+the analysis or updating the expectations based on statistical evidence
+\citep{grolemund_cognitive_2014}. These missing pieces of information
+provides little guidance for diagnosing issues in the analysis when the
+same methodology is applied to a new dataset that produces different
 outcomes. Similarly, when researchers with different background
-knowledge it becomes unclear whether variations in results arise from
-differences in the assumptions or in the analysis itself.
+knowledge it becomes unclear whether discrepancies arise from differing
+expectations or from the use of statistical techniques.
+
+One might gain insight into analysts' thought process by speaking with
+them directly or watching screencast videos they produce, such as,
+TidyTuesday screencast videos. However, direct conversation are not
+scalable and may not be feasible while creating educational videos
+requires significant effort from the researchers. Ideally, there could
+be a way to make these expectations explicit and accessible to others.
+Even better, if the encoding is machine-readable, we could analyze these
+expectations and learn from the analysis -- whether the checks also
+apply to other researchers analyzing new data in the same context,
+whether they reflect commo common practices in the field, or whether
+they are sepcific to the data or analysis at hand.
 
@@
 
+They shape how we interpret the results and assess whether they are
+consistent with existing knowledge or indicate the need for updates.
+\citep{grolemund_cognitive_2014}
+
 \citep{peng_diagnosing_2021} describes three pedagogical exercises of
@@
 data in hand, prompting more interest towards the data quality checks in
-the data analysis process. In R, there are some packages,
-like\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
+the data analysis process. In R, there are some packages, like
+\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
 provides basic data screening and reporting, while another class of
 packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
-\citep{validate}, and \texttt{pointblank} \citep{pointblank}, focuses on
+\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
 providing data validation tools, allowing users to define customized
@@
 
-Q: Whether we should formulate these concept with math notation?? A:
-only if it helps
-
-While data validation focuses on verifying or subsetting data based on
-predefined rules \citep{zio2015methodology}, data analysis checks
-specify assumptions made on the data required for the analysis process
-to obtain the reported results. These checks often involve evaluating
-aspects such as data distribution and outliers, relationships between
-variables, and the validity of model assumptions.
-
-However, there are other checks that reflect the mental process of of
-how data analysts diagnose the unexpected output that are not documented
-in this process. For example, when an unexpected output occurs, an
-analyst may check on whether a column in the data frame is between two
-values for data quality. However, it is not documented what motivates
-the analyst to conduct this check -- whether it also applies to other
-researchers analyzing new data in the same contexts, whether it is a
-common practice in the field, or whether it is a reaction to this
-particular data or scenario. Currently, most of these assumptions are
-not discussed in the publication or captured by tools themselves. While
-one might be able to infer some of the mental process of the analysts
-from external sources, such as talking to them or watching screencast
-videos produced by the analysts e.g.~TidyTuesday screencast videos,
-these insights are not systematically documented or made
-machine-readable. A question to consider here is whether it is possible
-to deliver this type of quality checks as structured data in a data
-analysis. This gap highlights the need for tools that provide higher
-level documentation of reasoning behind the checks, facilitating a more
-transparent and interpretable analysis process.
-
-An analysis plan is a set of analysis steps combined with expectations.
-Expectations represent our belief about certain aspects of the analysis,
-independent of the analysis itself. It can be divided into two types:
-\emph{outcome expectation} and \emph{plan expectation}. Outcome
-expectation refers to what we anticipate from the main result of the
-analysis based on prior knowledge. They shape how we interpret the
-results and assess whether they are consistent with existing knowledge
-or indicate the need for updates \citep{grolemund_cognitive_2014}. For
-example, in public health, prior research shows the average increase in
-mortality rate per unit increase in PM10 is about 0.50\%
-\citep{liu2019ambient}. This serves as an expectation for similar future
-studies. Plan expectations concern the intermediate steps within the
-analysis rather than the final outcome. They serve as checkpoints to
-detect deflection in the analysis process For example, we may expect
-temporal data to be ordered with no gaps and duplicates, or expect that
-temperature will be a significant covariate in the linear regression
-model of PM10 on mortality.
-
-Experienced analysts often have implicit expectation about the outcome
-and rely on a few ``directional signs'' to check when the outcome
-deviate from those expectation. However, these expectations are rarely
-made explicit within the analysis workflow. This makes it challenging
-for consumers of the analysis to evaluate the results, since it becomes
-difficult to disentangle whether discrepancies arise from differing
-expectations or from the use of statistical technique, without running
-the analysis themselves. Non-expert analysts, lacking prior knowledge or
-instinct, may not have clear expectations of the results. This can lead
-to reduced confidence of the analysis and makes it more difficult and
-time-consuming to diagnose the cause of the deviation when the results
-don't align with expectations. By explicitly formulating these
-expectations, an analysis plan can guide the analysis process,
-facilitate the communication and evaluate the validity of the results.
-
-Yet, making them explicit is crucial for
+Expectations represent our beliefs about certain aspects of the
+analysis, independent of the analysis itself. When outcomes deviate from
+these expectations, analysts often revisit the analysis cycle to
+identify potential issues, refine methods, or revise assumptions.
+Experienced analysts can typically identify issues quickly and correct
+them on the spot, but they often do so without discussing the underlying
+reasoning, making it harder for less experienced researchers to learn
+and master these skills. Here, we introduce the concept of
+\emph{analysis validation checks}, which frames these expectations or
+assumptions as explicit checks that returns a TRUE or FALSE output given
+the data analyzed at hand. Inspired by \emph{data validation checks}
+\citep{validate}, where ensure datasets meet expected format and
+quality, analysis validation checks reverse the approach: they validate
+the assumptions about the data necessary for the analysis to produce the
+reported results. This allows the concept to encompass a broad range of
+checks, such as data quality (i.e.~missing data, how the data is
+structured), data distribution and outliers, bivariate and multivariate
+relationships between variables, and contextual information. These
+externalized checks provide insights into analysts' thoughts process
+process and offer several benefits:
 
@@
 \item
-  helping junior researchers interpret results and diagnose the
-  unexpected,
+  help students and junior researchers develop skills to diagnose
+  unexpected analysis results,
 \item
-  providing checkpoints that support independent replication or
+  serve as clear checkpoints to support independent replication or
   application of methods to new data, and
 \item
-  aligning assumption across researchers from different backgrounds.
+  align assumptions among researchers from different domain backgrounds.
+\item
+  improve analysis transparency, reproducibility, and trustworthiness.
 \end{enumerate}

### 2024-12-06 index.qmd
rewrite Section 3.1

@@
 In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations.
-When outcomes deviate, they can quickly iterate through the analysis cycle to identify potential issues, refine methods, or revise assumptions.
-However, this iterative process they employ is rarely explicit, documented, or discussed in publications.
-As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence.
+When a result falls outside of this interval, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process.
+This mental process of where to diagnose unexpected outcomes is often difficult to trace and discuss in publications.
+As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014].
 These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
-Similarly, when researchers with different background knowledge it becomes unclear whether variations in results arise from differences in the assumptions or in the analysis itself.
+Similarly, when researchers with different background knowledge it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.
+
+One might gain insight into analysts' thought process by speaking with them directly or watching screencast videos they produce, such as, TidyTuesday screencast videos.
+However, direct conversation are not scalable and may not be feasible while creating educational videos requires significant effort from the researchers.
+Ideally, there could be a way to make these expectations explicit and accessible to others.
+Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are sepcific to the data or analysis at hand.
 
@@
 
+They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]
+
 [@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.
@@
 With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process.
-In R, there are some packages, like`skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g.
-`assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank], focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.
+In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g.
+`assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.
 
@@
 
-Q: Whether we should formulate these concept with math notation?? A: only if it helps
+<!-- We may anticipate certain interval on the main result of the analysis based on prior knowledge.
+For example, researchers in public health may expect an average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient] [TODO: check if this is correct].
+-->
 
-While data validation focuses on verifying or subsetting data based on predefined rules [@zio2015methodology], data analysis checks specify assumptions made on the data required for the analysis process to obtain the reported results.
-These checks often involve evaluating aspects such as data distribution and outliers, relationships between variables, and the validity of model assumptions.
+Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself.
+When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions.
+Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills.
+Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand.
+Inspired by *data validation checks* [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results.
+This allows the concept to encompass a broad range of checks, such as data quality (i.e.
+missing data, how the data is structured), data distribution and outliers, bivariate and multivariate relationships between variables, and contextual information.
+These externalized checks provide insights into analysts' thoughts process process and offer several benefits:
 
-However, there are other checks that reflect the mental process of of how data analysts diagnose the unexpected output that are not documented in this process.
-For example, when an unexpected output occurs, an analyst may check on whether a column in the data frame is between two values for data quality.
-However, it is not documented what motivates the analyst to conduct this check -- whether it also applies to other researchers analyzing new data in the same contexts, whether it is a common practice in the field, or whether it is a reaction to this particular data or scenario.
-Currently, most of these assumptions are not discussed in the publication or captured by tools themselves.
-While one might be able to infer some of the mental process of the analysts from external sources, such as talking to them or watching screencast videos produced by the analysts e.g.
-TidyTuesday screencast videos, these insights are not systematically documented or made machine-readable.
-A question to consider here is whether it is possible to deliver this type of quality checks as structured data in a data analysis.
-This gap highlights the need for tools that provide higher level documentation of reasoning behind the checks, facilitating a more transparent and interpretable analysis process.
-
-An analysis plan is a set of analysis steps combined with expectations.
-Expectations represent our belief about certain aspects of the analysis, independent of the analysis itself.
-It can be divided into two types: *outcome expectation* and *plan expectation*.
-Outcome expectation refers to what we anticipate from the main result of the analysis based on prior knowledge.
-They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates [@grolemund_cognitive_2014].
-For example, in public health, prior research shows the average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient].
-This serves as an expectation for similar future studies.
-Plan expectations concern the intermediate steps within the analysis rather than the final outcome.
-They serve as checkpoints to detect deflection in the analysis process For example, we may expect temporal data to be ordered with no gaps and duplicates, or expect that temperature will be a significant covariate in the linear regression model of PM10 on mortality.
-
-Experienced analysts often have implicit expectation about the outcome and rely on a few "directional signs" to check when the outcome deviate from those expectation.
-However, these expectations are rarely made explicit within the analysis workflow.
-This makes it challenging for consumers of the analysis to evaluate the results, since it becomes difficult to disentangle whether discrepancies arise from differing expectations or from the use of statistical technique, without running the analysis themselves.
-Non-expert analysts, lacking prior knowledge or instinct, may not have clear expectations of the results.
-This can lead to reduced confidence of the analysis and makes it more difficult and time-consuming to diagnose the cause of the deviation when the results don't align with expectations.
-By explicitly formulating these expectations, an analysis plan can guide the analysis process, facilitate the communication and evaluate the validity of the results.
-
-Yet, making them explicit is crucial for
-
-  1) helping junior researchers interpret results and diagnose the unexpected, 
-  2) providing checkpoints that support independent replication or application of methods to new data, and
-  3) aligning assumption across researchers from different backgrounds.
-
-<!-- The expectations can be thought of as a set of unit tests used to validate the results of data analysis.
-By specifying a range of values for these tests, multiple versions of the dataset can be generated to satisfy different sets of plan expectations.
-This allows us to present what we called the "result universe" -- the complete set of possible results that can be obtained from one data analysis process.
-By visualizing the result universe, data analysis consumers can observe how changes in expectations affect the results and the range of alternative outcomes that could arise under different conditions.
-This enables them to evaluate the outcomes based on their own plan expectations and gain a broader perspective on how the actual results produced by analysts fit within this spectrum of possibilities, promoting transparency and trust in analysis.
--->
+  1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
+  2) serve as clear checkpoints to support independent replication or application of methods to new data, and 
+  3) align assumptions among researchers from different domain backgrounds.
+  4) improve analysis transparency, reproducibility, and trustworthiness.
 

### 2024-12-06 _manuscript/_tex/index.tex
smooth abstract till Section 3.1

@@
 \begin{abstract}
-Unexpected results in data analysis can prompt researchers to examine
-data quality and analysis steps. While some researchers can typically
-diagnose issues effectively with a few checks, others may struggle to
-identify appropriate checks to diagnose the problem. {[}an example of
-check{]} These checks are often informal and difficult to trace and
-discuss in publications, resulting in others questioning the
-trustworthiness of the analysis. To address this, we formalize the
-informal checks into an \emph{analysis plan} that encompasses the
-analysis steps and (the unit tests): one test for whether the result
-meets expectations and multiple tests for checking the analysis. We then
-present a procedure to assess the quality of these unit tests based on
-their accuracy and redundancy on simulated versions of the original
-data. The accuracy is assessed using binary classification metrics,
-\emph{i.e.}, precision and recall, while redundancy is calculated using
-mutual information. This procedure can be used to conduct a sensitivity
-analysis, compare different analysis plans, and to identify the optimal
-cutoff point for the unit tests.
+In data analysis, unexpected results often prompt researchers to revisit
+their procedures to identify potential issues. While experienced
+researchers can typically diagnose problems by checking a few key
+assumptions, others may struggle to identify the root causes. These
+checked assumptions, or expectations, are typically informal, difficult
+to trace, and rarely discussed in publications. In this paper, we
+formalize these informal assumptions by framing them as binary
+\emph{analysis validation check}. We then introduce a procedure to
+quantify how violations of these checks may lead to unexpected results
+in the analysis. The procedure relies on simulations of the original
+data and evaluates boht accuracy and redundancy. Accuracy is calculated
+through binary classification metric, while redundancy is measured using
+mutual information. We demonstrate this approach with a toy example on
+the fitness step count data and a generalized linear model exmaple
+examining the effect of PM10 on mortality.
 \end{abstract}
@@
 whether they reflect commo common practices in the field, or whether
-they are sepcific to the data or analysis at hand.
+they are specific to the data or analysis at hand.
 
-In this paper, we frame these expectations as checks within an
-\emph{analysis plan}, which allows us to examine these implicit
-assumptions made about the data during analysis and to compare different
-checks for diagnosing unexpected outcomes. We then introduce a procedure
-to evaluate the quality of these checks, either individually or in
-combination through logic regression, based on their accuracy and
-redundancy across simulation of the original data. Accuracy is assessed
-using binary classification metrics, precision and recall, while
-redundancy is measured using mutual information. The proposed workflow
-offers a numerical guarantee, based on simulation, that the analysis
-will produce the expected results, assuming the data generating
-mechanism holds. {[}something about trustworthy data science{]}
+In this paper, we conceptualize these internal expectations and
+assumptions as \emph{analysis validation checks}, which allows us to
+examine these assumptions made during analysis to diagnose unexpected
+outcomes. We then introduce a procedure that provides a quantitative
+emasure on how violations in a tree of checks, derived from individual
+checks, will lead to an unexpected result. The procedure, based on
+simulations of the original data, calculates accuracy and redundancy.
+Accuracy is determined using binary classification metrics, precision
+and recall, from a logic regression fit while redundancy is measured
+using mutual information. The proposed workflow offers a numerical
+guarantee that the analysis will produce the expected results, assuming
+the data generating mechanism holds. {[}something about trustworthy data
+science{]}
 
@@
 unexpected outcomes and data quality checks. Section~\ref{sec-plan}
-introduces the concept of analysis plan, illustrated with a toy example
-on fitness step count. Section~\ref{sec-method} describes the procedure
-for selecting optimal check combinations to predict unexpected outcomes.
-Section~\ref{sec-pm10-mortality} applies this selection procedure to a
-linear regression example examining the effect of PM10 on mortality.
-Section~\ref{sec-discussion} discusses a few key considerations and
-Section~\ref{sec-conclusion} concludes the paper.
+introduces the concept of analysis validation check, illustrated with a
+toy example on fitness step count. Section~\ref{sec-method} describes
+the procedure that quantify how checks combined in logical operators can
+predict unexpected outcomes. Section~\ref{sec-pm10-mortality} applies
+this procedure to examine the effect of PM10 on mortality in public
+health in a generalized linear model. Section~\ref{sec-discussion}
+discusses a few key considerations and Section~\ref{sec-conclusion}
+concludes the paper.
 
@@
 
+\begin{itemize}
+\item
   They shape how we interpret the results and assess whether they are
@@
   \citep{grolemund_cognitive_2014}
-
+\item
   \citep{peng_diagnosing_2021} describes three pedagogical exercises of
   introducing diagnosing unexpected outcome into a classroom setting.
-
-TODO: what if the expectation is ``wrong''
+\item
+  what if the expectation is ``wrong''
+\end{itemize}
 
@@
 assumptions as explicit checks that returns a TRUE or FALSE output given
-the data analyzed at hand. Inspired by \emph{data validation checks}
+the data analyzed at hand. Inspired by data validation checks
 \citep{validate}, where ensure datasets meet expected format and
@@
 \item
-  serve as clear checkpoints to support independent replication or
-  application of methods to new data, and
+  serve as clear checkpoints to support replication or application of
+  methods to (new) data,
 \item
-  align assumptions among researchers from different domain backgrounds.
+  align assumptions among researchers from different domain backgrounds,
+  and
 \item
@@
 
-\section{Applications}\label{sec-examples}
+\section{Application}\label{sec-pm10-mortality}
 
@@
 process for analyzing the effect of PM10 on mortality. The example
-demonstrates how the test process can be used to select cutoff values in
-the unit tests and how the test can be used to diagnose an unexpected
+demonstrates how the procedure presented in Section~\ref{sec-method} can
+be used to select cutoff values in the checks to diagnose an unexpected
 PM10 coefficient from the generalized linear model.
-
-\subsection{Effect of PM10 on mortality}\label{sec-pm10-mortality}
 

### 2024-12-06 index.qmd
smooth abstract till Section 3.1

@@
 title: "My wonderful paper"
-abstract: "Unexpected results in data analysis can prompt researchers to examine data quality and analysis steps.
-While some researchers can typically diagnose issues effectively with a few checks, others may struggle to identify appropriate checks to diagnose the problem.
-[an example of check] These checks are often informal and difficult to trace and discuss in publications, resulting in others questioning the trustworthiness of the analysis.
-To address this, we formalize the informal checks into an *analysis plan* that encompasses the analysis steps and (the unit tests): one test for whether the result meets expectations and multiple tests for checking the analysis.
-We then present a procedure to assess the quality of these unit tests based on their accuracy and redundancy on simulated versions of the original data.
-The accuracy is assessed using binary classification metrics, *i.e.*, precision and recall, while redundancy is calculated using mutual information.
-This procedure can be used to conduct a sensitivity analysis, compare different analysis plans, and to identify the optimal cutoff point for the unit tests."
+abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues.
+While experienced researchers can typically diagnose problems by checking a few key assumptions, others may struggle to identify the root causes.
+These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications.
+In this paper, we formalize these informal assumptions by framing them as binary *analysis validation check*.
+We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis.
+The procedure relies on simulations of the original data and evaluates boht accuracy and redundancy.
+Accuracy is calculated through binary classification metric, while redundancy is measured using mutual information.
+We demonstrate this approach with a toy example on the fitness step count data and a generalized linear model exmaple examining the effect of PM10 on mortality.
+"
 author:
@@
 Ideally, there could be a way to make these expectations explicit and accessible to others.
-Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are sepcific to the data or analysis at hand.
+Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are specific to the data or analysis at hand.
 
-In this paper, we frame these expectations as checks within an *analysis plan*, which allows us to examine these implicit assumptions made about the data during analysis and to compare different checks for diagnosing unexpected outcomes.
-We then introduce a procedure to evaluate the quality of these checks, either individually or in combination through logic regression, based on their accuracy and redundancy across simulation of the original data.
-Accuracy is assessed using binary classification metrics, precision and recall, while redundancy is measured using mutual information.
-The proposed workflow offers a numerical guarantee, based on simulation, that the analysis will produce the expected results, assuming the data generating mechanism holds.
+In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine these assumptions made during analysis to diagnose unexpected outcomes.
+We then introduce a procedure that provides a quantitative emasure on how violations in a tree of checks, derived from individual checks, will lead to an unexpected result.
+The procedure, based on simulations of the original data, calculates accuracy and redundancy.
+Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit while redundancy is measured using mutual information.
+The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the data generating mechanism holds.
 [something about trustworthy data science]
@@
 The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks.
-@sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count.
-@sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes.
-@sec-pm10-mortality applies this selection procedure to a linear regression example examining the effect of PM10 on mortality.
+@sec-plan introduces the concept of analysis validation check, illustrated with a toy example on fitness step count.
+@sec-method describes the procedure that quantify how checks combined in logical operators can predict unexpected outcomes.
+@sec-pm10-mortality applies this procedure to examine the effect of PM10 on mortality in public health in a generalized linear model.
 @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.
@@
 
-They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]
+* They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]
 
-[@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.
+* [@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.
 
-TODO: what if the expectation is "wrong"
+* what if the expectation is "wrong"
 
@@
 
-<!-- We may anticipate certain interval on the main result of the analysis based on prior knowledge.
-For example, researchers in public health may expect an average increase in mortality rate per unit increase in PM10 is about 0.50% [@liu2019ambient] [TODO: check if this is correct].
--->
-
 Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself.
@@
 Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand.
-Inspired by *data validation checks* [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results.
+Inspired by data validation checks [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results.
 This allows the concept to encompass a broad range of checks, such as data quality (i.e.
@@
   1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
-  2) serve as clear checkpoints to support independent replication or application of methods to new data, and 
-  3) align assumptions among researchers from different domain backgrounds.
+  2) serve as clear checkpoints to support replication or application of methods to (new) data,
+  3) align assumptions among researchers from different domain backgrounds, and 
   4) improve analysis transparency, reproducibility, and trustworthiness.
@@
 
-# Applications {#sec-examples}
+# Application {#sec-pm10-mortality}
 
 A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality.
-The example demonstrates how the test process can be used to select cutoff values in the unit tests and how the test can be used to diagnose an unexpected PM10 coefficient from the generalized linear model.
-
-## Effect of PM10 on mortality {#sec-pm10-mortality}
+The example demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
 

### 2024-12-06 _manuscript/_tex/index.tex
section 3.2

@@
 \item
-  \emph{Check for TODOs}
+  \emph{Check for TODOs}, things inside ``{[}{]}'' are my comments
 \item
@@
 
-Consider a 30-day step count goal. Suppose you resolve to walk at least
-8,000 steps each day, using an app to record your daily step count. Your
-target average is 9,000 steps per day, with some ``lows'' day, where you
-walk around 4,000 steps and ``high'' days where you reach about 12,000
-steps. After 30 days, you check how many times your step count fall
-below 8,000 and aim for no more than five days under this threshold.
+Consider a 30-day step count experiment in public health. Testers are
+instructed to walk at least 8,000 steps each day, with an expected
+average of 9,000 steps, tracked by a step counter app. While there may
+be occasional ``low'' days due to factors such as forgetting to wear
+their watch or unfavorable weather condition limiting outdoor
+activities, there are also ``high'' days after an outdoor hike or
+intense workout. After 30 days, you review the data and expect no more
+than five days with step counts falling below the 8,000-step threshold
+{[}I think we need a bit more scenario to justify we expect no more than
+5 days, say it needs to have at least 25 days for it to be valid for the
+next stage research??{]}.
 
-To simulation this data, three normal distributions with different means
-are used for the daily step counts: \(\mathcal{N}(4000, 200)\) for low
-days, \(\mathcal{N}(12000, 200)\) for high days, and
+To simulation this data, three normal distributions are used for the
+daily step counts: \(\mathcal{N}(4000, 200)\) for low days,
+\(\mathcal{N}(12000, 200)\) for high days, and
 \(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
@@
 
-In this scenario, the outcome expectation is that the number of days
-with a step count below 8,000 will be no more than five. To diagnose
-potential reasons why this outcome expectation might fail, we can
-establish a few plan expectations. For example, if the average step
-count is too low, this may suggest there are too many low days,
-potentially lead to an unexpected outcome. Similarly, we can also check
-the quantile of the step count, if more than a third of the days fall
-below 8,000, this could indicate an excess of low-count days.
-Additionally, we may may expect the standard deviation of the step count
-not to be overly large.
-
-TODO: add one or two scenarios on why step count could go up or down: 1)
-you take off your watch in the middle of the day, 2) have an workout
-
-TODO: phrase it as Researcher measuring someone's step count rather than
-measuring your own step count
-
+In this scenario, the outcome is that the number of days with a step
+count below 8,000 being no more than five. To diagnose potential reasons
+why this outcome expectation might fail, we can establish a few analysis
+validation checks. For example, if the average step count is too low,
+this may suggest there are too many low days, potentially lead to the
+unexpected outcome. Similarly, we can also check the quantile of the
+step count, if more than a third of the days fall below 8,000, this
+could indicate an excess of low-count days. Additionally, we may may
+expect the standard deviation of the step count not to be overly large.
 These considerations yield the following three unit tests as plan
@@
 \item
-  test3: the test fails if the standard deviation of the step
-  countsexceeds 2,500.
+  test3: the test fails if the standard deviation of the step counts
+  exceeds 2,500.
 \end{itemize}

### 2024-12-06 index.qmd
section 3.2

@@
 
-* *Check for TODOs*
+* *Check for TODOs*, things inside "[]" are my comments
 - Literature review: Section *diagnosing unexpected outcomes in data analysis* 
@@
 
-Consider a 30-day step count goal.
-Suppose you resolve to walk at least 8,000 steps each day, using an app to record your daily step count.
-Your target average is 9,000 steps per day, with some "lows" day, where you walk around 4,000 steps and "high" days where you reach about 12,000 steps.
-After 30 days, you check how many times your step count fall below 8,000 and aim for no more than five days under this threshold.
+Consider a 30-day step count experiment in public health.
+Testers are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app.
+While there may be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather condition limiting outdoor activities, there are also "high" days after an outdoor hike or intense workout.
+After 30 days, you review the data and expect no more than five days with step counts falling below the 8,000-step threshold [I think we need a bit more scenario to justify we expect no more than 5 days, say it needs to have at least 25 days for it to be valid for the next stage research??].
 
-To simulation this data, three normal distributions with different means are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
+To simulation this data, three normal distributions are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
 The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$.
@@
 
-In this scenario, the outcome expectation is that the number of days with a step count below 8,000 will be no more than five.
-To diagnose potential reasons why this outcome expectation might fail, we can establish a few plan expectations.
-For example, if the average step count is too low, this may suggest there are too many low days, potentially lead to an unexpected outcome.
+In this scenario, the outcome is that the number of days with a step count below 8,000 being no more than five.
+To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks.
+For example, if the average step count is too low, this may suggest there are too many low days, potentially lead to the unexpected outcome.
 Similarly, we can also check the quantile of the step count, if more than a third of the days fall below 8,000, this could indicate an excess of low-count days.
 Additionally, we may may expect the standard deviation of the step count not to be overly large.
-
-TODO: add one or two scenarios on why step count could go up or down: 1) you take off your watch in the middle of the day, 2) have an workout
-
-TODO: phrase it as Researcher measuring someone's step count rather than measuring your own step count
-
 These considerations yield the following three unit tests as plan expectations:
@@
 -   test2: the test fails if the 30th percentile of the step counts is below 8,200
--   test3: the test fails if the standard deviation of the step countsexceeds 2,500.
+-   test3: the test fails if the standard deviation of the step counts exceeds 2,500.
 

### 2024-12-06 _manuscript/_tex/index.tex
section 4

@@
 in the analysis. The procedure relies on simulations of the original
-data and evaluates boht accuracy and redundancy. Accuracy is calculated
+data and evaluates both accuracy and redundancy. Accuracy is calculated
 through binary classification metric, while redundancy is measured using
@@
 \item
-  \emph{Check for TODOs}, things inside ``{[}{]}'' are my comments
+  \emph{Check for TODOs}, things inside ``{[}\ldots{]}'', other than
+  references, are my comments
 \item
@@
 apply to other researchers analyzing new data in the same context,
-whether they reflect commo common practices in the field, or whether
-they are specific to the data or analysis at hand.
+whether they reflect common practices in the field, or whether they are
+specific to the data or analysis at hand.
 
@@
 outcomes. We then introduce a procedure that provides a quantitative
-emasure on how violations in a tree of checks, derived from individual
+measure on how violations in a tree of checks, derived from individual
 checks, will lead to an unexpected result. The procedure, based on
@@
 
-In real-world applications, it is rare to create a set of unit tests can
-fully guarantee expected results. On one hand, there is the cost of
-effort involved in manually developing all these tests; on the other,
-there is the inherent complexity of the problem. (can we - if we're
-thriving for detecting 95\% of the cause?). However, the quality of unit
-tests can be evaluated by simulated data. One set of tests is considered
-better than another if a small set of tests can reliably detect
-unexpected outcomes, which brings two criteria in the evaluation metric:
-accuracy and parsimony. Figure~\ref{fig-metric-calc} illustrates the
-workflow for calculating the metrics.
+While some checks are crucial and can lead to unexpected outcomes,
+others may be tangential to the root cause. In this section, we propose
+a procedure to measure the violation of checks, combined using logical
+operators, contribute to an unexpected outcome. A small set of
+independent checks is considered effective if it translates to
+unexpected outcomes and Figure~\ref{fig-metric-calc} provides an
+overview of the calculation.
 
@@
 
-\caption{\label{fig-metric-calc}this is the cap}
+\caption{\label{fig-metric-calc}{[}this is the cap and the plot needs
+polish{]}}
 
@@
 
-Accuracy refers to a set of tests' ability to accurately detect
-unexpected outcomes while minimizing false positives and false
-negatives. A false positive can indicate (caution or skepticism on
-checking the data), whereas a false negative suggests the tests may lack
-sensitivity to unexpected outcomes. To model the relationship between
-the plan and outcome expectation (binary-binary), a logic regression
-model is used \citep{ruczinski_logic_2003}. Originally developed for SNP
+From the simulated data, the accuracy branch refers to a set of checks'
+ability to accurately detect unexpected outcomes while minimizing false
+positives and false negatives. While a false positive can raise caution
+or skepticism on the data, both false positive and false negative
+suggest the checks are over or lack sensitivity to unexpected outcomes.
+To incorporate the effect of multiple checks on the outcome, a logic
+regression model \citep{ruczinski_logic_2003} is fitted to the checks
+and whether the outcome is unexpected. Originally developed for SNP
 microarray data, logic regression constructs Boolean combinations of
-binary variables to solve regression problems {[}more introduction on
-logic regression{]}. Compared to other tree-based methods or machine
-learning methods for binary-binary prediction, logic regression produces
-Boolean combinations, or meta rules, that combines unit tests to solve
-the prediction problem. {[}need rewrite here{]}. The performance of the
-tests can be evaluated using precision and recall metrics derived from
-the confusion matrix of the logic regression prediction
+binary variables, in a tree structure, to predict both binary and
+numerical outcomes. Compared to other tree-based methods for
+binary-binary prediction, the Boolean combinations from the logic
+regression model produce a tree structure that can be directly
+interpreted as the possible combination of checks leading to an
+unexpected outcome, without the need to invert the tree as required in
+classic tree-based methods. The model is then used to predict the
+outcome expectation based on the checks, and the prediction is compared
+to the actual outcome expectation to calculate the precision and recall
+of the checks.
 
-\begin{itemize}
-\tightlist
-\item
-  precision: the proportion of correctly identified unexpected results
-  (true positives) out of all the predicted unexpected results (true
-  positives + false positives)
-\item
-  recall: the proportion of correctly identified unexpected results
-  (true positives) out of all the actual unexpected results (true
-  positives + false negatives)
-\end{itemize}
-
-The second criteria is parsimony in the tests. While tests may score
-high on accuracy, they may be less effective at explaining the reasons
-behind unexpected results. This could happen if a set of tests are all
-tangentially related to the cause of the unexpected results, but none
-addressing the root cause. It may also occur if the tests are correlated
-with one another, leading to redundancy.
+While tests may score high on accuracy, they may be less effective at
+explaining the reasons behind unexpected results. This could happen if a
+set of tests are all tangentially related to the cause of the unexpected
+results, but none addressing the root cause. It may also occur if the
+tests are correlated with one another, leading to redundancy.
 
@@
 This concept extends naturally to multiple variables through total
-correlation {[}ref{]}, \(C(X_1, X_2, \cdots, X_n)\), which captures
-redundancy across a set of \(n\) variables:
+correlation, \(C(X_1, X_2, \cdots, X_n)\), which captures redundancy
+across a set of \(n\) variables:
 
@@
 A high mutual information value indicates redundancy among the tests,
-while a low value suggests that the tests are independent and provide
+while a low value suggests that the tests are independent and provides
 unique information to diagnose the unexpected outcome. To standardize
@@
 To combine precision, recall, and independence into a single metric,
-various mathematical means, such as arithmetic mean, harmonic mean, and
-quadratic mean, can be used. The differences among these means are
+mathematical means, such as arithmetic mean, harmonic mean, and
+quadratic mean can be used. The differences among these means are
 minimal when the three metrics are similar. However, as the differences
@@
 \subsection{Toy example revisited}\label{toy-example-revisited}
-
-In the step count example, we can use the logic regression model to
-evaluate the quality of the unit tests. The logic regression model is
-fitted to the three unit tests (test1, test2, test3) and the outcome
-expectation (unexpect) as the response variable. The model is then used
-to predict the outcome expectation based on the unit tests. The
-prediction is then compared to the actual outcome expectation to
-calculate the precision and recall of the tests. The independence of the
-tests is also calculated to assess the redundancy of the tests.
-Figure~\ref{fig-logic-reg} presents the suggested logic regression model
-as a combination of test 1 and test 3 with an OR operator.
-
-Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
-independence for the three individual tests and the combined test rule
-(test1 OR test3) from the logic regression. The harmonic and arithmetic
-means are included to evaluate the quality of the unit tests in
-diagnosing unexpected step counts. The results show that the combined
-test rule (test1 OR test3) has the highest precision, recall, and
-independence, suggesting that it is the most effective test for
-diagnosing unexpected step counts. We also include the metric calculated
-from fitting a regression tree model to the data to compare the
-performance of the logic regression model. The regression tree produces
-a similar model of first split on test1 and then split on test3, and
-results in the same accuracy and overall score as the logic regression
-model.
-
-top-down and bottom up (regression tree + logic tree): more naturally
-useful summary of the say it is organized. (put down the plot)
 
@@
 \end{longtable}
+
+In the step count example, the logic regression model is fitted to the
+three unit tests to generate the prediction for calculating precision
+and recall. Figure~\ref{fig-logic-reg} shows the chosen logic regression
+model: a combination of test 1 and test 3 with an OR operator and
+Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
+independence for the three individual tests and the combined test rule
+(test1 OR test3) from the logic regression. We also include the metric
+calculated from fitting a regression tree model to the data to compare
+the performance of the logic regression model. The harmonic and
+arithmetic means are included to combine the three measures. The results
+show that the two tests produced by the logic regression can accurately
+predict 82.1\% of all \emph{predicted unexpected cases}, and 83.6\%
+cases of all \emph{actual unexpected results} in the simulation data.
+The regression tree produces a similar model of first split on test1 and
+then split on test3, and results in the same accuracy and overall score
+as the logic regression model.
 

### 2024-12-06 index.qmd
section 4

@@
 We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis.
-The procedure relies on simulations of the original data and evaluates boht accuracy and redundancy.
+The procedure relies on simulations of the original data and evaluates both accuracy and redundancy.
 Accuracy is calculated through binary classification metric, while redundancy is measured using mutual information.
@@
 
-* *Check for TODOs*, things inside "[]" are my comments
+* *Check for TODOs*, things inside "[...]", other than references, are my comments
 - Literature review: Section *diagnosing unexpected outcomes in data analysis* 
@@
 Ideally, there could be a way to make these expectations explicit and accessible to others.
-Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect commo common practices in the field, or whether they are specific to the data or analysis at hand.
+Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
 
 In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine these assumptions made during analysis to diagnose unexpected outcomes.
-We then introduce a procedure that provides a quantitative emasure on how violations in a tree of checks, derived from individual checks, will lead to an unexpected result.
+We then introduce a procedure that provides a quantitative measure on how violations in a tree of checks, derived from individual checks, will lead to an unexpected result.
 The procedure, based on simulations of the original data, calculates accuracy and redundancy.
@@
 
-In real-world applications, it is rare to create a set of unit tests can fully guarantee expected results.
-On one hand, there is the cost of effort involved in manually developing all these tests; on the other, there is the inherent complexity of the problem.
-(can we - if we're thriving for detecting 95% of the cause?).
-However, the quality of unit tests can be evaluated by simulated data.
-One set of tests is considered better than another if a small set of tests can reliably detect unexpected outcomes, which brings two criteria in the evaluation metric: accuracy and parsimony.
-@fig-metric-calc illustrates the workflow for calculating the metrics.
+While some checks are crucial and can lead to unexpected outcomes, others may be tangential to the root cause.
+In this section, we propose a procedure to measure the violation of checks, combined using logical operators, contribute to an unexpected outcome.
+A small set of independent checks is considered effective if it translates to unexpected outcomes and @fig-metric-calc provides an overview of the calculation.
 
@@
 #| label: fig-metric-calc
-#| fig-cap: "this is the cap"
+#| fig-cap: "[this is the cap and the plot needs polish]"
 knitr::include_graphics(here::here("figures", "metric-calc.png"))
@@
 
-Accuracy refers to a set of tests' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
-A false positive can indicate (caution or skepticism on checking the data), whereas a false negative suggests the tests may lack sensitivity to unexpected outcomes.
-To model the relationship between the plan and outcome expectation (binary-binary), a logic regression model is used [@ruczinski_logic_2003].
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables to solve regression problems \[more introduction on logic regression\].
-Compared to other tree-based methods or machine learning methods for binary-binary prediction, logic regression produces Boolean combinations, or meta rules, that combines unit tests to solve the prediction problem.
-\[need rewrite here\].
-The performance of the tests can be evaluated using precision and recall metrics derived from the confusion matrix of the logic regression prediction
+From the simulated data, the accuracy branch refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
+While a false positive can raise caution or skepticism on the data, both false positive and false negative suggest the checks are over or lack sensitivity to unexpected outcomes.
+To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the checks and whether the outcome is unexpected.
+Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, in a tree structure, to predict both binary and numerical outcomes.
+Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based methods.
+The model is then used to predict the outcome expectation based on the checks, and the prediction is compared to the actual outcome expectation to calculate the precision and recall of the checks.
 
--   precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives)
--   recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives)
+<!-- -   precision: the proportion of correctly identified unexpected results (true positives) out of all the predicted unexpected results (true positives + false positives) -->
+<!-- -   recall: the proportion of correctly identified unexpected results (true positives) out of all the actual unexpected results (true positives + false negatives) -->
 
-The second criteria is parsimony in the tests.
 While tests may score high on accuracy, they may be less effective at explaining the reasons behind unexpected results.
@@
 
-This concept extends naturally to multiple variables through total correlation \[ref\], $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
+This concept extends naturally to multiple variables through total correlation, $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
 
@@
 
-A high mutual information value indicates redundancy among the tests, while a low value suggests that the tests are independent and provide unique information to diagnose the unexpected outcome.
+A high mutual information value indicates redundancy among the tests, while a low value suggests that the tests are independent and provides unique information to diagnose the unexpected outcome.
 To standardize this measure, the total correlation *per observation* is calculated, and an independence score, ranging between 0 and 1, is defined as 1 - mutual information.
 
-To combine precision, recall, and independence into a single metric, various mathematical means, such as arithmetic mean, harmonic mean, and quadratic mean, can be used.
+To combine precision, recall, and independence into a single metric, mathematical means, such as arithmetic mean, harmonic mean, and quadratic mean can be used.
 The differences among these means are minimal when the three metrics are similar.
@@
 ## Toy example revisited
-
-In the step count example, we can use the logic regression model to evaluate the quality of the unit tests.
-The logic regression model is fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable.
-The model is then used to predict the outcome expectation based on the unit tests.
-The prediction is then compared to the actual outcome expectation to calculate the precision and recall of the tests.
-The independence of the tests is also calculated to assess the redundancy of the tests.
-@fig-logic-reg presents the suggested logic regression model as a combination of test 1 and test 3 with an OR operator.
-
-@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
-The harmonic and arithmetic means are included to evaluate the quality of the unit tests in diagnosing unexpected step counts.
-The results show that the combined test rule (test1 OR test3) has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected step counts.
-We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
-The regression tree produces a similar model of first split on test1 and then split on test3, and results in the same accuracy and overall score as the logic regression model.
-
-top-down and bottom up (regression tree + logic tree): more naturally useful summary of the say it is organized. (put down the plot)
-
-<!-- * provide interpretation at different scenarios: -->
-
-<!--   1) one test is flagged, the prediction is as expected:  --> <!--   2) multiple tests are flagged, the prediction is unexpected,  --> <!--   3) no test is flagged, the prediction is unexpected,  --> <!--   4) no test is flagged, the prediction is as expected -->
 
@@
 ```
+
+In the step count example, the logic regression model is fitted to the three unit tests to generate the prediction for calculating precision and recall.
+@fig-logic-reg shows the chosen logic regression model: a combination of test 1 and test 3 with an OR operator and @tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
+We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
+The harmonic and arithmetic means are included to combine the three measures.
+The results show that the two tests produced by the logic regression can accurately predict 82.1% of all *predicted unexpected cases*, and 83.6% cases of all *actual unexpected results* in the simulation data.
+The regression tree produces a similar model of first split on test1 and then split on test3, and results in the same accuracy and overall score as the logic regression model.
+
+<!-- * provide interpretation at different scenarios: -->
+
+<!--   1) one test is flagged, the prediction is as expected:  --> <!--   2) multiple tests are flagged, the prediction is unexpected,  --> <!--   3) no test is flagged, the prediction is unexpected,  --> <!--   4) no test is flagged, the prediction is as expected -->
 

### 2024-12-06 _manuscript/_tex/index.tex
section 5

@@
 steps across 300 simulated 30-day periods. The orange bars indicate
-instances where the count exeeds five days, representing an unexpected
+instances where the count exceeds five days, representing an unexpected
 outcome in this scenario.}
@@
 \caption{\label{fig-logic-reg}Logic regression model fitted to the three
-unit tests (test1, test2, test3) and the outcome expectation (unexpect)
-as the response variable. The model suggests using an OR rule to combine
-test1 and test3 to predict the outcome expectation.}
+unit tests (test1, test2, test3). The model suggests using an OR rule to
+combine test1 and test3 to predict the outcome expectation.}
 
@@
 
-A regression example is produced to illustrate the test selection
-process for analyzing the effect of PM10 on mortality. The example
-demonstrates how the procedure presented in Section~\ref{sec-method} can
-be used to select cutoff values in the checks to diagnose an unexpected
-PM10 coefficient from the generalized linear model.
-
 Consider a generalized linear model (GLM) to study the effect of PM10 on
@@
 Analysts may expect a PM10 coefficient between {[}0, 0.005{]} after
-considering the temperature confounding {[}TODO: reference{]}. This
-expectation can be framed into a check that fails, labelled as 1, if the
-PM10 coefficient is outside the range {[}0, 0.005{]}. Multiple factors
-can impact the PM10 coefficient, such as the sample size, the strength
-of the correlation between mortality and PM10, and the strength of the
-correlation between mortality and temperature. Analysts may expect a
-reasonable sample size to ensure the reliability of the coefficient
-estimate. Outliers in the three variables can also leverage the
-coefficient. While these are possible factors that could affect the
-analysis result, it is not clear the cutoff values for these checks to
-determine a failure. Here a list of checks are created in
-Table~\ref{tbl-checks}.
+considering the temperature confounding {[}TODO: reference{]}. The
+example demonstrates how the procedure presented in
+Section~\ref{sec-method} can be used to select cutoff values in the
+checks to diagnose an unexpected PM10 coefficient from the generalized
+linear model.
+
+This PM10 coefficient expectation can be framed into a check that fails,
+labelled as 1, if the PM10 coefficient is outside the range {[}0,
+0.005{]} and 0 if it is within. Multiple factors can affect the PM10
+coefficient, such as the sample size, the strength of the correlation
+between mortality and PM10, and the strength of the correlation between
+mortality and temperature. Analysts may expect a reasonable sample size
+to ensure the reliability of the coefficient estimate. Outliers in the
+three variables can also leverage the coefficient. While these are
+possible factors that could affect the analysis result, it is not clear
+the cutoff values for these checks to determine a failure. Here we
+create a list of checks in Table~\ref{tbl-checks} with varied cutoff
+values for each:
 
@@
 variable with qq plot presented in Figure~\ref{fig-dist-fit} to evaluate
-the fit. AIC suggests a negative binomial distributio nfor mortality, a
+the fit. AIC suggests a negative binomial distribution for mortality, a
 beta distribution for PM10 (multiple by 100 to obtain the original
 scale), and a Weibull distribution for temperature. To include the
-effect of outlier, we add a single outlier to the data for mortality and
-PM10 {[}more details{]}.
+effect of outlier, we add a single outlier to the data for both variable
+mortality and PM10.
 
@@
 \caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
-twelve unit tests and the outcome expectation (unexpect) as the response
-variable. The model suggests the relationship:
-(\(\text{cor}(\text{m}, \text{tmp}) > -0.35\) AND
-\(\text{smpl\_size} \le 600\)) OR
-\(\text{cor}(\text{m},  \text{PM10}) < - 0.05\)}
+fourteen unit tests and the outcome expectation (unexpected) as the
+response variable. The model suggests the relationship: (sample size
+\emph{greater than} 200 AND mortality-PM10 correlation \textless{}
+-0.04) AND (mortality-temperature correlation \textgreater{} -0.35 OR
+there exist mortality outlier) to predict the unexpected PM10
+coefficient. {[}sorry the proper plot is not ready{]}.}
 
@@
 tests (1: sample size, 2: mortality-PM10 correlation, 3:
-mortality-temperature correlation), and the combined test rule 4:
-(sample size AND mortality-temperature correlation) OR mortality-PM10
-correlation.}
+mortality-temperature correlation, 4: mortality outlier), and the
+combined test rule 5: (sample size AND mortality-PM10 correlation) AND
+(mortality-temperature correlation OR mortality outlier.}
 
@@
 A logic regression is fitted to predict whether the PM10 coefficient is
-unexpected (outside the range of {[}0, 0.005{]}) using the checks listed
-in Table~\ref{tbl-checks}. Precision, recall, and independence score,
-along with their harmonic and arithmetic mean are calculated.
+unexpected using the checks listed in Table~\ref{tbl-checks} and
 Figure~\ref{fig-linear-reg-tree} shows the logic regression tree from
-the fitted model and Table~\ref{tbl-linear-reg} prints the numerical
-summary of four selected single test and their combined test found by
-the logic regression. The logic regression model picks up the following
-cutoff value for each type of check:
+the fitted model. Precision, recall, and independence score, along with
+their harmonic and arithmetic mean are calculated in
+Table~\ref{tbl-linear-reg}. The logic regression model picks up the
+following cutoff value for each type of check:
 
@@
 on mortality or correlation between mortality and temperature. This
-combined check rule generates a 0.76 precision and a 0.88 recall for
-predicting the unexpected PM10 coefficient. The single check,
+combined check rule generates a precision of 0.76 and a recall of 0.88
+for predicting the unexpected PM10 coefficient. The single check,
 \texttt{cor(m,\ PM10)\ \textless{}\ -0.03}, is also powerful with a high

### 2024-12-06 index.qmd
section 5

@@
 #| fig-cap: Number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
-The orange bars indicate instances where the count exeeds five days, representing an unexpected outcome in this scenario.
+The orange bars indicate instances where the count exceeds five days, representing an unexpected outcome in this scenario.
 set.seed(1234)
@@
 #| fig.heigh: 2
-#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3) and the outcome expectation (unexpect) as the response variable.
-The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
+#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3). The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
 fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
@@
 
-A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality.
+Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study].
+Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference].
 The example demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
 
-Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study].
-Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference].
-This expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005].
-Multiple factors can impact the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
+This PM10 coefficient expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005] and 0 if it is within.
+Multiple factors can affect the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
 Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate.
@@
 While these are possible factors that could affect the analysis result, it is not clear the cutoff values for these checks to determine a failure.
-Here a list of checks are created in @tbl-checks.
+Here we create a list of checks in @tbl-checks with varied cutoff values for each:
 
@@
 AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit.
-AIC suggests a negative binomial distributio nfor mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature.
-To include the effect of outlier, we add a single outlier to the data for mortality and PM10 [more details].
+AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature.
+To include the effect of outlier, we add a single outlier to the data for both variable mortality and PM10.
 
@@
                                 seed = 6, nleaves = 4)
-plot_logicreg(fit) 
+plot(fit) 
 ```
-<!-- The caption with some LaTeX like this $\lambda$. -->
 
-Logic regression model fitted to the twelve unit tests and the outcome expectation (unexpect) as the response variable. The model suggests the relationship:
-($\text{cor}(\text{m}, \text{tmp}) > -0.35$ AND $\text{smpl\_size} \le 600$) OR $\text{cor}(\text{m},  \text{PM10}) < - 0.05$
+Logic regression model fitted to the fourteen unit tests and the outcome expectation (unexpected) as the response variable. The model suggests the relationship:
+(sample size *greater than* 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
+[sorry the proper plot is not ready].
 :::
@@
   kable(digits = 3, escape = FALSE,
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation), and the combined test rule 4: (sample size AND mortality-temperature correlation) OR mortality-PM10 correlation.")
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation,  4: mortality outlier), and the combined test rule 5: (sample size AND mortality-PM10 correlation) AND (mortality-temperature correlation OR mortality outlier.")
 
@@
 
-A logic regression is fitted to predict whether the PM10 coefficient is unexpected (outside the range of [0, 0.005]) using the checks listed in @tbl-checks.
-Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated.
-@fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of four selected single test and their combined test found by the logic regression.
+A logic regression is fitted to predict whether the PM10 coefficient is unexpected using the checks listed in @tbl-checks and @fig-linear-reg-tree shows the logic regression tree from the fitted model.
+Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in  @tbl-linear-reg.
 The logic regression model picks up the following cutoff value for each type of check:
@@
 The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature.
-This combined check rule generates a `r round(as.numeric(df[5, "precision"]), 3)` precision and a `r round(as.numeric(df[5, "recall"]), 3)` recall for predicting the unexpected PM10 coefficient.
+This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient.
 The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.

### 2024-12-09 index.qmd
Intro + related work

@@
 abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues.
-While experienced researchers can typically diagnose problems by checking a few key assumptions, others may struggle to identify the root causes.
+While experienced researchers can often quickly diagnose problems by checking a few key assumptions, others may struggle to identify the root causes.
 These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications.
-In this paper, we formalize these informal assumptions by framing them as binary *analysis validation check*.
+In this paper, we formalize these informal assumptions by framing them as binary *analysis validation checks*.
 We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis.
 The procedure relies on simulations of the original data and evaluates both accuracy and redundancy.
-Accuracy is calculated through binary classification metric, while redundancy is measured using mutual information.
-We demonstrate this approach with a toy example on the fitness step count data and a generalized linear model exmaple examining the effect of PM10 on mortality.
+Accuracy is calculated through a binary classification metric, while redundancy is measured using mutual information.
+We demonstrate this approach with a toy example based on fitness step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality.
 "
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
 library(tidyverse)
@@
 
--   TODO: emphasize trustworthy data science since it is the theme of the special issue
-
 In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations.
@@
 This mental process of where to diagnose unexpected outcomes is often difficult to trace and discuss in publications.
-As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned -- achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014].
+As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014].
 These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
-Similarly, when researchers with different background knowledge it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.
+Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.
 
-One might gain insight into analysts' thought process by speaking with them directly or watching screencast videos they produce, such as, TidyTuesday screencast videos.
-However, direct conversation are not scalable and may not be feasible while creating educational videos requires significant effort from the researchers.
-Ideally, there could be a way to make these expectations explicit and accessible to others.
-Even better, if the encoding is machine-readable, we could analyze these expectations and learn from the analysis -- whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
+One might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g.
+@gu2024data].
+However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the researchers.
+Ideally, there could be a way to make expectations about data analysis explicit and accessible to others.
+Even better, if the encoding were machine-readable, we could analyze these expectations and learn from the analysis itself.
+For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
 
-In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine these assumptions made during analysis to diagnose unexpected outcomes.
-We then introduce a procedure that provides a quantitative measure on how violations in a tree of checks, derived from individual checks, will lead to an unexpected result.
-The procedure, based on simulations of the original data, calculates accuracy and redundancy.
-Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit while redundancy is measured using mutual information.
-The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the data generating mechanism holds.
-[something about trustworthy data science]
+The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
+While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis.
+Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing.
+Thus, a reader looking at published code can often be left with many questions about why certain choices were made.
+Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].
 
-The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks.
-@sec-plan introduces the concept of analysis validation check, illustrated with a toy example on fitness step count.
-@sec-method describes the procedure that quantify how checks combined in logical operators can predict unexpected outcomes.
-@sec-pm10-mortality applies this procedure to examine the effect of PM10 on mortality in public health in a generalized linear model.
+In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine the assumptions made during an analysis and to diagnose unexpected outcomes.
+We then introduce a procedure that provides a quantitative measure of how violations in a tree of analysis checks, derived from individual checks, will lead to an unexpected result.
+The procedure, based on simulations of the original data, calculates the accuracy and redundancy of the analysis checks.
+Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit [@ruczinski_logic_2003], while redundancy is measured using mutual information.
+The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.
+
+The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks.
+@sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on fitness step count data.
+@sec-method describes the procedure that quantifies how analysis validation checks combined using logical operators can predict unexpected outcomes in an analysis.
+@sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality.
 @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.
 
-# Literature review {#sec-lit-review}
+# Related Work {#sec-lit-review}
 
@@
 
-* They shape how we interpret the results and assess whether they are consistent with existing knowledge or indicate the need for updates. [@grolemund_cognitive_2014]
+The concept of framing data analysis as a sense-making process was originally presented by [@grolemund_cognitive_2014] based on seminal work by [@wild1999statistical].
+Key to any sense-making process is a model for the world (i.e.
+expectations for what we might observe) and observed data with which we can compare our expectations.
+If there is a significant deviation between what we observe and our expectations, then a data analysis must determine what is causing that deviation.
+A naive approach would be to update our model for the world to match the data, under the assumption that the initial expectation was incorrect.
+However, experienced analysts know that the reality can be more nuanced than that, with errors occurring in data collection or data processing that can have an impact on final results.
 
-* [@peng_diagnosing_2021] describes three pedagogical exercises of introducing diagnosing unexpected outcome into a classroom setting.
+The skill of diagnosing unexpected data analysis results is not one that has received significant attention in the statistics literature.
+While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of an unexpected analysis result are typically not presented [@peng2022perspective].
+[@peng_diagnosing_2021] proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers.
+They provide a systematic approach involving working backwards from the analysis result to identify potential causes.
+There are parallels here to the concept of debugging and testing in software engineering [@donoghue2021teaching].
+For example, [@li2019towards] found that experienced engineers were generally able to identify problems in code faster than novices, and that the ability to debug code required knowledge that cut across different domains.
 
-* what if the expectation is "wrong"
+If it is true that the speed with which data analysts can identify problems with an analysis is related to their experience working with a given type of data, then there is perhaps room to improve the analytic process by externalizing the aspects that an analyst learns through experience.
+That way, inexperienced analysts could examine the thought process of an experienced analyst and learn to identify factors that can cause unexpected results to occur.
 
@@
 
-## Framing expectations as checks
-
-Expectations represent our beliefs about certain aspects of the analysis, independent of the analysis itself.
+Expectations represent our understanding of certain aspects of the analysis and the data, independent of the analysis itself.
 When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions.
 Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills.
-Here, we introduce the concept of *analysis validation checks*, which frames these expectations or assumptions as explicit checks that returns a TRUE or FALSE output given the data analyzed at hand.
-Inspired by data validation checks [ @validate], where ensure datasets meet expected format and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the reported results.
-This allows the concept to encompass a broad range of checks, such as data quality (i.e.
-missing data, how the data is structured), data distribution and outliers, bivariate and multivariate relationships between variables, and contextual information.
-These externalized checks provide insights into analysts' thoughts process process and offer several benefits:
+Here, we introduce the concept of **analysis validation checks**, which frames these expectations or assumptions as explicit checks that return a TRUE or FALSE output given the data analyzed at hand.
+Inspired by the concept of data validation checks (@validate), which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*.
+The focus on expected results allows the concept of analysis validation checks to encompass a broad range of checks, such as data quality (i.e.
+missing data, how the data are structured), data distribution and outliers, bivariate and multivariate relationships between variables, and other contextual information.
+These externalized checks provide insights into analysts' thought processes and offer the following benefits:
 
-  1) help students and junior researchers develop skills to diagnose unexpected analysis results, 
-  2) serve as clear checkpoints to support replication or application of methods to (new) data,
-  3) align assumptions among researchers from different domain backgrounds, and 
-  4) improve analysis transparency, reproducibility, and trustworthiness.
+  1. Serve as clear checkpoints to support the replication or application of methods to (new) data by programmatically communicating the requirements or assumptions made of the data;
+  3. Align assumptions among researchers from different domain backgrounds who may have different expectations about the data;
+  4. Improve analysis transparency, reproducibility, and trustworthiness by externalizing a key part of the analysis process; and
+  5. Quantify the effectiveness of analysis checks for predicting the expected outcome (see @sec-method);
+
+In addition to the above benefits, the development and publication of analysis checks has the potential to help students, inexperienced analysts, and junior researchers develop the skills needed to diagnose unexpected analysis results for a given type of data because the assumptions made about the data are made transparent.
+The analysis checks can serve as a basis for new analysts to have conversations about the data they are analyzing and to develop a better understanding of the potential data generation process.
 

### 2024-12-09 index.qmd
Toy example

@@
 
-A substantial body of literature has addressed the definition of data quality [@8642813, more] and developed frameworks, that includes dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
-These frameworks are often used information system and database management and support business decision-making in the industry.
-For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and reusability in future studies [ref].
-With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the data producers or collectors who have the full knowledge of data in hand, prompting more interest towards the data quality checks in the data analysis process.
-In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], provides basic data screening and reporting, while another class of packages, e.g.
+A substantial body of literature has addressed the definition of data quality [@8642813, more] and has developed frameworks that include dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
+These frameworks are often used in  information systems and database management and support business decision-making in various industries.
+For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and re-usability in future studies [ref].
+With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the producers or collectors of the data who would have the full knowledge of data in hand, prompting more interest towards data quality checks in the data analysis process.
+In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], that provide basic data screening and reporting tools, while another class of packages, e.g.
 `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.
+
+The literature on data quality typically focuses on the intrinsic or inherent quality of the data themselves, rather than the data's relationship to any specific data analysis.
+So for example, if a column in a data table is expecting numerical data, but we observe a character value in one of the entries, then that occurrence would trigger some sort of data quality check.
+This type of quality check can be triggered without any knowledge of what the data will ultimately be used for.
+However, for a given analysis, we may require specific aspects of the data to be true because they affect the result being computed.
+Conversely, certain types of poor quality data may have little impact on the ultimate result of an analysis (e.g.
+data that are missing completely at random).
+Defining data quality in terms of what may affect a specific analysis outcome or result has the potential to open new avenues for defining data checks and for building algorithms for optimizing the collection of checks defined for a specific analysis.
 
@@
 
-Expectations represent our understanding of certain aspects of the analysis and the data, independent of the analysis itself.
-When outcomes deviate from these expectations, analysts often revisit the analysis cycle to identify potential issues, refine methods, or revise assumptions.
+Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself.
+When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions.
 Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills.
-Here, we introduce the concept of **analysis validation checks**, which frames these expectations or assumptions as explicit checks that return a TRUE or FALSE output given the data analyzed at hand.
-Inspired by the concept of data validation checks (@validate), which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*.
+
+Here, we introduce the concept of **analysis validation checks**, which frame these expectations or assumptions as explicit checks that return a TRUE or FALSE result given the data analyzed at hand.
+Inspired by the concept of data validation checks (@validate), which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst.
 The focus on expected results allows the concept of analysis validation checks to encompass a broad range of checks, such as data quality (i.e.
 missing data, how the data are structured), data distribution and outliers, bivariate and multivariate relationships between variables, and other contextual information.
-These externalized checks provide insights into analysts' thought processes and offer the following benefits:
+
+Our proposed analysis validation checks provide insights into an analyst's thought process and offer the following benefits:
 
@@
 
-## A toy example
+## A Toy Example {#sec-toy}
 
 Consider a 30-day step count experiment in public health.
-Testers are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app.
-While there may be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather condition limiting outdoor activities, there are also "high" days after an outdoor hike or intense workout.
-After 30 days, you review the data and expect no more than five days with step counts falling below the 8,000-step threshold [I think we need a bit more scenario to justify we expect no more than 5 days, say it needs to have at least 25 days for it to be valid for the next stage research??].
+Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app.
+After 30 days, we review the data and examine the number of steps taken each day.
+With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.
+We may also expect "high" days recorded after an outdoor hike or intense workout.
+Given the requirements of the study, we expect that for a given subject there would be no more than five days with step counts falling below the 8,000-step threshold ("invalid days").
+If there were more than five invalid days, the data from that subject would not be usable for the study.
 
-To simulation this data, three normal distributions are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
-The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$.
-@fig-step-count displays the number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
-
-In this scenario, the outcome is that the number of days with a step count below 8,000 being no more than five.
-To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks.
-For example, if the average step count is too low, this may suggest there are too many low days, potentially lead to the unexpected outcome.
+In this scenario, the outcome is that the number of days with a step count below 8,000 and the expectation is that this outcome would take a value no more than 5.
+If the expectation is violated, we can consider the potential configurations of the data that could lead to such an anomaly.
+To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
+For example, if the average step count is too low, this may suggest that the subject is incapable of taking the required number of steps, potentially leading to the unexpected outcome.
 Similarly, we can also check the quantile of the step count, if more than a third of the days fall below 8,000, this could indicate an excess of low-count days.
 Additionally, we may may expect the standard deviation of the step count not to be overly large.
-These considerations yield the following three unit tests as plan expectations:
+These considerations yield the following three analysis validation checks:
 
 -   test1: the test fails if the mean step count is below 8,200
--   test2: the test fails if the 30th percentile of the step counts is below 8,200
--   test3: the test fails if the standard deviation of the step counts exceeds 2,500.
+-   test2: the test fails if the 30th percentile of the observed step counts is below 8,200
+-   test3: the test fails if the standard deviation of the observed step counts exceeds 2,500.
+
+The cutoff values chosen for these tests would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.
 
@@
 
+To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
+The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$.
+@fig-step-count displays the number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
+
 # Method {#sec-method}
 
-## A workflow to assess the quality of unit tests
-
-While some checks are crucial and can lead to unexpected outcomes, others may be tangential to the root cause.
-In this section, we propose a procedure to measure the violation of checks, combined using logical operators, contribute to an unexpected outcome.
+While some checks may be crucial and directly indicate an unexpected outcome, others may be tangential to the problem at hand and not indicate a root cause of an unexpected outcome.
+In this section, we propose a procedure to measure the effectiveness of checks that, when combined using logical operators, contribute to an unexpected outcome.
 A small set of independent checks is considered effective if it translates to unexpected outcomes and @fig-metric-calc provides an overview of the calculation.
@@
 From the simulated data, the accuracy branch refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
-While a false positive can raise caution or skepticism on the data, both false positive and false negative suggest the checks are over or lack sensitivity to unexpected outcomes.
-To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the checks and whether the outcome is unexpected.
+While a false positive can raise caution or skepticism on the data, the presence of both false positives and false negatives suggest that the checks are overly sensitive or lack sensitivity to unexpected outcomes.
+
+To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks using the indicator of an unexpected result as the outcome in the model.
 Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, in a tree structure, to predict both binary and numerical outcomes.
-Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based methods.
-The model is then used to predict the outcome expectation based on the checks, and the prediction is compared to the actual outcome expectation to calculate the precision and recall of the checks.
+Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods.
+The logic regression model is then used to predict the analysis result based on the values of checks, and the prediction is compared to the actual analysis result in order to calculate the precision and recall of the checks.
 
@@
 
-While tests may score high on accuracy, they may be less effective at explaining the reasons behind unexpected results.
-This could happen if a set of tests are all tangentially related to the cause of the unexpected results, but none addressing the root cause.
-It may also occur if the tests are correlated with one another, leading to redundancy.
+While tests may score high on accuracy, they may be less effective at explaining the various reasons behind unexpected results.
+This could happen if, for example, a set of tests are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
+It may also occur if the tests are highly correlated with one another, leading to redundancy.
 
@@
 
-To combine precision, recall, and independence into a single metric, mathematical means, such as arithmetic mean, harmonic mean, and quadratic mean can be used.
+To combine precision, recall, and independence into a single metric, we can combine the three scores using the arithmetic mean, harmonic mean, or quadratic mean.
 The differences among these means are minimal when the three metrics are similar.
@@
 
-## Toy example revisited
+## Toy Example Revisited
+
+Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, which is whether the number of days with fewer than 8,000 steps is greater than 5.
+The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics.
+@fig-logic-reg shows the best-fitting logic regression model, which is a combination of test1 ("mean step count is below 8,200") and test3 ("standard deviation exceeds 2,500") with an OR operator.
+In other words, if either test1 or test3 is true, then we would predict an unexpected outcome in the analysis (i.e.
+the number of days with fewer than 8,000 steps is greater than 5).
 
@@
 ```
+
+@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
+We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
+The harmonic and arithmetic means are included to combine the three measures.
+The results show that the two tests produced by the logic regression can accurately predict 83.6% cases of all *actual unexpected results* in the simulation data.
+Furthermore, 82.1% of all *predicted unexpected results* were in fact observed to be unexpected.
 
@@
 
-In the step count example, the logic regression model is fitted to the three unit tests to generate the prediction for calculating precision and recall.
-@fig-logic-reg shows the chosen logic regression model: a combination of test 1 and test 3 with an OR operator and @tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
-We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
-The harmonic and arithmetic means are included to combine the three measures.
-The results show that the two tests produced by the logic regression can accurately predict 82.1% of all *predicted unexpected cases*, and 83.6% cases of all *actual unexpected results* in the simulation data.
-The regression tree produces a similar model of first split on test1 and then split on test3, and results in the same accuracy and overall score as the logic regression model.
+For comparison, the regression tree produces a similar prediction to the logic regression, by first splitting on test1 and then splitting on test3, and results in the same accuracy and overall score as the logic regression model.
+However, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
+The logic regression tree is also directly comparable to other diagnostic techniques, which we discuss further in @sec-discussion.
 

### 2024-12-09 index.qmd
Discussion

@@
 In this section, we propose a procedure to measure the effectiveness of checks that, when combined using logical operators, contribute to an unexpected outcome.
-A small set of independent checks is considered effective if it translates to unexpected outcomes and @fig-metric-calc provides an overview of the calculation.
+A small set of independent checks is considered effective if it translates to unexpected outcomes.
+
+The approach relies on the use of simulated datasets that are generated based on the analyst's knowledge and assumptions about the data generation mechanism.
+Datasets are simulated to have the same structure and characteristics of the observed data that will be analyzed at some point in the future.
+For each simulated dataset, we can apply a collection of analysis validation checks to the dataset and record which ones were TRUE and which were FALSE.
+We can also compute the outcome of the analysis to see whether the outcome is unexpected in a given simulated datsets.
+We can then generate many datasets, each time applying the analysis validation checks and computing the outcome.
+After generating many datasets, we can relate the patterns in the analysis validation checks to the likely that the outcome will be unexpected in a given dataset.
+@fig-metric-calc provides an overview of the process.
 
@@
 
-Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study].
-Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference].
-The example demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
+In the study of the health effects of outdoor air pollution, one area of interest is the association between short-term, day-to-day changes in particulate matter air pollution and daily mortality counts.
+Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine].
+For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
 
-This PM10 coefficient expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005] and 0 if it is within.
-Multiple factors can affect the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
+The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10.
+Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between [0, 0.005], after adjusting for daily temperature [@samet200fine; @welty2005acute].
+Observing an estimated coefficient outside of this interval would be highly unusual and would warrant a serious re-examination of the analysis process.
+Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval [0, 0.005].
+In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
+
+This PM10 coefficient expectation can be framed as an analysis check that fails, labelled as 1, if the estimate of the PM10 coefficient (adjusted for temperature) is outside the range [0, 0.005], and 0 if it is within.
+Multiple factors can affect the estimated PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
 Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate.
 Outliers in the three variables can also leverage the coefficient.
-While these are possible factors that could affect the analysis result, it is not clear the cutoff values for these checks to determine a failure.
-Here we create a list of checks in @tbl-checks with varied cutoff values for each:
+While these are possible factors that could affect the analysis result, it is not clear what the cutoff values for these checks should be to determine a failure.
+Here we consider a list of checks in @tbl-checks with varied cutoff values for each:
 
@@
 
-To generate replicate of the data, we first simulate the correlation matrix of the three variables in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size.
+## Data Simulation
+
+To generate replicates of the dataset, we first generate the correlation matrix of the three variables (PM10, mortality, and temperature) in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size.
 The multivariate normal distribution is transformed using the normal CDF before the inverse CDF of the assumed distributions of the three variables is applied.
 To determine the appropriate distribution of each variable, various distributions are fitted and compared.
-This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$.
-AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit.
-AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature.
-To include the effect of outlier, we add a single outlier to the data for both variable mortality and PM10.
+This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for PM10 and temperature; and beta for PM10 after rescaling the data to $[0,1]$.
+
+To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed dataset to refine our simulations.
+AIC is used to determine the best distribution fit for each variable with the QQ-plot presented in @fig-dist-fit to evaluate the fit.
+AIC suggests a negative binomial distribution for mortality, a beta distribution for PM10 (multiple by 100 to recover the original scale), and a Weibull distribution for temperature.
+To include the potential effect of outliers, we add a single outlier to the data for both the mortality and PM10 variables.
 
@@
 
+A logic regression is fitted using all variations of the tests in @tbl-checks to predict whether the PM10 coefficient is unexpected.
+@fig-linear-reg-tree shows the optimal logic regression tree from the fitted model.
+Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in @tbl-linear-reg.
+
 ::: {#fig-linear-reg-tree}
@@
 #| fig.width: 8
-#| fig.heigh: 3
+#| fig.height: 4
 fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
@@
 
-Logic regression model fitted to the fourteen unit tests and the outcome expectation (unexpected) as the response variable. The model suggests the relationship:
-(sample size *greater than* 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
+Logic regression model fitted to the fourteen unit tests and the outcome expectation (unexpected) as the response variable.
+The model suggests the relationship: (sample size *greater than* 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
 [sorry the proper plot is not ready].
@@
 
-A logic regression is fitted to predict whether the PM10 coefficient is unexpected using the checks listed in @tbl-checks and @fig-linear-reg-tree shows the logic regression tree from the fitted model.
-Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in  @tbl-linear-reg.
-The logic regression model picks up the following cutoff value for each type of check:
+As indicated in @fig-linear-reg-tree, the logic regression model picks up the following cutoff values for each type of check:
 
 -   sample size *larger than* 200
--   mortality-temperature correlation greater than -0.35
--   mortality-PM10 correlation less than -0.04
--   mortality contain outliers that are detected by the univariate outlier detection
+-   mortality-temperature correlation greater than $-0.35$
+-   mortality-PM10 correlation less than $-0.04$
+-   mortality data contains outliers that are detected by the univariate outlier detection
 
 <!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->
-
 The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature.
@@
 
+As shown in @fig-linear-reg-tree, there is no single analysis check in the tree that predicts an unexpected outcome.
+rather at least three checks in the tree must be TRUE in order for the model to predict an unexpected outcome.
+Given the high independence of the tests (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.
+
 # Discussion {#sec-discussion}
+
+In this paper we have developed an approach to using analysis validation checks to externalize the assumptions about the data and analysis tools made during the data analysis process.
+These checks can serve as a useful summary of the analyst's thought process and can desscribe how characterstics of the data may lead to unexpected outcomes.
+Using logic regression, we can develop a graphical summary of the analysis validation checks as well as use the logic regression fitting process to choose the optimal set of checks.
+The logic regression model can also be used to develop summaries of the precision and recall of the collection of analysis validation checks in predicting the likelihood of an unexpected outcome.
+We demonstrated our method on an example relating daily mortality to outdoor air pollution data.
+
+An interesting connection can be drawn between our logic regression trees and a tool used in systems engineering known as a fault tree.
+A fault tree is used for conducting a structured risk assessment and has a long history in aviation, aerospace, and nuclear power applications [@vesely1981fault].
+A fault tree is a graphical tool that describes the possible combinations of causes and effects that lead to an anomaly.
+At the top of the tree is a description of an anomaly.
+The subsequent branches of the tree below the top event indicate possible causes of the event immediately above it in the tree.
+The tree can then be built recursively until we reach a root cause that cannot be further investigated.
+Each level of the tree is connected together using logic gates such as AND and OR gates.
+The leaf nodes of the tree indicate the root causes that may lead to an anomaly.
+While the logic regression tress are not identical to fault trees, they share many properties, such as the tree-based structure and the indicator of root causes at the leaf nodes.
+Perhaps more critically, both serve as graphical summaries of the assumptions made in a problem and the specific violations of those assumptions that could lead to an unexpected result.
+While fault trees are often used to discover the root cause of an anomaly after it occurs, an important use case for fault trees is to develop a comprehensive understanding of a system *before* an anomaly occurs [@michael2002fault].
 

### 2024-12-10 _manuscript/_tex/index.tex
render pdf

@@
 their procedures to identify potential issues. While experienced
-researchers can typically diagnose problems by checking a few key
+researchers can often quickly diagnose problems by checking a few key
 assumptions, others may struggle to identify the root causes. These
@@
 formalize these informal assumptions by framing them as binary
-\emph{analysis validation check}. We then introduce a procedure to
+\emph{analysis validation checks}. We then introduce a procedure to
 quantify how violations of these checks may lead to unexpected results
@@
 data and evaluates both accuracy and redundancy. Accuracy is calculated
-through binary classification metric, while redundancy is measured using
-mutual information. We demonstrate this approach with a toy example on
-the fitness step count data and a generalized linear model exmaple
-examining the effect of PM10 on mortality.
+through a binary classification metric, while redundancy is measured
+using mutual information. We demonstrate this approach with a toy
+example based on fitness step count data and a generalized linear model
+example examining the effect of particulate matter air pollution on
+daily mortality.
 \end{abstract}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  TODO: emphasize trustworthy data science since it is the theme of the
-  special issue
-\end{itemize}
-
 In data analysis, experienced researchers often rely on their prior
@@
 typically presented with the final outcomes of the analysis cycle where
-the results and expectations are aligned -- achieved either by refining
+the results and expectations are aligned, achieved either by refining
 the analysis or updating the expectations based on statistical evidence
@@
 outcomes. Similarly, when researchers with different background
-knowledge it becomes unclear whether discrepancies arise from differing
-expectations or from the use of statistical techniques.
+knowledge view results that they find to be unexpected, it becomes
+unclear whether discrepancies arise from differing expectations or from
+the use of statistical techniques.
 
-One might gain insight into analysts' thought process by speaking with
-them directly or watching screencast videos they produce, such as,
-TidyTuesday screencast videos. However, direct conversation are not
-scalable and may not be feasible while creating educational videos
-requires significant effort from the researchers. Ideally, there could
-be a way to make these expectations explicit and accessible to others.
-Even better, if the encoding is machine-readable, we could analyze these
-expectations and learn from the analysis -- whether the checks also
-apply to other researchers analyzing new data in the same context,
-whether they reflect common practices in the field, or whether they are
-specific to the data or analysis at hand.
+One might gain insight into analysts' thought processes by speaking with
+them directly or watching them work via screencast videos they produce,
+such as, TidyTuesday screencast videos or think-aloud type studies
+\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
+not scalable and may not always be feasible; creating educational
+screencast videos requires significant effort from the researchers.
+Ideally, there could be a way to make expectations about data analysis
+explicit and accessible to others. Even better, if the encoding were
+machine-readable, we could analyze these expectations and learn from the
+analysis itself. For example, we could answer questions about whether
+the checks also apply to other researchers analyzing new data in the
+same context, whether they reflect common practices in the field, or
+whether they are specific to the data or analysis at hand.
+
+The externalization of the data analysis process is a practice that has
+potential to improve the trustworthiness of analyses in general and the
+trustworthiness of subsequent products, such as machine learning models,
+that may be built on such analyses. While publication of analysis code
+and data is now a common a requirement for the sake of reproducibility
+\citep{peng2011reproducible}, the publication code alone is often
+insufficient for understanding the thought process behind an analysis.
+Code corresponding to published analyses often reflect the final
+decisions made about analysis and do not reveal the decisions or
+assumptions made about the data processing. Thus, a reader looking at
+published code can often be left with many questions about why certain
+choices were made. Developing an approach to reveal some of this process
+without requiring a reader to essentially reconstruct the analysis
+process from the raw data would provide an improved basis for trusting
+an analysis result \citep{peng2021reproducible}.
 
@@
 assumptions as \emph{analysis validation checks}, which allows us to
-examine these assumptions made during analysis to diagnose unexpected
-outcomes. We then introduce a procedure that provides a quantitative
-measure on how violations in a tree of checks, derived from individual
-checks, will lead to an unexpected result. The procedure, based on
-simulations of the original data, calculates accuracy and redundancy.
-Accuracy is determined using binary classification metrics, precision
-and recall, from a logic regression fit while redundancy is measured
-using mutual information. The proposed workflow offers a numerical
-guarantee that the analysis will produce the expected results, assuming
-the data generating mechanism holds. {[}something about trustworthy data
-science{]}
+examine the assumptions made during an analysis and to diagnose
+unexpected outcomes. We then introduce a procedure that provides a
+quantitative measure of how violations in a tree of analysis checks,
+derived from individual checks, will lead to an unexpected result. The
+procedure, based on simulations of the original data, calculates the
+accuracy and redundancy of the analysis checks. Accuracy is determined
+using binary classification metrics, precision and recall, from a logic
+regression fit \citep{ruczinski_logic_2003}, while redundancy is
+measured using mutual information. The proposed workflow offers a
+numerical guarantee that the analysis will produce the expected results,
+assuming the assumptions about the data generating mechanism hold.
 
 The rest of the paper is organized as follows:
-Section~\ref{sec-lit-review} reviews the concept of diagnosing
-unexpected outcomes and data quality checks. Section~\ref{sec-plan}
-introduces the concept of analysis validation check, illustrated with a
-toy example on fitness step count. Section~\ref{sec-method} describes
-the procedure that quantify how checks combined in logical operators can
-predict unexpected outcomes. Section~\ref{sec-pm10-mortality} applies
-this procedure to examine the effect of PM10 on mortality in public
-health in a generalized linear model. Section~\ref{sec-discussion}
-discusses a few key considerations and Section~\ref{sec-conclusion}
-concludes the paper.
+Section~\ref{sec-lit-review} reviews the concepts of diagnosing
+unexpected outcomes and general data quality checks.
+Section~\ref{sec-plan} introduces the concept of analysis validation
+checks, illustrated with a toy example based on fitness step count data.
+Section~\ref{sec-method} describes the procedure that quantifies how
+analysis validation checks combined using logical operators can predict
+unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
+applies this procedure to a larger example that estimates the effect of
+particulate matter air pollution on daily mortality.
+Section~\ref{sec-discussion} discusses a few key considerations and
+Section~\ref{sec-conclusion} concludes the paper.
 
-\section{Literature review}\label{sec-lit-review}
+\section{Related Work}\label{sec-lit-review}
 
@@
 
-\begin{itemize}
-\item
-  They shape how we interpret the results and assess whether they are
-  consistent with existing knowledge or indicate the need for updates.
-  \citep{grolemund_cognitive_2014}
-\item
-  \citep{peng_diagnosing_2021} describes three pedagogical exercises of
-  introducing diagnosing unexpected outcome into a classroom setting.
-\item
-  what if the expectation is ``wrong''
-\end{itemize}
+The concept of framing data analysis as a sense-making process was
+originally presented by \citep{grolemund_cognitive_2014} based on
+seminal work by \citep{wild1999statistical}. Key to any sense-making
+process is a model for the world (i.e.~expectations for what we might
+observe) and observed data with which we can compare our expectations.
+If there is a significant deviation between what we observe and our
+expectations, then a data analysis must determine what is causing that
+deviation. A naive approach would be to update our model for the world
+to match the data, under the assumption that the initial expectation was
+incorrect. However, experienced analysts know that the reality can be
+more nuanced than that, with errors occurring in data collection or data
+processing that can have an impact on final results.
+
+The skill of diagnosing unexpected data analysis results is not one that
+has received significant attention in the statistics literature. While
+the concept of diagnosis is often embedded in model checking or data
+visualization techniques, systematic approaches to identifying the root
+cause of an unexpected analysis result are typically not presented
+\citep{peng2022perspective}. \citep{peng_diagnosing_2021} proposed a
+series of exercises for training students in data analysis to diagnose
+different kinds of analysis problems such as coding errors or outliers.
+They provide a systematic approach involving working backwards from the
+analysis result to identify potential causes. There are parallels here
+to the concept of debugging and testing in software engineering
+\citep{donoghue2021teaching}. For example, \citep{li2019towards} found
+that experienced engineers were generally able to identify problems in
+code faster than novices, and that the ability to debug code required
+knowledge that cut across different domains.
+
+If it is true that the speed with which data analysts can identify
+problems with an analysis is related to their experience working with a
+given type of data, then there is perhaps room to improve the analytic
+process by externalizing the aspects that an analyst learns through
+experience. That way, inexperienced analysts could examine the thought
+process of an experienced analyst and learn to identify factors that can
+cause unexpected results to occur.
 
@@
 A substantial body of literature has addressed the definition of data
-quality \citep[more]{8642813} and developed frameworks, that includes
+quality \citep[more]{8642813} and has developed frameworks that include
 dimensions, attributes, and measures to evaluate and improve data
@@
 \citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
-These frameworks are often used information system and database
-management and support business decision-making in the industry. For
-research purposes, high-quality data ensures the credibility of
-scientific findings and supports reproducibility and reusability in
+These frameworks are often used in information systems and database
+management and support business decision-making in various industries.
+For research purposes, high-quality data ensures the credibility of
+scientific findings and supports reproducibility and re-usability in
 future studies {[}ref{]}. With the growing prevalence of open data in
 scientific research, the consumers or users of the data typically are no
-longer the data producers or collectors who have the full knowledge of
-data in hand, prompting more interest towards the data quality checks in
-the data analysis process. In R, there are some packages, like
+longer the producers or collectors of the data who would have the full
+knowledge of data in hand, prompting more interest towards data quality
+checks in the data analysis process. In R, there are some packages, like
 \texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
-provides basic data screening and reporting, while another class of
-packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
-\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
-providing data validation tools, allowing users to define customized
-data quality checks based on the applications.
+that provide basic data screening and reporting tools, while another
+class of packages, e.g.~\texttt{assertr} \citep{assertr},
+\texttt{validate} \citep{validate}, and \texttt{pointblank}
+\citep{pointblank} focuses on providing data validation tools, allowing
+users to define customized data quality checks based on the
+applications.
+
+The literature on data quality typically focuses on the intrinsic or
+inherent quality of the data themselves, rather than the data's
+relationship to any specific data analysis. So for example, if a column
+in a data table is expecting numerical data, but we observe a character
+value in one of the entries, then that occurrence would trigger some
+sort of data quality check. This type of quality check can be triggered
+without any knowledge of what the data will ultimately be used for.
+However, for a given analysis, we may require specific aspects of the
+data to be true because they affect the result being computed.
+Conversely, certain types of poor quality data may have little impact on
+the ultimate result of an analysis (e.g.~data that are missing
+completely at random). Defining data quality in terms of what may affect
+a specific analysis outcome or result has the potential to open new
+avenues for defining data checks and for building algorithms for
+optimizing the collection of checks defined for a specific analysis.
 
@@
 
-\subsection{Framing expectations as
-checks}\label{framing-expectations-as-checks}
+Expectations represent our understanding of certain aspects of the
+analysis and the data, independent of the results of the analysis
+itself. When observed outcomes deviate from these expectations, analysts
+often revisit the analysis process to identify potential issues, refine
+methods, or revise assumptions. Experienced analysts can typically
+identify issues quickly and correct them on the spot, but they often do
+so without discussing the underlying reasoning, making it harder for
+less experienced researchers to learn and master these skills.
 
-Expectations represent our beliefs about certain aspects of the
-analysis, independent of the analysis itself. When outcomes deviate from
-these expectations, analysts often revisit the analysis cycle to
-identify potential issues, refine methods, or revise assumptions.
-Experienced analysts can typically identify issues quickly and correct
-them on the spot, but they often do so without discussing the underlying
-reasoning, making it harder for less experienced researchers to learn
-and master these skills. Here, we introduce the concept of
-\emph{analysis validation checks}, which frames these expectations or
-assumptions as explicit checks that returns a TRUE or FALSE output given
-the data analyzed at hand. Inspired by data validation checks
-\citep{validate}, where ensure datasets meet expected format and
-quality, analysis validation checks reverse the approach: they validate
-the assumptions about the data necessary for the analysis to produce the
-reported results. This allows the concept to encompass a broad range of
-checks, such as data quality (i.e.~missing data, how the data is
-structured), data distribution and outliers, bivariate and multivariate
-relationships between variables, and contextual information. These
-externalized checks provide insights into analysts' thoughts process
-process and offer several benefits:
+Here, we introduce the concept of \textbf{analysis validation checks},
+which frame these expectations or assumptions as explicit checks that
+return a TRUE or FALSE result given the data analyzed at hand. Inspired
+by the concept of data validation checks (\citet{validate}), which are
+designed to ensure that datasets meet expected formats and quality,
+analysis validation checks reverse the approach: they validate the
+assumptions about the data necessary for the analysis to produce the
+\emph{expected results}, as defined by the analyst. The focus on
+expected results allows the concept of analysis validation checks to
+encompass a broad range of checks, such as data quality (i.e.~missing
+data, how the data are structured), data distribution and outliers,
+bivariate and multivariate relationships between variables, and other
+contextual information.
+
+Our proposed analysis validation checks provide insights into an
+analyst's thought process and offer the following benefits:
 
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi})}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  help students and junior researchers develop skills to diagnose
-  unexpected analysis results,
+  Serve as clear checkpoints to support the replication or application
+  of methods to (new) data by programmatically communicating the
+  requirements or assumptions made of the data;
 \item
-  serve as clear checkpoints to support replication or application of
-  methods to (new) data,
+  Align assumptions among researchers from different domain backgrounds
+  who may have different expectations about the data;
 \item
-  align assumptions among researchers from different domain backgrounds,
-  and
+  Improve analysis transparency, reproducibility, and trustworthiness by
+  externalizing a key part of the analysis process; and
 \item
-  improve analysis transparency, reproducibility, and trustworthiness.
+  Quantify the effectiveness of analysis checks for predicting the
+  expected outcome (see Section~\ref{sec-method});
 \end{enumerate}
 
-\subsection{A toy example}\label{a-toy-example}
+In addition to the above benefits, the development and publication of
+analysis checks has the potential to help students, inexperienced
+analysts, and junior researchers develop the skills needed to diagnose
+unexpected analysis results for a given type of data because the
+assumptions made about the data are made transparent. The analysis
+checks can serve as a basis for new analysts to have conversations about
+the data they are analyzing and to develop a better understanding of the
+potential data generation process.
 
-Consider a 30-day step count experiment in public health. Testers are
+\subsection{A Toy Example}\label{sec-toy}
+
+Consider a 30-day step count experiment in public health. Subjects are
 instructed to walk at least 8,000 steps each day, with an expected
-average of 9,000 steps, tracked by a step counter app. While there may
-be occasional ``low'' days due to factors such as forgetting to wear
-their watch or unfavorable weather condition limiting outdoor
-activities, there are also ``high'' days after an outdoor hike or
-intense workout. After 30 days, you review the data and expect no more
-than five days with step counts falling below the 8,000-step threshold
-{[}I think we need a bit more scenario to justify we expect no more than
-5 days, say it needs to have at least 25 days for it to be valid for the
-next stage research??{]}.
-
-To simulation this data, three normal distributions are used for the
-daily step counts: \(\mathcal{N}(4000, 200)\) for low days,
-\(\mathcal{N}(12000, 200)\) for high days, and
-\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
-days can be simulated from a Poisson distribution with \(\lambda = 4\).
-Figure~\ref{fig-step-count} displays the number of days with fewer than
-8,000 steps across 300 simulated 30-day periods.
+average of 9,000 steps, tracked by a step counter app. After 30 days, we
+review the data and examine the number of steps taken each day. With
+data of this nature, we may expect there to be occasional ``low'' days
+due to factors such as forgetting to wear their watch or unfavorable
+weather conditions limiting outdoor activities. We may also expect
+``high'' days recorded after an outdoor hike or intense workout. Given
+the requirements of the study, we expect that for a given subject there
+would be no more than five days with step counts falling below the
+8,000-step threshold (``invalid days''). If there were more than five
+invalid days, the data from that subject would not be usable for the
+study.
 
 In this scenario, the outcome is that the number of days with a step
-count below 8,000 being no more than five. To diagnose potential reasons
-why this outcome expectation might fail, we can establish a few analysis
-validation checks. For example, if the average step count is too low,
-this may suggest there are too many low days, potentially lead to the
-unexpected outcome. Similarly, we can also check the quantile of the
-step count, if more than a third of the days fall below 8,000, this
-could indicate an excess of low-count days. Additionally, we may may
-expect the standard deviation of the step count not to be overly large.
-These considerations yield the following three unit tests as plan
-expectations:
+count below 8,000 and the expectation is that this outcome would take a
+value no more than 5. If the expectation is violated, we can consider
+the potential configurations of the data that could lead to such an
+anomaly. To diagnose potential reasons why this outcome expectation
+might fail, we can establish a few analysis validation checks in
+anticipation of seeing the data. For example, if the average step count
+is too low, this may suggest that the subject is incapable of taking the
+required number of steps, potentially leading to the unexpected outcome.
+Similarly, we can also check the quantile of the step count, if more
+than a third of the days fall below 8,000, this could indicate an excess
+of low-count days. Additionally, we may may expect the standard
+deviation of the step count not to be overly large. These considerations
+yield the following three analysis validation checks:
 
@@
 \item
-  test2: the test fails if the 30th percentile of the step counts is
-  below 8,200
+  test2: the test fails if the 30th percentile of the observed step
+  counts is below 8,200
 \item
-  test3: the test fails if the standard deviation of the step counts
-  exceeds 2,500.
+  test3: the test fails if the standard deviation of the observed step
+  counts exceeds 2,500.
 \end{itemize}
+
+The cutoff values chosen for these tests would presumably be chosen
+based on prior experience with these kinds of data, but could also be
+optimized using the method presented in the next section.
 
@@
 
+To simulate this data, three normal distributions are used for the daily
+step counts: \(\mathcal{N}(4000, 200)\) for low days,
+\(\mathcal{N}(12000, 200)\) for high days, and
+\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
+days can be simulated from a Poisson distribution with \(\lambda = 4\).
+Figure~\ref{fig-step-count} displays the number of days with fewer than
+8,000 steps across 300 simulated 30-day periods.
+
 \section{Method}\label{sec-method}
 
-\subsection{A workflow to assess the quality of unit
-tests}\label{a-workflow-to-assess-the-quality-of-unit-tests}
+While some checks may be crucial and directly indicate an unexpected
+outcome, others may be tangential to the problem at hand and not
+indicate a root cause of an unexpected outcome. In this section, we
+propose a procedure to measure the effectiveness of checks that, when
+combined using logical operators, contribute to an unexpected outcome. A
+small set of independent checks is considered effective if it translates
+to unexpected outcomes.
 
-While some checks are crucial and can lead to unexpected outcomes,
-others may be tangential to the root cause. In this section, we propose
-a procedure to measure the violation of checks, combined using logical
-operators, contribute to an unexpected outcome. A small set of
-independent checks is considered effective if it translates to
-unexpected outcomes and Figure~\ref{fig-metric-calc} provides an
-overview of the calculation.
+The approach relies on the use of simulated datasets that are generated
+based on the analyst's knowledge and assumptions about the data
+generation mechanism. Datasets are simulated to have the same structure
+and characteristics of the observed data that will be analyzed at some
+point in the future. For each simulated dataset, we can apply a
+collection of analysis validation checks to the dataset and record which
+ones were TRUE and which were FALSE. We can also compute the outcome of
+the analysis to see whether the outcome is unexpected in a given
+simulated datsets. We can then generate many datasets, each time
+applying the analysis validation checks and computing the outcome. After
+generating many datasets, we can relate the patterns in the analysis
+validation checks to the likely that the outcome will be unexpected in a
+given dataset. Figure~\ref{fig-metric-calc} provides an overview of the
+process.
 
@@
 positives and false negatives. While a false positive can raise caution
-or skepticism on the data, both false positive and false negative
-suggest the checks are over or lack sensitivity to unexpected outcomes.
+or skepticism on the data, the presence of both false positives and
+false negatives suggest that the checks are overly sensitive or lack
+sensitivity to unexpected outcomes.
+
 To incorporate the effect of multiple checks on the outcome, a logic
-regression model \citep{ruczinski_logic_2003} is fitted to the checks
-and whether the outcome is unexpected. Originally developed for SNP
-microarray data, logic regression constructs Boolean combinations of
-binary variables, in a tree structure, to predict both binary and
-numerical outcomes. Compared to other tree-based methods for
-binary-binary prediction, the Boolean combinations from the logic
-regression model produce a tree structure that can be directly
-interpreted as the possible combination of checks leading to an
-unexpected outcome, without the need to invert the tree as required in
-classic tree-based methods. The model is then used to predict the
-outcome expectation based on the checks, and the prediction is compared
-to the actual outcome expectation to calculate the precision and recall
-of the checks.
+regression model \citep{ruczinski_logic_2003} is fitted to the analysis
+validation checks using the indicator of an unexpected result as the
+outcome in the model. Originally developed for SNP microarray data,
+logic regression constructs Boolean combinations of binary variables, in
+a tree structure, to predict both binary and numerical outcomes.
+Compared to other tree-based methods for binary-binary prediction, the
+Boolean combinations from the logic regression model produce a tree
+structure that can be directly interpreted as the possible combination
+of checks leading to an unexpected outcome, without the need to invert
+the tree as required in classic tree-based recursive partitioning
+methods. The logic regression model is then used to predict the analysis
+result based on the values of checks, and the prediction is compared to
+the actual analysis result in order to calculate the precision and
+recall of the checks.
 
 While tests may score high on accuracy, they may be less effective at
-explaining the reasons behind unexpected results. This could happen if a
-set of tests are all tangentially related to the cause of the unexpected
-results, but none addressing the root cause. It may also occur if the
-tests are correlated with one another, leading to redundancy.
+explaining the various reasons behind unexpected results. This could
+happen if, for example, a set of tests are all tangentially related to
+the cause of the unexpected results, but none addresses the root cause.
+It may also occur if the tests are highly correlated with one another,
+leading to redundancy.
 
@@
 
-To combine precision, recall, and independence into a single metric,
-mathematical means, such as arithmetic mean, harmonic mean, and
-quadratic mean can be used. The differences among these means are
-minimal when the three metrics are similar. However, as the differences
-among the metrics increases, the harmonic mean tends to produce the
-smallest overall score, as it penalizes low values, while the quadratic
-mean tends to produce the largest score by rewarding higher values more.
-For simple interpretation of the score, the arithmetic mean is
-preferred, while in applications where the difference between precision,
-recall, and independence need to be penalized or rewarded more, the
-harmonic and quadratic mean should be considered.
+To combine precision, recall, and independence into a single metric, we
+can combine the three scores using the arithmetic mean, harmonic mean,
+or quadratic mean. The differences among these means are minimal when
+the three metrics are similar. However, as the differences among the
+metrics increases, the harmonic mean tends to produce the smallest
+overall score, as it penalizes low values, while the quadratic mean
+tends to produce the largest score by rewarding higher values more. For
+simple interpretation of the score, the arithmetic mean is preferred,
+while in applications where the difference between precision, recall,
+and independence need to be penalized or rewarded more, the harmonic and
+quadratic mean should be considered.
 
-\subsection{Toy example revisited}\label{toy-example-revisited}
+\subsection{Toy Example Revisited}\label{toy-example-revisited}
+
+Returning to the step count example introduced in Section~\ref{sec-toy},
+the logic regression model is fitted to the three analysis checks
+described previously to generate the prediction of the unexpected
+outcome, which is whether the number of days with fewer than 8,000 steps
+is greater than 5. The predictions from the logic regression model can
+be compared with the simulated true outcome for calculating the
+precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
+best-fitting logic regression model, which is a combination of test1
+(``mean step count is below 8,200'') and test3 (``standard deviation
+exceeds 2,500'') with an OR operator. In other words, if either test1 or
+test3 is true, then we would predict an unexpected outcome in the
+analysis (i.e.~the number of days with fewer than 8,000 steps is greater
+than 5).
 
@@
 \end{figure}%
+
+Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
+independence for the three individual tests and the combined test rule
+(test1 OR test3) from the logic regression. We also include the metric
+calculated from fitting a regression tree model to the data to compare
+the performance of the logic regression model. The harmonic and
+arithmetic means are included to combine the three measures. The results
+show that the two tests produced by the logic regression can accurately
+predict 83.6\% cases of all \emph{actual unexpected results} in the
+simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
+results} were in fact observed to be unexpected.
 
@@
 
-In the step count example, the logic regression model is fitted to the
-three unit tests to generate the prediction for calculating precision
-and recall. Figure~\ref{fig-logic-reg} shows the chosen logic regression
-model: a combination of test 1 and test 3 with an OR operator and
-Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
-independence for the three individual tests and the combined test rule
-(test1 OR test3) from the logic regression. We also include the metric
-calculated from fitting a regression tree model to the data to compare
-the performance of the logic regression model. The harmonic and
-arithmetic means are included to combine the three measures. The results
-show that the two tests produced by the logic regression can accurately
-predict 82.1\% of all \emph{predicted unexpected cases}, and 83.6\%
-cases of all \emph{actual unexpected results} in the simulation data.
-The regression tree produces a similar model of first split on test1 and
-then split on test3, and results in the same accuracy and overall score
-as the logic regression model.
+For comparison, the regression tree produces a similar prediction to the
+logic regression, by first splitting on test1 and then splitting on
+test3, and results in the same accuracy and overall score as the logic
+regression model. However, we argue that the logic regression tree shown
+in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
+because it provides a direct representation of which combinations of
+analysis checks lead to unexpected outcomes. The logic regression tree
+is also directly comparable to other diagnostic techniques, which we
+discuss further in Section~\ref{sec-discussion}.
 
@@
 
-Consider a generalized linear model (GLM) to study the effect of PM10 on
-mortality {[}TODO: provide more context of the mortality-PM10 study{]}.
-Analysts may expect a PM10 coefficient between {[}0, 0.005{]} after
-considering the temperature confounding {[}TODO: reference{]}. The
-example demonstrates how the procedure presented in
-Section~\ref{sec-method} can be used to select cutoff values in the
-checks to diagnose an unexpected PM10 coefficient from the generalized
-linear model.
+In the study of the health effects of outdoor air pollution, one area of
+interest is the association between short-term, day-to-day changes in
+particulate matter air pollution and daily mortality counts. Substantial
+work has been done to study this question and to date, there appears to
+be strong evidence of an association between particulate matter less
+than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
+mortality from all non-accidental causes \citep{samet2000fine}. For our
+second example, we use the problem of studying PM10 and mortality along
+with data from the National Morbidity, Mortality, and Air Pollution
+Study (NMMAPS) to demonstrate how our analysis validation checks
+described in Section~\ref{sec-method} can be applied.
 
-This PM10 coefficient expectation can be framed into a check that fails,
-labelled as 1, if the PM10 coefficient is outside the range {[}0,
-0.005{]} and 0 if it is within. Multiple factors can affect the PM10
+The typical approach to studying the association between PM10 and
+mortality is to apply a generalized linear model with a Poisson link to
+relate daily mortality counts to daily measures of PM10. Based on
+previous work and the range of effect sizes published in the literature,
+an analyst might expect the coefficient for PM10 in this GLM to lie
+between {[}0, 0.005{]}, after adjusting for daily temperature
+\citep{samet200fine, welty2005acute}. Observing an estimated coefficient
+outside of this interval would be highly unusual and would warrant a
+serious re-examination of the analysis process. Therefore, our
+unexpected outcome in this analysis will be the binary indicator of
+whether the estimated coefficient for PM10 in the GLM lies outside of
+the interval {[}0, 0.005{]}. In addition to providing a more substantial
+problem for our methods, this example also demonstrates how the
+procedure presented in Section~\ref{sec-method} can be used to select
+cutoff values in the analysis checks to diagnose an unexpected PM10
+coefficient from the generalized linear model.
+
+This PM10 coefficient expectation can be framed as an analysis check
+that fails, labelled as 1, if the estimate of the PM10 coefficient
+(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
+it is within. Multiple factors can affect the estimated PM10
 coefficient, such as the sample size, the strength of the correlation
@@
 possible factors that could affect the analysis result, it is not clear
-the cutoff values for these checks to determine a failure. Here we
-create a list of checks in Table~\ref{tbl-checks} with varied cutoff
-values for each:
+what the cutoff values for these checks should be to determine a
+failure. Here we consider a list of checks in Table~\ref{tbl-checks}
+with varied cutoff values for each:
 
@@
 
-To generate replicate of the data, we first simulate the correlation
-matrix of the three variables in a grid and then use a Gaussian copula
-to generate a multivariate normal distribution based on the specified
-correlation matrix and sample size. The multivariate normal distribution
-is transformed using the normal CDF before the inverse CDF of the
-assumed distributions of the three variables is applied. To determine
-the appropriate distribution of each variable, various distributions are
-fitted and compared. This includes poisson and negative binomial for
-mortality; gamma, log-normal, exponential, weibull, and normal for pm10
-and temperature; and beta for pm10 after rescaling the data to
-\([0-1]\). AIC is used to determine the best distribution fit for each
-variable with qq plot presented in Figure~\ref{fig-dist-fit} to evaluate
-the fit. AIC suggests a negative binomial distribution for mortality, a
-beta distribution for PM10 (multiple by 100 to obtain the original
-scale), and a Weibull distribution for temperature. To include the
-effect of outlier, we add a single outlier to the data for both variable
-mortality and PM10.
+\subsection{Data Simulation}\label{data-simulation}
+
+To generate replicates of the dataset, we first generate the correlation
+matrix of the three variables (PM10, mortality, and temperature) in a
+grid and then use a Gaussian copula to generate a multivariate normal
+distribution based on the specified correlation matrix and sample size.
+The multivariate normal distribution is transformed using the normal CDF
+before the inverse CDF of the assumed distributions of the three
+variables is applied. To determine the appropriate distribution of each
+variable, various distributions are fitted and compared. This includes
+poisson and negative binomial for mortality; gamma, log-normal,
+exponential, weibull, and normal for PM10 and temperature; and beta for
+PM10 after rescaling the data to \([0,1]\).
+
+To ensure a reasonable likeness to data that might be used in such an
+analysis, we use characteristics of the observed dataset to refine our
+simulations. AIC is used to determine the best distribution fit for each
+variable with the QQ-plot presented in Figure~\ref{fig-dist-fit} to
+evaluate the fit. AIC suggests a negative binomial distribution for
+mortality, a beta distribution for PM10 (multiple by 100 to recover the
+original scale), and a Weibull distribution for temperature. To include
+the potential effect of outliers, we add a single outlier to the data
+for both the mortality and PM10 variables.
 
@@
 \end{figure}%
+
+A logic regression is fitted using all variations of the tests in
+Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
+unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
+regression tree from the fitted model. Precision, recall, and
+independence score, along with their harmonic and arithmetic mean are
+calculated in Table~\ref{tbl-linear-reg}.
 
@@
 
-A logic regression is fitted to predict whether the PM10 coefficient is
-unexpected using the checks listed in Table~\ref{tbl-checks} and
-Figure~\ref{fig-linear-reg-tree} shows the logic regression tree from
-the fitted model. Precision, recall, and independence score, along with
-their harmonic and arithmetic mean are calculated in
-Table~\ref{tbl-linear-reg}. The logic regression model picks up the
-following cutoff value for each type of check:
+As indicated in Figure~\ref{fig-linear-reg-tree}, the logic regression
+model picks up the following cutoff values for each type of check:
 
@@
 \item
-  mortality-temperature correlation greater than -0.35
+  mortality-temperature correlation greater than \(-0.35\)
 \item
-  mortality-PM10 correlation less than -0.04
+  mortality-PM10 correlation less than \(-0.04\)
 \item
-  mortality contain outliers that are detected by the univariate outlier
-  detection
+  mortality data contains outliers that are detected by the univariate
+  outlier detection
 \end{itemize}
@@
 
+As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
+analysis check in the tree that predicts an unexpected outcome. rather
+at least three checks in the tree must be TRUE in order for the model to
+predict an unexpected outcome. Given the high independence of the tests
+(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
+only likely after multiple anomalies are observed in the data.
+
 \section{Discussion}\label{sec-discussion}
+
+In this paper we have developed an approach to using analysis validation
+checks to externalize the assumptions about the data and analysis tools
+made during the data analysis process. These checks can serve as a
+useful summary of the analyst's thought process and can desscribe how
+characterstics of the data may lead to unexpected outcomes. Using logic
+regression, we can develop a graphical summary of the analysis
+validation checks as well as use the logic regression fitting process to
+choose the optimal set of checks. The logic regression model can also be
+used to develop summaries of the precision and recall of the collection
+of analysis validation checks in predicting the likelihood of an
+unexpected outcome. We demonstrated our method on an example relating
+daily mortality to outdoor air pollution data.
+
+An interesting connection can be drawn between our logic regression
+trees and a tool used in systems engineering known as a fault tree. A
+fault tree is used for conducting a structured risk assessment and has a
+long history in aviation, aerospace, and nuclear power applications
+\citep{vesely1981fault}. A fault tree is a graphical tool that describes
+the possible combinations of causes and effects that lead to an anomaly.
+At the top of the tree is a description of an anomaly. The subsequent
+branches of the tree below the top event indicate possible causes of the
+event immediately above it in the tree. The tree can then be built
+recursively until we reach a root cause that cannot be further
+investigated. Each level of the tree is connected together using logic
+gates such as AND and OR gates. The leaf nodes of the tree indicate the
+root causes that may lead to an anomaly. While the logic regression
+tress are not identical to fault trees, they share many properties, such
+as the tree-based structure and the indicator of root causes at the leaf
+nodes. Perhaps more critically, both serve as graphical summaries of the
+assumptions made in a problem and the specific violations of those
+assumptions that could lead to an unexpected result. While fault trees
+are often used to discover the root cause of an anomaly after it occurs,
+an important use case for fault trees is to develop a comprehensive
+understanding of a system \emph{before} an anomaly occurs
+\citep{michael2002fault}.
 

### 2024-12-10 index.qmd
add title

@@
 ---
-title: "My wonderful paper"
+title: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
 abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues.

### 2024-12-10 _manuscript/_tex/index.tex
add title
finish discussion

@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \usepackage{orcidlink}
@@
 \hypersetup{
-  pdftitle={My wonderful paper},
+  pdftitle={Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks},
   pdfauthor={H. Sherry Zhang; Roger D. Peng},
@@
 
-\title{My wonderful paper}
+\title{Inside Out: Externalizing Assumptions in Data Analysis as
+Validation Checks}
 \author{H. Sherry Zhang$\textsuperscript{1}$, Roger D.
@@
 
-\setstretch{2}
-README:
-
-\begin{itemize}
-\tightlist
-\item
-  \emph{Check for TODOs}, things inside ``{[}\ldots{]}'', other than
-  references, are my comments
-\item
-  Literature review: Section \emph{diagnosing unexpected outcomes in
-  data analysis}
-\item
-  \emph{Discussion} and \emph{Conclusion}
-\item
-  In Section \emph{Application}: provide more context on the
-  PM10-mortality study and add reference of the {[}0, 0.005{]} PM10
-  coefficient
-\end{itemize}
-
-\newpage
-
+\setstretch{1.15}
 \section{Introduction}\label{introduction}
@@
 
-TODO
+Visualization methods are also valuable tools for assessing data
+assumptions and can potentially be formalized as analysis validation
+checks. For instance, plotting a variable's distribution using a
+histogram, density plot, or bee swarm plot can reveal outliers or
+deviations from normality. These visualizations could be reframed as
+checks that fail when the data does not conform to the expected
+distribution. However, translating visualization results into binary
+checks that return (TRUE, FALSE) remains an open challenge, requiring
+either manual verification or the development of automated methods to
+interpret visualization outputs.
 
-\begin{itemize}
-\item
-  how to systematically simulate data is still unknown
-\item
-  plotting is a critical way to check data and they can still be frame
-  into checks. i.e.~the density/ histogram suggests there are outliers.
-  It is a open problem to how to encode the visualization into checks.
-\item
-  currently no automated way to generate checks. It is interesting to
-  see how check generation can be automated, although it requires the
-  inputs from experts across a wide array of common scenarios.
-\item
-  checks are also closely related to the concept of unit tests in
-  software engineering. While unit tests are designed to isolate and
-  test specific aspects of the code, it is difficult for analysis
-  validation check to do so.
-\item
-  There are cost and benefit on setting expectation on different
-  granularity. At the lowest level, one may have a plan for each data
-  entry and every data handling steps. This requires more work from the
-  analysts and may not be practical in practice. For more complex
-  analyses, analysts may divide the analysis into sections and set
-  expectations for each. They can then focus on the specific sections
-  flagged by the tests and sub-divide the sections to set expectation
-  and diagnose the analysis in a hierarchical manner.
-\end{itemize}
+Systematically generating simulated data is a key component of our
+approach. In the PM10 example, the inverse transform method is used to
+preserve the correlation structure among mortality, PM10, and
+temperature. However, simulation can become complex when additional
+restrictions are imposed. In such cases, techniques like the
+acceptance-reject method or permutation may be used to generate the
+data.
+
+Analysis validation checks are closely related to the concept of unit
+testing in software engineering. While unit tests isolate and test
+specific lines of the code, analysis validation checks focus on the
+assumptions underlying the analysis rather than the explicit code
+itself. Moreover, while software testing is deterministic, with clear
+rules for determining failure, analysis validation checks are
+probabilistic. As a result, an analysis may fail several assumption
+checks yet produce an expected outcome, or pass all checks but yield an
+unexpected result.
 

### 2024-12-10 index.qmd
finish discussion

@@
    fontsize: 12pt
-   linestretch: 2
+   linestretch: 1.15
    notebook-links: false
@@
 ```
-
-README:
-
-* *Check for TODOs*, things inside "[...]", other than references, are my comments
-- Literature review: Section *diagnosing unexpected outcomes in data analysis* 
-- *Discussion* and *Conclusion*
-- In Section *Application*: provide more context on the PM10-mortality study and add reference of the [0, 0.005] PM10 coefficient
-
-\newpage
 
@@
 
-TODO
+Visualization methods are also valuable tools for assessing data assumptions and can potentially be formalized as analysis validation checks.
+For instance, plotting a variable's distribution using a histogram, density plot, or bee swarm plot can reveal outliers or deviations from normality.
+These visualizations could be reframed as checks that fail when the data does not conform to the expected distribution.
+However, translating visualization results into binary checks that return (TRUE, FALSE) remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs.
 
--   how to systematically simulate data is still unknown
+Systematically generating simulated data is a key component of our approach.
+In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature.
+However, simulation can become complex when additional restrictions are imposed.
+In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.
 
--   plotting is a critical way to check data and they can still be frame into checks.
-i.e.
-the density/ histogram suggests there are outliers.
-It is a open problem to how to encode the visualization into checks.
-
--   currently no automated way to generate checks. It is interesting to see how check generation can be automated, although it requires the inputs from experts across a wide array of common scenarios.
-
-- checks are also closely related to the concept of unit tests in software engineering.
-While unit tests are designed to isolate and test specific aspects of the code, it is difficult for analysis validation check to do so.
-
--   There are cost and benefit on setting expectation on different granularity.
-At the lowest level, one may have a plan for each data entry and every data handling steps.
-This requires more work from the analysts and may not be practical in practice.
-For more complex analyses, analysts may divide the analysis into sections and set expectations for each.
-They can then focus on the specific sections flagged by the tests and sub-divide the sections to set expectation and diagnose the analysis in a hierarchical manner.
+Analysis validation checks are closely related to the concept of unit testing in software engineering.
+While unit tests isolate and test specific lines of the code, analysis validation checks focus on the assumptions underlying the analysis rather than the explicit code itself.
+Moreover, while software testing is deterministic, with clear rules for determining failure, analysis validation checks are probabilistic.
+As a result, an analysis may fail several assumption checks yet produce an expected outcome, or pass all checks but yield an unexpected result.
 

### 2024-12-10 _manuscript/_tex/index.tex
test -> check

@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage{orcidlink}
@@
 \item
-  test1: the test fails if the mean step count is below 8,200
+  check 1: the check fails if the mean step count is below 8,200
 \item
-  test2: the test fails if the 30th percentile of the observed step
+  check 2: the check fails if the 30th percentile of the observed step
   counts is below 8,200
 \item
-  test3: the test fails if the standard deviation of the observed step
-  counts exceeds 2,500.
+  check 3: the check fails if the standard deviation of the observed
+  step counts exceeds 2,500.
 \end{itemize}
 
-The cutoff values chosen for these tests would presumably be chosen
+The cutoff values chosen for these checks would presumably be chosen
 based on prior experience with these kinds of data, but could also be
@@
 
-While tests may score high on accuracy, they may be less effective at
+While checks may score high on accuracy, they may be less effective at
 explaining the various reasons behind unexpected results. This could
-happen if, for example, a set of tests are all tangentially related to
+happen if, for example, a set of checks are all tangentially related to
 the cause of the unexpected results, but none addresses the root cause.
-It may also occur if the tests are highly correlated with one another,
+It may also occur if the checks are highly correlated with one another,
 leading to redundancy.
@@
 
-A high mutual information value indicates redundancy among the tests,
-while a low value suggests that the tests are independent and provides
+A high mutual information value indicates redundancy among the checks,
+while a low value suggests that the checks are independent and provides
 unique information to diagnose the unexpected outcome. To standardize
@@
 precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
-best-fitting logic regression model, which is a combination of test1
-(``mean step count is below 8,200'') and test3 (``standard deviation
-exceeds 2,500'') with an OR operator. In other words, if either test1 or
-test3 is true, then we would predict an unexpected outcome in the
+best-fitting logic regression model, which is a combination of check 1
+(``mean step count is below 8,200'') and check 3 (``standard deviation
+exceeds 2,500'') with an OR operator. In other words, if either check 1
+or check 3 is true, then we would predict an unexpected outcome in the
 analysis (i.e.~the number of days with fewer than 8,000 steps is greater
@@
 \caption{\label{fig-logic-reg}Logic regression model fitted to the three
-unit tests (test1, test2, test3). The model suggests using an OR rule to
-combine test1 and test3 to predict the outcome expectation.}
+unit tests (check 1, check 2, check 3). The model suggests using an OR
+rule to combine check 1 and check 3 to predict the outcome expectation.}
 
@@
 Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
-independence for the three individual tests and the combined test rule
-(test1 OR test3) from the logic regression. We also include the metric
-calculated from fitting a regression tree model to the data to compare
-the performance of the logic regression model. The harmonic and
+independence for the three individual checks and the combined check rule
+(check 1 OR check 3) from the logic regression. We also include the
+metric calculated from fitting a regression tree model to the data to
+compare the performance of the logic regression model. The harmonic and
 arithmetic means are included to combine the three measures. The results
-show that the two tests produced by the logic regression can accurately
+show that the two checks produced by the logic regression can accurately
 predict 83.6\% cases of all \emph{actual unexpected results} in the
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2424}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1515}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1061}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1970}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1364}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2468}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1299}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0909}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1688}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1169}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1429}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1039}}@{}}
 
 \caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
-parsimony (independence) metrics for each individual unit test and for
-the combined test rule (test1 OR test3) derived from the logic
+parsimony (independence) metrics for each individual check and for the
+combined check rule (check 1 OR check 3) derived from the logic
 regression model. The harmonic and arithmetic means of the three metrics
-are included to evaluate the quality of the unit tests in diagnosing
+are included to evaluate the quality of the checks in diagnosing
 unexpected step counts (more than five days with fewer than 8,000
@@
 arithmetic
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+checks
 \end{minipage} \\
@@
 \endlastfoot
-test1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 \\
-test2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 \\
-test3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 \\
-test1 OR test3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 \\
-regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 \\
+1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 & check 1 \\
+2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 & check 2 \\
+3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 & check 3 \\
+check 1 OR check 3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 & NA \\
+regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 & NA \\
 
@@
 For comparison, the regression tree produces a similar prediction to the
-logic regression, by first splitting on test1 and then splitting on
-test3, and results in the same accuracy and overall score as the logic
+logic regression, by first splitting on check 1 and then splitting on
+check 3, and results in the same accuracy and overall score as the logic
 regression model. However, we argue that the logic regression tree shown
@@
 an analyst might expect the coefficient for PM10 in this GLM to lie
-between {[}0, 0.005{]}, after adjusting for daily temperature
+between \([0, 0.005]\), after adjusting for daily temperature
 \citep{samet200fine, welty2005acute}. Observing an estimated coefficient
@@
 whether the estimated coefficient for PM10 in the GLM lies outside of
-the interval {[}0, 0.005{]}. In addition to providing a more substantial
+the interval \([0, 0.005]\). In addition to providing a more substantial
 problem for our methods, this example also demonstrates how the
@@
 
-A logic regression is fitted using all variations of the tests in
+A logic regression is fitted using all variations of the checks in
 Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
@@
 \caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
-fourteen unit tests and the outcome expectation (unexpected) as the
-response variable. The model suggests the relationship: (sample size
+fourteen checks and the outcome expectation (unexpected) as the response
+variable. The model suggests the relationship: (sample size
 \emph{greater than} 200 AND mortality-PM10 correlation \textless{}
@@
 \begin{longtable}[]{@{}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0882}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1471}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1029}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1765}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1912}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1324}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1618}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.5368}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1053}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0947}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1158}}@{}}
 
@@
 parsimony (independence) metrics derived from the logic regression
-model, along with harmonic and arithmetic means, for individual unit
-tests (1: sample size, 2: mortality-PM10 correlation, 3:
-mortality-temperature correlation, 4: mortality outlier), and the
-combined test rule 5: (sample size AND mortality-PM10 correlation) AND
-(mortality-temperature correlation OR mortality outlier.}
+model, along with harmonic and arithmetic means, for individual checks
+(1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature
+correlation, 4: mortality outlier), and the combined check rule 5:
+(sample size AND mortality-PM10 correlation) AND (mortality-temperature
+correlation OR mortality outlier.}
 
@@
 \toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedleft
-tests
+\begin{minipage}[b]{\linewidth}\raggedright
+check
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-overlapping
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-independence
+indep.
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \endlastfoot
-1 & 0.087 & 0.215 & 0 & 1 & 0.175 & 0.434 \\
-2 & 0.988 & 0.610 & 0 & 1 & 0.822 & 0.866 \\
-3 & 0.392 & 0.581 & 0 & 1 & 0.569 & 0.658 \\
-4 & 0.649 & 0.641 & 0 & 1 & 0.732 & 0.763 \\
-5 & 0.760 & 0.880 & 0 & 1 & 0.869 & 0.880 \\
+check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
+0.434 \\
+check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
+0.866 \\
+check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
+0.658 \\
+check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
+(not check 1 AND check 2) AND (check 3 OR check 4) & 0.760 & 0.880 & 1 &
+0.869 & 0.880 \\
 
@@
 at least three checks in the tree must be TRUE in order for the model to
-predict an unexpected outcome. Given the high independence of the tests
+predict an unexpected outcome. Given the high independence of the checks
 (Table~\ref{tbl-linear-reg}), this suggests that unexpected results are

### 2024-12-10 index.qmd
test -> check

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 library(tidyverse)
@@
 
--   test1: the test fails if the mean step count is below 8,200
--   test2: the test fails if the 30th percentile of the observed step counts is below 8,200
--   test3: the test fails if the standard deviation of the observed step counts exceeds 2,500.
+-   check 1: the check fails if the mean step count is below 8,200
+-   check 2: the check fails if the 30th percentile of the observed step counts is below 8,200
+-   check 3: the check fails if the standard deviation of the observed step counts exceeds 2,500.
 
-The cutoff values chosen for these tests would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.
+The cutoff values chosen for these checks would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.
 
@@
 
-While tests may score high on accuracy, they may be less effective at explaining the various reasons behind unexpected results.
-This could happen if, for example, a set of tests are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
-It may also occur if the tests are highly correlated with one another, leading to redundancy.
+While checks may score high on accuracy, they may be less effective at explaining the various reasons behind unexpected results.
+This could happen if, for example, a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
+It may also occur if the checks are highly correlated with one another, leading to redundancy.
 
@@
 
-A high mutual information value indicates redundancy among the tests, while a low value suggests that the tests are independent and provides unique information to diagnose the unexpected outcome.
+A high mutual information value indicates redundancy among the checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
 To standardize this measure, the total correlation *per observation* is calculated, and an independence score, ranging between 0 and 1, is defined as 1 - mutual information.
@@
 The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics.
-@fig-logic-reg shows the best-fitting logic regression model, which is a combination of test1 ("mean step count is below 8,200") and test3 ("standard deviation exceeds 2,500") with an OR operator.
-In other words, if either test1 or test3 is true, then we would predict an unexpected outcome in the analysis (i.e.
+@fig-logic-reg shows the best-fitting logic regression model, which is a combination of check 1 ("mean step count is below 8,200") and check 3 ("standard deviation exceeds 2,500") with an OR operator.
+In other words, if either check 1 or check 3 is true, then we would predict an unexpected outcome in the analysis (i.e.
 the number of days with fewer than 8,000 steps is greater than 5).
@@
 #| fig.heigh: 2
-#| fig-cap: Logic regression model fitted to the three unit tests (test1, test2, test3). The model suggests using an OR rule to combine test1 and test3 to predict the outcome expectation.
+#| fig-cap: Logic regression model fitted to the three unit tests (check 1, check 2, check 3). The model suggests using an OR rule to combine check 1 and check 3 to predict the outcome expectation.
 fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
@@
 
-@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual tests and the combined test rule (test1 OR test3) from the logic regression.
+@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual checks and the combined check rule (check 1 OR check 3) from the logic regression.
 We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
 The harmonic and arithmetic means are included to combine the three measures.
-The results show that the two tests produced by the logic regression can accurately predict 83.6% cases of all *actual unexpected results* in the simulation data.
+The results show that the two checks produced by the logic regression can accurately predict 83.6% cases of all *actual unexpected results* in the simulation data.
 Furthermore, 82.1% of all *predicted unexpected results* were in fact observed to be unexpected.
@@
             calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "tests") |> 
-  mutate(tests = c("test1", "test2", "test3")) |>
+  mutate(checks = c("check 1", "check 2", "check 3")) |>
   bind_rows(augment(fit) |> 
@@
               calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
-              mutate(tests = "test1 OR test3") |> 
+              mutate(tests = "check 1 OR check 3") |> 
               dplyr::select(-overlapping)) |> 
@@
   kable(digits = 3, 
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual unit test and for the combined test rule (test1 OR test3) derived from the logic regression model.
-The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the unit tests in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and for the combined check rule (check 1 OR check 3) derived from the logic regression model.
+The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
 ```
 
-For comparison, the regression tree produces a similar prediction to the logic regression, by first splitting on test1 and then splitting on test3, and results in the same accuracy and overall score as the logic regression model.
+For comparison, the regression tree produces a similar prediction to the logic regression, by first splitting on check 1 and then splitting on check 3, and results in the same accuracy and overall score as the logic regression model.
 However, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
@@
 The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10.
-Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between [0, 0.005], after adjusting for daily temperature [@samet200fine; @welty2005acute].
+Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet200fine; @welty2005acute].
 Observing an estimated coefficient outside of this interval would be highly unusual and would warrant a serious re-examination of the analysis process.
-Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval [0, 0.005].
+Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval $[0, 0.005]$.
 In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
@@
 
-A logic regression is fitted using all variations of the tests in @tbl-checks to predict whether the PM10 coefficient is unexpected.
+A logic regression is fitted using all variations of the checks in @tbl-checks to predict whether the PM10 coefficient is unexpected.
 @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model.
@@
 ```{r}
-#| fig.width: 8
+#| fig.width: 9
 #| fig.height: 4
@@
                                 seed = 6, nleaves = 4)
-plot(fit) 
+plot_logicreg(fit) 
 ```
 
-Logic regression model fitted to the fourteen unit tests and the outcome expectation (unexpected) as the response variable.
+Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable.
 The model suggests the relationship: (sample size *greater than* 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
@@
 
-tbl <- tibble(tests = c(1,2,3,4, 5)) |> bind_cols(df |> dplyr::select(-id)) 
+tbl <- tibble(check = c("check 1: sample size <= 200", "check 2: cor(m, PM10) < -0.03",
+                        "check 3: cor(m, tmp) > -0.35", "check 4: mortality_outlier", 
+                        "(not check 1 AND check 2) AND (check 3 OR check 4)")) |> bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)
+
 tbl |> 
   kable(digits = 3, escape = FALSE,
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation,  4: mortality outlier), and the combined test rule 5: (sample size AND mortality-PM10 correlation) AND (mortality-temperature correlation OR mortality outlier.")
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual checks (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation,  4: mortality outlier), and the combined check rule 5: (sample size AND mortality-PM10 correlation) AND (mortality-temperature correlation OR mortality outlier.")
 
@@
 rather at least three checks in the tree must be TRUE in order for the model to predict an unexpected outcome.
-Given the high independence of the tests (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.
+Given the high independence of the checks (@tbl-linear-reg), this suggests that unexpected results are only likely after multiple anomalies are observed in the data.
 

### 2024-12-12 _manuscript/_tex/index.tex
revamp toy example

@@
 particulate matter air pollution on daily mortality.
-Section~\ref{sec-discussion} discusses a few key considerations and
-Section~\ref{sec-conclusion} concludes the paper.
+Section~\ref{sec-discussion} summarises teh paper and discusses a few
+key considerations.
 
@@
 ``high'' days recorded after an outdoor hike or intense workout. Given
-the requirements of the study, we expect that for a given subject there
-would be no more than five days with step counts falling below the
-8,000-step threshold (``invalid days''). If there were more than five
-invalid days, the data from that subject would not be usable for the
-study.
-
-In this scenario, the outcome is that the number of days with a step
-count below 8,000 and the expectation is that this outcome would take a
-value no more than 5. If the expectation is violated, we can consider
-the potential configurations of the data that could lead to such an
-anomaly. To diagnose potential reasons why this outcome expectation
-might fail, we can establish a few analysis validation checks in
-anticipation of seeing the data. For example, if the average step count
-is too low, this may suggest that the subject is incapable of taking the
-required number of steps, potentially leading to the unexpected outcome.
-Similarly, we can also check the quantile of the step count, if more
-than a third of the days fall below 8,000, this could indicate an excess
-of low-count days. Additionally, we may may expect the standard
+the requirements of the study, we expect that for a given subject the
+average step count would be between \([8500, 9500]\).
+
+To diagnose potential reasons why this outcome expectation might fail,
+we can establish a few analysis validation checks in anticipation of
+seeing the data. For example, we can check the quantile of the step
+count, if more than a third of the days fall below 8,000, or more than
+two-thirds exceed 10,000 steps, this could indicate an excess of
+low-count or high-count days. Similarly, we may may expect the standard
 deviation of the step count not to be overly large. These considerations
-yield the following three analysis validation checks:
+yield the following three analysis validation checks that \emph{fails}
+when:
 
@@
 \item
-  check 1: the check fails if the mean step count is below 8,200
-\item
-  check 2: the check fails if the 30th percentile of the observed step
-  counts is below 8,200
-\item
-  check 3: the check fails if the standard deviation of the observed
-  step counts exceeds 2,500.
+  check 1: the 60\% quantile of the observed step counts is greater than
+  10,000
+\item
+  check 2: the 40\% quantile of the observed step counts is less than
+  8,000, and
+\item
+  check 3: the standard deviation of the observed step counts exceeds
+  2,500.
 \end{itemize}
@@
 
-\caption{\label{fig-step-count}Number of days with fewer than 8,000
-steps across 300 simulated 30-day periods. The orange bars indicate
-instances where the count exceeds five days, representing an unexpected
-outcome in this scenario.}
+\caption{\label{fig-step-count}Average step count across 300 simulated
+30-day periods. The orange points indicate instances where the average
+step count fails outside the {[}8,500, 9,500{]} interval, representing
+an unexpected outcome in this scenario.}
 
@@
 To simulate this data, three normal distributions are used for the daily
-step counts: \(\mathcal{N}(4000, 200)\) for low days,
+step counts: \(\mathcal{N}(6000, 200)\) for low days,
 \(\mathcal{N}(12000, 200)\) for high days, and
 \(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
-days can be simulated from a Poisson distribution with \(\lambda = 4\).
-Figure~\ref{fig-step-count} displays the number of days with fewer than
-8,000 steps across 300 simulated 30-day periods.
+days can be simulated from a Poisson distribution with \(\lambda = 8\).
+Figure~\ref{fig-step-count} displays average step count across 300
+simulated 30-day periods.
 
@@
 regression model \citep{ruczinski_logic_2003} is fitted to the analysis
-validation checks using the indicator of an unexpected result as the
-outcome in the model. Originally developed for SNP microarray data,
-logic regression constructs Boolean combinations of binary variables, in
-a tree structure, to predict both binary and numerical outcomes.
+validation checks. Originally developed for SNP microarray data, logic
+regression constructs Boolean combinations of binary variables,
+\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict both binary and
+numerical outcomes. In our case,
+
+\begin{itemize}
+\tightlist
+\item
+  the outcome, \(Y\), is a binary variable indicating whether the result
+  is \emph{unexpected} (1) or \emph{expected} (0),
+\item
+  \(X_1, X_2, \cdots, X_n\) represent the binary analysis validation
+  checks, where each check is labeled as 1 if it fails and 0 if it
+  passes.
+\end{itemize}
+
+For classification problems, the logic regression model is reduced to
+the following form:
+
+\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\] The logical operations of
+the binary variables \(X_1, X_2, \cdots, X_n\), including AND
+(\(\land\)), OR (\(\lor\)), and NOT (\(\neg\)) and an example of such
+logic combination can be \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\).
+Simulated annealing is used to explore the search space by minimizing
+the mis-classification error. The following 6 moves are permitted during
+the search (as detailed in Figure 2 of \citet{ruczinski_logic_2003}): 1)
+replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4)
+pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
+
 Compared to other tree-based methods for binary-binary prediction, the
@@
 described previously to generate the prediction of the unexpected
-outcome, which is whether the number of days with fewer than 8,000 steps
-is greater than 5. The predictions from the logic regression model can
+outcome, whether the average number of step falls within the {[}8,500,
+9,500{]} interval. The predictions from the logic regression model can
 be compared with the simulated true outcome for calculating the
 precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
-best-fitting logic regression model, which is a combination of check 1
-(``mean step count is below 8,200'') and check 3 (``standard deviation
-exceeds 2,500'') with an OR operator. In other words, if either check 1
-or check 3 is true, then we would predict an unexpected outcome in the
-analysis (i.e.~the number of days with fewer than 8,000 steps is greater
-than 5).
+best-fitting logic regression model as
+
+(quantile(step, 0.6) \textgreater{} 10,000 OR quantile(step, 0.4)
+\textless{} 8,000) AND sd(step) \textless{} 2,500
+
+In other words, we would predict an unexpected outcome in the analysis
+if the standard deviation of the step count is not too large (2,500) and
+either the 60\% quantile of the step count exceeds 10,000 or the 40\%
+quantile of the step count falls below 8,000.
 
@@
 \caption{\label{fig-logic-reg}Logic regression model fitted to the three
-unit tests (check 1, check 2, check 3). The model suggests using an OR
-rule to combine check 1 and check 3 to predict the outcome expectation.}
+checks. The model suggests the rule (quantile(step, 0.6) \textgreater{}
+10,000 OR quantile(step, 0.4) \textless{} 8,000) AND sd(step)
+\textless{} 2,500}
 
@@
 Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
-independence for the three individual checks and the combined check rule
-(check 1 OR check 3) from the logic regression. We also include the
-metric calculated from fitting a regression tree model to the data to
-compare the performance of the logic regression model. The harmonic and
-arithmetic means are included to combine the three measures. The results
-show that the two checks produced by the logic regression can accurately
-predict 83.6\% cases of all \emph{actual unexpected results} in the
-simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
-results} were in fact observed to be unexpected.
+independence for the three individual checks and the check rule found by
+the logic regression. The harmonic and arithmetic means are included to
+combine the three measures. The results show that the two checks
+produced by the logic regression can accurately predict 86.1\% cases of
+all \emph{actual unexpected results} in the simulation data.
+Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
+fact observed to be unexpected.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.2468}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1299}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0909}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1688}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1169}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1429}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1039}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.5368}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1053}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0947}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1158}}@{}}
 
 \caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
-parsimony (independence) metrics for each individual check and for the
-combined check rule (check 1 OR check 3) derived from the logic
-regression model. The harmonic and arithmetic means of the three metrics
-are included to evaluate the quality of the checks in diagnosing
+parsimony (independence) metrics for each individual check and the logic
+regression check rule. The harmonic and arithmetic means of the three
+metrics are included to evaluate the quality of the checks in diagnosing
 unexpected step counts (more than five days with fewer than 8,000
@@
 \begin{minipage}[b]{\linewidth}\raggedright
-tests
+checks
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-independence
+indep.
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
@@
 arithmetic
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-checks
 \end{minipage} \\
@@
 \endlastfoot
-1 & 0.482 & 0.964 & 1.000 & 0.730 & 0.815 & check 1 \\
-2 & 0.214 & 1.000 & 1.000 & 0.450 & 0.738 & check 2 \\
-3 & 0.589 & 0.805 & 1.000 & 0.762 & 0.798 & check 3 \\
-check 1 OR check 3 & 0.821 & 0.836 & 0.999 & 0.879 & 0.886 & NA \\
-regression tree & 0.821 & 0.836 & 1.000 & 0.879 & 0.886 & NA \\
+check 1: q(step, 0.6) \textgreater{} 10000 & 0.319 & 0.575 & 1.000 &
+0.511 & 0.631 \\
+check 2: q(step, 0.4) \textless{} 8000 & 0.264 & 0.613 & 1.000 & 0.467 &
+0.626 \\
+check 3: sd(step) \textgreater{} 2500 & 0.153 & 0.289 & 1.000 & 0.273 &
+0.481 \\
+logic regression: (check 1 OR check 2) AND check 3 & 0.431 & 0.861 &
+0.998 & 0.669 & 0.763 \\
+regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774 \\
 
@@
 For comparison, the regression tree produces a similar prediction to the
-logic regression, by first splitting on check 1 and then splitting on
-check 3, and results in the same accuracy and overall score as the logic
-regression model. However, we argue that the logic regression tree shown
+logic regression, however, we argue that the logic regression tree shown
 in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.5368}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1053}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0947}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1158}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.6106}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0885}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0796}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0973}}@{}}
 
@@
 check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
-(not check 1 AND check 2) AND (check 3 OR check 4) & 0.760 & 0.880 & 1 &
-0.869 & 0.880 \\
+logic regression: (not check 1 AND check 2) AND (check 3 OR check 4) &
+0.760 & 0.880 & 1 & 0.869 & 0.880 \\
 
@@
 
-\section{Conclusion}\label{sec-conclusion}
-
-TODO
-
 \section{Acknowledgement}\label{acknowledgement}

### 2024-12-12 index.qmd
revamp toy example

@@
 library(infotheo)
+library(patchwork)
 pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
@@
 @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality.
-@sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.
+@sec-discussion summarises teh paper and discusses a few key considerations.
 
@@
 We may also expect "high" days recorded after an outdoor hike or intense workout.
-Given the requirements of the study, we expect that for a given subject there would be no more than five days with step counts falling below the 8,000-step threshold ("invalid days").
-If there were more than five invalid days, the data from that subject would not be usable for the study.
+Given the requirements of the study, we expect that for a given subject
+the average step count would be between $[8500, 9500]$.
 
-In this scenario, the outcome is that the number of days with a step count below 8,000 and the expectation is that this outcome would take a value no more than 5.
-If the expectation is violated, we can consider the potential configurations of the data that could lead to such an anomaly.
 To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
-For example, if the average step count is too low, this may suggest that the subject is incapable of taking the required number of steps, potentially leading to the unexpected outcome.
-Similarly, we can also check the quantile of the step count, if more than a third of the days fall below 8,000, this could indicate an excess of low-count days.
-Additionally, we may may expect the standard deviation of the step count not to be overly large.
-These considerations yield the following three analysis validation checks:
+For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than two-thirds exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
+Similarly, we may may expect the standard deviation of the step count not to be overly large.
+These considerations yield the following three analysis validation checks that *fails* when:
 
--   check 1: the check fails if the mean step count is below 8,200
--   check 2: the check fails if the 30th percentile of the observed step counts is below 8,200
--   check 3: the check fails if the standard deviation of the observed step counts exceeds 2,500.
+-   check 1: the 60% quantile of the observed step counts is greater than 10,000
+-   check 2: the 40% quantile of the observed step counts is less than 8,000, and
+-   check 3: the standard deviation of the observed step counts exceeds 2,500.
 
@@
 ```{r}
+#| fig-height: 3
 #| label: fig-step-count
-#| fig-cap: Number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
-The orange bars indicate instances where the count exceeds five days, representing an unexpected outcome in this scenario.
+#| fig-cap: Average step count across 300 simulated 30-day periods.
+The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario.
 set.seed(1234)
@@
   mutate(
-    small_n = rpois(1, 4),
-    large_n = rpois(1, 4),
+    small_n = rpois(1, 8),
+    large_n = rpois(1, 8),
     norm_sd = rlogis(1, 800, scale = 100),
     norm_n = 30 - small_n - large_n,
-    step = list(round(c(rnorm(small_n, 4000, 200), 
+    step = list(round(c(rnorm(small_n, 6000, 200), 
                         rnorm(large_n, 12000, 200), 
@@
     ), 0)),
-    less_than_8000 = sum(step < 8000), 
-    unexpect = ifelse(less_than_8000 > 5, 1, 0),
-    test1 = ifelse(mean(step) < 8200, 1, 0),
-    test2 = ifelse(quantile(step, 0.3, na.rm = TRUE) < 8300, 1, 0),
-    test3 = ifelse(sd(step) > 2500, 1, 0),
+    mean = mean(step, na.rm = TRUE),
+    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
+    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
+    `q(step, 0.4) < 8000` = ifelse(quantile(step, 0.4, na.rm = TRUE) < 8000, 1, 0),
+    `sd(step) > 2500` = ifelse(sd(step) > 2500, 1, 0),
   ) |>
@@
 
-# step_count |> 
-#   dplyr::select(id, step) |> 
-#   unnest(step) |>
-#   mutate(day = rep(1:30, 300)) |> 
-#   ggplot(aes(x = step, y = id)) + 
-#   geom_point(alpha = 0.5) + 
-#   theme_bw()
-
 step_count |> 
-  ggplot(aes(x = less_than_8000, fill = as.factor(unexpect))) + 
-  geom_histogram(binwidth = 0.5) + 
-  scale_x_continuous(breaks = seq(0, 13, 1)) +
-  scale_fill_brewer(palette = "Dark2") + 
-  xlab("Number of days with step count less than 8000") + 
+  ggplot(aes(x = mean, y = 1, color = as.factor(unexpect))) + 
+  ggbeeswarm::geom_quasirandom() + 
+  scale_y_continuous(breaks = seq(0, 13, 1)) +
+  scale_color_brewer(palette = "Dark2") + 
+  xlab("Average number of steps") + 
+  ylab("") + 
   theme_bw() + 
-  theme(legend.position = "none")
+  theme(legend.position = "none", panel.grid.minor = element_blank(), axis.text.y = element_blank())
 ```
 
-To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(4000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
-The number of low and high days can be simulated from a Poisson distribution with $\lambda = 4$.
-@fig-step-count displays the number of days with fewer than 8,000 steps across 300 simulated 30-day periods.
+To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
+The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$.
+@fig-step-count displays average step count across 300 simulated 30-day periods.
 
@@
 
-To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks using the indicator of an unexpected result as the outcome in the model.
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, in a tree structure, to predict both binary and numerical outcomes.
+To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks.
+Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict both binary and numerical outcomes.
+In our case,
+
+* the outcome, $Y$, is a binary variable indicating whether the result is *unexpected* (1) or *expected* (0),
+* $X_1, X_2, \cdots, X_n$ represent the binary analysis validation checks, where each check is labeled as 1 if it fails and 0 if it passes.
+
+For classification problems, the logic regression model is reduced to the following form:
+
+$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$
+The logical operations of the binary variables $X_1, X_2, \cdots, X_n$, including AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of such logic combination can be $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
+Simulated annealing is used to explore the search space by minimizing the mis-classification error.
+The following 6 moves are permitted during the search (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
+
 Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods.
@@
 
-Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, which is whether the number of days with fewer than 8,000 steps is greater than 5.
+Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, whether the average number of step falls within the [8,500, 9,500] interval.
 The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics.
-@fig-logic-reg shows the best-fitting logic regression model, which is a combination of check 1 ("mean step count is below 8,200") and check 3 ("standard deviation exceeds 2,500") with an OR operator.
-In other words, if either check 1 or check 3 is true, then we would predict an unexpected outcome in the analysis (i.e.
-the number of days with fewer than 8,000 steps is greater than 5).
+@fig-logic-reg shows the best-fitting logic regression model as
+
+(quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500
+
+In other words, we would predict an unexpected outcome in the analysis if the standard deviation of the step count is not too large (2,500) and either the 60% quantile of the step count exceeds 10,000 or the 40% quantile of the step count falls below 8,000.
 
@@
 #| label: fig-logic-reg
-#| fig.width: 6
+#| fig.width: 8
 #| fig.heigh: 2
-#| fig-cap: Logic regression model fitted to the three unit tests (check 1, check 2, check 3). The model suggests using an OR rule to combine check 1 and check 3 to predict the outcome expectation.
-fit <- step_count |> fit_logic_reg(unexpect, test1:test3, seed = 1, nleaves = 4)
+#| fig-cap: Logic regression model fitted to the three checks. The model suggests the rule (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500
+fit <- step_count |> fit_logic_reg(unexpect, `q(step, 0.6) > 10000`:`sd(step) > 2500`, seed = 1, nleaves = 3)
 plot_logicreg(fit)
@@
 ```{r}
-fit_regtree <- rpart(unexpect ~ ., data = step_count |> dplyr::select(unexpect:test3), method = "class", 
+fit_regtree <- rpart(unexpect ~ ., data = step_count |> dplyr::select(unexpect:`sd(step) > 2500`), method = "class", 
                      control = rpart.control(cp = 0.01)) 
@@
   calc_miscla_rate(unexpect, .fitted) |> 
-  mutate(overlapping = multiinformation(step_count[,c(9, 11)])/nrow(step_count), 
+  mutate(overlapping = multiinformation(step_count[,c(8, 11)])/nrow(step_count), 
          independence = 1 - overlapping) |> 
@@
 
-@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual checks and the combined check rule (check 1 OR check 3) from the logic regression.
-We also include the metric calculated from fitting a regression tree model to the data to compare the performance of the logic regression model.
+@tbl-logic-reg presents the calculated precision, recall, and independence for the three individual checks and the check rule found by the logic regression.
 The harmonic and arithmetic means are included to combine the three measures.
-The results show that the two checks produced by the logic regression can accurately predict 83.6% cases of all *actual unexpected results* in the simulation data.
-Furthermore, 82.1% of all *predicted unexpected results* were in fact observed to be unexpected.
+The results show that the two checks produced by the logic regression can accurately predict 86.1% cases of all *actual unexpected results* in the simulation data.
+Furthermore, 43.1% of all *predicted unexpected results* were in fact observed to be unexpected.
 
@@
 #| label: tbl-logic-reg
-list(tibble(.fitted = step_count$test1, unexpect = step_count$unexpect),
-     tibble(.fitted = step_count$test2, unexpect = step_count$unexpect),
-     tibble(.fitted = step_count$test3, unexpect = step_count$unexpect)
+list(tibble(.fitted = step_count$`q(step, 0.6) > 10000`, unexpect = step_count$unexpect),
+     tibble(.fitted = step_count$`q(step, 0.4) < 8000`, unexpect = step_count$unexpect),
+     tibble(.fitted = step_count$`sd(step) > 2500`, unexpect = step_count$unexpect)
 ) |>
@@
             mutate(independence = 1) |> 
-            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "tests") |> 
-  mutate(checks = c("check 1", "check 2", "check 3")) |>
+            calc_metrics(metrics = c("harmonic", "arithmetic"))) |> 
+  mutate(checks = c("check 1: q(step, 0.6) > 10000", "check 2: q(step, 0.4) < 8000", "check 3: sd(step) > 2500")) |>
   bind_rows(augment(fit) |> 
@@
               calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
-              mutate(tests = "check 1 OR check 3") |> 
+              mutate(checks = "logic regression: (check 1 OR check 2) AND check 3") |> 
               dplyr::select(-overlapping)) |> 
@@
               dplyr::select(-overlapping) |>
-              mutate(tests = "regression tree")) |>
+              mutate(checks = "regression tree")) |>
+  dplyr::select(checks, precision:arithmetic) |> 
+  dplyr::rename(`indep.` = independence) |> 
   kable(digits = 3, 
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and for the combined check rule (check 1 OR check 3) derived from the logic regression model.
+        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and the logic regression check rule.
 The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
@@
 
-For comparison, the regression tree produces a similar prediction to the logic regression, by first splitting on check 1 and then splitting on check 3, and results in the same accuracy and overall score as the logic regression model.
-However, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
+For comparison, the regression tree produces a similar prediction to the logic regression, however, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
 The logic regression tree is also directly comparable to other diagnostic techniques, which we discuss further in @sec-discussion.
@@
                         "check 3: cor(m, tmp) > -0.35", "check 4: mortality_outlier", 
-                        "(not check 1 AND check 2) AND (check 3 OR check 4)")) |> bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)
+                        "logic regression: (not check 1 AND check 2) AND (check 3 OR check 4)")) |> 
+  bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)
 
@@
 
-# Conclusion {#sec-conclusion}
-
-TODO
-
 # Acknowledgement

### 2024-12-12 index.qmd
fix a reference

@@
 The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10.
-Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet200fine; @welty2005acute].
+Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet2000fine; @welty2005acute].
 Observing an estimated coefficient outside of this interval would be highly unusual and would warrant a serious re-examination of the analysis process.

### 2024-12-12 _manuscript/_tex/index.tex
fix a reference
small changes here and there " q

@@
 
-\caption{\label{fig-step-count}Average step count across 300 simulated
-30-day periods. The orange points indicate instances where the average
-step count fails outside the {[}8,500, 9,500{]} interval, representing
-an unexpected outcome in this scenario.}
+\caption{\label{fig-step-count}Beeswarm plot of average step counts
+across 300 simulated 30-day periods. Each point represents the average
+step count from one simulation. While similar to a boxplot or violin
+plot, the beeswarm plot also displays the distribution of each
+individual data point. The orange points indicate instances where the
+average step count fails outside the {[}8,500, 9,500{]} interval,
+representing an unexpected outcome in this scenario.}
 
@@
 checks. The model suggests the rule (quantile(step, 0.6) \textgreater{}
-10,000 OR quantile(step, 0.4) \textless{} 8,000) AND sd(step)
-\textless{} 2,500}
+10,000 OR quantile(step, 0.4) \textless{} 8,000) AND (NOT sd(step)
+\textgreater{} 2,500). The NOT operator applied to
+\texttt{sd(step)\ \textgreater{}\ 2,500)} is colored with a black
+background to distinguish it from other checks.}
 
@@
 between \([0, 0.005]\), after adjusting for daily temperature
-\citep{samet200fine, welty2005acute}. Observing an estimated coefficient
-outside of this interval would be highly unusual and would warrant a
-serious re-examination of the analysis process. Therefore, our
+\citep{samet2000fine, welty2005acute}. Observing an estimated
+coefficient outside of this interval would be highly unusual and would
+warrant a serious re-examination of the analysis process. Therefore, our
 unexpected outcome in this analysis will be the binary indicator of
@@
 fourteen checks and the outcome expectation (unexpected) as the response
-variable. The model suggests the relationship: (sample size
-\emph{greater than} 200 AND mortality-PM10 correlation \textless{}
--0.04) AND (mortality-temperature correlation \textgreater{} -0.35 OR
-there exist mortality outlier) to predict the unexpected PM10
-coefficient. {[}sorry the proper plot is not ready{]}.}
+variable. The model suggests the relationship: (NOT sample size less
+than or equal to 200 AND mortality-PM10 correlation \textless{} -0.04)
+AND (mortality-temperature correlation \textgreater{} -0.35 OR there
+exist mortality outlier) to predict the unexpected PM10 coefficient. The
+NOT operator applied to \texttt{smpl\_size\ \textless{}=\ 200} is
+colored with a black background to distinguish it from other checks.}
 
@@
 made during the data analysis process. These checks can serve as a
-useful summary of the analyst's thought process and can desscribe how
-characterstics of the data may lead to unexpected outcomes. Using logic
+useful summary of the analyst's thought process and can describe how
+characteristics of the data may lead to unexpected outcomes. Using logic
 regression, we can develop a graphical summary of the analysis
@@
 histogram, density plot, or bee swarm plot can reveal outliers or
-deviations from normality. These visualizations could be reframed as
+deviations from normality. These visualizations could be re-framed as
 checks that fail when the data does not conform to the expected
@@
 either manual verification or the development of automated methods to
-interpret visualization outputs.
+interpret visualization outputs. An existing example of visual test is
+the R package \texttt{vdiffr} \citep{vdiffr} for software unit testing.
+The package saves a template plot and compares it to the current plot to
+determine whether the unit tests passes or fails.
 

### 2024-12-12 index.qmd
small changes here and there " q

@@
 #| label: fig-step-count
-#| fig-cap: Average step count across 300 simulated 30-day periods.
+#| fig-cap: Beeswarm plot of average step counts across 300 simulated 30-day periods.
+Each point represents the average step count from one simulation.
+While similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point.
 The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario.
@@
 #| fig.heigh: 2
-#| fig-cap: Logic regression model fitted to the three checks. The model suggests the rule (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500
+#| fig-cap: Logic regression model fitted to the three checks.
+The model suggests the rule (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND (NOT sd(step) > 2,500).
+The NOT operator applied to `sd(step) > 2,500)` is colored with a black background to distinguish it from other checks.
 fit <- step_count |> fit_logic_reg(unexpect, `q(step, 0.6) > 10000`:`sd(step) > 2500`, seed = 1, nleaves = 3)
@@
 Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable.
-The model suggests the relationship: (sample size *greater than* 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
-[sorry the proper plot is not ready].
+The model suggests the relationship: (NOT sample size less than or equal to 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
+The NOT operator applied to `smpl_size <= 200` is colored with a black background to distinguish it from other checks.
 :::
@@
 In this paper we have developed an approach to using analysis validation checks to externalize the assumptions about the data and analysis tools made during the data analysis process.
-These checks can serve as a useful summary of the analyst's thought process and can desscribe how characterstics of the data may lead to unexpected outcomes.
+These checks can serve as a useful summary of the analyst's thought process and can describe how characteristics of the data may lead to unexpected outcomes.
 Using logic regression, we can develop a graphical summary of the analysis validation checks as well as use the logic regression fitting process to choose the optimal set of checks.
@@
 For instance, plotting a variable's distribution using a histogram, density plot, or bee swarm plot can reveal outliers or deviations from normality.
-These visualizations could be reframed as checks that fail when the data does not conform to the expected distribution.
+These visualizations could be re-framed as checks that fail when the data does not conform to the expected distribution.
 However, translating visualization results into binary checks that return (TRUE, FALSE) remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs.
+An existing example of visual test is the R package `vdiffr` [@vdiffr] for software unit testing.
+The package saves a template plot and compares it to the current plot to determine whether the unit tests passes or fails.
 

### 2024-12-12 _manuscript/_tex/index.tex
add practical use of avc

@@
 
+With the analysis validation checks proposed in this paper, analysts may
+perform these checks \emph{ex ante} to identify and address potential
+issues before proceeding with the analysis or use them \emph{ex post} to
+diagnose the analysis when an unexpected outcome occurs. Theses checks
+act as checkpoints designed by analysts to flag potential issues,
+clarify assumptions made during the analysis, and enhance transparency
+and communication among stakeholders. They also support students and
+junior researchers develop the skills needed to independently diagnose
+issues in analyses.
+
 An interesting connection can be drawn between our logic regression

### 2024-12-12 index.qmd
add practical use of avc

@@
 
-<!-- * provide interpretation at different scenarios: -->
-
-<!--   1) one test is flagged, the prediction is as expected:  --> <!--   2) multiple tests are flagged, the prediction is unexpected,  --> <!--   3) no test is flagged, the prediction is unexpected,  --> <!--   4) no test is flagged, the prediction is as expected -->
-
 # Application {#sec-pm10-mortality}
@@
 
+With the analysis validation checks proposed in this paper, analysts may perform these checks *ex ante* to identify and address potential issues before proceeding with the analysis or use them *ex post* to diagnose the analysis when an unexpected outcome occurs.
+Theses checks act as checkpoints designed by analysts to flag potential issues, clarify assumptions made during the analysis, and enhance transparency and communication among stakeholders.
+They also support students and junior researchers develop the skills needed to independently diagnose issues in analyses.
+
 An interesting connection can be drawn between our logic regression trees and a tool used in systems engineering known as a fault tree.

### 2024-12-13 _manuscript/_tex/index.tex
illustration and cite veridical ds

@@
 The externalization of the data analysis process is a practice that has
-potential to improve the trustworthiness of analyses in general and the
-trustworthiness of subsequent products, such as machine learning models,
-that may be built on such analyses. While publication of analysis code
-and data is now a common a requirement for the sake of reproducibility
+potential to improve the trustworthiness of analyses in general
+\citep{yu2024veridical62} and the trustworthiness of subsequent
+products, such as machine learning models, that may be built on such
+analyses. While publication of analysis code and data is now a common a
+requirement for the sake of reproducibility
 \citep{peng2011reproducible}, the publication code alone is often
@@
 
-\includegraphics[width=6.73in,height=\textheight]{figures/metric-calc.png}
+\includegraphics[width=1\textwidth,height=\textheight]{figures/metric-calc.png}
 
@@
 
-\caption{\label{fig-metric-calc}{[}this is the cap and the plot needs
-polish{]}}
+\caption{\label{fig-metric-calc}Overview of the procedure to quantify
+the effectiveness of analysis validation checks. The procedure involves
+simulating replicates of the data, applying the anlysis, and running the
+analysis validation checks on each replicate to determine whether the
+outcome meets the expectation (the unexpected column) and whether each
+check passes (check 1, check 2, \ldots). The simulation results are then
+passed to two branches: the accuracy branch calculates the precision and
+recall of the checks from a logic regression prediction, while the
+redundancy branch to calculate the mutual information and independence
+of the checks. These three metrics are combined to produce a single
+metric that quantifies the effectiveness of the checks in diagnosing
+unexpected outcomes in the analysis.}
 

### 2024-12-13 index.qmd
illustration and cite veridical ds

@@
 
-The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
+The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical62] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
 While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis.
@@
 #| label: fig-metric-calc
-#| fig-cap: "[this is the cap and the plot needs polish]"
+#| out.width: 100%
+#| fig-cap: "Overview of the procedure to quantify the effectiveness of analysis validation checks.
+The procedure involves simulating replicates of the data, applying the anlysis, and running the analysis validation checks on each replicate to determine whether the outcome meets the expectation (the unexpected column) and whether each check passes (check 1, check 2, ...).
+The simulation results are then passed to two branches: the accuracy branch calculates the precision and recall of the checks from a logic regression prediction, while the redundancy branch to calculate the mutual information and independence of the checks.
+These three metrics are combined to produce a single metric that quantifies the effectiveness of the checks in diagnosing unexpected outcomes in the analysis."
 knitr::include_graphics(here::here("figures", "metric-calc.png"))

### 2024-12-13 _manuscript/_tex/index.tex
more references plus repo name

@@
 potential to improve the trustworthiness of analyses in general
-\citep{yu2024veridical62} and the trustworthiness of subsequent
-products, such as machine learning models, that may be built on such
-analyses. While publication of analysis code and data is now a common a
+\citep{yu2024veridical} and the trustworthiness of subsequent products,
+such as machine learning models, that may be built on such analyses.
+While publication of analysis code and data is now a common a
 requirement for the sake of reproducibility
@@
 The concept of framing data analysis as a sense-making process was
-originally presented by \citep{grolemund_cognitive_2014} based on
-seminal work by \citep{wild1999statistical}. Key to any sense-making
+originally presented by \citet{grolemund_cognitive_2014} based on
+seminal work by \citet{wild1999statistical}. Key to any sense-making
 process is a model for the world (i.e.~expectations for what we might
@@
 cause of an unexpected analysis result are typically not presented
-\citep{peng2022perspective}. \citep{peng_diagnosing_2021} proposed a
+\citep{peng2022perspective}. \citet{peng_diagnosing_2021} proposed a
 series of exercises for training students in data analysis to diagnose
@@
 to the concept of debugging and testing in software engineering
-\citep{donoghue2021teaching}. For example, \citep{li2019towards} found
+\citep{donoghue2021teaching}. For example, \citet{li2019towards} found
 that experienced engineers were generally able to identify problems in
@@
 A substantial body of literature has addressed the definition of data
-quality \citep[more]{8642813} and has developed frameworks that include
-dimensions, attributes, and measures to evaluate and improve data
-quality
-\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
+quality and has developed frameworks that include dimensions,
+attributes, and measures to evaluate and improve data quality
+\citep{8642813, batini2009methodologies, cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
 These frameworks are often used in information systems and database
 management and support business decision-making in various industries.
-For research purposes, high-quality data ensures the credibility of
-scientific findings and supports reproducibility and re-usability in
-future studies {[}ref{]}. With the growing prevalence of open data in
-scientific research, the consumers or users of the data typically are no
-longer the producers or collectors of the data who would have the full
-knowledge of data in hand, prompting more interest towards data quality
-checks in the data analysis process. In R, there are some packages, like
-\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
-that provide basic data screening and reporting tools, while another
-class of packages, e.g.~\texttt{assertr} \citep{assertr},
-\texttt{validate} \citep{validate}, and \texttt{pointblank}
-\citep{pointblank} focuses on providing data validation tools, allowing
-users to define customized data quality checks based on the
-applications.
+For example, such data validation systems are applied in the Google
+Tensorflow framework \citep{polyzotis2019data} and Amazon
+\citep{schelter2018automating}. For research purposes, high-quality data
+ensures the credibility of scientific findings and supports
+reproducibility and re-usability in future studies. With the growing
+prevalence of open data in scientific research, the consumers or users
+of the data typically are no longer the producers or collectors of the
+data who would have the full knowledge of data in hand, prompting more
+interest towards data quality checks in the data analysis process. In R,
+there are some packages, like \texttt{skimr} \citep{skimr} and
+\texttt{dataMaid} \citep{dataMaid}, that provide basic data screening
+and reporting tools, while another class of packages,
+e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
+\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
+providing data validation tools, allowing users to define customized
+data quality checks based on the applications.
 
@@
 return a TRUE or FALSE result given the data analyzed at hand. Inspired
-by the concept of data validation checks (\citet{validate}), which are
+by the concept of data validation checks \citep{validate}, which are
 designed to ensure that datasets meet expected formats and quality,

### 2024-12-13 index.qmd
more references plus repo name

@@
 
-The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical62] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
+The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
 While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis.
@@
 
-The concept of framing data analysis as a sense-making process was originally presented by [@grolemund_cognitive_2014] based on seminal work by [@wild1999statistical].
+The concept of framing data analysis as a sense-making process was originally presented by @grolemund_cognitive_2014 based on seminal work by @wild1999statistical.
 Key to any sense-making process is a model for the world (i.e.
@@
 While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of an unexpected analysis result are typically not presented [@peng2022perspective].
-[@peng_diagnosing_2021] proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers.
+@peng_diagnosing_2021 proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers.
 They provide a systematic approach involving working backwards from the analysis result to identify potential causes.
 There are parallels here to the concept of debugging and testing in software engineering [@donoghue2021teaching].
-For example, [@li2019towards] found that experienced engineers were generally able to identify problems in code faster than novices, and that the ability to debug code required knowledge that cut across different domains.
+For example, @li2019towards found that experienced engineers were generally able to identify problems in code faster than novices, and that the ability to debug code required knowledge that cut across different domains.
 
@@
 
-A substantial body of literature has addressed the definition of data quality [@8642813, more] and has developed frameworks that include dimensions, attributes, and measures to evaluate and improve data quality [@cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
+A substantial body of literature has addressed the definition of data quality and has developed frameworks that include dimensions, attributes, and measures to evaluate and improve data quality [@8642813; @batini2009methodologies; @cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
 These frameworks are often used in  information systems and database management and support business decision-making in various industries.
-For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and re-usability in future studies [ref].
+For example, such data validation systems are applied in the Google Tensorflow framework [@polyzotis2019data] and Amazon [@schelter2018automating].
+For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and re-usability in future studies.
 With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the producers or collectors of the data who would have the full knowledge of data in hand, prompting more interest towards data quality checks in the data analysis process.
@@
 Here, we introduce the concept of **analysis validation checks**, which frame these expectations or assumptions as explicit checks that return a TRUE or FALSE result given the data analyzed at hand.
-Inspired by the concept of data validation checks (@validate), which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst.
+Inspired by the concept of data validation checks [@validate], which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst.
 The focus on expected results allows the concept of analysis validation checks to encompass a broad range of checks, such as data quality (i.e.

### 2024-12-17 index.qmd
Minor changes

@@
 In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations.
-When a result falls outside of this interval, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process.
-This mental process of where to diagnose unexpected outcomes is often difficult to trace and discuss in publications.
+When a result falls outside of this expectation, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process.
+This mental process of where to diagnose unexpected outcomes is often difficult to trace and document in publications.
 As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014].
-These missing pieces of information provides little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
+These missing pieces of information provide little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
 Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.
@@
 The skill of diagnosing unexpected data analysis results is not one that has received significant attention in the statistics literature.
-While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of an unexpected analysis result are typically not presented [@peng2022perspective].
+While the concept of diagnosis is often embedded in model checking or data visualization techniques, systematic approaches to identifying the root cause of a general unexpected analysis result are typically not presented [@peng2022perspective].
+Furthermore, if interesting information is discovered through model checking or data visualization, there is no formal way to document such discoveries for the next analyst to consider.
 @peng_diagnosing_2021 proposed a series of exercises for training students in data analysis to diagnose different kinds of analysis problems such as coding errors or outliers.
@@
 When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions.
-Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, making it harder for less experienced researchers to learn and master these skills.
+Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, limiting transparency and making it harder for less experienced researchers to learn and master these skills.
 
@@
 
-Consider a 30-day step count experiment in public health.
+Consider a 30-day step count experiment in a public health setting.
 Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app.
@@
 To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
-For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than two-thirds exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
+For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
 Similarly, we may may expect the standard deviation of the step count not to be overly large.
-These considerations yield the following three analysis validation checks that *fails* when:
+These considerations yield the following three analysis validation checks that *fail* when:
 
@@
 
-From the simulated data, the accuracy branch refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
-While a false positive can raise caution or skepticism on the data, the presence of both false positives and false negatives suggest that the checks are overly sensitive or lack sensitivity to unexpected outcomes.
+From the simulated data, the accuracy branch in @fig-metric-calc refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
+While a false positive can raise caution or skepticism about the data, the presence of both false positives and false negatives suggests that the checks are overly sensitive or lack sensitivity to unexpected outcomes, respectively.
 
@@
 
-While checks may score high on accuracy, they may be less effective at explaining the various reasons behind unexpected results.
+While checks may score high on predictive accuracy, they may be less effective at explaining the various reasons behind unexpected results.
 This could happen if, for example, a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
@@
 For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
+The dataset is from New York City, and contains daily PM10, mortality, and temperature values from 1992--2000.
 
@@
 | Sample size less than or equal to 800 |
-| Mortality-PM10 correlation less than -0.03 |
-| Mortality-PM10 correlation less than -0.04 |
-| Mortality-PM10 correlation less than -0.05 |
-| Mortality-PM10 correlation less than -0.06 |
-| Mortality-temperature correlation greater than -0.3 | 
-| Mortality-temperature correlation greater than -0.35 | 
-| Mortality-temperature correlation greater than -0.4 | 
-| Mortality-temperature correlation greater than -0.45 | 
+| Mortality-PM10 correlation less than $-$0.03 |
+| Mortality-PM10 correlation less than $-$0.04 |
+| Mortality-PM10 correlation less than $-$0.05 |
+| Mortality-PM10 correlation less than $-$0.06 |
+| Mortality-temperature correlation greater than $-$0.3 | 
+| Mortality-temperature correlation greater than $-$0.35 | 
+| Mortality-temperature correlation greater than $-$0.4 | 
+| Mortality-temperature correlation greater than $-$0.45 | 
 | Outlier(s) are presented in the variable PM10 | 
@@
 
-To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed dataset to refine our simulations.
+To ensure a reasonable likeness to data that might be used in such an analysis, we use characteristics of the observed New York City dataset to refine our simulations.
 AIC is used to determine the best distribution fit for each variable with the QQ-plot presented in @fig-dist-fit to evaluate the fit.
@@
 We demonstrated our method on an example relating daily mortality to outdoor air pollution data.
+The results from that example could be used to inform future analyses of air pollution and health data, perhaps in other cities or locations.
 
@@
 
-Systematically generating simulated data is a key component of our approach.
+Systematically generating realistic simulated data is a key component of our approach and is deserving of further consideration.
 In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature.
-However, simulation can become complex when additional restrictions are imposed.
+However, simulation can become complex when additional restrictions are imposed or when a greater range of scenarios is desired.
 In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.
@@
 As a result, an analysis may fail several assumption checks yet produce an expected outcome, or pass all checks but yield an unexpected result.
+
+Communicating the process of data analysis is a key element to providing transparency, improving reproducibility, and building trust with a range of audiences.
+The analysis validation checks described here provide a general way to encode the assumptions that an data analyst makes about the data and statistical tools applied to the data.
+These assumptions can be studied or challenged, depending on the specific analyst's perspective, and can serve as a roadmap for diagnosing unexpected results.
 

### 2024-12-19 make_nmmaps_data2.R
Add city/name map

@@ -36,3 +36,10 @@ for(city in cities) {
 
   write_csv(dat, file = sprintf("data/pm10-%s.csv", city))
 }
+
+
+meta |>
+  select(city, cityname) |>
+  distinct() |>
+  filter(city %in% cities) |>
+  write_csv("city_name_map.csv")

### 2024-12-19 _manuscript/_tex/index.tex
new table results

@@
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
 % Set default figure placement to htbp
-\makeatletter
 \def\fps@figure{htbp}
@@
 
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
 \usepackage[]{natbib}
@@
 knowledge or domain expertise to quickly assess whether results align
-with their expectations. When a result falls outside of this interval,
-it prompts the researchers to investigate backwards on the data quality,
-the analysis steps, or the assumptions made during the analysis process.
-This mental process of where to diagnose unexpected outcomes is often
-difficult to trace and discuss in publications. As a result, readers are
-typically presented with the final outcomes of the analysis cycle where
-the results and expectations are aligned, achieved either by refining
-the analysis or updating the expectations based on statistical evidence
-\citep{grolemund_cognitive_2014}. These missing pieces of information
-provides little guidance for diagnosing issues in the analysis when the
-same methodology is applied to a new dataset that produces different
-outcomes. Similarly, when researchers with different background
-knowledge view results that they find to be unexpected, it becomes
-unclear whether discrepancies arise from differing expectations or from
-the use of statistical techniques.
+with their expectations. When a result falls outside of this
+expectation, it prompts the researchers to investigate backwards on the
+data quality, the analysis steps, or the assumptions made during the
+analysis process. This mental process of where to diagnose unexpected
+outcomes is often difficult to trace and document in publications. As a
+result, readers are typically presented with the final outcomes of the
+analysis cycle where the results and expectations are aligned, achieved
+either by refining the analysis or updating the expectations based on
+statistical evidence \citep{grolemund_cognitive_2014}. These missing
+pieces of information provide little guidance for diagnosing issues in
+the analysis when the same methodology is applied to a new dataset that
+produces different outcomes. Similarly, when researchers with different
+background knowledge view results that they find to be unexpected, it
+becomes unclear whether discrepancies arise from differing expectations
+or from the use of statistical techniques.
 
@@
 visualization techniques, systematic approaches to identifying the root
-cause of an unexpected analysis result are typically not presented
-\citep{peng2022perspective}. \citet{peng_diagnosing_2021} proposed a
-series of exercises for training students in data analysis to diagnose
-different kinds of analysis problems such as coding errors or outliers.
-They provide a systematic approach involving working backwards from the
-analysis result to identify potential causes. There are parallels here
-to the concept of debugging and testing in software engineering
+cause of a general unexpected analysis result are typically not
+presented \citep{peng2022perspective}. Furthermore, if interesting
+information is discovered through model checking or data visualization,
+there is no formal way to document such discoveries for the next analyst
+to consider. \citet{peng_diagnosing_2021} proposed a series of exercises
+for training students in data analysis to diagnose different kinds of
+analysis problems such as coding errors or outliers. They provide a
+systematic approach involving working backwards from the analysis result
+to identify potential causes. There are parallels here to the concept of
+debugging and testing in software engineering
 \citep{donoghue2021teaching}. For example, \citet{li2019towards} found
@@
 identify issues quickly and correct them on the spot, but they often do
-so without discussing the underlying reasoning, making it harder for
-less experienced researchers to learn and master these skills.
+so without discussing the underlying reasoning, limiting transparency
+and making it harder for less experienced researchers to learn and
+master these skills.
 
@@
 
-Consider a 30-day step count experiment in public health. Subjects are
-instructed to walk at least 8,000 steps each day, with an expected
-average of 9,000 steps, tracked by a step counter app. After 30 days, we
-review the data and examine the number of steps taken each day. With
-data of this nature, we may expect there to be occasional ``low'' days
-due to factors such as forgetting to wear their watch or unfavorable
-weather conditions limiting outdoor activities. We may also expect
-``high'' days recorded after an outdoor hike or intense workout. Given
-the requirements of the study, we expect that for a given subject the
-average step count would be between \([8500, 9500]\).
+Consider a 30-day step count experiment in a public health setting.
+Subjects are instructed to walk at least 8,000 steps each day, with an
+expected average of 9,000 steps, tracked by a step counter app. After 30
+days, we review the data and examine the number of steps taken each day.
+With data of this nature, we may expect there to be occasional ``low''
+days due to factors such as forgetting to wear their watch or
+unfavorable weather conditions limiting outdoor activities. We may also
+expect ``high'' days recorded after an outdoor hike or intense workout.
+Given the requirements of the study, we expect that for a given subject
+the average step count would be between \([8500, 9500]\).
 
@@
 seeing the data. For example, we can check the quantile of the step
-count, if more than a third of the days fall below 8,000, or more than
-two-thirds exceed 10,000 steps, this could indicate an excess of
-low-count or high-count days. Similarly, we may may expect the standard
-deviation of the step count not to be overly large. These considerations
-yield the following three analysis validation checks that \emph{fails}
-when:
+count, if more than a third of the days fall below 8,000, or more than a
+third exceed 10,000 steps, this could indicate an excess of low-count or
+high-count days. Similarly, we may may expect the standard deviation of
+the step count not to be overly large. These considerations yield the
+following three analysis validation checks that \emph{fail} when:
 
@@
 
-\includegraphics{index_files/figure-pdf/fig-step-count-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-step-count-1.pdf}}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{figures/metric-calc.png}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{figures/metric-calc.png}
 
@@
 
-From the simulated data, the accuracy branch refers to a set of checks'
-ability to accurately detect unexpected outcomes while minimizing false
-positives and false negatives. While a false positive can raise caution
-or skepticism on the data, the presence of both false positives and
-false negatives suggest that the checks are overly sensitive or lack
-sensitivity to unexpected outcomes.
+From the simulated data, the accuracy branch in
+Figure~\ref{fig-metric-calc} refers to a set of checks' ability to
+accurately detect unexpected outcomes while minimizing false positives
+and false negatives. While a false positive can raise caution or
+skepticism about the data, the presence of both false positives and
+false negatives suggests that the checks are overly sensitive or lack
+sensitivity to unexpected outcomes, respectively.
 
@@
 
-While checks may score high on accuracy, they may be less effective at
-explaining the various reasons behind unexpected results. This could
-happen if, for example, a set of checks are all tangentially related to
-the cause of the unexpected results, but none addresses the root cause.
-It may also occur if the checks are highly correlated with one another,
-leading to redundancy.
+While checks may score high on predictive accuracy, they may be less
+effective at explaining the various reasons behind unexpected results.
+This could happen if, for example, a set of checks are all tangentially
+related to the cause of the unexpected results, but none addresses the
+root cause. It may also occur if the checks are highly correlated with
+one another, leading to redundancy.
 
@@
 
-\includegraphics{index_files/figure-pdf/fig-logic-reg-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-logic-reg-1.pdf}}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.5368}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1053}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0947}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1158}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.5368}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1053}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0947}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1158}}@{}}
 
@@
 Study (NMMAPS) to demonstrate how our analysis validation checks
-described in Section~\ref{sec-method} can be applied.
+described in Section~\ref{sec-method} can be applied. The dataset is
+from New York City, and contains daily PM10, mortality, and temperature
+values from 1992--2000.
 
@@
 Sample size less than or equal to 800 \\
-Mortality-PM10 correlation less than -0.03 \\
-Mortality-PM10 correlation less than -0.04 \\
-Mortality-PM10 correlation less than -0.05 \\
-Mortality-PM10 correlation less than -0.06 \\
-Mortality-temperature correlation greater than -0.3 \\
-Mortality-temperature correlation greater than -0.35 \\
-Mortality-temperature correlation greater than -0.4 \\
-Mortality-temperature correlation greater than -0.45 \\
+Mortality-PM10 correlation less than \$-\$0.03 \\
+Mortality-PM10 correlation less than \$-\$0.04 \\
+Mortality-PM10 correlation less than \$-\$0.05 \\
+Mortality-PM10 correlation less than \$-\$0.06 \\
+Mortality-temperature correlation greater than \$-\$0.3 \\
+Mortality-temperature correlation greater than \$-\$0.35 \\
+Mortality-temperature correlation greater than \$-\$0.4 \\
+Mortality-temperature correlation greater than \$-\$0.45 \\
 Outlier(s) are presented in the variable PM10 \\
@@
 To ensure a reasonable likeness to data that might be used in such an
-analysis, we use characteristics of the observed dataset to refine our
-simulations. AIC is used to determine the best distribution fit for each
-variable with the QQ-plot presented in Figure~\ref{fig-dist-fit} to
-evaluate the fit. AIC suggests a negative binomial distribution for
-mortality, a beta distribution for PM10 (multiple by 100 to recover the
-original scale), and a Weibull distribution for temperature. To include
-the potential effect of outliers, we add a single outlier to the data
-for both the mortality and PM10 variables.
+analysis, we use characteristics of the observed New York City dataset
+to refine our simulations. AIC is used to determine the best
+distribution fit for each variable with the QQ-plot presented in
+Figure~\ref{fig-dist-fit} to evaluate the fit. AIC suggests a negative
+binomial distribution for mortality, a beta distribution for PM10
+(multiple by 100 to recover the original scale), and a Weibull
+distribution for temperature. To include the potential effect of
+outliers, we add a single outlier to the data for both the mortality and
+PM10 variables.
 
@@
 
-\includegraphics{index_files/figure-pdf/fig-dist-fit-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-dist-fit-1.pdf}}
 
@@
 
-\includegraphics{index_files/figure-pdf/unnamed-chunk-8-1.pdf}
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/unnamed-chunk-8-1.pdf}}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.6106}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0885}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0619}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0796}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0973}}@{}}
+  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.6106}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0885}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0796}}
+  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0973}}@{}}
 
@@
 
+\subsection{subsection}\label{subsection}
+
+\begin{table}
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}{lrrrrrrr}
+\toprule
+City & \makecell[l]{PM10\\Coefficient} & \makecell[l]{Outcome\\Unexpected} & \makecell[l]{Predicted\\Unexpected} & \makecell[l]{Correlation\\Mortality\\PM10} & \makecell[l]{Correlation\\Mortality\\Temperature} & \makecell[l]{Mortality\\Outlier} & \makecell[l]{PM10\\Outlier}\\
+\midrule
+Atlanta & -0.0008 & T & T & --0.26 (T) & --0.16 (T) & F & F\\
+Austin & -0.0030 & T & T & --0.12 (T) & --0.11 (T) & F & T\\
+Baltimore & 0.0018 & F & F & --0.21 (T) & 0.01 (F) & T & T\\
+Boston & 0.0019 & F & F & --0.16 (T) & 0.02 (F) & F & T\\
+Buffalo & 0.0028 & F & F & --0.23 (T) & --0.02 (F) & F & T\\
+\addlinespace
+Chicago & 0.0011 & F & F & --0.38 (F) & --0.02 (F) & T & T\\
+Cincinnati & 0.0011 & F & T & --0.27 (T) & --0.06 (T) & T & T\\
+Cleveland & 0.0010 & F & F & --0.30 (F) & --0.02 (F) & T & T\\
+Columbus & 0.0004 & F & T & --0.20 (T) & --0.07 (T) & F & T\\
+Denver & 0.0010 & F & F & --0.23 (T) & 0.06 (F) & F & T\\
+\addlinespace
+Detroit & 0.0011 & F & F & --0.30 (T) & 0.04 (F) & F & T\\
+Dallas/Fort Worth & 0.0016 & F & T & --0.30 (F) & --0.03 (T) & T & T\\
+Honolulu & 0.0006 & F & F & --0.17 (T) & 0.05 (F) & T & T\\
+Houston & 0.0007 & F & F & --0.22 (T) & --0.03 (F) & F & T\\
+Indianapolis & 0.0014 & F & F & --0.13 (T) & 0.02 (F) & F & T\\
+\addlinespace
+Kansas City & 0.0022 & F & F & --0.20 (T) & 0.02 (F) & F & T\\
+Los Angeles & 0.0011 & F & F & --0.47 (F) & --0.01 (F) & T & T\\
+Las Vegas & 0.0008 & F & F & --0.27 (T) & 0.05 (F) & T & T\\
+Memphis & -0.0007 & T & T & --0.16 (T) & --0.10 (T) & F & T\\
+Miami & -0.0004 & T & F & --0.16 (T) & --0.02 (F) & F & T\\
+\addlinespace
+Milwaukee & 0.0026 & F & F & --0.22 (T) & 0.09 (F) & F & T\\
+Minneapolis/St. Paul & 0.0009 & F & F & --0.30 (F) & --0.01 (F) & T & T\\
+New York & 0.0022 & F & F & --0.46 (F) & --0.01 (F) & T & T\\
+Oakland & 0.0008 & F & F & --0.28 (T) & 0.04 (F) & F & T\\
+Orlando & -0.0008 & T & T & --0.21 (T) & --0.05 (T) & F & T\\
+\addlinespace
+Philadelphia & 0.0021 & F & F & --0.27 (T) & 0.11 (F) & T & T\\
+Phoenix & 0.0009 & F & F & --0.44 (F) & 0.07 (F) & T & T\\
+Pittsburgh & 0.0009 & F & T & --0.36 (F) & --0.08 (T) & T & T\\
+Riverside & 0.0001 & F & T & --0.27 (T) & --0.11 (T) & T & T\\
+Sacramento & 0.0005 & F & F & --0.30 (T) & 0.05 (F) & T & T\\
+\addlinespace
+Salt Lake City & -0.0004 & T & T & --0.19 (T) & --0.03 (T) & F & T\\
+San Antonio & -0.0021 & T & T & --0.22 (T) & --0.16 (T) & F & T\\
+San Bernardino & 0.0006 & F & T & --0.28 (T) & --0.09 (T) & T & T\\
+San Diego & 0.0011 & F & F & --0.33 (F) & 0.03 (F) & T & T\\
+San Jose & 0.0006 & F & F & --0.23 (T) & 0.08 (F) & F & T\\
+\addlinespace
+Seattle & 0.0003 & F & F & --0.24 (T) & 0.06 (F) & F & T\\
+Santa Ana/Anaheim & 0.0005 & F & T & --0.26 (T) & --0.04 (T) & T & T\\
+St. Petersburg & 0.0012 & F & F & --0.31 (F) & 0.01 (F) & F & T\\
+Tampa & -0.0003 & T & T & --0.20 (T) & --0.04 (T) & F & T\\
+Tucson & 0.0028 & F & F & --0.31 (F) & 0.18 (F) & T & T\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
 \section{Discussion}\label{sec-discussion}
@@
 unexpected outcome. We demonstrated our method on an example relating
-daily mortality to outdoor air pollution data.
+daily mortality to outdoor air pollution data. The results from that
+example could be used to inform future analyses of air pollution and
+health data, perhaps in other cities or locations.
 
@@
 
-Systematically generating simulated data is a key component of our
-approach. In the PM10 example, the inverse transform method is used to
-preserve the correlation structure among mortality, PM10, and
-temperature. However, simulation can become complex when additional
-restrictions are imposed. In such cases, techniques like the
-acceptance-reject method or permutation may be used to generate the
-data.
+Systematically generating realistic simulated data is a key component of
+our approach and is deserving of further consideration. In the PM10
+example, the inverse transform method is used to preserve the
+correlation structure among mortality, PM10, and temperature. However,
+simulation can become complex when additional restrictions are imposed
+or when a greater range of scenarios is desired. In such cases,
+techniques like the acceptance-reject method or permutation may be used
+to generate the data.
 
@@
 
+Communicating the process of data analysis is a key element to providing
+transparency, improving reproducibility, and building trust with a range
+of audiences. The analysis validation checks described here provide a
+general way to encode the assumptions that an data analyst makes about
+the data and statistical tools applied to the data. These assumptions
+can be studied or challenged, depending on the specific analyst's
+perspective, and can serve as a roadmap for diagnosing unexpected
+results.
+
 \section{Acknowledgement}\label{acknowledgement}

### 2024-12-19 index.qmd
new table results

@@
 
+## subsection
+
+```{r}
+pm10 <- read_csv("data/pm10.csv") |> filter(!is.na(pm10)) |> filter(!is.na(temp))
+files <- list.files(path = "data", pattern = "csv", full.names = TRUE)[-1]
+pred_res <- files |> map_dfr(function(raw){
+  data <- read_csv(raw) |> filter(!is.na(pm10)) |> filter(!is.na(temp))
+  res_df <- glm(mortality ~ pm10 + temp, family = "poisson", data = data)$coefficients |>
+    as_tibble_row()
+
+  res_df |>
+      mutate(
+        cor_m_tmp = cor(data$mortality, data$temp),
+        cor_m_pm10 = cor(data$mortality, data$pm10),
+        cor_tmp_pm10 = cor(data$temp, data$pm10),
+        unexpect = !between(pm10, 0, 0.005),
+        mort_outlier = any(scale(data$mortality) > 4),
+        pm10_outlier = any(scale(data$pm10) > 4),
+        lgl_cor_m_tmp = cor(data$mortality, data$temp) > -0.3,
+        lgl_cor_m_pm10 = cor(data$mortality, data$pm10) < -0.03,
+        .pred = ((mort_outlier &  pm10_outlier) | (lgl_cor_m_tmp)) & (lgl_cor_m_pm10),
+        ss = nrow(data)
+        )
+    })
+
+library(readr)
+library(kableExtra)
+city_name_map <- read_csv(here::here("data/city_name_map.csv"))
+tibble(city = str_remove(files, "data/pm10-") |> str_remove(".csv")) |> bind_cols(pred_res) |>
+  left_join(city_name_map) |> 
+  filter(city != "data/pm10") |> 
+  dplyr::select(cityname, pm10, unexpect, .pred, cor_m_tmp, lgl_cor_m_tmp,
+                cor_m_pm10, lgl_cor_m_pm10, mort_outlier, pm10_outlier) |>
+  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
+         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
+         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
+         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
+  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
+                ~ifelse(.x, "T", "F"))) |>
+  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
+         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
+  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |>
+  kable(format = "latex", booktabs = TRUE, escape = F, align = "lrrrrrrr",
+        col.names = linebreak(c("City", "PM10\nCoefficient", "Outcome\nUnexpected", "Predicted\nUnexpected", 
+                                "Correlation\nMortality\nPM10", "Correlation\nMortality\nTemperature",
+                                "Mortality\nOutlier", "PM10\nOutlier"))) |>
+  kableExtra::kable_styling(latex_options="scale_down") 
+```
+
 # Discussion {#sec-discussion}

### 2024-12-19 index.qmd
Add paragraphs on data

--- 2024-12-17 index.qmd
@@
 For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
-The dataset is from New York City, and contains daily PM10, mortality, and temperature values from 1992--2000.
+The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000.
+We then will apply the analysis checks to similar data from 39 other cities from the NMMAPS study.
 
@@
 Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet2000fine; @welty2005acute].
-Observing an estimated coefficient outside of this interval would be highly unusual and would warrant a serious re-examination of the analysis process.
+Note that the relatively simple modeling approach being used here is primarily for demonstration; typically far more sophisticated semi-parametric approaches are used in the literature [@peng2006model].
+
+For the purposes of this application, observing an estimated coefficient outside of the expected interval is considered unusual and might warrant a re-examination of the analysis process.
 Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval $[0, 0.005]$.
@@
 
-| the check fails if ... | 
+| The check fails if ... | 
 |--------------| 
@@
 | Sample size less than or equal to 800 |
-| Mortality-PM10 correlation less than $-$0.03 |
-| Mortality-PM10 correlation less than $-$0.04 |
-| Mortality-PM10 correlation less than $-$0.05 |
-| Mortality-PM10 correlation less than $-$0.06 |
-| Mortality-temperature correlation greater than $-$0.3 | 
-| Mortality-temperature correlation greater than $-$0.35 | 
-| Mortality-temperature correlation greater than $-$0.4 | 
-| Mortality-temperature correlation greater than $-$0.45 | 
+| Mortality-PM10 correlation less than $-0.03$ |
+| Mortality-PM10 correlation less than $-0.04$ |
+| Mortality-PM10 correlation less than $-0.05$ |
+| Mortality-PM10 correlation less than $-0.06$ |
+| Mortality-temperature correlation greater than $-0.3$ | 
+| Mortality-temperature correlation greater than $-0.35$ | 
+| Mortality-temperature correlation greater than $-0.4$ | 
+| Mortality-temperature correlation greater than $-0.45$ | 
 | Outlier(s) are presented in the variable PM10 | 
@@
 
+## Additional Cities
+
+In order to test our development of analysis validation checks based on the New York City data, we incorporate data from 39 additional cities from the original NMMAPS study and apply our checks to those data.
+These cities (along with New York City) represent approximately the largest 40 metropolitan areas in the United States.
+The complete list of cities used are presented in @tbl-all-cities.
+For each city, we had daily data on all-cause mortality, PM10, and average temperature for the years 1992--2000.
+
+```{r}
+#| label: tbl-all-cities
+
+pm10 <- read_csv("data/pm10.csv") |> filter(!is.na(pm10)) |> filter(!is.na(temp))
+files <- list.files(path = "data", pattern = "csv", full.names = TRUE)[-1]
+pred_res <- files |> map_dfr(function(raw){
+  data <- read_csv(raw) |> filter(!is.na(pm10)) |> filter(!is.na(temp))
+  res_df <- glm(mortality ~ pm10 + temp, family = "poisson", data = data)$coefficients |>
+    as_tibble_row()
+
+  res_df |>
+      mutate(
+        cor_m_tmp = cor(data$mortality, data$temp),
+        cor_m_pm10 = cor(data$mortality, data$pm10),
+        cor_tmp_pm10 = cor(data$temp, data$pm10),
+        unexpect = !between(pm10, 0, 0.005),
+        mort_outlier = any(scale(data$mortality) > 4),
+        pm10_outlier = any(scale(data$pm10) > 4),
+        lgl_cor_m_tmp = cor(data$mortality, data$temp) > -0.3,
+        lgl_cor_m_pm10 = cor(data$mortality, data$pm10) < -0.03,
+        .pred = ((mort_outlier &  pm10_outlier) | (lgl_cor_m_tmp)) & (lgl_cor_m_pm10),
+        ss = nrow(data)
+        )
+    })
+
+library(readr)
+library(kableExtra)
+city_name_map <- read_csv(here::here("data/city_name_map.csv"))
+tibble(city = str_remove(files, "data/pm10-") |> str_remove(".csv")) |> bind_cols(pred_res) |>
+  left_join(city_name_map) |> 
+  filter(city != "data/pm10") |> 
+  dplyr::select(cityname, pm10, unexpect, .pred, cor_m_tmp, lgl_cor_m_tmp,
+                cor_m_pm10, lgl_cor_m_pm10, mort_outlier, pm10_outlier) |>
+  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
+         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
+         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
+         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
+  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
+                ~ifelse(.x, "T", "F"))) |>
+  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
+         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
+  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |>
+  kable(format = "latex", booktabs = TRUE, escape = F, 
+        align = "lrccrrcc",
+        col.names = linebreak(c("City", "PM10\nCoefficient", "Outcome\nUnexpected", "Predicted\nUnexpected", 
+                                "Correlation\nMortality\nPM10", "Correlation\nMortality\nTemperature",
+                                "Mortality\nOutlier", "PM10\nOutlier"))) |>
+  kableExtra::kable_styling(latex_options="scale_down") 
+```
+
+We applied each of the analysis validation checks and the fitted logic regression model separately to each of the cities' data to make predictions of whether the outcome will be unexpected or not.
+We also applied the generalized linear model to the data from each of the cities to estimate the association between PM10 and mortality in each city.
+The estimates of the association and the results of each of the analysis validation checks for each city are shown in @tbl-all-cities.
+
+[Sherry: Summarize the results from the 40 cities, along with 2x2 table of outcome vs. predicted]
+
 # Discussion {#sec-discussion}

### 2024-12-20 _manuscript/_tex/index.tex
summary + 2 by 2 table

@@
 Study (NMMAPS) to demonstrate how our analysis validation checks
-described in Section~\ref{sec-method} can be applied. The dataset is
-from New York City, and contains daily PM10, mortality, and temperature
-values from 1992--2000.
+described in Section~\ref{sec-method} can be applied. The dataset that
+we use to develop our simulations and analysis checks is from New York
+City, and contains daily PM10, all-cause (non-accidental) mortality, and
+average temperature values from 1992--2000. We then will apply the
+analysis checks to similar data from 39 other cities from the NMMAPS
+study.
 
@@
 between \([0, 0.005]\), after adjusting for daily temperature
-\citep{samet2000fine, welty2005acute}. Observing an estimated
-coefficient outside of this interval would be highly unusual and would
-warrant a serious re-examination of the analysis process. Therefore, our
-unexpected outcome in this analysis will be the binary indicator of
-whether the estimated coefficient for PM10 in the GLM lies outside of
-the interval \([0, 0.005]\). In addition to providing a more substantial
-problem for our methods, this example also demonstrates how the
-procedure presented in Section~\ref{sec-method} can be used to select
-cutoff values in the analysis checks to diagnose an unexpected PM10
-coefficient from the generalized linear model.
+\citep{samet2000fine, welty2005acute}. Note that the relatively simple
+modeling approach being used here is primarily for demonstration;
+typically far more sophisticated semi-parametric approaches are used in
+the literature \citep{peng2006model}.
+
+For the purposes of this application, observing an estimated coefficient
+outside of the expected interval is considered unusual and might warrant
+a re-examination of the analysis process. Therefore, our unexpected
+outcome in this analysis will be the binary indicator of whether the
+estimated coefficient for PM10 in the GLM lies outside of the interval
+\([0, 0.005]\). In addition to providing a more substantial problem for
+our methods, this example also demonstrates how the procedure presented
+in Section~\ref{sec-method} can be used to select cutoff values in the
+analysis checks to diagnose an unexpected PM10 coefficient from the
+generalized linear model.
 
@@
 \toprule\noalign{}
-the check fails if \ldots{} \\
+The check fails if \ldots{} \\
 \midrule\noalign{}
@@
 Sample size less than or equal to 800 \\
-Mortality-PM10 correlation less than \$-\$0.03 \\
-Mortality-PM10 correlation less than \$-\$0.04 \\
-Mortality-PM10 correlation less than \$-\$0.05 \\
-Mortality-PM10 correlation less than \$-\$0.06 \\
-Mortality-temperature correlation greater than \$-\$0.3 \\
-Mortality-temperature correlation greater than \$-\$0.35 \\
-Mortality-temperature correlation greater than \$-\$0.4 \\
-Mortality-temperature correlation greater than \$-\$0.45 \\
+Mortality-PM10 correlation less than \(-0.03\) \\
+Mortality-PM10 correlation less than \(-0.04\) \\
+Mortality-PM10 correlation less than \(-0.05\) \\
+Mortality-PM10 correlation less than \(-0.06\) \\
+Mortality-temperature correlation greater than \(-0.3\) \\
+Mortality-temperature correlation greater than \(-0.35\) \\
+Mortality-temperature correlation greater than \(-0.4\) \\
+Mortality-temperature correlation greater than \(-0.45\) \\
 Outlier(s) are presented in the variable PM10 \\
@@
 
-\subsection{subsection}\label{subsection}
+\subsection{Additional Cities}\label{additional-cities}
+
+In order to test our development of analysis validation checks based on
+the New York City data, we incorporate data from 39 additional cities
+from the original NMMAPS study and apply our checks to those data. These
+cities (along with New York City) represent approximately the largest 40
+metropolitan areas in the United States. The complete list of cities
+used are presented in Table~\ref{tbl-all-cities}. For each city, we had
+daily data on all-cause mortality, PM10, and average temperature for the
+years 1992--2000.
 
 \begin{table}
+
+\caption{\label{tbl-all-cities}}
+
+\centering{
+
 \centering
 \resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
-\begin{tabular}{lrrrrrrr}
+\begin{tabular}{lrccrrcc}
 \toprule
@@
 \end{tabular}}
-\end{table}
+
+}
+
+\end{table}%
+
+We applied each of the analysis validation checks and the fitted logic
+regression model separately to each of the cities' data to make
+predictions of whether the outcome will be unexpected or not. We also
+applied the generalized linear model to the data from each of the cities
+to estimate the association between PM10 and mortality in each city. The
+estimates of the association and the results of each of the analysis
+validation checks for each city are shown in Table~\ref{tbl-all-cities}.
+
+\begin{table}
+
+\caption{\label{tbl-accuracy}}
+
+\centering{
+
+\caption{A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.}
+\centering
+\begin{tabular}[t]{lrr}
+\toprule
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{Prediction} \\
+\cmidrule(l{3pt}r{3pt}){2-3}
+Actual & FALSE & TRUE\\
+\midrule
+FALSE & 25 & 7\\
+TRUE & 1 & 7\\
+\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+Table~\ref{tbl-accuracy} summaries the prediction accuracy of the logic
+regression model for the 40 cities. The model correctly predicts
+unexpected outcomes in 32 cities, producing an accuracy of 80\%. Among
+the 8 cities with an unexpected outcome (PM10 coefficient outside {[}0,
+0.005{]}), the model correctly identifies 7 (Atlanta, Austin, Memphis,
+Orlando, Salt Lake City, San Antonio, Tampa), giving a recall of 88\%.
+These cities are flagged due to failure in both mortality-PM10
+correlation and mortality-temperature correlation checks. Out of the 14
+cities with positive predictions from the model, 7 cities (Cincinnati,
+Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
+Santa Ana/Anaheim) were a false positive, resulting in a precision of
+50\%.
+
+{[}Sherry: Summarize the results from the 40 cities, along with 2x2
+table of outcome vs.~predicted{]}
 

### 2024-12-20 index.qmd
summary + 2 by 2 table

@@
 library(patchwork)
+library(readr)
+library(kableExtra)
+library(scales)
 pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
@@
     })
+city_name_map <- read_csv(here::here("data/city_name_map.csv"))
 
-library(readr)
-library(kableExtra)
-city_name_map <- read_csv(here::here("data/city_name_map.csv"))
-tibble(city = str_remove(files, "data/pm10-") |> str_remove(".csv")) |> bind_cols(pred_res) |>
+df <- tibble(city = str_remove(files, "data/pm10-") |> str_remove(".csv")) |> bind_cols(pred_res) |>
   left_join(city_name_map) |> 
@@
   dplyr::select(cityname, pm10, unexpect, .pred, cor_m_tmp, lgl_cor_m_tmp,
-                cor_m_pm10, lgl_cor_m_pm10, mort_outlier, pm10_outlier) |>
+                cor_m_pm10, lgl_cor_m_pm10, mort_outlier, pm10_outlier)
+
+df |>
   mutate(pm10 = format(round(pm10, 4), nsmall = 4),
@@
 
-[Sherry: Summarize the results from the 40 cities, along with 2x2 table of outcome vs. predicted]
+```{r}
+#| label: tbl-accuracy
+df |> 
+  dplyr::select(unexpect, .pred) |> 
+  table() |> 
+  as_tibble() |> 
+  rename(Actual = unexpect) |> 
+  pivot_wider(names_from = .pred, values_from = n) |> 
+  kable(caption = "A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.", 
+        format = "latex", booktabs = TRUE) |> 
+   add_header_above(c("", "Prediction"=2), escape = FALSE) 
+```
+
+@tbl-accuracy summaries the prediction accuracy of the logic regression model for the 40 cities.
+The model correctly predicts `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy of `r label_percent()(accurate/nrow(df))`.
+Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r df |> filter(unexpect, .pred) |> pull(cityname)`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)`.
+These cities are flagged due to failure in both mortality-PM10 correlation and mortality-temperature correlation checks.
+Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were a false positive, resulting in a precision of `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`.
 

### 2024-12-20 index.qmd

@@
 The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
-While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the publication code alone is often insufficient for understanding the thought process behind an analysis.
+While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis.
 Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing.
@@
 To quantify redundancy, the concept of mutual information is used.
-Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the KL-distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:
+Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the Kullback-Leibler distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:
 
@@
 In order to test our development of analysis validation checks based on the New York City data, we incorporate data from 39 additional cities from the original NMMAPS study and apply our checks to those data.
-These cities (along with New York City) represent approximately the largest 40 metropolitan areas in the United States.
+These cities (along with New York City) represent approximately the largest 40 metropolitan areas in the United States by population.
 The complete list of cities used are presented in @tbl-all-cities.
@@
 #| label: tbl-all-cities
+#| tbl-cap: Results from full analysis of 40 cities from the NMMAPS air pollution and mortality study. In the table, a "T" indicates that a check failed and an "F" indicates that a check passed.
 
@@
   filter(city != "data/pm10") |> 
-  dplyr::select(cityname, pm10, unexpect, .pred, cor_m_tmp, lgl_cor_m_tmp,
-                cor_m_pm10, lgl_cor_m_pm10, mort_outlier, pm10_outlier)
+  dplyr::select(cityname, pm10, unexpect, .pred, 
+                cor_m_tmp, lgl_cor_m_tmp,                
+                cor_m_pm10, lgl_cor_m_pm10, 
+                mort_outlier, pm10_outlier)
 
@@
         align = "lrccrrcc",
-        col.names = linebreak(c("City", "PM10\nCoefficient", "Outcome\nUnexpected", "Predicted\nUnexpected", 
-                                "Correlation\nMortality\nPM10", "Correlation\nMortality\nTemperature",
-                                "Mortality\nOutlier", "PM10\nOutlier"))) |>
+        col.names = linebreak(c("City", 
+                                "PM10\nEstimated\nCoefficient",
+                                "Outcome\nObserved\nUnexpected", 
+                                "Logic Regression\nPredicted\nUnexpected", 
+                                "Correlation:\nMortality-\nTemperature",
+                                "Correlation:\nMortality-\nPM10",
+                                "Mortality\nOutlier", 
+                                "PM10\nOutlier"))) |>
   kableExtra::kable_styling(latex_options="scale_down") 
@@
 We applied each of the analysis validation checks and the fitted logic regression model separately to each of the cities' data to make predictions of whether the outcome will be unexpected or not.
-We also applied the generalized linear model to the data from each of the cities to estimate the association between PM10 and mortality in each city.
+We also applied the generalized linear model to the data from each of the cities to estimate the association between PM10 and mortality in each city, adjusted for temperature.
 The estimates of the association and the results of each of the analysis validation checks for each city are shown in @tbl-all-cities.
@@
 #| label: tbl-accuracy
+#| tbl-cap: Summary of the observed and the predicted unexpected PM10 coefficient results from the 40 NMMAPS cities using the logic regression model.
+
 df |> 
@@
   as_tibble() |> 
-  rename(Actual = unexpect) |> 
+  rename(Observed = unexpect) |> 
   pivot_wider(names_from = .pred, values_from = n) |> 
-  kable(caption = "A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.", 
-        format = "latex", booktabs = TRUE) |> 
-   add_header_above(c("", "Prediction"=2), escape = FALSE) 
+  kable(format = "latex", 
+        booktabs = TRUE) |> 
+   add_header_above(c("", "Predicted"=2), 
+                    escape = FALSE) 
 ```
 
-@tbl-accuracy summaries the prediction accuracy of the logic regression model for the 40 cities.
-The model correctly predicts `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy of `r label_percent()(accurate/nrow(df))`.
-Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r df |> filter(unexpect, .pred) |> pull(cityname)`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)`.
-These cities are flagged due to failure in both mortality-PM10 correlation and mortality-temperature correlation checks.
-Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were a false positive, resulting in a precision of `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`.
+```{r,include=FALSE}
+cities_accurate <- df |> 
+  filter(unexpect, .pred) |> 
+  pull(cityname)
+cities_missed <- df |> 
+  filter(unexpect, !.pred) |> 
+  pull(cityname)
+```
+
+It is clear from @tbl-all-cities that some of the estimated PM10 coefficients are outside the expected range of [0, 0.005].
+Eight of the 40 cities had negative estimated coefficients and were therefore considered unexpected by our original criterion.
+(It is perhaps worth noting that none of the unexpected outcomes was in the positive direction.) @tbl-accuracy summarizes the prediction accuracy of the logic regression model for the 40 cities.
+The model correctly predicts the PM10 coefficient status of `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy rate of `r label_percent()(accurate/nrow(df))`.
+Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model).
+These cities were flagged due to failures in both the mortality-PM10 correlation and mortality-temperature correlation analysis checks.
+Out of the `r df |> filter(.pred) |> nrow()` cities with positive predictions from the model, `r df |> filter(unexpect != .pred, !unexpect) |> nrow()` cities (`r df |> filter(unexpect != .pred, !unexpect) |> pull(cityname)`) were false positives, resulting in a precision of only `r pres <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(precision)); label_percent()(pres)`.
+This precision value is substantially lower than what was estimated by the simulation procedure (see @tbl-linear-reg) and suggests that the simulation does not adequately capture some features of the data generation process.
 
@@
 They also support students and junior researchers develop the skills needed to independently diagnose issues in analyses.
+
+In @sec-pm10-mortality we used the analysis checks in a diagnostic manner to examine the eight cities whose PM10 coefficients were unexpected.
+There, we found that seven of the cities had a mortality-PM10 correlation that was more negative than expected while also having a mortality-temperature correlation that was somewhat larger (less negative) than expected.
+What this result implies for the broader analysis depends on a number of factors, but the interplay between mortality, PM10, and temperature may warrant further investigation [see e.g.
+@welty2005acute].
+Note that simply because the PM10 coefficient estimates were unexpected by our criterion, we do not mean to imply that they are ``wrong" in any sense.
+Indeed, negative estimated coefficients have been found in other studies of this nature [@bell2004ozone].
+Rather, in this type of analysis, it may be that the interval for expected results needs to be revised.
+Further, it may be necessary to incorporate other outputs from the analysis, such as measures of uncertainty.
 
@@
 Systematically generating realistic simulated data is a key component of our approach and is deserving of further consideration.
-In the PM10 example, the inverse transform method is used to preserve the correlation structure among mortality, PM10, and temperature.
-However, simulation can become complex when additional restrictions are imposed or when a greater range of scenarios is desired.
+In the PM10 example in @sec-pm10-mortality, the inverse transform method was used to preserve the correlation structure among mortality, PM10, and temperature.
+However, the simulation process can become complex when additional restrictions are imposed or when a greater range of scenarios is desired.
 In such cases, techniques like the acceptance-reject method or permutation may be used to generate the data.
+The results in @sec-pm10-mortality on the broader NMMAPS dataset suggest that our simulation procedure may have been inadequate in reflecting the range of possible configurations that the data could take.
+In particular, our split-data approach, using New York City to guide the simulation and then applying the analysis checks to other cities, may not have been ideal.
+It may be worth exploring some recent work in data thinning [@neufeld2024data] and data fission [@leiner2023data] to create training datasets that are more representative of future data.
 

### 2024-12-20 _manuscript/_tex/index.tex
update application results

@@
 requirement for the sake of reproducibility
-\citep{peng2011reproducible}, the publication code alone is often
+\citep{peng2011reproducible}, the published code alone is often
 insufficient for understanding the thought process behind an analysis.
@@
 Mutual information \(I(x, y)\) measures the amount of information shared
-between two random variables and is defined as the KL-distance
-\(D(p \parallel q)\) between the joint distribution of the two variables
-and the product of the marginal distributions:
+between two random variables and is defined as the Kullback-Leibler
+distance \(D(p \parallel q)\) between the joint distribution of the two
+variables and the product of the marginal distributions:
 
@@
 \textgreater{} 2,500). The NOT operator applied to
-\texttt{sd(step)\ \textgreater{}\ 2,500)} is colored with a black
+\texttt{sd(step)\ \textgreater{}\ 2,500} is colored with a black
 background to distinguish it from other checks.}
@@
 the logic regression. The harmonic and arithmetic means are included to
-combine the three measures. The results show that the two checks
+combine the three measures. The results show that the three checks
 produced by the logic regression can accurately predict 86.1\% cases of
@@
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.5368}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1053}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0737}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0947}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.1158}}@{}}
+\begin{table}
 
@@
 
-\tabularnewline
-
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-checks
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-precision
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-recall
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-indep.
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-harmonic
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-arithmetic
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-check 1: q(step, 0.6) \textgreater{} 10000 & 0.319 & 0.575 & 1.000 &
-0.511 & 0.631 \\
-check 2: q(step, 0.4) \textless{} 8000 & 0.264 & 0.613 & 1.000 & 0.467 &
-0.626 \\
-check 3: sd(step) \textgreater{} 2500 & 0.153 & 0.289 & 1.000 & 0.273 &
-0.481 \\
-logic regression: (check 1 OR check 2) AND check 3 & 0.431 & 0.861 &
-0.998 & 0.669 & 0.763 \\
-regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774 \\
-
-\end{longtable}
+\centering{
+
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}{>{\raggedright\arraybackslash}p{17em}rrrrr}
+\toprule
+checks & precision & recall & indep. & harmonic & arithmetic\\
+\midrule
+Check 1: q(step, 0.6) > 10000 & 0.319 & 0.575 & 1.000 & 0.511 & 0.631\\
+Check 2: q(step, 0.4) < 8000 & 0.264 & 0.613 & 1.000 & 0.467 & 0.626\\
+Check 3: sd(step) > 2500 & 0.153 & 0.289 & 1.000 & 0.273 & 0.481\\
+Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.998 & 0.669 & 0.763\\
+Regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774\\
+\bottomrule
+\end{tabular}}
+
+}
+
+\end{table}%
 
@@
 it is within. Multiple factors can affect the estimated PM10
-coefficient, such as the sample size, the strength of the correlation
-between mortality and PM10, and the strength of the correlation between
-mortality and temperature. Analysts may expect a reasonable sample size
-to ensure the reliability of the coefficient estimate. Outliers in the
-three variables can also leverage the coefficient. While these are
-possible factors that could affect the analysis result, it is not clear
-what the cutoff values for these checks should be to determine a
-failure. Here we consider a list of checks in Table~\ref{tbl-checks}
-with varied cutoff values for each:
+coefficient, such as the strength of the correlation between mortality
+and PM10, or between mortality and temperature. Outliers in the three
+variables can also leverage the coefficient. While these are possible
+factors that could affect the analysis result, it is not clear what the
+cutoff values for these checks should be to determine a failure. Here we
+consider a list of checks in Table~\ref{tbl-checks} with varied cutoff
+values for each:
 
@@
 \toprule\noalign{}
-The check fails if \ldots{} \\
+The check fails (encoded as 1) if \ldots{} \\
 \midrule\noalign{}
@@
 \endlastfoot
-Sample size less than or equal to 200 \\
-Sample size less than or equal to 400 \\
-Sample size less than or equal to 600 \\
-Sample size less than or equal to 800 \\
+Mortality-PM10 correlation less than \(-0.05\) \\
 Mortality-PM10 correlation less than \(-0.03\) \\
-Mortality-PM10 correlation less than \(-0.04\) \\
-Mortality-PM10 correlation less than \(-0.05\) \\
-Mortality-PM10 correlation less than \(-0.06\) \\
+Mortality-PM10 correlation greater than \(0.03\) \\
+Mortality-PM10 correlation greater than \(0.05\) \\
 Mortality-temperature correlation greater than \(-0.3\) \\
@@
 Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
-unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
-regression tree from the fitted model. Precision, recall, and
-independence score, along with their harmonic and arithmetic mean are
-calculated in Table~\ref{tbl-linear-reg}.
+unexpected. Given the imbalance of outcome (expected vs.~unexpected),
+inverse weights proportional to the number of observations in each
+outcome are applied during the logic regression fit.
+Figure~\ref{fig-linear-reg-tree} shows the optimal logic regression tree
+from the fitted model. Precision, recall, and independence score, along
+with their harmonic and arithmetic mean are calculated in
+Table~\ref{tbl-linear-reg}.
 
@@
 fourteen checks and the outcome expectation (unexpected) as the response
-variable. The model suggests the relationship: (NOT sample size less
-than or equal to 200 AND mortality-PM10 correlation \textless{} -0.04)
-AND (mortality-temperature correlation \textgreater{} -0.35 OR there
-exist mortality outlier) to predict the unexpected PM10 coefficient. The
-NOT operator applied to \texttt{smpl\_size\ \textless{}=\ 200} is
-colored with a black background to distinguish it from other checks.}
+variable. The model suggests the relationship: (mortality-PM10
+correlation \textless{} -0.03) AND (mortality-temperature correlation
+\textgreater{} -0.3 OR (there exist mortality outlier AND there exist
+PM10 outlier)) to predict the unexpected PM10 coefficient.}
 
@@
 
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.6106}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0885}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0619}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0796}}
-  >{\raggedleft\arraybackslash}p{(\linewidth - 10\tabcolsep) * \real{0.0973}}@{}}
+\begin{table}
 
 \caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
-parsimony (independence) metrics derived from the logic regression
-model, along with harmonic and arithmetic means, for individual checks
-(1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature
-correlation, 4: mortality outlier), and the combined check rule 5:
-(sample size AND mortality-PM10 correlation) AND (mortality-temperature
-correlation OR mortality outlier.}
-
-\tabularnewline
-
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-check
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-precision
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-recall
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-indep.
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-harmonic
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
-arithmetic
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-check 1: sample size \textless= 200 & 0.087 & 0.215 & 1 & 0.175 &
-0.434 \\
-check 2: cor(m, PM10) \textless{} -0.03 & 0.988 & 0.610 & 1 & 0.822 &
-0.866 \\
-check 3: cor(m, tmp) \textgreater{} -0.35 & 0.392 & 0.581 & 1 & 0.569 &
-0.658 \\
-check 4: mortality\_outlier & 0.649 & 0.641 & 1 & 0.732 & 0.763 \\
-logic regression: (not check 1 AND check 2) AND (check 3 OR check 4) &
-0.760 & 0.880 & 1 & 0.869 & 0.880 \\
-
-\end{longtable}
+parsimony (independence) metrics derived from the logic regression model
+(last row) and individual checks, along with harmonic and arithmetic
+means.}
+
+\centering{
+
+\centering
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
+\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
+\toprule
+check & precision & recall & indep. & harmonic & arithmetic\\
+\midrule
+Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1 & 0.763 & 0.815\\
+Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1 & 0.472 & 0.612\\
+Check 3: mortality outlier & 0.302 & 0.775 & 1 & 0.536 & 0.692\\
+Check 4: PM10 outlier & 0.305 & 0.766 & 1 & 0.537 & 0.690\\
+Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.722 & 0.876 & 1 & 0.851 & 0.866\\
+\bottomrule
+\end{tabular}}
+
+}
+
+\end{table}%
 
@@
 \item
-  sample size \emph{larger than} 200
-\item
-  mortality-temperature correlation greater than \(-0.35\)
-\item
-  mortality-PM10 correlation less than \(-0.04\)
+  mortality-PM10 correlation less than \(-0.03\)
+\item
+  mortality-temperature correlation greater than \(-0.3\)
+\item
+  PM10 data contains outliers that are detected by the univariate
+  outlier detection
 \item
@@
 on mortality or correlation between mortality and temperature. This
-combined check rule generates a precision of 0.76 and a recall of 0.88
-for predicting the unexpected PM10 coefficient. The single check,
-\texttt{cor(m,\ PM10)\ \textless{}\ -0.03}, is also powerful with a high
-precision of 0.988, but the low recall value of 0.61 suggests its high
-false positive rate, as compared to the combined rule suggested by the
-logic regression.
+combined check rule generates a precision of 0.722 and a recall of 0.876
+for predicting the unexpected PM10 coefficient.
 
@@
 cities (along with New York City) represent approximately the largest 40
-metropolitan areas in the United States. The complete list of cities
-used are presented in Table~\ref{tbl-all-cities}. For each city, we had
-daily data on all-cause mortality, PM10, and average temperature for the
-years 1992--2000.
+metropolitan areas in the United States by population. The complete list
+of cities used are presented in Table~\ref{tbl-all-cities}. For each
+city, we had daily data on all-cause mortality, PM10, and average
+temperature for the years 1992--2000.
 
@@
 
-\caption{\label{tbl-all-cities}}
+\caption{\label{tbl-all-cities}Results from full analysis of 40 cities
+from the NMMAPS air pollution and mortality study. In the table, a ``T''
+indicates that a check failed and an ``F'' indicates that a check
+passed.}
 
@@
 \toprule
-City & \makecell[l]{PM10\\Coefficient} & \makecell[l]{Outcome\\Unexpected} & \makecell[l]{Predicted\\Unexpected} & \makecell[l]{Correlation\\Mortality\\PM10} & \makecell[l]{Correlation\\Mortality\\Temperature} & \makecell[l]{Mortality\\Outlier} & \makecell[l]{PM10\\Outlier}\\
+City & \makecell[l]{PM10\\Estimated\\Coefficient} & \makecell[l]{Outcome\\Observed\\Unexpected} & \makecell[l]{Logic Regression\\Predicted\\Unexpected} & \makecell[l]{Correlation:\\Mortality-\\Temperature} & \makecell[l]{Correlation:\\Mortality-\\PM10} & \makecell[l]{Mortality\\Outlier} & \makecell[l]{PM10\\Outlier}\\
 \midrule
@@
 applied the generalized linear model to the data from each of the cities
-to estimate the association between PM10 and mortality in each city. The
-estimates of the association and the results of each of the analysis
-validation checks for each city are shown in Table~\ref{tbl-all-cities}.
+to estimate the association between PM10 and mortality in each city,
+adjusted for temperature. The estimates of the association and the
+results of each of the analysis validation checks for each city are
+shown in Table~\ref{tbl-all-cities}.
 
@@
 
-\caption{\label{tbl-accuracy}}
+\caption{\label{tbl-accuracy}Summary of the observed and the predicted
+unexpected PM10 coefficient results from the 40 NMMAPS cities using the
+logic regression model.}
 
@@
 
-\caption{A 2x2 table of the outcome and the predicted unexpected results from the 40 cities.}
-\centering
-\begin{tabular}[t]{lrr}
+\begin{tabular}{lrr}
 \toprule
-\multicolumn{1}{c}{} & \multicolumn{2}{c}{Prediction} \\
+\multicolumn{1}{c}{} & \multicolumn{2}{c}{Predicted} \\
 \cmidrule(l{3pt}r{3pt}){2-3}
-Actual & FALSE & TRUE\\
+Observed & FALSE & TRUE\\
 \midrule
@@
 
-Table~\ref{tbl-accuracy} summaries the prediction accuracy of the logic
-regression model for the 40 cities. The model correctly predicts
-unexpected outcomes in 32 cities, producing an accuracy of 80\%. Among
-the 8 cities with an unexpected outcome (PM10 coefficient outside {[}0,
-0.005{]}), the model correctly identifies 7 (Atlanta, Austin, Memphis,
-Orlando, Salt Lake City, San Antonio, Tampa), giving a recall of 88\%.
-These cities are flagged due to failure in both mortality-PM10
-correlation and mortality-temperature correlation checks. Out of the 14
-cities with positive predictions from the model, 7 cities (Cincinnati,
+It is clear from Table~\ref{tbl-all-cities} that some of the estimated
+PM10 coefficients are outside the expected range of {[}0, 0.005{]}.
+Eight of the 40 cities had negative estimated coefficients and were
+therefore considered unexpected by our original criterion. (It is
+perhaps worth noting that none of the unexpected outcomes was in the
+positive direction.) Table~\ref{tbl-accuracy} summarizes the prediction
+accuracy of the logic regression model for the 40 cities. The model
+correctly predicts the PM10 coefficient status of 32 cities, producing
+an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome
+(PM10 coefficient outside {[}0, 0.005{]}), the model correctly
+identifies 7 (Atlanta, Austin, Memphis, Orlando, Salt Lake City, San
+Antonio, Tampa), giving a recall of 88\% (only Miami was not properly
+classified by the logic regression model). These cities were flagged due
+to failures in both the mortality-PM10 correlation and
+mortality-temperature correlation analysis checks. Out of the 14 cities
+with positive predictions from the model, 7 cities (Cincinnati,
 Columbus, Dallas/Fort Worth, Pittsburgh, Riverside, San Bernardino,
-Santa Ana/Anaheim) were a false positive, resulting in a precision of
-50\%.
-
-{[}Sherry: Summarize the results from the 40 cities, along with 2x2
-table of outcome vs.~predicted{]}
+Santa Ana/Anaheim) were false positives, resulting in a precision of
+only 50\%. This precision value is substantially lower than what was
+estimated by the simulation procedure (see Table~\ref{tbl-linear-reg})
+and suggests that the simulation does not adequately capture some
+features of the data generation process.
 
@@
 issues in analyses.
+
+In Section~\ref{sec-pm10-mortality} we used the analysis checks in a
+diagnostic manner to examine the eight cities whose PM10 coefficients
+were unexpected. There, we found that seven of the cities had a
+mortality-PM10 correlation that was more negative than expected while
+also having a mortality-temperature correlation that was somewhat larger
+(less negative) than expected. What this result implies for the broader
+analysis depends on a number of factors, but the interplay between
+mortality, PM10, and temperature may warrant further investigation
+\citep[see e.g.][]{welty2005acute}. Note that simply because the PM10
+coefficient estimates were unexpected by our criterion, we do not mean
+to imply that they are ``wrong'' in any sense. Indeed, negative
+estimated coefficients have been found in other studies of this nature
+\citep{bell2004ozone}. Rather, in this type of analysis, it may be that
+the interval for expected results needs to be revised. Further, it may
+be necessary to incorporate other outputs from the analysis, such as
+measures of uncertainty.
 
@@
 our approach and is deserving of further consideration. In the PM10
-example, the inverse transform method is used to preserve the
-correlation structure among mortality, PM10, and temperature. However,
-simulation can become complex when additional restrictions are imposed
-or when a greater range of scenarios is desired. In such cases,
-techniques like the acceptance-reject method or permutation may be used
-to generate the data.
+example in Section~\ref{sec-pm10-mortality}, the inverse transform
+method was used to preserve the correlation structure among mortality,
+PM10, and temperature. However, the simulation process can become
+complex when additional restrictions are imposed or when a greater range
+of scenarios is desired. In such cases, techniques like the
+acceptance-reject method or permutation may be used to generate the
+data. The results in Section~\ref{sec-pm10-mortality} on the broader
+NMMAPS dataset suggest that our simulation procedure may have been
+inadequate in reflecting the range of possible configurations that the
+data could take. In particular, our split-data approach, using New York
+City to guide the simulation and then applying the analysis checks to
+other cities, may not have been ideal. It may be worth exploring some
+recent work in data thinning \citep{neufeld2024data} and data fission
+\citep{leiner2023data} to create training datasets that are more
+representative of future data.
 

### 2024-12-20 index.qmd
update application results

@@
 The model suggests the rule (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND (NOT sd(step) > 2,500).
-The NOT operator applied to `sd(step) > 2,500)` is colored with a black background to distinguish it from other checks.
+The NOT operator applied to `sd(step) > 2,500` is colored with a black background to distinguish it from other checks.
 fit <- step_count |> fit_logic_reg(unexpect, `q(step, 0.6) > 10000`:`sd(step) > 2500`, seed = 1, nleaves = 3)
@@
 The harmonic and arithmetic means are included to combine the three measures.
-The results show that the two checks produced by the logic regression can accurately predict 86.1% cases of all *actual unexpected results* in the simulation data.
+The results show that the three checks produced by the logic regression can accurately predict 86.1% cases of all *actual unexpected results* in the simulation data.
 Furthermore, 43.1% of all *predicted unexpected results* were in fact observed to be unexpected.
@@
 #| label: tbl-logic-reg
+#| tbl-cap: "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and the logic regression check rule.
+The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps)."
 list(tibble(.fitted = step_count$`q(step, 0.6) > 10000`, unexpect = step_count$unexpect),
@@
             calc_metrics(metrics = c("harmonic", "arithmetic"))) |> 
-  mutate(checks = c("check 1: q(step, 0.6) > 10000", "check 2: q(step, 0.4) < 8000", "check 3: sd(step) > 2500")) |>
+  mutate(checks = c("Check 1: q(step, 0.6) > 10000", "Check 2: q(step, 0.4) < 8000", "Check 3: sd(step) > 2500")) |>
   bind_rows(augment(fit) |> 
-              calc_miscla_rate(unexpect, .fitted) |> 
+              calc_miscla_rate(.fitted, unexpect) |> 
               calc_independence() |> 
               calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
-              mutate(checks = "logic regression: (check 1 OR check 2) AND check 3") |> 
+              mutate(checks = "Logic regression: (check 1 OR check 2) AND (not check 3)") |> 
               dplyr::select(-overlapping)) |> 
@@
               dplyr::select(-overlapping) |>
-              mutate(checks = "regression tree")) |>
+              mutate(checks = "Regression tree")) |>
   dplyr::select(checks, precision:arithmetic) |> 
   dplyr::rename(`indep.` = independence) |> 
-  kable(digits = 3, 
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and the logic regression check rule.
-The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps).")
+  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE) |> 
+  kableExtra::kable_styling(latex_options="scale_down") |> 
+  column_spec(1, width = '17em')
 ```
@@
 This PM10 coefficient expectation can be framed as an analysis check that fails, labelled as 1, if the estimate of the PM10 coefficient (adjusted for temperature) is outside the range [0, 0.005], and 0 if it is within.
-Multiple factors can affect the estimated PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature.
-Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate.
+Multiple factors can affect the estimated PM10 coefficient, such as the strength of the correlation between mortality and PM10, or between mortality and temperature.
 Outliers in the three variables can also leverage the coefficient.
@@
 
-| The check fails if ... | 
+| The check fails (encoded as 1) if ... | 
 |--------------| 
-| Sample size less than or equal to 200 |
-| Sample size less than or equal to 400 |
-| Sample size less than or equal to 600 |
-| Sample size less than or equal to 800 |
+| Mortality-PM10 correlation less than $-0.05$ |
 | Mortality-PM10 correlation less than $-0.03$ |
-| Mortality-PM10 correlation less than $-0.04$ |
-| Mortality-PM10 correlation less than $-0.05$ |
-| Mortality-PM10 correlation less than $-0.06$ |
+| Mortality-PM10 correlation greater than $0.03$ |
+| Mortality-PM10 correlation greater than $0.05$ |
 | Mortality-temperature correlation greater than $-0.3$ | 
@@
 # cor(pm10$temp, pm10$pm10)
-corr_grid <- expand.grid(seq(0.01, 0.2, by = 0.02),                         
-                         seq(0.25, 0.5, by = 0.05),                           
+corr_grid <- expand.grid(#seq(0.01, 0.2, by = 0.02),
+                         seq(-0.1, 0.1, by = 0.02),
+                         seq(0.1, 0.5, by = 0.05),
                          seq(0.25, 0.45, by = 0.05))
@@
   # correlation between mortality and temp are negative - r2
-  cor_matrix <- matrix(c(1, -r1, -r2,                          
-                         -r1, 1, r3,                          
+  cor_matrix <- matrix(c(1, r1, -r2,
+                         r1, 1, r3,
                          -r2, r3, 1), nrow = 3, byrow = TRUE)
@@
 corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]
-sample_size <- c(200, 400, 600, 800, 1000)
+sample_size <- c(500, 3000)
 outlier <- c(TRUE, FALSE)
@@
     unexpect = ifelse(between(coef, 0, 0.005), 0, 1),
-    `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
-    `smpl_size <= 400` = ifelse(sample_size <= 400, 1, 0),
-    `smpl_size <= 600` = ifelse(sample_size <= 600, 1, 0),
-    `smpl_size <= 800` = ifelse(sample_size <= 800, 1, 0),
+    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
     `cor(m, PM10) < -0.03` = ifelse(mor_pm10_correlation < -0.03, 1, 0),
-    `cor(m, PM10) < -0.04` = ifelse(mor_pm10_correlation < -0.04, 1, 0),
-    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
-    `cor(m, PM10) < -0.06` = ifelse(mor_pm10_correlation < -0.06, 1, 0),
+    `cor(m, PM10) > 0.03` = ifelse(mor_pm10_correlation > 0.03, 1, 0),
+    `cor(m, PM10) > 0.05` = ifelse(mor_pm10_correlation > 0.05, 1, 0),
     `cor(m, tmp) > -0.3` = ifelse(mor_temp_correlation > -0.3, 1, 0),
@@
   rowwise() |>
-  mutate(`PM10 outlier` = ifelse(any(scale(data$pm10) > 4), 1, 0),
-         `mortality_outlier` = ifelse(any(scale(data$mortality) > 4), 1, 0)
+  mutate(PM10_outlier = ifelse(any(scale(data$pm10) > 4), 1, 0),
+         mortality_outlier = ifelse(any(scale(data$mortality) > 4), 1, 0)
   ) |>
   ungroup()
+
+unexpect_df <- tibble(
+  unexpect = c(0, 1),
+  n = pm10_df$unexpect |> table() |> as.numeric() |> rev(),
+  wgt = n/sum(n))
+wgt_vec <- pm10_df |> dplyr::select(unexpect) |> left_join(unexpect_df) |> pull(wgt)
 ```
@@
 A logic regression is fitted using all variations of the checks in @tbl-checks to predict whether the PM10 coefficient is unexpected.
+Given the imbalance of outcome (expected vs.
+unexpected), inverse weights proportional to the number of observations in each outcome are applied during the logic regression fit.
 @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model.
@@
 ```{r}
-#| fig.width: 9
+#| fig.width: 12
 #| fig.height: 4
-fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
-                                seed = 6, nleaves = 4)
+fit <- pm10_df |> fit_logic_reg(unexpect, `cor(m, PM10) < -0.05`:mortality_outlier,
+                                seed = 7, nleaves = 4, wgt = wgt_vec)
 plot_logicreg(fit)
@@
 Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable.
-The model suggests the relationship: (NOT sample size less than or equal to 200 AND mortality-PM10 correlation < -0.04) AND (mortality-temperature correlation > -0.35 OR there exist mortality outlier) to predict the unexpected PM10 coefficient.
-The NOT operator applied to `smpl_size <= 200` is colored with a black background to distinguish it from other checks.
+The model suggests the relationship:  (mortality-PM10 correlation < -0.03) AND (mortality-temperature correlation > -0.3 OR (there exist mortality outlier AND there exist PM10 outlier)) to predict the unexpected PM10 coefficient.
 :::
@@
 #| label: tbl-linear-reg
-library(kableExtra)
+#| tbl-cap: "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model (last row) and individual checks, along with harmonic and arithmetic means."
 df <- list(
-  tibble(.fitted = pm10_df$`smpl_size <= 200`, 
-         unexpect = pm10_df$unexpect),
   tibble(.fitted = pm10_df$`cor(m, PM10) < -0.03`, 
          unexpect = pm10_df$unexpect),
-  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.35`, 
+  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.3`, 
          unexpect = pm10_df$unexpect),
@@
          unexpect = pm10_df$unexpect),
+  tibble(.fitted = pm10_df$PM10_outlier, 
+         unexpect = pm10_df$unexpect),
   augment(fit)) |>
-  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
+  map_dfr(~.x |> calc_miscla_rate(.fitted, unexpect) |> 
             mutate(overlapping = 0, independence = 1) |> 
@@
 
-tbl <- tibble(check = c("check 1: sample size <= 200", "check 2: cor(m, PM10) < -0.03",
-                        "check 3: cor(m, tmp) > -0.35", "check 4: mortality_outlier", 
-                        "logic regression: (not check 1 AND check 2) AND (check 3 OR check 4)")) |> 
+tbl <- tibble(check = c("Check 1: cor(m, PM10) < -0.03", "Check 2: cor(m, tmp) > -0.35",
+                        "Check 3: mortality outlier", "Check 4: PM10 outlier", 
+                        "Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4))")) |> 
   bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)
@@
 tbl |> 
-  kable(digits = 3, escape = FALSE,
-        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual checks (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation,  4: mortality outlier), and the combined check rule 5: (sample size AND mortality-PM10 correlation) AND (mortality-temperature correlation OR mortality outlier.")
+  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE) |> 
+  kableExtra::kable_styling(latex_options="scale_down") |> 
+  column_spec(1, width = '18em')
 
@@
 
--   sample size *larger than* 200
--   mortality-temperature correlation greater than $-0.35$
--   mortality-PM10 correlation less than $-0.04$
+-   mortality-PM10 correlation less than $-0.03$
+-   mortality-temperature correlation greater than $-0.3$
+-   PM10 data contains outliers that are detected by the univariate outlier detection
 -   mortality data contains outliers that are detected by the univariate outlier detection
 
-<!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->
 The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature.
 This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient.
-The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.
 

### 2024-12-23 _manuscript/_tex/index.tex
proofread Section 1-2

@@
 In data analysis, unexpected results often prompt researchers to revisit
-their procedures to identify potential issues. While experienced
-researchers can often quickly diagnose problems by checking a few key
-assumptions, others may struggle to identify the root causes. These
+their procedures to identify potential issues. While some researchers
+may struggle to identify the root causes, experienced researchers can
+often quickly diagnose problems by checking a few key assumptions. These
 checked assumptions, or expectations, are typically informal, difficult
 to trace, and rarely discussed in publications. In this paper, we
-formalize these informal assumptions by framing them as binary
-\emph{analysis validation checks}. We then introduce a procedure to
-quantify how violations of these checks may lead to unexpected results
-in the analysis. The procedure relies on simulations of the original
-data and evaluates both accuracy and redundancy. Accuracy is calculated
-through a binary classification metric, while redundancy is measured
-using mutual information. We demonstrate this approach with a toy
-example based on fitness step count data and a generalized linear model
-example examining the effect of particulate matter air pollution on
-daily mortality.
+introduce the term \emph{analysis validation checks} to formalize and
+externalize these informal assumptions. We then introduce a procedure to
+identify a subset of checks that best predict the occurrence of
+unexpected outcomes, based on simulations of the original data. The
+checks are evaluated in terms of accuracy, determined by binary
+classification metrics, and redundancy, which measures the shared
+information among checks using mutual information. We demonstrate this
+approach with a toy example using step count data and a generalized
+linear model example examining the effect of particulate matter air
+pollution on daily mortality.
 \end{abstract}
@@
 
-In data analysis, experienced researchers often rely on their prior
-knowledge or domain expertise to quickly assess whether results align
-with their expectations. When a result falls outside of this
-expectation, it prompts the researchers to investigate backwards on the
-data quality, the analysis steps, or the assumptions made during the
-analysis process. This mental process of where to diagnose unexpected
-outcomes is often difficult to trace and document in publications. As a
-result, readers are typically presented with the final outcomes of the
-analysis cycle where the results and expectations are aligned, achieved
-either by refining the analysis or updating the expectations based on
-statistical evidence \citep{grolemund_cognitive_2014}. These missing
-pieces of information provide little guidance for diagnosing issues in
-the analysis when the same methodology is applied to a new dataset that
-produces different outcomes. Similarly, when researchers with different
-background knowledge view results that they find to be unexpected, it
-becomes unclear whether discrepancies arise from differing expectations
-or from the use of statistical techniques.
-
-One might gain insight into analysts' thought processes by speaking with
+In data analysis, researchers often rely on their prior knowledge or
+domain expertise to form expectations and assess whether results align
+with them. When results deviate from these expectations, experienced
+researchers check assumptions made about the data and during the
+analysis, but these assumptions checked are often difficult to trace and
+are seldom documented in publications. While publication of analysis
+code and data is now a common a requirement for the sake of
+reproducibility \citep{peng2011reproducible}, the published code alone
+is often insufficient for understanding the thought process behind an
+analysis. Code corresponding to published analyses often reflect the
+final decisions made about analysis and do not reveal the decisions or
+assumptions made about the data during the analysis. Thus, a reader
+looking at published code can often be left with many questions about
+why certain choices were made.
+
+We might gain insight into analysts' thought processes by speaking with
 them directly or watching them work via screencast videos they produce,
@@
 not scalable and may not always be feasible; creating educational
-screencast videos requires significant effort from the researchers.
-Ideally, there could be a way to make expectations about data analysis
-explicit and accessible to others. Even better, if the encoding were
-machine-readable, we could analyze these expectations and learn from the
-analysis itself. For example, we could answer questions about whether
-the checks also apply to other researchers analyzing new data in the
-same context, whether they reflect common practices in the field, or
-whether they are specific to the data or analysis at hand.
-
-The externalization of the data analysis process is a practice that has
-potential to improve the trustworthiness of analyses in general
-\citep{yu2024veridical} and the trustworthiness of subsequent products,
-such as machine learning models, that may be built on such analyses.
-While publication of analysis code and data is now a common a
-requirement for the sake of reproducibility
-\citep{peng2011reproducible}, the published code alone is often
-insufficient for understanding the thought process behind an analysis.
-Code corresponding to published analyses often reflect the final
-decisions made about analysis and do not reveal the decisions or
-assumptions made about the data processing. Thus, a reader looking at
-published code can often be left with many questions about why certain
-choices were made. Developing an approach to reveal some of this process
-without requiring a reader to essentially reconstruct the analysis
-process from the raw data would provide an improved basis for trusting
-an analysis result \citep{peng2021reproducible}.
-
-In this paper, we conceptualize these internal expectations and
-assumptions as \emph{analysis validation checks}, which allows us to
-examine the assumptions made during an analysis and to diagnose
-unexpected outcomes. We then introduce a procedure that provides a
-quantitative measure of how violations in a tree of analysis checks,
-derived from individual checks, will lead to an unexpected result. The
-procedure, based on simulations of the original data, calculates the
-accuracy and redundancy of the analysis checks. Accuracy is determined
-using binary classification metrics, precision and recall, from a logic
-regression fit \citep{ruczinski_logic_2003}, while redundancy is
+screencast videos requires significant effort from the analysts.
+Ideally, there could be a way to explicitly communciate these throught
+processes in the data analysis to others. Even better, if these
+expectations were machine-readable, we could analyze them and learn from
+the analysis itself. For example, we could answer questions about
+whether the checks also apply to other researchers analyzing new data in
+the same context, whether they reflect common practices in the field, or
+whether they are specific to the data or analysis at hand. Externalizing
+these thought processes is a practice that has potential to improve the
+trustworthiness of the analyses \citep{yu2024veridical} in general and
+the trustworthiness of subsequent products, such as machine learning
+models, that may be built on such analyses. Developing an approach to
+reveal some of this process without requiring a reader to essentially
+reconstruct the analysis process from the raw data would provide an
+improved basis for trusting an analysis result
+\citep{peng2021reproducible}.
+
+In this paper, we conceptualize these inexplicit expectations as
+\emph{analysis validation checks}, which allows us to examine the
+assumptions made about the data during an analysis. We then introduce a
+procedure to identify the subset of checks that best predict the
+occurrence of unexpected outcomes. The procedure, based on simulations
+of the original data, compares the accuracy and redundancy of different
+combinations of analysis validation checks. Accuracy is determined using
+binary classification metrics (precision and recall) from a logic
+regression model \citep{ruczinski_logic_2003}, while redundancy is
 measured using mutual information. The proposed workflow offers a
@@
 Section~\ref{sec-plan} introduces the concept of analysis validation
-checks, illustrated with a toy example based on fitness step count data.
-Section~\ref{sec-method} describes the procedure that quantifies how
-analysis validation checks combined using logical operators can predict
-unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
-applies this procedure to a larger example that estimates the effect of
-particulate matter air pollution on daily mortality.
-Section~\ref{sec-discussion} summarises teh paper and discusses a few
-key considerations.
+checks, illustrated with a toy example based on step count data.
+Section~\ref{sec-method} describes the procedure that selects the best
+subset of checks for understanding the occurrence of unexpected
+outcomes. Section~\ref{sec-pm10-mortality} applies this procedure to a
+larger example that estimates the effect of particulate matter air
+pollution on daily mortality. Section~\ref{sec-discussion} summarises
+the paper and discusses a few key considerations.
 
@@
 
-A substantial body of literature has addressed the definition of data
-quality and has developed frameworks that include dimensions,
-attributes, and measures to evaluate and improve data quality
-\citep{8642813, batini2009methodologies, cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
-These frameworks are often used in information systems and database
-management and support business decision-making in various industries.
-For example, such data validation systems are applied in the Google
-Tensorflow framework \citep{polyzotis2019data} and Amazon
-\citep{schelter2018automating}. For research purposes, high-quality data
-ensures the credibility of scientific findings and supports
-reproducibility and re-usability in future studies. With the growing
-prevalence of open data in scientific research, the consumers or users
-of the data typically are no longer the producers or collectors of the
-data who would have the full knowledge of data in hand, prompting more
-interest towards data quality checks in the data analysis process. In R,
-there are some packages, like \texttt{skimr} \citep{skimr} and
-\texttt{dataMaid} \citep{dataMaid}, that provide basic data screening
-and reporting tools, while another class of packages,
-e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
-\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
-providing data validation tools, allowing users to define customized
-data quality checks based on the applications.
+The concept of data quality has been studied in the literature, with
+earlier work focusing on defining frameworks -- such as dimensions,
+attributes, and measures -- to improve data quality in database and
+information systems
+\citep{wang1996beyond, batini2009methodologies, 6204995, woodall2014classification, cai2015challenges, 8642813}.
+More recently, data validation has been incorporated into frameworks
+like Google TensorFlow \citep{polyzotis2019data} to ensure the quality
+of data for training machine learning models, as well as for supporting
+business decision-making \citep{schelter2018automating}. With the
+increasing availability of open data in scientific research, the users
+of the data are often not the original data collector and may not be
+aware of all the details or nuances of the data. This encourages
+researchers to conduct data quality checks before beginning the analytic
+process, helping to avoid unexpected discoveries later on. In R,
+packages like \texttt{skimr} \citep{skimr} and \texttt{dataMaid}
+\citep{dataMaid} provide basic data screening and data quality reports,
+while packages such as \texttt{assertr} \citep{assertr},
+\texttt{validate} \citep{validate}, and \texttt{pointblank}
+\citep{pointblank} focuse on providing infustructures that allow users
+to define customized data quality checks.
 

### 2024-12-23 index.qmd
proofread Section 1-2

@@
 abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues.
-While experienced researchers can often quickly diagnose problems by checking a few key assumptions, others may struggle to identify the root causes.
+While some researchers may struggle to identify the root causes, experienced researchers can often quickly diagnose problems by checking a few key assumptions.
 These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications.
-In this paper, we formalize these informal assumptions by framing them as binary *analysis validation checks*.
-We then introduce a procedure to quantify how violations of these checks may lead to unexpected results in the analysis.
-The procedure relies on simulations of the original data and evaluates both accuracy and redundancy.
-Accuracy is calculated through a binary classification metric, while redundancy is measured using mutual information.
-We demonstrate this approach with a toy example based on fitness step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality.
+In this paper, we introduce the term *analysis validation checks* to formalize and externalize these informal assumptions.
+We then introduce a procedure to identify a subset of checks that best predict the occurrence of unexpected outcomes, based on simulations of the original data.
+The checks are evaluated in terms of accuracy, determined by binary classification metrics, and redundancy, which measures the shared information among checks using mutual information.
+We demonstrate this approach with a toy example using step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality.
 "
@@
 
-In data analysis, experienced researchers often rely on their prior knowledge or domain expertise to quickly assess whether results align with their expectations.
-When a result falls outside of this expectation, it prompts the researchers to investigate backwards on the data quality, the analysis steps, or the assumptions made during the analysis process.
-This mental process of where to diagnose unexpected outcomes is often difficult to trace and document in publications.
-As a result, readers are typically presented with the final outcomes of the analysis cycle where the results and expectations are aligned, achieved either by refining the analysis or updating the expectations based on statistical evidence [@grolemund_cognitive_2014].
-These missing pieces of information provide little guidance for diagnosing issues in the analysis when the same methodology is applied to a new dataset that produces different outcomes.
-Similarly, when researchers with different background knowledge view results that they find to be unexpected, it becomes unclear whether discrepancies arise from differing expectations or from the use of statistical techniques.
+In data analysis, researchers often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them.
+When results deviate from these expectations, experienced researchers check assumptions made about the data and during the analysis, but these assumptions checked are often difficult to trace and are seldom documented in publications.
+While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis.
+Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data during the analysis.
+Thus, a reader looking at published code can often be left with many questions about why certain choices were made.
 
-One might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g.
+We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g.
 @gu2024data].
-However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the researchers.
-Ideally, there could be a way to make expectations about data analysis explicit and accessible to others.
-Even better, if the encoding were machine-readable, we could analyze these expectations and learn from the analysis itself.
+However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts.
+Ideally, there could be a way to explicitly communciate these throught processes in the data analysis to others.
+Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself.
 For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
-
-The externalization of the data analysis process is a practice that has potential to improve the trustworthiness of analyses in general [@yu2024veridical] and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
-While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis.
-Code corresponding to published analyses often reflect the final decisions made about analysis and do not reveal the decisions or assumptions made about the data processing.
-Thus, a reader looking at published code can often be left with many questions about why certain choices were made.
+Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
 Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].
 
-In this paper, we conceptualize these internal expectations and assumptions as *analysis validation checks*, which allows us to examine the assumptions made during an analysis and to diagnose unexpected outcomes.
-We then introduce a procedure that provides a quantitative measure of how violations in a tree of analysis checks, derived from individual checks, will lead to an unexpected result.
-The procedure, based on simulations of the original data, calculates the accuracy and redundancy of the analysis checks.
-Accuracy is determined using binary classification metrics, precision and recall, from a logic regression fit [@ruczinski_logic_2003], while redundancy is measured using mutual information.
+In this paper, we conceptualize these inexplicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis.
+We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes.
+The procedure, based on simulations of the original data, compares the accuracy and redundancy of different combinations of analysis validation checks.
+Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while redundancy is measured using mutual information.
 The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.
@@
 The rest of the paper is organized as follows: @sec-lit-review reviews the concepts of diagnosing unexpected outcomes and general data quality checks.
-@sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on fitness step count data.
-@sec-method describes the procedure that quantifies how analysis validation checks combined using logical operators can predict unexpected outcomes in an analysis.
+@sec-plan introduces the concept of analysis validation checks, illustrated with a toy example based on step count data.
+@sec-method describes the procedure that selects the best subset of checks for understanding the occurrence of unexpected outcomes.
 @sec-pm10-mortality applies this procedure to a larger example that estimates the effect of particulate matter air pollution on daily mortality.
-@sec-discussion summarises teh paper and discusses a few key considerations.
+@sec-discussion summarises the paper and discusses a few key considerations.
 
@@
 
-A substantial body of literature has addressed the definition of data quality and has developed frameworks that include dimensions, attributes, and measures to evaluate and improve data quality [@8642813; @batini2009methodologies; @cai2015challenges; @wang1996beyond; @6204995; @woodall2014classification].
-These frameworks are often used in  information systems and database management and support business decision-making in various industries.
-For example, such data validation systems are applied in the Google Tensorflow framework [@polyzotis2019data] and Amazon [@schelter2018automating].
-For research purposes, high-quality data ensures the credibility of scientific findings and supports reproducibility and re-usability in future studies.
-With the growing prevalence of open data in scientific research, the consumers or users of the data typically are no longer the producers or collectors of the data who would have the full knowledge of data in hand, prompting more interest towards data quality checks in the data analysis process.
-In R, there are some packages, like `skimr` [@skimr] and `dataMaid` [@dataMaid], that provide basic data screening and reporting tools, while another class of packages, e.g.
-`assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuses on providing data validation tools, allowing users to define customized data quality checks based on the applications.
+The concept of data quality has been studied in the literature, with earlier work focusing on defining frameworks -- such as dimensions, attributes, and measures -- to improve data quality in database and information systems [@wang1996beyond; @batini2009methodologies; @6204995; @woodall2014classification; @cai2015challenges; @8642813].
+More recently, data validation has been incorporated into frameworks like Google TensorFlow [@polyzotis2019data] to ensure the quality of data for training machine learning models, as well as for supporting business decision-making [@schelter2018automating].
+With the increasing availability of open data in scientific research, the users of the data are often not the original data collector and may not be aware of all the details or nuances of the data.
+This encourages researchers to conduct data quality checks before beginning the analytic process, helping to avoid unexpected discoveries later on.
+In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuse on providing infustructures that allow users to define customized data quality checks.
 

### 2024-12-24 _manuscript/_tex/index.tex
section 3

@@
 
-In data analysis, researchers often rely on their prior knowledge or
-domain expertise to form expectations and assess whether results align
-with them. When results deviate from these expectations, experienced
-researchers check assumptions made about the data and during the
-analysis, but these assumptions checked are often difficult to trace and
-are seldom documented in publications. While publication of analysis
-code and data is now a common a requirement for the sake of
-reproducibility \citep{peng2011reproducible}, the published code alone
-is often insufficient for understanding the thought process behind an
-analysis. Code corresponding to published analyses often reflect the
-final decisions made about analysis and do not reveal the decisions or
+In data analysis, analysts often rely on their prior knowledge or domain
+expertise to form expectations and assess whether results align with
+them. When results deviate from these expectations, experienced analysts
+check assumptions made about the data and during the analysis, but often
+do so without discussing the underlying reasoning in the publications.
+While publication of analysis code and data is now a common a
+requirement for the sake of reproducibility
+\citep{peng2011reproducible}, the published code alone is often
+insufficient for understanding the thought process behind an analysis.
+Code corresponding to published analyses often reflect the final
+decisions made about analysis and do not reveal the decisions or
 assumptions made about the data during the analysis. Thus, a reader
@@
 
-Expectations represent our understanding of certain aspects of the
-analysis and the data, independent of the results of the analysis
-itself. When observed outcomes deviate from these expectations, analysts
-often revisit the analysis process to identify potential issues, refine
-methods, or revise assumptions. Experienced analysts can typically
-identify issues quickly and correct them on the spot, but they often do
-so without discussing the underlying reasoning, limiting transparency
-and making it harder for less experienced researchers to learn and
-master these skills.
-
-Here, we introduce the concept of \textbf{analysis validation checks},
-which frame these expectations or assumptions as explicit checks that
-return a TRUE or FALSE result given the data analyzed at hand. Inspired
-by the concept of data validation checks \citep{validate}, which are
-designed to ensure that datasets meet expected formats and quality,
-analysis validation checks reverse the approach: they validate the
-assumptions about the data necessary for the analysis to produce the
-\emph{expected results}, as defined by the analyst. The focus on
-expected results allows the concept of analysis validation checks to
-encompass a broad range of checks, such as data quality (i.e.~missing
-data, how the data are structured), data distribution and outliers,
-bivariate and multivariate relationships between variables, and other
-contextual information.
+Analysis validation checks are assumptions made by the analysts during
+the analysis, framed as explicit checks that return a binary TRUE or
+FALSE value based on the data. Inspired by the concept of data
+validation checks \citep{validate}, which are designed to ensure that
+datasets meet expected quality before the analysis begins, analysis
+validation checks reverse the approach: they validate the assumptions
+about the data necessary for the analysis to produce the \emph{expected
+results}, as defined by the analyst. The focus on expected results
+allows analysis validation checks to encompass a wide range of checks,
+such as data quality, variable distributions and outliers, bivariate and
+multivariate relationships, and contextual information relevant to the
+analysis.
 
@@
 In addition to the above benefits, the development and publication of
-analysis checks has the potential to help students, inexperienced
+analysis checks have the potential to help students, inexperienced
 analysts, and junior researchers develop the skills needed to diagnose
@@
 Subjects are instructed to walk at least 8,000 steps each day, with an
-expected average of 9,000 steps, tracked by a step counter app. After 30
-days, we review the data and examine the number of steps taken each day.
-With data of this nature, we may expect there to be occasional ``low''
-days due to factors such as forgetting to wear their watch or
-unfavorable weather conditions limiting outdoor activities. We may also
-expect ``high'' days recorded after an outdoor hike or intense workout.
-Given the requirements of the study, we expect that for a given subject
-the average step count would be between \([8500, 9500]\).
-
-To diagnose potential reasons why this outcome expectation might fail,
-we can establish a few analysis validation checks in anticipation of
-seeing the data. For example, we can check the quantile of the step
-count, if more than a third of the days fall below 8,000, or more than a
-third exceed 10,000 steps, this could indicate an excess of low-count or
-high-count days. Similarly, we may may expect the standard deviation of
-the step count not to be overly large. These considerations yield the
-following three analysis validation checks that \emph{fail} when:
+expected average of 9,000 steps, tracked by a step counter app. With
+data of this nature, we may expect there to be occasional ``low'' days
+due to factors such as unfavorable weather conditions limiting outdoor
+activities. We may also expect ``high'' days recorded after an outdoor
+activities or intense workout. Given the requirements of the study, we
+form our expectation as follows:
+
+\begin{quote}
+Expectation: The average step count of a given subject is between
+\([8,500, 9,500]\)
+\end{quote}
+
+To diagnose potential reasons why this expectation might fail, we can
+establish a few analysis validation checks in anticipation of seeing the
+data. For example, we can check the quantile of the step count, if more
+than a third of the days fall below 8,000, or more than a third exceed
+10,000 steps, this could indicate an excess of low-count or high-count
+days. Similarly, we may may expect the standard deviation of the step
+count not to be overly large. These considerations yield the following
+three analysis validation checks that \emph{fail} when:
 
@@
 \item
-  check 1: the 60\% quantile of the observed step counts is greater than
+  Check 1: the 60\% quantile of the observed step counts is greater than
   10,000
 \item
-  check 2: the 40\% quantile of the observed step counts is less than
+  Check 2: the 40\% quantile of the observed step counts is less than
   8,000, and
 \item
-  check 3: the standard deviation of the observed step counts exceeds
+  Check 3: the standard deviation of the observed step counts exceeds
   2,500.
@@
 
-The cutoff values chosen for these checks would presumably be chosen
-based on prior experience with these kinds of data, but could also be
-optimized using the method presented in the next section.
-
-\phantomsection\label{cell-fig-step-count}
-\begin{figure}[H]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-step-count-1.pdf}}
-
-}
-
-\caption{\label{fig-step-count}Beeswarm plot of average step counts
-across 300 simulated 30-day periods. Each point represents the average
-step count from one simulation. While similar to a boxplot or violin
-plot, the beeswarm plot also displays the distribution of each
-individual data point. The orange points indicate instances where the
-average step count fails outside the {[}8,500, 9,500{]} interval,
-representing an unexpected outcome in this scenario.}
-
-\end{figure}%
+The cutoff values for these checks would presumably be chosen based on
+prior experience with these kinds of data, but could also be optimized
+using the method presented in the next section.
 
 To simulate this data, three normal distributions are used for the daily
-step counts: \(\mathcal{N}(6000, 200)\) for low days,
-\(\mathcal{N}(12000, 200)\) for high days, and
-\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
+step counts: \(\mathcal{N}(6,000, 200)\) for low days,
+\(\mathcal{N}(12,000, 200)\) for high days, and
+\(\mathcal{N}(9,000, 300)\) for typical days. The number of low and high
 days can be simulated from a Poisson distribution with \(\lambda = 8\).
@@
 simulated 30-day periods.
+
+\phantomsection\label{cell-fig-step-count}
+\begin{figure}[H]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/fig-step-count-1.pdf}}
+
+}
+
+\caption{\label{fig-step-count}Beeswarm plot of average step counts
+across 300 simulated 30-day periods. Each point represents the average
+step count from one simulation. Similar to a boxplot or violin plot, the
+beeswarm plot also displays the distribution of each individual data
+point. The orange points indicate instances where the average step count
+fails outside the {[}8,500, 9,500{]} interval, representing an
+unexpected outcome in this scenario.}
+
+\end{figure}%
 

### 2024-12-24 index.qmd
section 3

@@
 
-In data analysis, researchers often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them.
-When results deviate from these expectations, experienced researchers check assumptions made about the data and during the analysis, but these assumptions checked are often difficult to trace and are seldom documented in publications.
+In data analysis, analysts often rely on their prior knowledge or domain expertise to form expectations and assess whether results align with them.
+When results deviate from these expectations, experienced analysts check assumptions made about the data and during the analysis, but often do so without discussing the underlying reasoning in the publications.
 While publication of analysis code and data is now a common a requirement for the sake of reproducibility [@peng2011reproducible], the published code alone is often insufficient for understanding the thought process behind an analysis.
@@
 
-Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself.
+<!-- Expectations represent our understanding of certain aspects of the analysis and the data, independent of the results of the analysis itself.
 When observed outcomes deviate from these expectations, analysts often revisit the analysis process to identify potential issues, refine methods, or revise assumptions.
 Experienced analysts can typically identify issues quickly and correct them on the spot, but they often do so without discussing the underlying reasoning, limiting transparency and making it harder for less experienced researchers to learn and master these skills.
+-->
 
-Here, we introduce the concept of **analysis validation checks**, which frame these expectations or assumptions as explicit checks that return a TRUE or FALSE result given the data analyzed at hand.
-Inspired by the concept of data validation checks [@validate], which are designed to ensure that datasets meet expected formats and quality, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst.
-The focus on expected results allows the concept of analysis validation checks to encompass a broad range of checks, such as data quality (i.e.
-missing data, how the data are structured), data distribution and outliers, bivariate and multivariate relationships between variables, and other contextual information.
+Analysis validation checks are assumptions made by the analysts during the analysis, framed as explicit checks that return a binary TRUE or FALSE value based on the data.
+Inspired by the concept of data validation checks [@validate], which are designed to ensure that datasets meet expected quality before the analysis begins, analysis validation checks reverse the approach: they validate the assumptions about the data necessary for the analysis to produce the *expected results*, as defined by the analyst.
+The focus on expected results allows analysis validation checks to encompass a wide range of checks, such as data quality, variable distributions and outliers, bivariate and multivariate relationships, and contextual information relevant to the analysis.
 
@@
 
-In addition to the above benefits, the development and publication of analysis checks has the potential to help students, inexperienced analysts, and junior researchers develop the skills needed to diagnose unexpected analysis results for a given type of data because the assumptions made about the data are made transparent.
+In addition to the above benefits, the development and publication of analysis checks have the potential to help students, inexperienced analysts, and junior researchers develop the skills needed to diagnose unexpected analysis results for a given type of data because the assumptions made about the data are made transparent.
 The analysis checks can serve as a basis for new analysts to have conversations about the data they are analyzing and to develop a better understanding of the potential data generation process.
@@
 Subjects are instructed to walk at least 8,000 steps each day, with an expected average of 9,000 steps, tracked by a step counter app.
-After 30 days, we review the data and examine the number of steps taken each day.
-With data of this nature, we may expect there to be occasional "low" days due to factors such as forgetting to wear their watch or unfavorable weather conditions limiting outdoor activities.
-We may also expect "high" days recorded after an outdoor hike or intense workout.
-Given the requirements of the study, we expect that for a given subject
-the average step count would be between $[8500, 9500]$.
+With data of this nature, we may expect there to be occasional "low" days due to factors such as unfavorable weather conditions limiting outdoor activities.
+We may also expect "high" days recorded after an outdoor activities or intense workout.
+Given the requirements of the study, we form our expectation as follows:
 
-To diagnose potential reasons why this outcome expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
+  > Expectation: The average step count of a given subject is between $[8,500, 9,500]$
+
+To diagnose potential reasons why this expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
 For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
@@
 
--   check 1: the 60% quantile of the observed step counts is greater than 10,000
--   check 2: the 40% quantile of the observed step counts is less than 8,000, and
--   check 3: the standard deviation of the observed step counts exceeds 2,500.
+-   Check 1: the 60% quantile of the observed step counts is greater than 10,000
+-   Check 2: the 40% quantile of the observed step counts is less than 8,000, and
+-   Check 3: the standard deviation of the observed step counts exceeds 2,500.
 
-The cutoff values chosen for these checks would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.
+The cutoff values for these checks would presumably be chosen based on prior experience with these kinds of data, but could also be optimized using the method presented in the next section.
+
+To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6,000, 200)$ for low days, $\mathcal{N}(12,000, 200)$ for high days, and $\mathcal{N}(9,000, 300)$ for typical days.
+The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$.
+@fig-step-count displays average step count across 300 simulated 30-day periods.
 
@@
 Each point represents the average step count from one simulation.
-While similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point.
+Similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point.
 The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario.
@@
 ```
-
-To simulate this data, three normal distributions are used for the daily step counts: $\mathcal{N}(6000, 200)$ for low days, $\mathcal{N}(12000, 200)$ for high days, and $\mathcal{N}(9000, 300)$ for typical days.
-The number of low and high days can be simulated from a Poisson distribution with $\lambda = 8$.
-@fig-step-count displays average step count across 300 simulated 30-day periods.
 

### 2024-12-24 _manuscript/_tex/index.tex
section 4

@@
 checks are evaluated in terms of accuracy, determined by binary
-classification metrics, and redundancy, which measures the shared
-information among checks using mutual information. We demonstrate this
-approach with a toy example using step count data and a generalized
-linear model example examining the effect of particulate matter air
-pollution on daily mortality.
+classification metrics, and independence, which measures the shared
+information among checks. We demonstrate this approach with a toy
+example using step count data and a generalized linear model example
+examining the effect of particulate matter air pollution on daily
+mortality.
 \end{abstract}
@@
 occurrence of unexpected outcomes. The procedure, based on simulations
-of the original data, compares the accuracy and redundancy of different
-combinations of analysis validation checks. Accuracy is determined using
-binary classification metrics (precision and recall) from a logic
-regression model \citep{ruczinski_logic_2003}, while redundancy is
-measured using mutual information. The proposed workflow offers a
-numerical guarantee that the analysis will produce the expected results,
-assuming the assumptions about the data generating mechanism hold.
+of the original data, compares the accuracy and independence of
+different combinations of analysis validation checks. Accuracy is
+determined using binary classification metrics (precision and recall)
+from a logic regression model \citep{ruczinski_logic_2003}, while
+independence is measures the shared information among checks. The
+proposed workflow offers a numerical guarantee that the analysis will
+produce the expected results, assuming the assumptions about the data
+generating mechanism hold.
 
@@
 
-While some checks may be crucial and directly indicate an unexpected
-outcome, others may be tangential to the problem at hand and not
-indicate a root cause of an unexpected outcome. In this section, we
-propose a procedure to measure the effectiveness of checks that, when
-combined using logical operators, contribute to an unexpected outcome. A
-small set of independent checks is considered effective if it translates
-to unexpected outcomes.
-
-The approach relies on the use of simulated datasets that are generated
-based on the analyst's knowledge and assumptions about the data
-generation mechanism. Datasets are simulated to have the same structure
-and characteristics of the observed data that will be analyzed at some
-point in the future. For each simulated dataset, we can apply a
-collection of analysis validation checks to the dataset and record which
-ones were TRUE and which were FALSE. We can also compute the outcome of
-the analysis to see whether the outcome is unexpected in a given
-simulated datsets. We can then generate many datasets, each time
-applying the analysis validation checks and computing the outcome. After
-generating many datasets, we can relate the patterns in the analysis
-validation checks to the likely that the outcome will be unexpected in a
-given dataset. Figure~\ref{fig-metric-calc} provides an overview of the
-process.
+While not all analysis validation checks are equally important, some may
+be more useful at diagnosing unexpected outcomes than others. This
+section introduces a procedure to identify a subset of analysis
+validation checks that best predict the occurence of unexpected
+outcomes.
+
+The approach uses simulated datasets that have the same structure and
+characteristics of the observed data to be analyzed in the future. For
+each simulated dataset, analysis validation checks are applied to record
+their TRUE or FALSE values. The outcome of the analysis is also computed
+to determine whether it is unexpected (1) or expected (0). By generating
+many datasets, each time applying the analysis validation checks and
+computing the outcome, we can relate the patterns in the checks to the
+likely that the outcome will be unexpected. Figure~\ref{fig-metric-calc}
+provides an overview of the process.
 
@@
 
+\subsection{Accuracy}\label{accuracy}
+
 From the simulated data, the accuracy branch in
-Figure~\ref{fig-metric-calc} refers to a set of checks' ability to
-accurately detect unexpected outcomes while minimizing false positives
-and false negatives. While a false positive can raise caution or
-skepticism about the data, the presence of both false positives and
-false negatives suggests that the checks are overly sensitive or lack
-sensitivity to unexpected outcomes, respectively.
-
-To incorporate the effect of multiple checks on the outcome, a logic
-regression model \citep{ruczinski_logic_2003} is fitted to the analysis
-validation checks. Originally developed for SNP microarray data, logic
-regression constructs Boolean combinations of binary variables,
-\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict both binary and
-numerical outcomes. In our case,
-
-\begin{itemize}
-\tightlist
-\item
-  the outcome, \(Y\), is a binary variable indicating whether the result
-  is \emph{unexpected} (1) or \emph{expected} (0),
-\item
-  \(X_1, X_2, \cdots, X_n\) represent the binary analysis validation
-  checks, where each check is labeled as 1 if it fails and 0 if it
-  passes.
-\end{itemize}
-
-For classification problems, the logic regression model is reduced to
-the following form:
-
-\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\] The logical operations of
-the binary variables \(X_1, X_2, \cdots, X_n\), including AND
-(\(\land\)), OR (\(\lor\)), and NOT (\(\neg\)) and an example of such
-logic combination can be \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\).
-Simulated annealing is used to explore the search space by minimizing
-the mis-classification error. The following 6 moves are permitted during
-the search (as detailed in Figure 2 of \citet{ruczinski_logic_2003}): 1)
-replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4)
-pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
-
-Compared to other tree-based methods for binary-binary prediction, the
-Boolean combinations from the logic regression model produce a tree
-structure that can be directly interpreted as the possible combination
-of checks leading to an unexpected outcome, without the need to invert
-the tree as required in classic tree-based recursive partitioning
-methods. The logic regression model is then used to predict the analysis
-result based on the values of checks, and the prediction is compared to
-the actual analysis result in order to calculate the precision and
-recall of the checks.
-
-While checks may score high on predictive accuracy, they may be less
-effective at explaining the various reasons behind unexpected results.
-This could happen if, for example, a set of checks are all tangentially
-related to the cause of the unexpected results, but none addresses the
-root cause. It may also occur if the checks are highly correlated with
-one another, leading to redundancy.
-
-To quantify redundancy, the concept of mutual information is used.
-Mutual information \(I(x, y)\) measures the amount of information shared
-between two random variables and is defined as the Kullback-Leibler
-distance \(D(p \parallel q)\) between the joint distribution of the two
-variables and the product of the marginal distributions:
-
-\[I(x,y) = D\big(p(x,y) \parallel p(x)p(y)\big) = \sum_x \sum_y p(x,y) \log \frac{p(x,y)}{p(x)p(y)}\]
-
-This concept extends naturally to multiple variables through total
-correlation, \(C(X_1, X_2, \cdots, X_n)\), which captures redundancy
-across a set of \(n\) variables:
+Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
+detect unexpected outcomes accurately. To relate multiple checks to the
+outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
+Originally developed for SNP microarray data, logic regression
+constructs Boolean combinations of binary variables,
+\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
+regression outcomes. The logical operations allowed to construct
+\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
+(\(\neg\)) and an example of the Boolean combination of
+\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
+our binary-binary classification problem, the logic regression model can
+be written as follows:
+
+\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\]
+
+where the outcome, \(Y\), indicates whether the outcome is
+\emph{unexpected} (1) or \emph{expected} (0), and the analysis
+validation checks, \(X_1, X_2, \cdots, X_n\), are labeled as 1 if they
+fail and 0 if they pass.
+
+This objective function is optimized using simulated annealing algorithm
+to minimize the misclassification error and the following 6 moves are
+permitted during the optimization to grow or purne the tree (as detailed
+in Figure 2 of \citet{ruczinski_logic_2003}): 1) replacing a leaf node,
+2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5)
+splitting a leaf node, and 6) deleting a leaf node. Predictions from the
+logic regression model is then used, along with the observed outcome, to
+calculate the precision and recall of the optimized Boolean combination
+of checks. Compared to other tree-based methods for binary-binary
+prediction, the Boolean combinations from the logic regression model
+produce a tree structure that can be directly interpreted as the
+possible combination of checks leading to an unexpected outcome, without
+the need to invert the tree as required in classic tree-based recursive
+partitioning methods.
+
+\subsection{Independence}\label{independence}
+
+While checks may score high on predictive accuracy, they may be
+correlated and less useful to provide actions to act on the analysis.
+This could happen if a set of checks are all tangentially related to the
+cause of the unexpected results, but none addresses the root cause. The
+independence score in Figure~\ref{fig-metric-calc} is defined as 1 -
+total correlation per observation.
+
+Total correlation is a generalization of mutual information that
+measures the amount of shared information across multiple variables
+multiple variables based on Kullback-Leibler distance:
 
@@
 
-A high mutual information value indicates redundancy among the checks,
-while a low value suggests that the checks are independent and provides
-unique information to diagnose the unexpected outcome. To standardize
-this measure, the total correlation \emph{per observation} is
-calculated, and an independence score, ranging between 0 and 1, is
-defined as 1 - mutual information.
-
-To combine precision, recall, and independence into a single metric, we
-can combine the three scores using the arithmetic mean, harmonic mean,
-or quadratic mean. The differences among these means are minimal when
-the three metrics are similar. However, as the differences among the
-metrics increases, the harmonic mean tends to produce the smallest
-overall score, as it penalizes low values, while the quadratic mean
-tends to produce the largest score by rewarding higher values more. For
-simple interpretation of the score, the arithmetic mean is preferred,
-while in applications where the difference between precision, recall,
-and independence need to be penalized or rewarded more, the harmonic and
+A high total correlation value indicates overlapping of information
+among the analysis validation checks, while a low value suggests that
+the checks are independent and provides unique information to diagnose
+the unexpected outcome. To standardize this measure, the total
+correlation \emph{per observation} is used in calculating the
+independence score.
+
+The three metrics (precision, recall, and independence) can be combiend
+into a single metric using the arithmetic mean, harmonic mean, or
+quadratic mean. The differences among these means are minimal when the
+three metrics are similar. However, as the differences among the metrics
+increases, the harmonic mean tends to produce the smallest overall
+score, as it penalizes low values, while the quadratic mean tends to
+produce the largest score by rewarding higher values more. For simple
+interpretation of the score, the arithmetic mean is preferred, while in
+applications where the difference between precision, recall, and
+independence need to be penalized or rewarded more, the harmonic and
 quadratic mean should be considered.
@@
 Returning to the step count example introduced in Section~\ref{sec-toy},
-the logic regression model is fitted to the three analysis checks
-described previously to generate the prediction of the unexpected
-outcome, whether the average number of step falls within the {[}8,500,
-9,500{]} interval. The predictions from the logic regression model can
-be compared with the simulated true outcome for calculating the
-precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
-best-fitting logic regression model as
-
+the logic regression model is fitted to the three validation checks
+described previously to predict the outcome, whether the average number
+of step falls within the {[}8,500, 9,500{]} interval.
+Figure~\ref{fig-logic-reg} shows the best-fitting logic regression model
+as
+
+\begin{quote}
 (quantile(step, 0.6) \textgreater{} 10,000 OR quantile(step, 0.4)
-\textless{} 8,000) AND sd(step) \textless{} 2,500
-
-In other words, we would predict an unexpected outcome in the analysis
-if the standard deviation of the step count is not too large (2,500) and
-either the 60\% quantile of the step count exceeds 10,000 or the 40\%
-quantile of the step count falls below 8,000.
+\textless{} 8,000) AND (NOT sd(step) \textgreater{} 2,500)
+\end{quote}
+
+We would predict an unexpected outcome in the analysis if the standard
+deviation of the step count is not too large (2,500) and either the 60\%
+quantile of the step count exceeds 10,000 or the 40\% quantile of the
+step count falls below 8,000.
 
@@
 produced by the logic regression can accurately predict 86.1\% cases of
-all \emph{actual unexpected results} in the simulation data.
-Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
-fact observed to be unexpected.
+all \emph{predicted unexpected results} in the simulation data.
+Furthermore, 43.1\% of all \emph{actual unexpected results} were in fact
+observed to be unexpected.
 
@@
 
-\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
-parsimony (independence) metrics for each individual check and the logic
-regression check rule. The harmonic and arithmetic means of the three
-metrics are included to evaluate the quality of the checks in diagnosing
-unexpected step counts (more than five days with fewer than 8,000
-steps).}
+\caption{\label{tbl-logic-reg}Precision, recall, and independence
+calculated for each individual check and the logic regression check
+rule. The harmonic and arithmetic means of the three metrics are
+included to evaluate the quality of the checks in diagnosing unexpected
+step counts (more than five days with fewer than 8,000 steps).}
 
@@
 \toprule
-checks & precision & recall & indep. & harmonic & arithmetic\\
+Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
 \midrule
@@
 \toprule
-check & precision & recall & indep. & harmonic & arithmetic\\
+Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
 \midrule

### 2024-12-24 index.qmd
section 4

@@
 We then introduce a procedure to identify a subset of checks that best predict the occurrence of unexpected outcomes, based on simulations of the original data.
-The checks are evaluated in terms of accuracy, determined by binary classification metrics, and redundancy, which measures the shared information among checks using mutual information.
+The checks are evaluated in terms of accuracy, determined by binary classification metrics, and independence, which measures the shared information among checks.
 We demonstrate this approach with a toy example using step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality.
@@
 We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes.
-The procedure, based on simulations of the original data, compares the accuracy and redundancy of different combinations of analysis validation checks.
-Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while redundancy is measured using mutual information.
+The procedure, based on simulations of the original data, compares the accuracy and independence of different combinations of analysis validation checks.
+Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence is measures the shared information among checks.
 The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.
@@
 
-While some checks may be crucial and directly indicate an unexpected outcome, others may be tangential to the problem at hand and not indicate a root cause of an unexpected outcome.
-In this section, we propose a procedure to measure the effectiveness of checks that, when combined using logical operators, contribute to an unexpected outcome.
-A small set of independent checks is considered effective if it translates to unexpected outcomes.
+While not all analysis validation checks are equally important, some may be more useful at diagnosing unexpected outcomes than others.
+This section introduces a procedure to identify a subset of analysis validation checks that best predict the occurence of unexpected outcomes.
 
-The approach relies on the use of simulated datasets that are generated based on the analyst's knowledge and assumptions about the data generation mechanism.
-Datasets are simulated to have the same structure and characteristics of the observed data that will be analyzed at some point in the future.
-For each simulated dataset, we can apply a collection of analysis validation checks to the dataset and record which ones were TRUE and which were FALSE.
-We can also compute the outcome of the analysis to see whether the outcome is unexpected in a given simulated datsets.
-We can then generate many datasets, each time applying the analysis validation checks and computing the outcome.
-After generating many datasets, we can relate the patterns in the analysis validation checks to the likely that the outcome will be unexpected in a given dataset.
+The approach uses simulated datasets that have the same structure and characteristics of the observed data to be analyzed in the future.
+For each simulated dataset, analysis validation checks are applied to record  their TRUE or FALSE values.
+The outcome of the analysis is also computed to determine whether it is unexpected (1) or expected (0).
+By generating many datasets, each time applying the analysis validation checks and computing the outcome, we can relate the patterns in the checks to the likely that the outcome will be unexpected.
 @fig-metric-calc provides an overview of the process.
@@
 
-From the simulated data, the accuracy branch in @fig-metric-calc refers to a set of checks' ability to accurately detect unexpected outcomes while minimizing false positives and false negatives.
-While a false positive can raise caution or skepticism about the data, the presence of both false positives and false negatives suggests that the checks are overly sensitive or lack sensitivity to unexpected outcomes, respectively.
+## Accuracy
 
-To incorporate the effect of multiple checks on the outcome, a logic regression model [@ruczinski_logic_2003] is fitted to the analysis validation checks.
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict both binary and numerical outcomes.
-In our case,
-
-* the outcome, $Y$, is a binary variable indicating whether the result is *unexpected* (1) or *expected* (0),
-* $X_1, X_2, \cdots, X_n$ represent the binary analysis validation checks, where each check is labeled as 1 if it fails and 0 if it passes.
-
-For classification problems, the logic regression model is reduced to the following form:
+From the simulated data, the accuracy branch in @fig-metric-calc refers to the set of checks' ability to detect unexpected outcomes accurately.
+To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used.
+Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes.
+The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
+In our binary-binary classification problem, the logic regression model can be written as follows:
 
 $$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$
-The logical operations of the binary variables $X_1, X_2, \cdots, X_n$, including AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of such logic combination can be $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
-Simulated annealing is used to explore the search space by minimizing the mis-classification error.
-The following 6 moves are permitted during the search (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
 
+where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_n$, are labeled as 1 if they fail and 0 if they pass.
+
+This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or purne the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
+Predictions from the logic regression model is then used, along with the observed outcome, to calculate the precision and recall of the optimized Boolean combination of checks.
 Compared to other tree-based methods for binary-binary prediction, the Boolean combinations from the logic regression model produce a tree structure that can be directly interpreted as the possible combination of checks leading to an unexpected outcome, without the need to invert the tree as required in classic tree-based recursive partitioning methods.
-The logic regression model is then used to predict the analysis result based on the values of checks, and the prediction is compared to the actual analysis result in order to calculate the precision and recall of the checks.
 
@@
 
-While checks may score high on predictive accuracy, they may be less effective at explaining the various reasons behind unexpected results.
-This could happen if, for example, a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
-It may also occur if the checks are highly correlated with one another, leading to redundancy.
+## Independence
 
-To quantify redundancy, the concept of mutual information is used.
-Mutual information $I(x, y)$ measures the amount of information shared between two random variables and is defined as the Kullback-Leibler distance $D(p \parallel q)$ between the joint distribution of the two variables and the product of the marginal distributions:
+While checks may score high on predictive accuracy, they may be correlated and less useful to provide actions to act on the analysis.
+This could happen if a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
+The independence score in @fig-metric-calc is defined as 1 - total correlation per observation.
 
-$$I(x,y) = D\big(p(x,y) \parallel p(x)p(y)\big) = \sum_x \sum_y p(x,y) \log \frac{p(x,y)}{p(x)p(y)}$$
-
-This concept extends naturally to multiple variables through total correlation, $C(X_1, X_2, \cdots, X_n)$, which captures redundancy across a set of $n$ variables:
+Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables multiple variables based on Kullback-Leibler distance:
 
@@
 
-A high mutual information value indicates redundancy among the checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
-To standardize this measure, the total correlation *per observation* is calculated, and an independence score, ranging between 0 and 1, is defined as 1 - mutual information.
+A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
+To standardize this measure, the total correlation *per observation* is used in calculating the independence score.
 
-To combine precision, recall, and independence into a single metric, we can combine the three scores using the arithmetic mean, harmonic mean, or quadratic mean.
+The three metrics (precision, recall, and independence) can be combiend into a single metric using the arithmetic mean, harmonic mean, or quadratic mean.
 The differences among these means are minimal when the three metrics are similar.
@@
 
-Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three analysis checks described previously to generate the prediction of the unexpected outcome, whether the average number of step falls within the [8,500, 9,500] interval.
-The predictions from the logic regression model can be compared with the simulated true outcome for calculating the precision and recall metrics.
+Returning to the step count example introduced in @sec-toy, the logic regression model is fitted to the three validation checks described previously to predict the outcome, whether the average number of step falls within the [8,500, 9,500] interval.
 @fig-logic-reg shows the best-fitting logic regression model as
 
-(quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND sd(step) < 2,500
+  > (quantile(step, 0.6) > 10,000 OR quantile(step, 0.4) < 8,000) AND (NOT sd(step) > 2,500)
 
-In other words, we would predict an unexpected outcome in the analysis if the standard deviation of the step count is not too large (2,500) and either the 60% quantile of the step count exceeds 10,000 or the 40% quantile of the step count falls below 8,000.
+We would predict an unexpected outcome in the analysis if the standard deviation of the step count is not too large (2,500) and either the 60% quantile of the step count exceeds 10,000 or the 40% quantile of the step count falls below 8,000.
 
@@
 The harmonic and arithmetic means are included to combine the three measures.
-The results show that the three checks produced by the logic regression can accurately predict 86.1% cases of all *actual unexpected results* in the simulation data.
-Furthermore, 43.1% of all *predicted unexpected results* were in fact observed to be unexpected.
+The results show that the three checks produced by the logic regression can accurately predict 86.1% cases of all *predicted unexpected results* in the simulation data.
+Furthermore, 43.1% of all *actual unexpected results* were in fact observed to be unexpected.
 
@@
 #| label: tbl-logic-reg
-#| tbl-cap: "Accuracy (precision and recall) and parsimony (independence) metrics for each individual check and the logic regression check rule.
+#| tbl-cap: "Precision, recall, and independence calculated for each individual check and the logic regression check rule.
 The harmonic and arithmetic means of the three metrics are included to evaluate the quality of the checks in  diagnosing unexpected step counts (more than five days with fewer than 8,000 steps)."
@@
   dplyr::rename(`indep.` = independence) |> 
-  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE) |> 
+  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE, 
+        col.names = c("Checks", "Precision", "Recall", "Independence", "Harmonic", "Arithmetic")) |> 
   kableExtra::kable_styling(latex_options="scale_down") |> 
@@
 tbl |> 
-  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE) |> 
+  kable(digits = 3, escape = FALSE, format = "latex", booktab = TRUE,
+        col.names = c("Checks", "Precision", "Recall", "Independence", "Harmonic", "Arithmetic")) |> 
   kableExtra::kable_styling(latex_options="scale_down") |> 

### 2024-12-24 _manuscript/_tex/index.tex
section 5 and 6

@@
 Study (NMMAPS) to demonstrate how our analysis validation checks
-described in Section~\ref{sec-method} can be applied. The dataset that
-we use to develop our simulations and analysis checks is from New York
-City, and contains daily PM10, all-cause (non-accidental) mortality, and
-average temperature values from 1992--2000. We then will apply the
-analysis checks to similar data from 39 other cities from the NMMAPS
-study.
+described in Section~\ref{sec-method} can be applied. In addition to
+providing a more substantial problem for our methods, this example also
+demonstrates how the procedure presented in Section~\ref{sec-method} can
+be used to select cutoff values in the analysis checks to diagnose an
+unexpected PM10 coefficient from the generalized linear model. The
+dataset that we use to develop our simulations and analysis checks is
+from New York City, and contains daily PM10, all-cause (non-accidental)
+mortality, and average temperature values from 1992--2000. We then will
+apply the analysis checks to similar data from 39 other cities from the
+NMMAPS study.
 
@@
 
-For the purposes of this application, observing an estimated coefficient
-outside of the expected interval is considered unusual and might warrant
-a re-examination of the analysis process. Therefore, our unexpected
-outcome in this analysis will be the binary indicator of whether the
-estimated coefficient for PM10 in the GLM lies outside of the interval
-\([0, 0.005]\). In addition to providing a more substantial problem for
-our methods, this example also demonstrates how the procedure presented
-in Section~\ref{sec-method} can be used to select cutoff values in the
-analysis checks to diagnose an unexpected PM10 coefficient from the
-generalized linear model.
-
-This PM10 coefficient expectation can be framed as an analysis check
-that fails, labelled as 1, if the estimate of the PM10 coefficient
-(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
-it is within. Multiple factors can affect the estimated PM10
-coefficient, such as the strength of the correlation between mortality
-and PM10, or between mortality and temperature. Outliers in the three
-variables can also leverage the coefficient. While these are possible
-factors that could affect the analysis result, it is not clear what the
-cutoff values for these checks should be to determine a failure. Here we
-consider a list of checks in Table~\ref{tbl-checks} with varied cutoff
-values for each:
+Multiple factors can affect the estimated PM10 coefficient, such as the
+strength of the correlation between mortality and PM10, or between
+mortality and temperature. Outliers in the variables can also leverage
+the coefficient. While these are possible factors that could affect the
+analysis result, it is not clear what the cutoff values for these checks
+should be to determine a failure. Here we consider a list of checks in
+Table~\ref{tbl-checks} with varied cutoff values:
 
@@
 from the fitted model. Precision, recall, and independence score, along
-with their harmonic and arithmetic mean are calculated in
+with their harmonic and arithmetic mean are presented in
 Table~\ref{tbl-linear-reg}.
@@
 
-\caption{\label{tbl-linear-reg}Accuracy (precision and recall) and
-parsimony (independence) metrics derived from the logic regression model
-(last row) and individual checks, along with harmonic and arithmetic
-means.}
+\caption{\label{tbl-linear-reg}Precision, recall, and independence
+metrics derived from the logic regression model and each individual
+check, along with harmonic and arithmetic means.}
 
@@
 \item
-  PM10 data contains outliers that are detected by the univariate
-  outlier detection
+  PM10 data contains outliers detected by the univariate outlier
+  detection
 \item
-  mortality data contains outliers that are detected by the univariate
-  outlier detection
+  mortality data contains outliers detected by the univariate outlier
+  detection
 \end{itemize}
 
-The tree structure suggests checking mortality-PM10 correlation and a
-sample size larger than 200 with an additional check of either outlier
-on mortality or correlation between mortality and temperature. This
-combined check rule generates a precision of 0.722 and a recall of 0.876
-for predicting the unexpected PM10 coefficient.
-
-As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
-analysis check in the tree that predicts an unexpected outcome. rather
-at least three checks in the tree must be TRUE in order for the model to
-predict an unexpected outcome. Given the high independence of the checks
+The fitted logic regression model is
+
+\begin{quote}
+(cor(mortality, PM10) \textless{} -0.03) AND (cor(mortality, tmp)
+\textgreater{} -0.3 OR (mortality outlier AND PM10 outlier))
+\end{quote}
+
+This generates a precision of 0.722 and a recall of 0.876 for predicting
+the unexpected PM10 coefficient. As shown in
+Figure~\ref{fig-linear-reg-tree}, there is no single analysis check in
+the tree that predicts an unexpected outcome. rather at least three
+checks in the tree must be TRUE in order for the model to predict an
+unexpected outcome. Given the high independence of the checks
 (Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
@@
 correctly predicts the PM10 coefficient status of 32 cities, producing
-an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome
-(PM10 coefficient outside {[}0, 0.005{]}), the model correctly
-identifies 7 (Atlanta, Austin, Memphis, Orlando, Salt Lake City, San
-Antonio, Tampa), giving a recall of 88\% (only Miami was not properly
-classified by the logic regression model). These cities were flagged due
-to failures in both the mortality-PM10 correlation and
+an accuracy rate of 80\%. Among the 8 cities with an unexpected outcome,
+the model correctly identifies 7 (Atlanta, Austin, Memphis, Orlando,
+Salt Lake City, San Antonio, Tampa), giving a recall of 88\% (only Miami
+was not properly classified by the logic regression model). These cities
+were flagged due to failures in both the mortality-PM10 correlation and
 mortality-temperature correlation analysis checks. Out of the 14 cities
@@
 health data, perhaps in other cities or locations.
-
-With the analysis validation checks proposed in this paper, analysts may
-perform these checks \emph{ex ante} to identify and address potential
-issues before proceeding with the analysis or use them \emph{ex post} to
-diagnose the analysis when an unexpected outcome occurs. Theses checks
-act as checkpoints designed by analysts to flag potential issues,
-clarify assumptions made during the analysis, and enhance transparency
-and communication among stakeholders. They also support students and
-junior researchers develop the skills needed to independently diagnose
-issues in analyses.
 
@@
 deviations from normality. These visualizations could be re-framed as
-checks that fail when the data does not conform to the expected
-distribution. However, translating visualization results into binary
-checks that return (TRUE, FALSE) remains an open challenge, requiring
-either manual verification or the development of automated methods to
-interpret visualization outputs. An existing example of visual test is
-the R package \texttt{vdiffr} \citep{vdiffr} for software unit testing.
-The package saves a template plot and compares it to the current plot to
+analysis checks that fail when the data does not conform to the visual
+expectation. However, translating visualization results into binary
+checks remains an open challenge, requiring either manual verification
+or the development of automated methods to interpret visualization
+outputs. An existing example of visual test is the R package
+\texttt{vdiffr} \citep{vdiffr} for graphic software unit testing. The
+package saves a template plot and compares it to the current plot to
 determine whether the unit tests passes or fails.

### 2024-12-24 index.qmd
section 5 and 6

@@
 For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
+In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
 The dataset that we use to develop our simulations and analysis checks is from New York City, and contains daily PM10, all-cause (non-accidental) mortality, and average temperature values from 1992--2000.
@@
 
-For the purposes of this application, observing an estimated coefficient outside of the expected interval is considered unusual and might warrant a re-examination of the analysis process.
-Therefore, our unexpected outcome in this analysis will be the binary indicator of whether the estimated coefficient for PM10 in the GLM lies outside of the interval $[0, 0.005]$.
-In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
-
-This PM10 coefficient expectation can be framed as an analysis check that fails, labelled as 1, if the estimate of the PM10 coefficient (adjusted for temperature) is outside the range [0, 0.005], and 0 if it is within.
 Multiple factors can affect the estimated PM10 coefficient, such as the strength of the correlation between mortality and PM10, or between mortality and temperature.
-Outliers in the three variables can also leverage the coefficient.
+Outliers in the variables can also leverage the coefficient.
 While these are possible factors that could affect the analysis result, it is not clear what the cutoff values for these checks should be to determine a failure.
-Here we consider a list of checks in @tbl-checks with varied cutoff values for each:
+Here we consider a list of checks in @tbl-checks with varied cutoff values:
 
@@
 @fig-linear-reg-tree shows the optimal logic regression tree from the fitted model.
-Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated in @tbl-linear-reg.
+Precision, recall, and independence score, along with their harmonic and arithmetic mean are presented in @tbl-linear-reg.
 
@@
 #| fig.width: 12
-#| fig.height: 4
+#| fig.height: 6
 fit <- pm10_df |> fit_logic_reg(unexpect, `cor(m, PM10) < -0.05`:mortality_outlier,
@@
 #| label: tbl-linear-reg
-#| tbl-cap: "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model (last row) and individual checks, along with harmonic and arithmetic means."
+#| tbl-cap: "Precision, recall, and independence metrics derived from the logic regression model and each individual check, along with harmonic and arithmetic means."
 df <- list(
@@
 -   mortality-temperature correlation greater than $-0.3$
--   PM10 data contains outliers that are detected by the univariate outlier detection
--   mortality data contains outliers that are detected by the univariate outlier detection
+-   PM10 data contains outliers detected by the univariate outlier detection
+-   mortality data contains outliers detected by the univariate outlier detection
 
-The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature.
-This combined check rule generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient.
+The fitted logic regression model is
 
+  > (cor(mortality, PM10) < -0.03) AND (cor(mortality, tmp) > -0.3 OR (mortality outlier AND PM10 outlier))
+
+This generates a precision of `r round(as.numeric(df[5, "precision"]), 3)` and a recall of `r round(as.numeric(df[5, "recall"]), 3)` for predicting the unexpected PM10 coefficient.
 As shown in @fig-linear-reg-tree, there is no single analysis check in the tree that predicts an unexpected outcome.
@@
 The model correctly predicts the PM10 coefficient status of `r accurate <- df |> filter(unexpect == .pred) |> nrow(); accurate` cities, producing an accuracy rate of `r label_percent()(accurate/nrow(df))`.
-Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome (PM10 coefficient outside [0, 0.005]), the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model).
+Among the `r df |> filter(unexpect) |> nrow()` cities with an unexpected outcome, the model correctly identifies `r df |> filter(unexpect, .pred) |> nrow()` (`r cities_accurate`), giving a recall of `r recall <- as.numeric(df |> calc_miscla_rate(.pred, unexpect) |> pull(recall)); label_percent()(recall)` (only `r cities_missed` was not properly classified by the logic regression model).
 These cities were flagged due to failures in both the mortality-PM10 correlation and mortality-temperature correlation analysis checks.
@@
 The results from that example could be used to inform future analyses of air pollution and health data, perhaps in other cities or locations.
-
-With the analysis validation checks proposed in this paper, analysts may perform these checks *ex ante* to identify and address potential issues before proceeding with the analysis or use them *ex post* to diagnose the analysis when an unexpected outcome occurs.
-Theses checks act as checkpoints designed by analysts to flag potential issues, clarify assumptions made during the analysis, and enhance transparency and communication among stakeholders.
-They also support students and junior researchers develop the skills needed to independently diagnose issues in analyses.
 
@@
 For instance, plotting a variable's distribution using a histogram, density plot, or bee swarm plot can reveal outliers or deviations from normality.
-These visualizations could be re-framed as checks that fail when the data does not conform to the expected distribution.
-However, translating visualization results into binary checks that return (TRUE, FALSE) remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs.
-An existing example of visual test is the R package `vdiffr` [@vdiffr] for software unit testing.
+These visualizations could be re-framed as analysis checks that fail when the data does not conform to the visual expectation.
+However, translating visualization results into binary checks remains an open challenge, requiring either manual verification or the development of automated methods to interpret visualization outputs.
+An existing example of visual test is the R package `vdiffr` [@vdiffr] for graphic software unit testing.
 The package saves a template plot and compares it to the current plot to determine whether the unit tests passes or fails.

### 2025-07-09 _manuscript/revision/response.qmd
some revision

---
title: "Response to Reviewers JDS24112"
subtitle: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
author: "H. Sherry Zhang, Roger D. Peng"
date: "2025-07-08"
format: 
  html:
    toc: false
---

We would like to thank the reviewers and editor for taking the time to review our paper and provide constructive comments. Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in normal text and **our responses are in bold text**.

# Reviewer 1

## General comments

The manuscript is being considered for the special issue on Statistical Aspects of Trustworthy Machine Learning. However, the connection to the theme of the special issue feels somewhat underdeveloped. It would be helpful for the authors to clarify which aspect(s) of trustworthiness are being addressed. While the work appears to relate to robustness and reproducibility, these connections should be more clearly articulated and substantiated in the manuscript.

The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2). However, this assumption may not be realistic in many practical scenarios. Simulating data requires a data-generating model, but commonly used statistical methods such as generalized estimating equations (GEE) or the generalized method of moments (GMM) do not fully specify the underlying distribution or likelihood. Even when the mean structure is specified, there may be infinitely many likelihood models consistent with it. Simulating from only one such model risks overlooking a broad class of plausible alternatives. How do the authors propose to handle such cases? More broadly, generating synthetic data with properties similar to real data is a fundamental and challenging problem in data science practice. The authors may wish to comment on recent advances in this area that aim to address this challenge without fully specifying a probabilistic model.

The construction of validation checks appears to be a critical step in applying the proposed methodology. It would be valuable if the authors could offer some guidance or recommendations on the principles for designing these checks. Clarifying this aspect would enhance the practical utility of the approach and help readers apply it more effectively in real-world settings.

The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable. By definition, the metric seems to always take the value one in such cases, as observed in Tables 1 and 3. Including it in the total score may artificially inflate the overall score when compared to checks involving multiple components. Further, in the current examples, the multivariate checks consistently yield independence values close to one. Are there cases where this metric meaningfully deviates from one? Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology.

The writing is generally clear and easy to follow. However, there are a number of minor syntax errors that could be addressed with more careful proofreading. The use of AI-based writing assistance tools could be helpful in this regard and, in my view, would constitute an ethical and constructive application of such tools. **A: We have run though the spelling check on the article and fixed the spelling mistakes.**

## Specific Comments

1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful.

2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy.**

3.  Section 4.1, paragraph 1: X1, . . . , Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size. **A: easy. TODO**

4.  Section 4.2, end of paragraph 1: I assume the dash should be the minus sign. Having a minus sign in a sentence is confusing. This needs to be rephrased, perhaps just use “minus”.  **A: We have replaced the dash with the word "minus" in the text. **

5.  Section 4.2, paragraph 2: What is the range of total correlation? My understanding is that it is \[0, ∞). Even after standardizing to per observation, the range may still be not \[0, 1) as desired. **A: The range of total correlation is \[0, ∞), to obtain a measure in \[0, 1), we calculate total correlation per observation as the independence score. This is explained in the last sentence of the paragraph. **

6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) as TODO and close the sentence.**

7.  Table 2: How did the authors come up with these checks? Readers would appreciate the motivations of these checks. Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation).

8.  Section 5.1, paragraph 1: Gaussian copula does not allow tail dependence. So by imposing a Gaussian copula, it is assumed that there is no tail dependence in these three variables. As questioned in one of my general comments, this might not be desirable.

9.  Figure 4: Could add bands of 95% confidence intervals of the QQ-plot; the texts on top of each panel could be moved to the caption; the labels of the axis should clarify empirical versus theoretical.

10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **A: Fixed.**

11. Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis. **A: The intention of using "second example" is because we count the toy example as the first example. But to avoid this confusion, we have changed this to "In the following example".**

12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table.

13. Section 5: It’d be valuable to see some real improvements from the checks. For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks.

## Minor comments

-   Abstract: “simulations of the original data” is misleading. **A: we remove this information in this sentence for easier understanding of the sentence. **

-   Page 2, paragraph 1: “thought” process, not “throught” process. **A: Fixed.**

-   Page 2, paragraph 2: “is measures”, two verbs. **A: The redundant "is" has been removed**.

### 2025-07-09 _manuscript/_tex/index.tex
some revision

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 ]{interact}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
-\usepackage{setspace}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 }{}
+\usepackage{setspace}
 \makeatletter
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
 
@@
 \makeatother
-
-\usepackage[]{natbib}
-\bibliographystyle{plainnat}
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 screencast videos requires significant effort from the analysts.
-Ideally, there could be a way to explicitly communciate these throught
+Ideally, there could be a way to explicitly communicate these thought
 processes in the data analysis to others. Even better, if these
@@
 
-In this paper, we conceptualize these inexplicit expectations as
+In this paper, we conceptualize these implicit expectations as
 \emph{analysis validation checks}, which allows us to examine the
@@
 from a logic regression model \citep{ruczinski_logic_2003}, while
-independence is measures the shared information among checks. The
-proposed workflow offers a numerical guarantee that the analysis will
-produce the expected results, assuming the assumptions about the data
-generating mechanism hold.
+independence measures the shared information among checks. The proposed
+workflow offers a numerical guarantee that the analysis will produce the
+expected results, assuming the assumptions about the data generating
+mechanism hold.
 
@@
 \texttt{validate} \citep{validate}, and \texttt{pointblank}
-\citep{pointblank} focuse on providing infustructures that allow users
+\citep{pointblank} focus on providing infrastructures that allow users
 to define customized data quality checks.
@@
 section introduces a procedure to identify a subset of analysis
-validation checks that best predict the occurence of unexpected
+validation checks that best predict the occurrence of unexpected
 outcomes.
@@
 outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
-Originally developed for SNP microarray data, logic regression
+Originally developed for SNP micro-array data, logic regression
 constructs Boolean combinations of binary variables,
@@
 to minimize the misclassification error and the following 6 moves are
-permitted during the optimization to grow or purne the tree (as detailed
+permitted during the optimization to grow or prune the tree (as detailed
 in Figure 2 of \citet{ruczinski_logic_2003}): 1) replacing a leaf node,
@@
 cause of the unexpected results, but none addresses the root cause. The
-independence score in Figure~\ref{fig-metric-calc} is defined as 1 -
+independence score in Figure~\ref{fig-metric-calc} is defined as 1 minus
 total correlation per observation.
@@
 
-\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}\]
-
-A high total correlation value indicates overlapping of information
-among the analysis validation checks, while a low value suggests that
-the checks are independent and provides unique information to diagnose
-the unexpected outcome. To standardize this measure, the total
-correlation \emph{per observation} is used in calculating the
-independence score.
-
-The three metrics (precision, recall, and independence) can be combiend
+\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},\]
+
+where \(p(.)\) is the TODO. A high total correlation value indicates
+overlapping of information among the analysis validation checks, while a
+low value suggests that the checks are independent and provides unique
+information to diagnose the unexpected outcome. To standardize this
+measure, the total correlation \emph{per observation} is used in
+calculating the independence score.
+
+The three metrics (precision, recall, and independence) can be combined
 into a single metric using the arithmetic mean, harmonic mean, or
@@
 than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
-mortality from all non-accidental causes \citep{samet2000fine}. For our
-second example, we use the problem of studying PM10 and mortality along
-with data from the National Morbidity, Mortality, and Air Pollution
-Study (NMMAPS) to demonstrate how our analysis validation checks
-described in Section~\ref{sec-method} can be applied. In addition to
-providing a more substantial problem for our methods, this example also
-demonstrates how the procedure presented in Section~\ref{sec-method} can
-be used to select cutoff values in the analysis checks to diagnose an
-unexpected PM10 coefficient from the generalized linear model. The
-dataset that we use to develop our simulations and analysis checks is
-from New York City, and contains daily PM10, all-cause (non-accidental)
-mortality, and average temperature values from 1992--2000. We then will
-apply the analysis checks to similar data from 39 other cities from the
-NMMAPS study.
+mortality from all non-accidental causes \citep{samet2000fine}. In the
+following example, we use the problem of studying PM10 and mortality
+along with data from the National Morbidity, Mortality, and Air
+Pollution Study (NMMAPS) to demonstrate how our analysis validation
+checks described in Section~\ref{sec-method} can be applied. In addition
+to providing a more substantial problem for our methods, this example
+also demonstrates how the procedure presented in
+Section~\ref{sec-method} can be used to select cutoff values in the
+analysis checks to diagnose an unexpected PM10 coefficient from the
+generalized linear model. The dataset that we use to develop our
+simulations and analysis checks is from New York City, and contains
+daily PM10, all-cause (non-accidental) mortality, and average
+temperature values from 1992--2000. We then will apply the analysis
+checks to similar data from 39 other cities from the NMMAPS study.
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/unnamed-chunk-8-1.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/unnamed-chunk-3-1.pdf}}
 
@@
 variable. The model suggests the relationship: (mortality-PM10
-correlation \textless{} -0.03) AND (mortality-temperature correlation
-\textgreater{} -0.3 OR (there exist mortality outlier AND there exist
-PM10 outlier)) to predict the unexpected PM10 coefficient.}
+correlation \textless{} \(-0.03\)) AND (mortality-temperature
+correlation \textgreater{} \(-0.3\) OR (there exist mortality outlier
+AND there exist PM10 outlier)) to predict the unexpected PM10
+coefficient.}
 

### 2025-07-09 _manuscript/index.qmd
some revision

--- 2024-12-24 index.qmd
@@
 knitr::opts_chunk$set(echo = FALSE)
+#remotes::install_github("huizezhang-sherry/adtoolbox")
 library(tidyverse)
@@
 However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts.
-Ideally, there could be a way to explicitly communciate these throught processes in the data analysis to others.
+Ideally, there could be a way to explicitly communicate these thought processes in the data analysis to others.
 Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself.
@@
 
-In this paper, we conceptualize these inexplicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis.
+In this paper, we conceptualize these implicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis.
 We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes.
 The procedure, based on simulations of the original data, compares the accuracy and independence of different combinations of analysis validation checks.
-Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence is measures the shared information among checks.
+Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence measures the shared information among checks.
 The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.
@@
 This encourages researchers to conduct data quality checks before beginning the analytic process, helping to avoid unexpected discoveries later on.
-In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focuse on providing infustructures that allow users to define customized data quality checks.
+In R, packages like `skimr` [@skimr] and `dataMaid` [@dataMaid] provide basic data screening and data quality reports, while packages such as  `assertr` [@assertr], `validate` [@validate], and `pointblank` [@pointblank] focus on providing infrastructures that allow users to define customized data quality checks.
 
@@
 While not all analysis validation checks are equally important, some may be more useful at diagnosing unexpected outcomes than others.
-This section introduces a procedure to identify a subset of analysis validation checks that best predict the occurence of unexpected outcomes.
+This section introduces a procedure to identify a subset of analysis validation checks that best predict the occurrence of unexpected outcomes.
 
@@
 To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used.
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes.
+Originally developed for SNP micro-array data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes.
 The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
@@
 
-This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or purne the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
+This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or prune the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
 Predictions from the logic regression model is then used, along with the observed outcome, to calculate the precision and recall of the optimized Boolean combination of checks.
@@
 This could happen if a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
-The independence score in @fig-metric-calc is defined as 1 - total correlation per observation.
+The independence score in @fig-metric-calc is defined as 1 minus total correlation per observation.
 
@@
 
-$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}$$
+$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},$$
 
+where $p(.)$ is the TODO.
 A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
@@
 
-The three metrics (precision, recall, and independence) can be combiend into a single metric using the arithmetic mean, harmonic mean, or quadratic mean.
+The three metrics (precision, recall, and independence) can be combined into a single metric using the arithmetic mean, harmonic mean, or quadratic mean.
 The differences among these means are minimal when the three metrics are similar.
@@
 Substantial work has been done to study this question and to date, there appears to be strong evidence of an association between particulate matter less than 10 $\mu$g/m$^3$ in aerodynamic diameter (PM10) and daily mortality from all non-accidental causes [@samet2000fine].
-For our second example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
+In the following example, we use the problem of studying PM10 and mortality along with data from the National Morbidity, Mortality, and Air Pollution Study (NMMAPS) to demonstrate how our analysis validation checks described in @sec-method can be applied.
 In addition to providing a more substantial problem for our methods, this example also demonstrates how the procedure presented in @sec-method can be used to select cutoff values in the analysis checks to diagnose an unexpected PM10 coefficient from the generalized linear model.
@@
 Logic regression model fitted to the fourteen checks and the outcome expectation (unexpected) as the response variable.
-The model suggests the relationship:  (mortality-PM10 correlation < -0.03) AND (mortality-temperature correlation > -0.3 OR (there exist mortality outlier AND there exist PM10 outlier)) to predict the unexpected PM10 coefficient.
+The model suggests the relationship:  (mortality-PM10 correlation < $-0.03$) AND (mortality-temperature correlation > $-0.3$ OR (there exist mortality outlier AND there exist PM10 outlier)) to predict the unexpected PM10 coefficient.
 :::

### 2025-07-20 _manuscript/revision/response.qmd
more responses

@@
 date: "2025-07-08"
-format: 
-  html:
-    toc: false
+format: html
+bibliography: ../references.bib
 ---
@@
 
+1.
 The manuscript is being considered for the special issue on Statistical Aspects of Trustworthy Machine Learning.
@@
 While the work appears to relate to robustness and reproducibility, these connections should be more clearly articulated and substantiated in the manuscript.
+**A: We now discuss the connection between our work and the trustworthiness of machine learning in the third paragraph of the *Introduction*.
+Specifically, our work address the breakdown in trust caused by  violations of algorithmic assumptions -- corresponding to the third step in @broderick_toward_2023.
+In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcome.
+Our approach advocates for making these assumptions -- often simple or taken for granted -- explicit, in order to more efficiently diagnose unexpected outcomes.
+**
 
+2.
 The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2).
@@
 The authors may wish to comment on recent advances in this area that aim to address this challenge without fully specifying a probabilistic model.
+**A: The reviewer is correct that in certain scenarios (e.g.
+GEE and GMM), it can be challenging to accurately simulate dataset that resemble the observed data.
+However, the primary aim of our paper is not to recover the true data generating process or to find the model that best fits the data.
+Instead, our goal is to understand which characteristic of the data may lead to the unexpected results - a mismatch between the data and the model.
+Here, we hold the model fixed and examine the idiosyncrasies in the data that could causes such incompatibility.
+This contrasts with the approach of treating data as fixed and focus on improving the model.
+Our perspective is motivated by the applied studies, where certain modelling strategies are generally accepted for specific types of data.
+When unexpected outcomes arise, analysts often first investigate whether particular characteristics of the data may cause the results, rather than developing new models.
+In this context, the analysis validation checks we propose provide a pre-developed "diagnostic checks" to investigate the data.
+To note that, even when the results is as expected, we don't necessarily assume the model is correct.
+[More to address the second part]**
 
+<!-- Train AI model with synthetic data. Take the image and rotate 45 degree. - maybe talk about it in the discussion** -->
+
+3.
 The construction of validation checks appears to be a critical step in applying the proposed methodology.
@@
 
+**A: In the toy example section (Section 3.1), we include some explanations on the construction of validation checks: these are diagnostics analysts may naturally check when they receive an unexpected outcome.
+For example, when the mean of a vector falls outside an expected range, analysts may check the the five-number summary or plot the variable distribution.
+The validation checks are those diagnostics framed as binary predicates.
+**
+
+4.
 The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable.
@@
 Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology.
+**A: I need to investigate on this.**
 
+5.
 The writing is generally clear and easy to follow.
@@
 
-1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful.
+1.
+Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data?
+If so, this is a big difference from bootstrap samples and a remark would be useful.
+**A: The simulated dataset in our paper are not necessarily independent of the observed data, and we do not rely on bootstrap sampling at any point, although it can be a method to generate the simulated data.
+We have include a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.**
 
-2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy.**
+2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy. TODO**
 
-3.  Section 4.1, paragraph 1: X1, . . . , Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size. **A: easy. TODO**
+3.
+Section 4.1, paragraph 1: X1, .
+.
+.
+, Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size.
+**A: We have added the definition of $X_i$ as the binary predictors in Section 4.1 and replace $n$ with $k$ for the number of checks.
+**
 
@@
 This is explained in the last sentence of the paragraph.
-**
+TODO**
 
-6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) as TODO and close the sentence.**
+6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) and close the sentence.**
 
@@
 Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation).
+**A: Four mortality-pm10/ mortality-temperature tests are constructed because analysts may not be sure what the best cutoff point for the correlation tests would be.
+This example illustrates how the method developed can be used to determine the cutoff value for tests.**
 
@@
 As questioned in one of my general comments, this might not be desirable.
+**A: ask Roger** may or may not being able to simulate a particular characteristic of the data and generate
 
@@
 Figure 4: Could add bands of 95% confidence intervals of the QQ-plot; the texts on top of each panel could be moved to the caption; the labels of the axis should clarify empirical versus theoretical.
+**A: We have added the 95% confidence interval, remove the plot title to the caption, and refine the axis labels.**
 
@@
 Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis.
-**A: The intention of using "second example" is because we count the toy example as the first example.
-But to avoid this confusion, we have changed this to "In the following example".**
+**A: We use the wording "second example" because we count the toy example as the first example.
+But, to avoid this confusion, we have changed the wording to "In the following example".**
 
-12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table.
+12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table. **A: addressed in *Specific comments* #7. **
 
@@
 For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks.
+**A: The reviewer's comment assumes we need a better model when an unexpected outcome occurs, however, in practice, we always assume there is something issue with the data, rather than rushing into developing new methods.
+The analysis validation checks are used to motivate a better understanding of the data, not an ideal model (This point is also addressed in *General comments* #2).**
 
@@
 -   Page 2, paragraph 2: “is measures”, two verbs. **A: The redundant "is" has been removed**.
+
+# Reference

### 2025-07-20 _manuscript/_tex/index.tex
more responses

@@
 them directly or watching them work via screencast videos they produce,
-such as, TidyTuesday screencast videos or think-aloud type studies
+such as TidyTuesday screencast videos or think-aloud type studies
 \citep[e.g.][]{gu2024data}. However, direct observation of analysis is
@@
 whether they are specific to the data or analysis at hand. Externalizing
-these thought processes is a practice that has potential to improve the
-trustworthiness of the analyses \citep{yu2024veridical} in general and
-the trustworthiness of subsequent products, such as machine learning
-models, that may be built on such analyses. Developing an approach to
-reveal some of this process without requiring a reader to essentially
-reconstruct the analysis process from the raw data would provide an
-improved basis for trusting an analysis result
-\citep{peng2021reproducible}.
+these thought processes can improve the trustworthiness of the analyses
+\citep{yu2024veridical}.
+
+The trustworthiness of an analysis depends not only on the robustness of
+the methods used but also on a deep understanding of the data. This
+corresponds to the third step identified in
+\citet{broderick_toward_2023}, where trust can break down if the
+assumptions underlying an algorithm are not satisfied in practice. In
+the past, data were often collected directly by analysts through
+carefully designed experiments, however, contemporary analyses often
+rely on data curated by government or research institutions, where some
+data characteristics may be unknown to the analysts. When unexpected
+outcomes arise, they often later traced back to implicit assumptions
+made by the analysts - assumptions that were unknowingly violated by the
+idiosyncratic in the data. By externalizing these assumptions through
+analysis validation checks, we aim to make it easier to identify and
+diagnose such unexpected results.
 
@@
 establish a few analysis validation checks in anticipation of seeing the
-data. For example, we can check the quantile of the step count, if more
-than a third of the days fall below 8,000, or more than a third exceed
-10,000 steps, this could indicate an excess of low-count or high-count
-days. Similarly, we may may expect the standard deviation of the step
-count not to be overly large. These considerations yield the following
-three analysis validation checks that \emph{fail} when:
+data. {[}TODO: Add guidelines for designing the checks{]} For example,
+we can check the quantile of the step count, if more than a third of the
+days fall below 8,000, or more than a third exceed 10,000 steps, this
+could indicate an excess of low-count or high-count days. Similarly, we
+may may expect the standard deviation of the step count not to be overly
+large. These considerations yield the following three analysis
+validation checks that \emph{fail} when:
 
@@
 
-The approach uses simulated datasets that have the same structure and
-characteristics of the observed data to be analyzed in the future. For
-each simulated dataset, analysis validation checks are applied to record
-their TRUE or FALSE values. The outcome of the analysis is also computed
-to determine whether it is unexpected (1) or expected (0). By generating
-many datasets, each time applying the analysis validation checks and
-computing the outcome, we can relate the patterns in the checks to the
-likely that the outcome will be unexpected. Figure~\ref{fig-metric-calc}
-provides an overview of the process.
+The approach generates simulated datasets from the observed data to
+compute the outcome and the analysis validation checks (unexpected as 1
+and expected as 0). This then allows us to apply a logic regression tree
+to link the outcome and the check results to construct a Boolean
+expression to best predict the unexpected outcomes.
+Figure~\ref{fig-metric-calc} provides an overview of the process.
+
+The specific strategy to simulate the data depends on the problem and
+the nature of the data at hand. Simulated data can be generated using
+prior domain knowledge, summary statistics derived from the observed
+data, or statistical techniques such as training-testing splits or
+bootstrap sampling. While each simulated data may not be independent of
+the original data - as is the case with bootstrap samples - it should
+adequately capture the range of possible idiosyncrasies present in the
+data. Failure to do so can reduce the ability of the resulting Boolean
+expression to predict unexpected outcomes.
 
@@
 outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
-Originally developed for SNP micro-array data, logic regression
-constructs Boolean combinations of binary variables,
-\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
-regression outcomes. The logical operations allowed to construct
+Originally developed for SNP micro-array data, logic regression predict
+classification or regression outcomes by constructing Boolean
+combinations, \(\mathcal{L}(X_1, X_2, \cdots, X_k)\), where \(X_i\) are
+binary predictors. The logical operations allowed to construct
 \(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
@@
 
-\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\]
+\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)\]
 
@@
 \emph{unexpected} (1) or \emph{expected} (0), and the analysis
-validation checks, \(X_1, X_2, \cdots, X_n\), are labeled as 1 if they
+validation checks, \(X_1, X_2, \cdots, X_k\), are labeled as 1 if they
 fail and 0 if they pass.
@@
 Total correlation is a generalization of mutual information that
-measures the amount of shared information across multiple variables
-multiple variables based on Kullback-Leibler distance:
-
-\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},\]
-
-where \(p(.)\) is the TODO. A high total correlation value indicates
-overlapping of information among the analysis validation checks, while a
-low value suggests that the checks are independent and provides unique
-information to diagnose the unexpected outcome. To standardize this
-measure, the total correlation \emph{per observation} is used in
-calculating the independence score.
+measures the amount of shared information across multiple variables,
+based on Kullback-Leibler distance:
+
+\[C(X_1, X_2, \cdots, X_k) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_k} p(x_1, x_2, \cdots, x_k) \log \frac{p(x_1, x_2, \cdots, x_k)}{p(x_1)p(x_2) \cdots p(x_k)},\]
+
+where \(p(x_1, x_2, \cdots, x_k)\) is the joint distribution of the
+predictors and \(p(\cdot)\) is the individual distribution of the
+predictors. A high total correlation value indicates overlapping of
+information among the analysis validation checks, while a low value
+suggests that the checks are independent and provides unique information
+to diagnose the unexpected outcome. To standardize this measure, the
+total correlation \emph{per observation} is used in calculating the
+independence score.
 
@@
 
-\caption{\label{fig-dist-fit}QQ-plot of the distribution fit for
-mortality, PM10, and temperature based on the fitted distribution from
-the original data. The fitted distribution is compared to the observed
-data to assess the distribution fit.}
+\caption{\label{fig-dist-fit}QQ-plot of the goodness-of-fit of selected
+distributions used to generate simulated data for mortality, PM10, and
+temperature. Left: Mortality as a negative binomial distribution (size =
+74, mean = 183), Middle: PM10 is modeled using a rescaled beta
+distribution with shape parameters 4.21 and 11.67, multiplied by 10 to
+match the observed range. Right: Temperature is modeled using a Weibull
+distribution (shape = 3.8, scale = 61)}
 

### 2025-07-20 _manuscript/index.qmd
more responses

@@
 library(scales)
+library(qqplotr)
 pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
@@
 
-We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g.
+We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as TidyTuesday screencast videos or think-aloud type studies [e.g.
 @gu2024data].
@@
 For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
-Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
-Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].
+Externalizing these thought processes can improve the trustworthiness of the analyses [@yu2024veridical].
+
+The trustworthiness of an analysis depends not only on the robustness of the methods used but also on a deep understanding of the data.
+This corresponds to the third step identified in @broderick_toward_2023, where trust can break down if the assumptions underlying an algorithm are not satisfied in practice.
+In the past, data were often collected directly by analysts through carefully designed experiments, however, contemporary analyses often rely on data curated by government or research institutions, where some data characteristics may be unknown to the analysts.
+When unexpected outcomes arise, they often later traced back to implicit assumptions made by the analysts - assumptions that were unknowingly violated by the idiosyncratic in the data.
+By externalizing these assumptions through analysis validation checks, we aim to make it easier to identify and diagnose such unexpected results.
 
@@
 To diagnose potential reasons why this expectation might fail, we can establish a few analysis validation checks in anticipation of seeing the data.
-For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
+[TODO: Add guidelines for designing the checks] For example, we can check the quantile of the step count, if more than a third of the days fall below 8,000, or more than a third exceed 10,000 steps, this could indicate an excess of low-count or high-count days.
 Similarly, we may may expect the standard deviation of the step count not to be overly large.
@@
 
-The approach uses simulated datasets that have the same structure and characteristics of the observed data to be analyzed in the future.
-For each simulated dataset, analysis validation checks are applied to record  their TRUE or FALSE values.
-The outcome of the analysis is also computed to determine whether it is unexpected (1) or expected (0).
-By generating many datasets, each time applying the analysis validation checks and computing the outcome, we can relate the patterns in the checks to the likely that the outcome will be unexpected.
+The approach generates simulated datasets from the observed data to compute the outcome and the analysis validation checks (unexpected as 1 and expected as 0).
+This then allows us to apply a logic regression tree to link the outcome and the check results to construct a Boolean expression to best predict the unexpected outcomes.
 @fig-metric-calc provides an overview of the process.
+
+The specific strategy to simulate the data depends on the problem and the nature of the data at hand.
+Simulated data can be generated using prior domain knowledge, summary statistics derived from the observed data, or statistical techniques such as training-testing splits or bootstrap sampling.
+While each simulated data may not be independent of the original data - as is the case with bootstrap samples - it should adequately capture the range of possible idiosyncrasies present in the data.
+Failure to do so can reduce the ability of the resulting Boolean expression to predict unexpected outcomes.
 
@@
 To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used.
-Originally developed for SNP micro-array data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes.
+Originally developed for SNP micro-array data, logic regression predict classification or regression outcomes by constructing Boolean combinations, $\mathcal{L}(X_1, X_2, \cdots, X_k)$, where $X_i$ are binary predictors.
 The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
@@
 
-$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$
+$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)$$
 
-where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_n$, are labeled as 1 if they fail and 0 if they pass.
+where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_k$, are labeled as 1 if they fail and 0 if they pass.
 
@@
 
-Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables multiple variables based on Kullback-Leibler distance:
+Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables, based on Kullback-Leibler distance:
 
-$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},$$
+$$C(X_1, X_2, \cdots, X_k) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_k} p(x_1, x_2, \cdots, x_k) \log \frac{p(x_1, x_2, \cdots, x_k)}{p(x_1)p(x_2) \cdots p(x_k)},$$
 
-where $p(.)$ is the TODO.
+where $p(x_1, x_2, \cdots, x_k)$ is the joint distribution of the predictors and $p(\cdot)$ is the individual distribution of the predictors.
 A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
@@
 #| label: fig-dist-fit
-#| fig-cap: QQ-plot of the distribution fit for mortality, PM10, and temperature based on the fitted distribution from the original data.
-The fitted distribution is compared to the observed data to assess the distribution fit.
+#| fig-cap: "QQ-plot of the goodness-of-fit of selected distributions used to generate simulated data for mortality, PM10, and temperature.
+Left: Mortality as a negative binomial distribution (size = 74, mean = 183), Middle: PM10 is modeled using a rescaled beta distribution with shape parameters 4.21 and 11.67, multiplied by 10 to match the observed range.
+Right: Temperature is modeled using a Weibull distribution (shape = 3.8, scale = 61)"
 #| fig.height: 4
@@
 aa2 <- fitdistrplus::fitdist(pm10$mortality, "nbinom")
-p1 <- tibble(fitted= rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
-       observed= pm10$mortality) %>%
+p1 <- tibble(fitted = rnbinom(558, size = aa2$estimate[1], mu = aa2$estimate[2]),
+       observed = pm10$mortality) %>%
   ggplot(aes(sample = fitted)) +
-  stat_qq() +
-  stat_qq_line() + 
+  stat_qq_band(fill = "grey80") +
+  stat_qq_line(color = "grey30") +
+  stat_qq_point(size = 0.3) +
   theme_bw() + 
   theme(aspect.ratio = 1) + 
-  ggtitle("mortality: \nNB(size = 74, mu = 183)")
+  ggtitle("Mortality")
 
@@
   ggplot(aes(sample = fitted)) +
-  stat_qq() +
-  stat_qq_line() + 
+  stat_qq_band(fill = "grey80") +
+  stat_qq_line(color = "grey30") +
+  stat_qq_point(size = 0.3) +
   theme_bw() + 
   theme(aspect.ratio = 1) +
-  ggtitle("pm10: \n100 * beta(shp1 = 4.21, shp2 = 11.67)")
+  ggtitle("PM10")
 
@@
   ggplot(aes(sample = fitted)) +
-  stat_qq() +
-  stat_qq_line() + 
+  stat_qq_band(fill = "grey80") +
+  stat_qq_line(color = "grey30") +
+  stat_qq_point(size = 0.3) +
   theme_bw() + 
   theme(aspect.ratio = 1) +
-  ggtitle("temperature: \nweibull(shape = 3.8, scale = 61)")
+  ggtitle("Temperature")
 

### 2025-07-21 revision/response.qmd
revision done

--- 2025-07-20 _manuscript/revision/response.qmd
@@
 date: "2025-07-08"
-format: html
+format: pdf
 bibliography: ../references.bib
@@
 To note that, even when the results is as expected, we don't necessarily assume the model is correct.
-[More to address the second part]**
+For generating synthetic data with similar properties to real data, we discussed in the discussion the potential of using recent work in data thinning, data fission and differential privacy.
+**
 
@@
 Clarifying this aspect would enhance the practical utility of the approach and help readers apply it more effectively in real-world settings.
-
 **A: In the toy example section (Section 3.1), we include some explanations on the construction of validation checks: these are diagnostics analysts may naturally check when they receive an unexpected outcome.
@@
 Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology.
-**A: I need to investigate on this.**
+**A: After reviewing the definition of the independence metric, we find it is the previous version is less ideal and we have refined the definition.
+The amount of overlapping information among checks can be represented by the normalized total correlation, $C' = \frac{C(X_1, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}$, where$H(X_i)$ is the entropy of check $X_i$, $C(X_1, \cdots, X_k)$ and $H(X_1, \cdots, X_k)$ are the total correlation and joint entropy of checks $\{X_1, \cdots, X_k\}$, respectively.
+The amount of independent information is then $\eta = 1 - C' = \frac{ H(X_1, \cdots, X_k)}{\sum_{i = 1}^k}$.
+The independence metric is now defined as  $\frac{\eta - 1/k}{1 - 1/k}$ to scale $\eta$ to [0, 1], where 0 is attained when all checks are independent and 1 is attained when checks are all identical.
+A trivia case occurs when there is only one single check, then both the joint entropy and the sum of all entropies is the entropy of the check itself, hence giving an independence of 1.
+These changes have been reflected in Section 4.2 and in the examples.**
 
@@
 
-2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy. TODO**
+2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: Fixed.**
 
@@
 Even after standardizing to per observation, the range may still be not \[0, 1) as desired.
-**A: The range of total correlation is \[0, ∞), to obtain a measure in \[0, 1), we calculate total correlation per observation as the independence score.
-This is explained in the last sentence of the paragraph.
-TODO**
+**A: We have re-defined the independence metric (See *General comments* #4) and the metric is now bounded in [0, 1].
+**
 
@@
 As questioned in one of my general comments, this might not be desirable.
-**A: ask Roger** may or may not being able to simulate a particular characteristic of the data and generate
+**A: While using a different copula (t- copula or extreme value distribution copula) to better model the tail dependence could produce better simulated data to model temperature and mortality, high temperature does not necessarily correlates with high PM10 in air pollution literature.
+We have maintained the current simulation scheme but add a discussion on the matter in the discussion section.
+**
 

### 2025-07-21 index.qmd
revision done

@@
 
-We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as, TidyTuesday screencast videos or think-aloud type studies [e.g.
+We might gain insight into analysts' thought processes by speaking with them directly or watching them work via screencast videos they produce, such as TidyTuesday screencast videos or think-aloud type studies [e.g.
 @gu2024data].
 However, direct observation of analysis is not scalable and may not always be feasible; creating educational screencast videos requires significant effort from the analysts.
-Ideally, there could be a way to explicitly communciate these throught processes in the data analysis to others.
+Ideally, there could be a way to explicitly communicate these thought processes in the data analysis to others.
 Even better, if these expectations were machine-readable, we could analyze them and learn from the analysis itself.
 For example, we could answer questions about whether the checks also apply to other researchers analyzing new data in the same context, whether they reflect common practices in the field, or whether they are specific to the data or analysis at hand.
-Externalizing these thought processes is a practice that has potential to improve the trustworthiness of the analyses [@yu2024veridical] in general and the trustworthiness of subsequent products, such as machine learning models, that may be built on such analyses.
-Developing an approach to reveal some of this process without requiring a reader to essentially reconstruct the analysis process from the raw data would provide an improved basis for trusting an analysis result [@peng2021reproducible].
+Externalizing these thought processes can improve the trustworthiness of the analyses [@yu2024veridical].
 
-In this paper, we conceptualize these inexplicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis.
+The trustworthiness of an analysis depends not only on the robustness of the methods used but also on a deep understanding of the data.
+This corresponds to the third step identified in @broderick_toward_2023, where trust can break down if the assumptions underlying an algorithm are not satisfied in practice.
+In the past, data were often collected directly by analysts through carefully designed experiments, however, contemporary analyses often rely on data curated by government or research institutions, where some data characteristics may be unknown to the analysts.
+When unexpected outcomes arise, they often later traced back to implicit assumptions made by the analysts - assumptions that were unknowingly violated by the idiosyncratic in the data.
+By externalizing these assumptions through analysis validation checks, we aim to make it easier to identify and diagnose such unexpected results.
+
+In this paper, we conceptualize these implicit expectations as *analysis validation checks*, which allows us to examine the assumptions made about the data during an analysis.
 We then introduce a procedure to identify the subset of checks that best predict the occurrence of unexpected outcomes.
 The procedure, based on simulations of the original data, compares the accuracy and independence of different combinations of analysis validation checks.
-Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence is measures the shared information among checks.
+Accuracy is determined using binary classification metrics (precision and recall) from a logic regression model [@ruczinski_logic_2003], while independence measures the shared information among checks.
 The proposed workflow offers a numerical guarantee that the analysis will produce the expected results, assuming the assumptions about the data generating mechanism hold.
@@
 While not all analysis validation checks are equally important, some may be more useful at diagnosing unexpected outcomes than others.
-This section introduces a procedure to identify a subset of analysis validation checks that best predict the occurence of unexpected outcomes.
+This section introduces a procedure to identify a subset of analysis validation checks that best predict the occurrence of unexpected outcomes.
 
-The approach uses simulated datasets that have the same structure and characteristics of the observed data to be analyzed in the future.
-For each simulated dataset, analysis validation checks are applied to record  their TRUE or FALSE values.
-The outcome of the analysis is also computed to determine whether it is unexpected (1) or expected (0).
-By generating many datasets, each time applying the analysis validation checks and computing the outcome, we can relate the patterns in the checks to the likely that the outcome will be unexpected.
+The approach generates simulated datasets from the observed data to compute the outcome and the analysis validation checks (unexpected as 1 and expected as 0).
+This then allows us to apply a logic regression tree to link the outcome and the check results to construct a Boolean expression to best predict the unexpected outcomes.
 @fig-metric-calc provides an overview of the process.
+
+The specific strategy to simulate the data depends on the problem and the nature of the data at hand.
+Simulated data can be generated using prior domain knowledge, summary statistics derived from the observed data, or statistical techniques such as training-testing splits or bootstrap sampling.
+While each simulated data may not be independent of the original data - as is the case with bootstrap samples - it should adequately capture the range of possible idiosyncrasies present in the data.
+Failure to do so can reduce the ability of the resulting Boolean expression to predict unexpected outcomes.
 
@@
 To relate multiple checks to the outcome, a logic regression model [@ruczinski_logic_2003] is used.
-Originally developed for SNP microarray data, logic regression constructs Boolean combinations of binary variables, $\mathcal{L}(X_1, X_2, \cdots, X_n)$, to predict classification or regression outcomes.
+Originally developed for SNP micro-array data, logic regression predict classification or regression outcomes by constructing Boolean combinations, $\mathcal{L}(X_1, X_2, \cdots, X_k)$, where $X_i$ are binary predictors.
 The logical operations allowed to construct $\mathcal{L(\cdot)}$ includes AND ($\land$), OR ($\lor$), and NOT ($\neg$) and an example of the Boolean combination of $\mathcal{L(\cdot)}$ is $X_1 \text{ AND } (X_2 \text{ OR } X_3)$.
@@
 
-$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)$$
+$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)$$
 
-where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_n$, are labeled as 1 if they fail and 0 if they pass.
+where the outcome, $Y$, indicates whether the outcome is *unexpected* (1) or *expected* (0), and the analysis validation checks, $X_1, X_2, \cdots, X_k$, are labeled as 1 if they fail and 0 if they pass.
 
-This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or purne the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
+This objective function is optimized using simulated annealing algorithm to minimize the misclassification error and the following 6 moves are permitted during the optimization to grow or prune the tree (as detailed in Figure 2 of @ruczinski_logic_2003): 1) replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.
 Predictions from the logic regression model is then used, along with the observed outcome, to calculate the precision and recall of the optimized Boolean combination of checks.
@@
 This could happen if a set of checks are all tangentially related to the cause of the unexpected results, but none addresses the root cause.
-The independence score in @fig-metric-calc is defined as 1 - total correlation per observation.
+To quantify the redundancy among the checks, we use total correlation, $C$, a multivariate generalization of mutual information that captures the amount of shared information across two variables.
+We define the normalized total correlation, $C'$, as
 
-Total correlation is a generalization of mutual information that measures the amount of shared information across multiple variables multiple variables based on Kullback-Leibler distance:
+$$C'(X_1, X_2, \cdots, X_k) = \frac{C(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}, $$
+where $H(X_i)$ is the entropy of the $i$-th check and $H(X_1, X_2, \cdots, X_k)$ is the joint entropy of checks $\{X_1, X_2, \cdots, X_k\}$.
+The amount of independent information, $\eta$, can be defined as:
 
-$$C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}$$
+$$\eta(X_1, X_2, \cdots, X_k) = 1 - C'(X_1, X_2, \cdots, X_k) = \frac{H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}.$$
 
-A high total correlation value indicates overlapping of information among the analysis validation checks, while a low value suggests that the checks are independent and provides unique information to diagnose the unexpected outcome.
-To standardize this measure, the total correlation *per observation* is used in calculating the independence score.
+To scale this quantity to the range $[0,1]$ for consistency with other metrics (i.e., precision and recall), we define the independence metric as:
 
-The three metrics (precision, recall, and independence) can be combiend into a single metric using the arithmetic mean, harmonic mean, or quadratic mean.
+$$\text{independence} = \frac{\eta - 1/k}{1 - 1/k}= \frac{H(X_1, X_2, \cdots, X_k)/\sum_{i= 1}^K H(X_i) - 1/k}{1 - 1/k}.$$
+
+The metric attains a value of 1 indicates that all checks are independent: $H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)$ and provide unique information to diagnose the unexpected outcome (trivia case when there is only one check), and a value of 0 when the checks are identical: $H(X_1, X_2, \cdots, X_k)- \sum_{i= 1}^k H(X_i) = H(X_i)/ k H(X_i) = 1/k$.
+
+The three metrics (precision, recall, and independence) can be combined into a single metric using the arithmetic mean, harmonic mean, or quadratic mean.
 The differences among these means are minimal when the three metrics are similar.
@@
 regtree_df <- step_count |> dplyr::select(unexpect) |> bind_cols(pred_vec) |> 
+  calc_miscla_rate(unexpect, .fitted)
+
+sum_entropy <- lapply(as.list(step_count[,c(9:11)]),  function(x){infotheo::entropy(x, method = "emp")}) |> unlist() |> sum()
+multi_info <- infotheo::multiinformation(step_count[,c(9:11)])
+joint_entropy <- sum_entropy - multi_info
+ratio <- joint_entropy / sum_entropy
+
+regtree_df <- regtree_df |> 
+  mutate(independence =  (ratio - 1/3) / (1 - 1/3)) |> 
+  calc_metrics(metrics = c("harmonic", "arithmetic"))
+
+sum_entropy <- lapply(as.list(step_count[,c(9,11)]),  function(x){infotheo::entropy(x, method = "emp")}) |> unlist() |> sum()
+multi_info <- infotheo::multiinformation(step_count[,c(9,11)])
+joint_entropy <- sum_entropy - multi_info
+ratio <-  joint_entropy / sum_entropy
+
+another <- tibble(.fitted = as.vector(step_count[,9] & !step_count[,11]) |> as.numeric(), 
+                  unexpect = step_count$unexpect) |> 
   calc_miscla_rate(unexpect, .fitted) |>
-  mutate(overlapping = multiinformation(step_count[,c(8, 11)])/nrow(step_count), 
-         independence = 1 - overlapping) |> 
-  calc_metrics(metrics = c("harmonic", "arithmetic"))
+  mutate(independence = (ratio - 1/2) / (1 - 1/2)) |> 
+  calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
+  mutate(id = 6)
 ```
@@
               calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
-              mutate(checks = "Logic regression: (check 1 OR check 2) AND (not check 3)") |> 
-              dplyr::select(-overlapping)) |> 
+              mutate(checks = "Logic regression: (check 1 OR check 2) AND (not check 3)")) |> 
+  bind_rows(another  |>
+              mutate(checks = "Comparison: (check 1) AND (not check 3)")) |>
   bind_rows(regtree_df |>
-              dplyr::select(-overlapping) |>
               mutate(checks = "Regression tree")) |>
@@
   kableExtra::kable_styling(latex_options="scale_down") |> 
-  column_spec(1, width = '17em')
+  column_spec(1, width = '19em')
 ```
 
-For comparison, the regression tree produces a similar prediction to the logic regression, however, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
+For comparison, we include a simplified rule, (quantile(step, 0.6) > 10,000) AND (NOT sd(step) > 2,500).
+Although the checks on the 60th quantile and standard deviation are less correlated -- reflected in a higher independence score -- this pair is less accurate at predicting the unexpected outcome, resulting in lower harmonic and arithmetic mean scores.
+The regression tree produces a similar prediction to the logic regression, however, we argue that the logic regression tree shown in @fig-logic-reg is more interpretable for our purposes because it provides a direct representation of which combinations of analysis checks lead to unexpected outcomes.
 The logic regression tree is also directly comparable to other diagnostic techniques, which we discuss further in @sec-discussion.
@@
   tibble(.fitted = pm10_df$PM10_outlier, 
-         unexpect = pm10_df$unexpect),
-  augment(fit)) |>
+         unexpect = pm10_df$unexpect)) |>
   map_dfr(~.x |> calc_miscla_rate(.fitted, unexpect) |> 
-            mutate(overlapping = 0, independence = 1) |> 
-            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")
+            mutate(independence = 1) |> 
+            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id") |> 
+  mutate(id = as.numeric(id))
+
+reg <- fit |> augment() |>
+  calc_miscla_rate(unexpect, .fitted) |>
+  calc_independence() |> 
+  calc_metrics(metrics = c("harmonic", "arithmetic")) |> 
+  mutate(id = 5)
+
+df <- df |> bind_rows(reg)
 
@@
                         "Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4))")) |> 
-  bind_cols(df |> dplyr::select(-id, -overlapping)) |> rename(`indep.` = independence)
+  bind_cols(df |> dplyr::select(-id)) |> rename(`indep.` = independence)
 
@@
 In particular, our split-data approach, using New York City to guide the simulation and then applying the analysis checks to other cities, may not have been ideal.
-It may be worth exploring some recent work in data thinning [@neufeld2024data] and data fission [@leiner2023data] to create training datasets that are more representative of future data.
+It may be worth exploring some recent work in data thinning [@neufeld2024data], data fission [@leiner2023data], or differential privacy methods [@dong_gaussian_2022] to create training datasets that are more representative of future data.
+The Gaussian copula used in the example doesn't account for the tail dependence between the variables and other copula, e.g.
+t-copula or extreme value distribution copula, could also be explored to better capture the tail dependence.
 

### 2025-07-21 index.tex
revision done

--- 2025-07-20 _manuscript/_tex/index.tex
@@
 
-\phantomsection\label{cell-fig-step-count}
-\begin{figure}[H]
+\begin{verbatim}
+Orientation inferred to be along y-axis; override with
+`position_quasirandom(orientation = 'x')`
+\end{verbatim}
+
+\begin{figure}
 
@@
 
-\phantomsection\label{cell-fig-metric-calc}
-\begin{figure}[H]
+\begin{figure}
 
@@
 This could happen if a set of checks are all tangentially related to the
-cause of the unexpected results, but none addresses the root cause. The
-independence score in Figure~\ref{fig-metric-calc} is defined as 1 minus
-total correlation per observation.
-
-Total correlation is a generalization of mutual information that
-measures the amount of shared information across multiple variables,
-based on Kullback-Leibler distance:
-
-\[C(X_1, X_2, \cdots, X_k) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_k} p(x_1, x_2, \cdots, x_k) \log \frac{p(x_1, x_2, \cdots, x_k)}{p(x_1)p(x_2) \cdots p(x_k)},\]
-
-where \(p(x_1, x_2, \cdots, x_k)\) is the joint distribution of the
-predictors and \(p(\cdot)\) is the individual distribution of the
-predictors. A high total correlation value indicates overlapping of
-information among the analysis validation checks, while a low value
-suggests that the checks are independent and provides unique information
-to diagnose the unexpected outcome. To standardize this measure, the
-total correlation \emph{per observation} is used in calculating the
-independence score.
+cause of the unexpected results, but none addresses the root cause. To
+quantify the redundancy among the checks, we use total correlation,
+\(C\), a multivariate generalization of mutual information that captures
+the amount of shared information across two variables. We define the
+normalized total correlation, \(C'\), as
+
+\[C'(X_1, X_2, \cdots, X_k) = \frac{C(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}, \]
+where \(H(X_i)\) is the entropy of the \(i\)-th check and
+\(H(X_1, X_2, \cdots, X_k)\) is the joint entropy of checks
+\(\{X_1, X_2, \cdots, X_k\}\). The amount of independent information,
+\(\eta\), can be defined as:
+
+\[\eta(X_1, X_2, \cdots, X_k) = 1 - C'(X_1, X_2, \cdots, X_k) = \frac{H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}.\]
+
+To scale this quantity to the range \([0,1]\) for consistency with other
+metrics (i.e., precision and recall), we define the independence metric
+as:
+
+\[\text{independence} = \frac{\eta - 1/k}{1 - 1/k}= \frac{H(X_1, X_2, \cdots, X_k)/\sum_{i= 1}^K H(X_i) - 1/k}{1 - 1/k}.\]
+
+The metric attains a value of 1 indicates that all checks are
+independent: \(H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)\) and
+provide unique information to diagnose the unexpected outcome (trivia
+case when there is only one check), and a value of 0 when the checks are
+identical:
+\(H(X_1, X_2, \cdots, X_k)- \sum_{i= 1}^k H(X_i) = H(X_i)/ k H(X_i) = 1/k\).
 
@@
 
-\phantomsection\label{cell-fig-logic-reg}
-\begin{figure}[H]
+\begin{figure}
 
@@
 \resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
-\begin{tabular}{>{\raggedright\arraybackslash}p{17em}rrrrr}
+\begin{tabular}{>{\raggedright\arraybackslash}p{19em}rrrrr}
 \toprule
@@
 Check 3: sd(step) > 2500 & 0.153 & 0.289 & 1.000 & 0.273 & 0.481\\
-Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.998 & 0.669 & 0.763\\
-Regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774\\
+Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.818 & 0.637 & 0.703\\
+Comparison: (check 1) AND (not check 3) & 0.278 & 0.870 & 0.881 & 0.510 & 0.676\\
+\addlinespace
+Regression tree & 0.542 & 0.780 & 0.818 & 0.689 & 0.713\\
 \bottomrule
@@
 
-For comparison, the regression tree produces a similar prediction to the
-logic regression, however, we argue that the logic regression tree shown
-in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
-because it provides a direct representation of which combinations of
-analysis checks lead to unexpected outcomes. The logic regression tree
-is also directly comparable to other diagnostic techniques, which we
-discuss further in Section~\ref{sec-discussion}.
+For comparison, we include a simplified rule, (quantile(step, 0.6)
+\textgreater{} 10,000) AND (NOT sd(step) \textgreater{} 2,500). Although
+the checks on the 60th quantile and standard deviation are less
+correlated -- reflected in a higher independence score -- this pair is
+less accurate at predicting the unexpected outcome, resulting in lower
+harmonic and arithmetic mean scores. The regression tree produces a
+similar prediction to the logic regression, however, we argue that the
+logic regression tree shown in Figure~\ref{fig-logic-reg} is more
+interpretable for our purposes because it provides a direct
+representation of which combinations of analysis checks lead to
+unexpected outcomes. The logic regression tree is also directly
+comparable to other diagnostic techniques, which we discuss further in
+Section~\ref{sec-discussion}.
 
@@
 
-\phantomsection\label{cell-fig-dist-fit}
-\begin{figure}[H]
+\begin{figure}
 
@@
 \end{figure}%
+
+\begin{verbatim}
+Warning: There were 990 warnings in `mutate()`.
+The first warning was:
+i In argument: `fit = list(...)`.
+i In row 2.
+Caused by warning in `dpois()`:
+! non-integer x = 291.790133
+i Run `dplyr::last_dplyr_warnings()` to see the 989 remaining warnings.
+\end{verbatim}
+
+\begin{verbatim}
+Joining with `by = join_by(unexpect)`
+\end{verbatim}
 
@@
 \midrule
-Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1 & 0.763 & 0.815\\
-Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1 & 0.472 & 0.612\\
-Check 3: mortality outlier & 0.302 & 0.775 & 1 & 0.536 & 0.692\\
-Check 4: PM10 outlier & 0.305 & 0.766 & 1 & 0.537 & 0.690\\
-Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.722 & 0.876 & 1 & 0.851 & 0.866\\
+Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1.000 & 0.763 & 0.815\\
+Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1.000 & 0.472 & 0.612\\
+Check 3: mortality outlier & 0.302 & 0.775 & 1.000 & 0.536 & 0.692\\
+Check 4: PM10 outlier & 0.305 & 0.766 & 1.000 & 0.537 & 0.690\\
+Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.876 & 0.722 & 0.812 & 0.798 & 0.803\\
 \bottomrule
@@
 
-This generates a precision of 0.722 and a recall of 0.876 for predicting
+This generates a precision of 0.876 and a recall of 0.722 for predicting
 the unexpected PM10 coefficient. As shown in
@@
 temperature for the years 1992--2000.
+
+\begin{verbatim}
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 3288 Columns: 6
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr  (1): season
+dbl  (4): mortality, pm10, temp, dewpt
+date (1): date
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Rows: 40 Columns: 2
+-- Column specification --------------------------------------------------------
+Delimiter: ","
+chr (2): city, cityname
+
+i Use `spec()` to retrieve the full column specification for this data.
+i Specify the column types or set `show_col_types = FALSE` to quiet this message.
+Joining with `by = join_by(city)`
+\end{verbatim}
 
@@
 other cities, may not have been ideal. It may be worth exploring some
-recent work in data thinning \citep{neufeld2024data} and data fission
-\citep{leiner2023data} to create training datasets that are more
-representative of future data.
+recent work in data thinning \citep{neufeld2024data}, data fission
+\citep{leiner2023data}, or differential privacy methods
+\citep{dong_gaussian_2022} to create training datasets that are more
+representative of future data. The Gaussian copula used in the example
+doesn't account for the tail dependence between the variables and other
+copula, e.g.~t-copula or extreme value distribution copula, could also
+be explored to better capture the tail dependence.
 

### 2025-07-21 revision/response.qmd
format

--- 2025-07-09 _manuscript/revision/response.qmd
@@
 date: "2025-07-08"
-format: 
-  html:
-    toc: false
+format: pdf
 ---
@@
 Below is a point-by-point explanation of how we have changed the paper.
-Reviewers' comments are in normal text and **our responses are in bold text**.
+Reviewers' comments are in normal text and **\textcolor{purple}{our responses are in red}**.
 
@@
 
+**\textcolor{purple}{We now discuss the connection between our work and the trustworthiness of machine learning in the third paragraph of the Introduction.
+Specifically, our work address the breakdown in trust caused by  violations of algorithmic assumptions -- corresponding to the third step in Broderick et al.
+(2023).
+In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcome.
+Our approach advocates for making these assumptions -- often simple or taken for granted -- explicit, in order to more efficiently diagnose unexpected outcomes.}**
+
+2.
 The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2).
@@
 
+**\textcolor{purple}{The reviewer is correct that in certain scenarios (e.g.
+GEE and GMM), it can be challenging to accurately simulate dataset that resemble the observed data.
+However, the primary aim of our paper is not to recover the true data generating process or to find the model that best fits the data.
+Instead, our goal is to understand which characteristic of the data may lead to the unexpected results - a mismatch between the data and the model.
+Here, we hold the model fixed and examine the idiosyncrasies in the data that could causes such incompatibility.
+This contrasts with the approach of treating data as fixed and focus on improving the model.
+Our perspective is motivated by the applied studies, where certain modelling strategies are generally accepted for specific types of data.
+When unexpected outcomes arise, analysts often first investigate whether particular characteristics of the data may cause the results, rather than developing new models.
+In this context, the analysis validation checks we propose provide a pre-developed "diagnostic checks" to investigate the data.
+To note that, even when the results is as expected, we don't necessarily assume the model is correct.
+For generating synthetic data with similar properties to real data, we discussed in the discussion the potential of using recent work in data thinning, data fission and differential privacy.}**
+
+<!-- Train AI model with synthetic data. Take the image and rotate 45 degree. - maybe talk about it in the discussion** -->
+
+3.
 The construction of validation checks appears to be a critical step in applying the proposed methodology.
@@
 
+**\textcolor{purple}{In the toy example section (Section 3.1), we include some explanations on the construction of validation checks: these are diagnostics analysts may naturally check when they receive an unexpected outcome.
+For example, when the mean of a vector falls outside an expected range, analysts may check the the five-number summary or plot the variable distribution.
+The validation checks are those diagnostics framed as binary predicates.}**
+
+4.
 The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable.
@@
 
+**\textcolor{purple}{After reviewing the definition of the independence metric, we find it is the previous version is less ideal and we have refined the definition.
+The amount of overlapping information among checks can be represented by the normalized total correlation, $C' = \frac{C(X_1, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}$, where$H(X_i)$ is the entropy of check $X_i$, $C(X_1, \cdots, X_k)$ and $H(X_1, \cdots, X_k)$ are the total correlation and joint entropy of checks $\{X_1, \cdots, X_k\}$, respectively.
+The amount of independent information is then $\eta = 1 - C' = \frac{ H(X_1, \cdots, X_k)}{\sum_{i = 1}^k}$.
+The independence metric is now defined as  $\frac{\eta - 1/k}{1 - 1/k}$ to scale $\eta$ to [0, 1], where 0 is attained when all checks are independent and 1 is attained when checks are all identical.
+A trivia case occurs when there is only one single check, then both the joint entropy and the sum of all entropies is the entropy of the check itself, hence giving an independence of 1.
+These changes have been reflected in Section 4.2 and in the examples.}**
+
+5.
 The writing is generally clear and easy to follow.
@@
 The use of AI-based writing assistance tools could be helpful in this regard and, in my view, would constitute an ethical and constructive application of such tools.
-**A: We have run though the spelling check on the article and fixed the spelling mistakes.**
+
+**\textcolor{purple}{We have run though the spelling check on the article and fixed the spelling mistakes.}**
 
@@
 
-1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful.
+1.
+Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data?
+If so, this is a big difference from bootstrap samples and a remark would be useful.
+**\textcolor{purple}{The simulated dataset in our paper are not necessarily independent of the observed data, and we do not rely on bootstrap sampling at any point, although it can be a method to generate the simulated data.
+We have include a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.}**
 
-2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy.**
+2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **\textcolor{purple}{Fixed.}**
 
-3.  Section 4.1, paragraph 1: X1, . . . , Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size. **A: easy. TODO**
+3.
+Section 4.1, paragraph 1: X1, .
+.
+.
+, Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size.
+**\textcolor{purple}{We have added the definition of $X_i$ as the binary predictors in Section 4.1 and replace $n$ with $k$ for the number of checks.}**
 
@@
 This needs to be rephrased, perhaps just use “minus”.
-**A: We have replaced the dash with the word "minus" in the text.
-**
+**\textcolor{purple}{We have replaced the dash with the word *minus* in the text.}**
 
@@
 Even after standardizing to per observation, the range may still be not \[0, 1) as desired.
-**A: The range of total correlation is \[0, ∞), to obtain a measure in \[0, 1), we calculate total correlation per observation as the independence score.
-This is explained in the last sentence of the paragraph.
-**
+**\textcolor{purple}{We have re-defined the independence metric (See General comments 4) and the metric is now bounded in [0, 1].}**
 
-6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) as TODO and close the sentence.**
+6.  Continue: p is undefined; the equation needs to close with a period.  **\textcolor{purple}{We have added the definition of p(.) and close the sentence.}**
 
@@
 Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation).
+**\textcolor{purple}{Four mortality-pm10/ mortality-temperature tests are constructed because analysts may not be sure what the best cutoff point for the correlation tests would be.
+This example illustrates how the method developed can be used to determine the cutoff value for tests.}**
 
@@
 As questioned in one of my general comments, this might not be desirable.
+**\textcolor{purple}{While using a different copula (t- copula or extreme value distribution copula) to better model the tail dependence could produce better simulated data to model temperature and mortality, high temperature does not necessarily correlates with high PM10 in air pollution literature.
+We have maintained the current simulation scheme but add a discussion on the matter in the discussion section.}**
 
@@
 Figure 4: Could add bands of 95% confidence intervals of the QQ-plot; the texts on top of each panel could be moved to the caption; the labels of the axis should clarify empirical versus theoretical.
+**\textcolor{purple}{We have added the confidence interval, remove the plot title to the caption, and refine the axis labels.}**
 
-10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **A: Fixed.**
+10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **\textcolor{purple}{Fixed.}**
 
@@
 Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis.
-**A: The intention of using "second example" is because we count the toy example as the first example.
-But to avoid this confusion, we have changed this to "In the following example".**
+**\textcolor{purple}{We use the wording *second example* because we count the toy example as the first example.
+But, to avoid this confusion, we have changed the wording to *In the following example*.}**
 
-12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table.
+12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table. **\textcolor{purple}{Addressed in Specific comments 7.}**
 
@@
 For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks.
+**\textcolor{purple}{The reviewer's comment assumes we need a better model when an unexpected outcome occurs, however, in practice, we always assume there is something issue with the data, rather than rushing into developing new methods.
+The analysis validation checks are used to motivate a better understanding of the data, not an ideal model (This point is also addressed in General comments 2)}**
 
@@
 
--   Abstract: “simulations of the original data” is misleading. **A: we remove this information in this sentence for easier understanding of the sentence. **
+-   Abstract: “simulations of the original data” is misleading. **\textcolor{purple}{We remove this information in this sentence for easier understanding of the sentence.}**
 
--   Page 2, paragraph 1: “thought” process, not “throught” process. **A: Fixed.**
+-   Page 2, paragraph 1: “thought” process, not “throught” process. **\textcolor{purple}{Fixed.}**
 
--   Page 2, paragraph 2: “is measures”, two verbs. **A: The redundant "is" has been removed**.
+-   Page 2, paragraph 2: “is measures”, two verbs. **\textcolor{purple}{The redundant *is* has been removed.}**

### 2025-07-22 index.qmd
Minor edits to section 4.2

@@
 
-The metric attains a value of 1 indicates that all checks are independent: $H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)$ and provide unique information to diagnose the unexpected outcome (trivia case when there is only one check), and a value of 0 when the checks are identical: $H(X_1, X_2, \cdots, X_k)- \sum_{i= 1}^k H(X_i) = H(X_i)/ k H(X_i) = 1/k$.
+An independence metric value of 1 indicates that all checks are independent: $H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)$ and each check provides unique information to diagnose the unexpected outcome (the trivial case is when there is only one check).
+A value of 0 indicates the checks are all identical: $H(X_1, X_2, \cdots, X_k)- \sum_{i= 1}^k H(X_i) = H(X_i)/ k H(X_i) = 1/k$.
 

### 2025-07-22 index.tex
Minor edits to section 4.2

@@
-% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 ]{interact}
-\usepackage{xcolor}
+
 \usepackage{amsmath,amssymb}
-\setcounter{secnumdepth}{5}
+\usepackage{setspace}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 }{}
-\usepackage{setspace}
 \makeatletter
@@
 \makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\usepackage{longtable,booktabs,array}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
-
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-\usepackage[]{natbib}
-\bibliographystyle{plainnat}
 
@@
 \makeatother
+
+\usepackage[]{natbib}
+\bibliographystyle{plainnat}
 \usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
 
-The metric attains a value of 1 indicates that all checks are
-independent: \(H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)\) and
-provide unique information to diagnose the unexpected outcome (trivia
-case when there is only one check), and a value of 0 when the checks are
-identical:
+An independence metric value of 1 indicates that all checks are
+independent: \(H(X_1, X_2, \cdots, X_k)=\sum_{i= 1}^k H(X_i)\) and each
+check provides unique information to diagnose the unexpected outcome
+(the trivial case is when there is only one check). A value of 0
+indicates the checks are all identical:
 \(H(X_1, X_2, \cdots, X_k)- \sum_{i= 1}^k H(X_i) = H(X_i)/ k H(X_i) = 1/k\).
@@
 \midrule
-Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1.000 & 0.763 & 0.815\\
-Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1.000 & 0.472 & 0.612\\
-Check 3: mortality outlier & 0.302 & 0.775 & 1.000 & 0.536 & 0.692\\
-Check 4: PM10 outlier & 0.305 & 0.766 & 1.000 & 0.537 & 0.690\\
-Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.876 & 0.722 & 0.812 & 0.798 & 0.803\\
+Check 1: cor(m, PM10) < -0.03 & 0.518 & 0.888 & 1.000 & 0.740 & 0.802\\
+Check 2: cor(m, tmp) > -0.35 & 0.259 & 0.543 & 1.000 & 0.448 & 0.601\\
+Check 3: mortality outlier & 0.281 & 0.833 & 1.000 & 0.521 & 0.705\\
+Check 4: PM10 outlier & 0.289 & 0.871 & 1.000 & 0.535 & 0.720\\
+Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.874 & 0.667 & 0.779 & 0.764 & 0.773\\
 \bottomrule
@@
 
-This generates a precision of 0.876 and a recall of 0.722 for predicting
+This generates a precision of 0.874 and a recall of 0.667 for predicting
 the unexpected PM10 coefficient. As shown in

### 2025-07-22 revision/response.qmd

@@
 (2023).
-In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcome.
+In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcomes.
 Our approach advocates for making these assumptions -- often simple or taken for granted -- explicit, in order to more efficiently diagnose unexpected outcomes.}**
@@
 **\textcolor{purple}{The reviewer is correct that in certain scenarios (e.g.
-GEE and GMM), it can be challenging to accurately simulate dataset that resemble the observed data.
+GEE and GMM), it can be challenging to accurately simulate a dataset that resembles the observed data.
 However, the primary aim of our paper is not to recover the true data generating process or to find the model that best fits the data.
@@
 Here, we hold the model fixed and examine the idiosyncrasies in the data that could causes such incompatibility.
-This contrasts with the approach of treating data as fixed and focus on improving the model.
-Our perspective is motivated by the applied studies, where certain modelling strategies are generally accepted for specific types of data.
+This contrasts with the common approach of treating data as fixed and focusing on improving the model.
+Our perspective is motivated by applied studies, where certain modelling strategies are generally accepted for specific types of data.
 When unexpected outcomes arise, analysts often first investigate whether particular characteristics of the data may cause the results, rather than developing new models.
 In this context, the analysis validation checks we propose provide a pre-developed "diagnostic checks" to investigate the data.
-To note that, even when the results is as expected, we don't necessarily assume the model is correct.
-For generating synthetic data with similar properties to real data, we discussed in the discussion the potential of using recent work in data thinning, data fission and differential privacy.}**
+Note that even when the results are as-expected, \textit{we do not assume the model is correct}.
+For generating synthetic data with similar properties to real data, we agree that this is a challenging area worthy of further investigation.
+We have added material in the discussion regarding the potential of using recent work in data thinning, data fission, and differential privacy.}**
 
@@
 **\textcolor{purple}{After reviewing the definition of the independence metric, we find it is the previous version is less ideal and we have refined the definition.
-The amount of overlapping information among checks can be represented by the normalized total correlation, $C' = \frac{C(X_1, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}$, where$H(X_i)$ is the entropy of check $X_i$, $C(X_1, \cdots, X_k)$ and $H(X_1, \cdots, X_k)$ are the total correlation and joint entropy of checks $\{X_1, \cdots, X_k\}$, respectively.
+The amount of overlapping information among checks can be represented by the normalized total correlation, $C' = \frac{C(X_1, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}$, where $H(X_i)$ is the entropy of check $X_i$, $C(X_1, \cdots, X_k)$ and $H(X_1, \cdots, X_k)$ are the total correlation and joint entropy of checks $\{X_1, \cdots, X_k\}$, respectively.
 The amount of independent information is then $\eta = 1 - C' = \frac{ H(X_1, \cdots, X_k)}{\sum_{i = 1}^k}$.
 The independence metric is now defined as  $\frac{\eta - 1/k}{1 - 1/k}$ to scale $\eta$ to [0, 1], where 0 is attained when all checks are independent and 1 is attained when checks are all identical.
-A trivia case occurs when there is only one single check, then both the joint entropy and the sum of all entropies is the entropy of the check itself, hence giving an independence of 1.
+A trivial case occurs when there is only one single check, in which case both the joint entropy and the sum of all entropies is the entropy of the check itself, hence giving an independence of 1.
 These changes have been reflected in Section 4.2 and in the examples.}**
@@
 **\textcolor{purple}{The simulated dataset in our paper are not necessarily independent of the observed data, and we do not rely on bootstrap sampling at any point, although it can be a method to generate the simulated data.
-We have include a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.}**
+We have included a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.}**
 
@@
 As questioned in one of my general comments, this might not be desirable.
-**\textcolor{purple}{While using a different copula (t- copula or extreme value distribution copula) to better model the tail dependence could produce better simulated data to model temperature and mortality, high temperature does not necessarily correlates with high PM10 in air pollution literature.
-We have maintained the current simulation scheme but add a discussion on the matter in the discussion section.}**
+**\textcolor{purple}{While using a different copula (t- copula or extreme value distribution copula) to better model the tail dependence could produce better simulated data to model temperature and mortality, extreme high temperature does not always correlate with extreme high PM10 in the air pollution literature (hence its role as a confounder can vary).
+We have maintained the current simulation scheme but added a discussion on this matter in the discussion section.}**
 
@@
 For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks.
-**\textcolor{purple}{The reviewer's comment assumes we need a better model when an unexpected outcome occurs, however, in practice, we always assume there is something issue with the data, rather than rushing into developing new methods.
-The analysis validation checks are used to motivate a better understanding of the data, not an ideal model (This point is also addressed in General comments 2)}**
+**\textcolor{purple}{We appreciate the reviewer's suggestion, but would like to emphasize that our perspective here is slightly different.
+The reviewer's comment assumes we always need a better model when an unexpected outcome occurs.
+However, in practice, it is often likely that there is an issue with the data themselves, warranting at least an initial examination of the data collection or pre-processing aspects, not a revision of the model (although the model may be ultimately updated).
+Our motiviation in developing the analysis validation checks is primarily to develop a better understanding of the data, not to identify a best-fitting model (This point is also addressed in General comments 2)}**
 

### 2025-07-22 index.qmd
fix reference

@@
    notebook-links: false
+   keywords: diagnostic, logic regression, data analysis assumptions
 preamble: >
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
 #remotes::install_github("huizezhang-sherry/adtoolbox")
@@
 
-$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)$$
+$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k),$$
 
@@
 
-The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10.
+The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model (Poisson family with a log link) to relate daily mortality counts to daily measures of PM10.
 Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet2000fine; @welty2005acute].
@@
 
-tbl <- tibble(check = c("Check 1: cor(m, PM10) < -0.03", "Check 2: cor(m, tmp) > -0.35",
+tbl <- tibble(check = c("Check 1: cor(m, PM10) < -- 0.03", "Check 2: cor(m, tmp) > -- 0.35",
                         "Check 3: mortality outlier", "Check 4: PM10 outlier", 
@@
 
-  > (cor(mortality, PM10) < -0.03) AND (cor(mortality, tmp) > -0.3 OR (mortality outlier AND PM10 outlier))
+  > (cor(mortality, PM10) < -- 0.03) AND (cor(mortality, tmp) > -- 0.3 OR (mortality outlier AND PM10 outlier))
 
@@
 
-```{r}
+```{r message=FALSE}
 #| label: tbl-all-cities
@@
          cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
-         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
+         across(c(pm10, cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
   mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 

### 2025-07-22 index.tex
fix reference

@@
   pdfauthor={H. Sherry Zhang; Roger D. Peng},
+  pdfkeywords={diagnostic, logic regression, data analysis assumptions},
   colorlinks=true,
@@
 \end{abstract}
+\begin{keywords}
+\def\sep{;\ }
+diagnostic, logic regression, data analysis assumptions\sep 
+diagnostic, logic regression, data analysis assumptions
+\end{keywords}
 
@@
 
-\begin{verbatim}
-Orientation inferred to be along y-axis; override with
-`position_quasirandom(orientation = 'x')`
-\end{verbatim}
-
 \begin{figure}
@@
 
-\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)\]
+\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k),\]
 
@@
 The typical approach to studying the association between PM10 and
-mortality is to apply a generalized linear model with a Poisson link to
-relate daily mortality counts to daily measures of PM10. Based on
-previous work and the range of effect sizes published in the literature,
-an analyst might expect the coefficient for PM10 in this GLM to lie
-between \([0, 0.005]\), after adjusting for daily temperature
+mortality is to apply a generalized linear model (Poisson family with a
+log link) to relate daily mortality counts to daily measures of PM10.
+Based on previous work and the range of effect sizes published in the
+literature, an analyst might expect the coefficient for PM10 in this GLM
+to lie between \([0, 0.005]\), after adjusting for daily temperature
 \citep{samet2000fine, welty2005acute}. Note that the relatively simple
@@
 
-\begin{verbatim}
-Warning: There were 990 warnings in `mutate()`.
-The first warning was:
-i In argument: `fit = list(...)`.
-i In row 2.
-Caused by warning in `dpois()`:
-! non-integer x = 291.790133
-i Run `dplyr::last_dplyr_warnings()` to see the 989 remaining warnings.
-\end{verbatim}
-
-\begin{verbatim}
-Joining with `by = join_by(unexpect)`
-\end{verbatim}
-
 A logic regression is fitted using all variations of the checks in
@@
 \midrule
-Check 1: cor(m, PM10) < -0.03 & 0.518 & 0.888 & 1.000 & 0.740 & 0.802\\
-Check 2: cor(m, tmp) > -0.35 & 0.259 & 0.543 & 1.000 & 0.448 & 0.601\\
-Check 3: mortality outlier & 0.281 & 0.833 & 1.000 & 0.521 & 0.705\\
-Check 4: PM10 outlier & 0.289 & 0.871 & 1.000 & 0.535 & 0.720\\
-Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.874 & 0.667 & 0.779 & 0.764 & 0.773\\
+Check 1: cor(m, PM10) < -- 0.03 & 0.551 & 0.894 & 1.000 & 0.763 & 0.815\\
+Check 2: cor(m, tmp) > -- 0.35 & 0.281 & 0.556 & 1.000 & 0.472 & 0.612\\
+Check 3: mortality outlier & 0.302 & 0.775 & 1.000 & 0.536 & 0.692\\
+Check 4: PM10 outlier & 0.305 & 0.766 & 1.000 & 0.537 & 0.690\\
+Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.876 & 0.722 & 0.812 & 0.798 & 0.803\\
 \bottomrule
@@
 \begin{quote}
-(cor(mortality, PM10) \textless{} -0.03) AND (cor(mortality, tmp)
-\textgreater{} -0.3 OR (mortality outlier AND PM10 outlier))
+(cor(mortality, PM10) \textless{} -- 0.03) AND (cor(mortality, tmp)
+\textgreater{} -- 0.3 OR (mortality outlier AND PM10 outlier))
 \end{quote}
 
-This generates a precision of 0.874 and a recall of 0.667 for predicting
+This generates a precision of 0.876 and a recall of 0.722 for predicting
 the unexpected PM10 coefficient. As shown in
@@
 
-\begin{verbatim}
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 3288 Columns: 6
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr  (1): season
-dbl  (4): mortality, pm10, temp, dewpt
-date (1): date
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Rows: 40 Columns: 2
--- Column specification --------------------------------------------------------
-Delimiter: ","
-chr (2): city, cityname
-
-i Use `spec()` to retrieve the full column specification for this data.
-i Specify the column types or set `show_col_types = FALSE` to quiet this message.
-Joining with `by = join_by(city)`
-\end{verbatim}
-
 \begin{table}
@@
 \midrule
-Atlanta & -0.0008 & T & T & --0.26 (T) & --0.16 (T) & F & F\\
-Austin & -0.0030 & T & T & --0.12 (T) & --0.11 (T) & F & T\\
+Atlanta & --0.0008 & T & T & --0.26 (T) & --0.16 (T) & F & F\\
+Austin & --0.0030 & T & T & --0.12 (T) & --0.11 (T) & F & T\\
 Baltimore & 0.0018 & F & F & --0.21 (T) & 0.01 (F) & T & T\\
@@
 Las Vegas & 0.0008 & F & F & --0.27 (T) & 0.05 (F) & T & T\\
-Memphis & -0.0007 & T & T & --0.16 (T) & --0.10 (T) & F & T\\
-Miami & -0.0004 & T & F & --0.16 (T) & --0.02 (F) & F & T\\
+Memphis & --0.0007 & T & T & --0.16 (T) & --0.10 (T) & F & T\\
+Miami & --0.0004 & T & F & --0.16 (T) & --0.02 (F) & F & T\\
 \addlinespace
@@
 Oakland & 0.0008 & F & F & --0.28 (T) & 0.04 (F) & F & T\\
-Orlando & -0.0008 & T & T & --0.21 (T) & --0.05 (T) & F & T\\
+Orlando & --0.0008 & T & T & --0.21 (T) & --0.05 (T) & F & T\\
 \addlinespace
@@
 \addlinespace
-Salt Lake City & -0.0004 & T & T & --0.19 (T) & --0.03 (T) & F & T\\
-San Antonio & -0.0021 & T & T & --0.22 (T) & --0.16 (T) & F & T\\
+Salt Lake City & --0.0004 & T & T & --0.19 (T) & --0.03 (T) & F & T\\
+San Antonio & --0.0021 & T & T & --0.22 (T) & --0.16 (T) & F & T\\
 San Bernardino & 0.0006 & F & T & --0.28 (T) & --0.09 (T) & T & T\\
@@
 St. Petersburg & 0.0012 & F & F & --0.31 (F) & 0.01 (F) & F & T\\
-Tampa & -0.0003 & T & T & --0.20 (T) & --0.04 (T) & F & T\\
+Tampa & --0.0003 & T & T & --0.20 (T) & --0.04 (T) & F & T\\
 Tucson & 0.0028 & F & F & --0.31 (F) & 0.18 (F) & T & T\\

### 2025-10-02 index.qmd
minor revision

@@
 
+# Supplementary material {.unnumbered}
+
+The supplementary materials include a full script of the examples in the paper (`index.R`) and its output (`index.html`), the data used for in the examples in Section 5 (`data/`), the package source (`adtoolbox_0.1.0.tar.gz`), and a README.md file containing the install instructions for running the scripts.
+
 # References

### 2025-10-02 index.tex
minor revision

@@
 
+\section*{Supplementary material}\label{supplementary-material}
+\addcontentsline{toc}{section}{Supplementary material}
+
+The supplementary materials include a full script of the examples in the
+paper (\texttt{index.R}) and its output (\texttt{index.html}), the data
+used for in the examples in Section 5 (\texttt{data/}), the package
+source (\texttt{adtoolbox\_0.1.0.tar.gz}), and a README.md file
+containing the install instructions for running the scripts.
+
 \renewcommand\refname{References}

### 2025-10-02 supplementary/index.R

@@ -1,0 +1,357 @@
+#' # Reproducible script for *Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks*
+## ----setup, include=FALSE----------------------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, messages = FALSE)
+#remotes::install_github("huizezhang-sherry/adtoolbox")
+library(tidyverse)
+library(patchwork)
+library(MASS)
+library(ggh4x)
+library(adtoolbox)
+library(broom)
+library(knitr)
+library(rpart)
+library(infotheo)
+library(patchwork)
+library(readr)
+library(kableExtra)
+library(scales)
+library(qqplotr)
+pm10 <- read_csv(here::here("data", "pm10.csv")) |>
+  filter(!is.na(pm10)) |>
+  mutate(season = as.factor(season))
+
+
+## ----------------------------------------------------------------------------------------
+#| fig-height: 3
+#| label: fig-step-count
+#| fig-cap: Beeswarm plot of average step counts across 300 simulated 30-day periods. Each point represents the average step count from one simulation. Similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point. The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario.
+set.seed(1234)
+step_count <- tibble(id = 1:300) |>
+  rowwise() |>
+  mutate(
+    small_n = rpois(1, 8),
+    large_n = rpois(1, 8),
+    norm_sd = rlogis(1, 800, scale = 100),
+    norm_n = 30 - small_n - large_n,
+    step = list(round(c(rnorm(small_n, 6000, 200),
+                        rnorm(large_n, 12000, 200),
+                        rnorm(norm_n, 9000, 300)
+    ), 0)),
+    mean = mean(step, na.rm = TRUE),
+    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
+    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
[…]

### 2025-10-02 supplementary/index.R
supplementray materials as a zip

@@ -1,357 +1,0 @@
-#' # Reproducible script for *Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks*
-## ----setup, include=FALSE----------------------------------------------------------------
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, messages = FALSE)
-#remotes::install_github("huizezhang-sherry/adtoolbox")
-library(tidyverse)
-library(patchwork)
-library(MASS)
-library(ggh4x)
-library(adtoolbox)
-library(broom)
-library(knitr)
-library(rpart)
-library(infotheo)
-library(patchwork)
-library(readr)
-library(kableExtra)
-library(scales)
-library(qqplotr)
-pm10 <- read_csv(here::here("data", "pm10.csv")) |>
-  filter(!is.na(pm10)) |>
-  mutate(season = as.factor(season))
-
-
-## ----------------------------------------------------------------------------------------
-#| fig-height: 3
-#| label: fig-step-count
-#| fig-cap: Beeswarm plot of average step counts across 300 simulated 30-day periods. Each point represents the average step count from one simulation. Similar to a boxplot or violin plot, the beeswarm plot also displays the distribution of each individual data point. The orange points indicate instances where the average step count fails outside the [8,500, 9,500] interval, representing an unexpected outcome in this scenario.
-set.seed(1234)
-step_count <- tibble(id = 1:300) |>
-  rowwise() |>
-  mutate(
-    small_n = rpois(1, 8),
-    large_n = rpois(1, 8),
-    norm_sd = rlogis(1, 800, scale = 100),
-    norm_n = 30 - small_n - large_n,
-    step = list(round(c(rnorm(small_n, 6000, 200),
-                        rnorm(large_n, 12000, 200),
-                        rnorm(norm_n, 9000, 300)
-    ), 0)),
-    mean = mean(step, na.rm = TRUE),
-    unexpect = ifelse(between(mean, 8500, 9500), 0, 1),
-    `q(step, 0.6) > 10000` = ifelse(quantile(step, 0.6, na.rm = TRUE) > 10000, 1, 0),
[…]
